### 2019-07-12 paper.tex
intro outline and started methods section

%% Template article for Elsevier's document class `elsarticle'
%% with numbered style bibliographic references
\documentclass[preprint,12pt]{elsarticle}

\usepackage{setspace}
\doublespacing

% remove preprent footnote
\makeatletter
\def\ps@pprintTitle{%
 \let\@oddhead\@empty
 \let\@evenhead\@empty
 \def\@oddfoot{\centerline{\thepage}}%
 \let\@evenfoot\@oddfoot}
\makeatother

%% Use the option review to obtain double line spacing
%% \documentclass[preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times]{elsarticle}
%% \documentclass[final,1p,times,twocolumn]{elsarticle}
%% \documentclass[final,3p,times]{elsarticle}
%% \documentclass[final,3p,times,twocolumn]{elsarticle}
%% \documentclass[final,5p,times]{elsarticle}
%% \documentclass[final,5p,times,twocolumn]{elsarticle}

\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}

\biboptions{comma,round}

% shortcuts
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\T}{\intercal}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\bm}{\boldsymbol}  % bold maths symbols
\newcommand{\tl}{\tilde{\lambda}}   % thinned little lambda
\newcommand{\tL}{\tilde{\Lambda}}  % thinned big lambda

\begin{document}
\begin{frontmatter}
\title{Point transect distance sampling in inlabru}

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for the associated footnote;
%% use the fnref command within \author or \address for footnotes;
%% use the fntext command for the associated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for the associated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%%
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \address{Address\fnref{label3}}
%% \fntext[label3]{}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{<author name>}
%% \address[label1]{<address>}
%% \address[label2]{<address>}

\author{Andrew E Seaton}

\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}

\begin{abstract}
Point transect distance sampling in \texttt{inlabru}
\end{abstract}

\begin{keyword}
Distance sampling \sep Stochastic partial differential equations \sep Integrated nested Laplace approximation \sep Generalized additive model
\end{keyword}
\end{frontmatter}

\section*{Introduction}

Introduction outline:
\begin{enumerate}
	\item Estimation of abundance and spatial distribution of animals is an important topic
	\item Distance sampling is a family of methods that aim to do this whilst accounting for detectability of animals
	\item Distance sampling began life under a design-based inference paradigm
	\item Randomised survey design is used in two ways.  Firstly, to justify the assumption that the true distribution of animals is uniform with respect to distance from the observer.  Secondly, to construct Horvit-Thompson-like estimators of animal density, which are then converted to abundance estimates for regions of interest
	\item Interest in the spatial distribution of animals has led to the development of approaches that combine traditional distance sampling methods and spatially explicit generalized additive models
	\item In this approach the estimated detection probabilities for discrete spatial units are used as an offset in a generalized additive model where the response variable is a count of detections in each unit
	\item This two-stage approach needs a way to propogate uncertainty from the detection model to the spatial model
	\item Recently [[JOYCE PAPER]] presented a one-stage approach that simultaneously estimates the detection and spatial processes, avoids binning data into counts and removes the need to consider uncertainty propagation
	\item This approach considered line-transect distance sampling as thinned log-Gaussian Cox process
	\item To estimate the detection function [[JOYCE et al]] constructed a spline-like function using an SPDE approach and fit the model using \texttt{R-INLA}
	\item In this paper we show how to achieve a similar one-stage model for point transect distance sampling
	\item Point transect surveys require a different approach than line transects because the area surveyed increases with distance from observer
	\item In addition to accounting for this we present an alternative way to specify the detection function.  We replace the spline approach with a parametric family of detection functions such as half-normal and hazard rate functions
	\item  These detection functions are typically not linear in their parameters and so cannot be straightforwardly estimated using methods for additive linear predictors
	\item To solve this we implement an iterative approach based on a Taylor approximation of the detection function and fit the model using the \texttt{inlabru} package which extends the functionality of \texttt{R-INLA}
	\item This represents a novel development in distance sampling methods that allows simultaneous estimation of parametric detection functions and the spatial distribution of animals from point transect data
\end{enumerate}

\section*{Methods}

We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(s)$.  The log-Gaussian Cox process is a flexible approach that can include spatially structured random effects on the intensity process to account for unexplained heterogeneity not captured by fixed effect covariates.

\sloppy For the case with imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected})$. A key property of the log-Gaussian Cox process is that a realisation thinned with thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.

Standard distance sampling methodology specifies $g(s)$ as a function that decays with increasing distance from the observer.  For example, if $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated from the number of observed distances.  The parameter $\sigma$ can only be estimated if some assumption is made about the intensity of the animal locations.  Without such an assumption detectability and intensity are confounded.  The standard assumption of distance sampling is that the intensity is constant with respect to changes in $r(s)$.  Thus any observed differences from uniformity can be attributed to detectability, and not variation in the intensity.

A point transect distance sampling survey consists of a set of $K$ sampled regions.  The $k$-th sampled region we denote $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region $\Omega = \cup_{k=1}^K \Omega_k$.  For simplicity we assume that all sampled regions are discs with radius $W$ and $\Omega_j \cap \Omega_k = \emptyset$ for all $j \neq k$.  The thinning probability function $g(s)$ is then defined relative to the positions of the surveyed regions.  For any area not surveyed the probability of observing a point is $0$, i.e. $g(s) = 0$ for any $s \notin \cup_{k=1}^K \Omega_k$.  Let the probability of observing a point $s \in \Omega_k$ be denoted $g_k(s)$.  Since the surveyed regions are non-overlapping each location $s$ is associated with only one thinning probability function $g_k$.  For example, for an observer at location $s_k \in \Omega_k$, the half-normal detection function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$. The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation although we do not consider this here.  The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for any $s \in \Omega_k$.

The likelihood of observing points at locations $\bm{Y} = (s_1, \ldots, s_n)$ is
\begin{equation}
\pi(\bm{Y} | \theta, \phi) = \exp\left(-\int_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
\end{equation}

\end{document}

### 2019-07-19 paper.tex
methods section nearly finished

@@
 
-\begin{keyword}
-Distance sampling \sep Stochastic partial differential equations \sep Integrated nested Laplace approximation \sep Generalized additive model
-\end{keyword}
+%\begin{keyword}
+%Distance sampling \sep Stochastic partial differential equations \sep Integrated nested Laplace approximation \sep Generalized additive model
+%\end{keyword}
 \end{frontmatter}
@@
 	\item Estimation of abundance and spatial distribution of animals is an important topic
-	\item Distance sampling is a family of methods that aim to do this whilst accounting for detectability of animals
+	\item Distance sampling is a family of methods that aim to do this whilst accounting for imperfect detectability of animals
 	\item Distance sampling began life under a design-based inference paradigm
@@
 	\item Interest in the spatial distribution of animals has led to the development of approaches that combine traditional distance sampling methods and spatially explicit generalized additive models
-	\item In this approach the estimated detection probabilities for discrete spatial units are used as an offset in a generalized additive model where the response variable is a count of detections in each unit
+	\item In this approach the estimated detection probabilities for discrete spatial units are used as an offset in a generalized additive model where the response variable is a count of detections within each unit
 	\item This two-stage approach needs a way to propogate uncertainty from the detection model to the spatial model
@@
 
-\section*{Methods}
+\section*{Distance Sampling as a Thinned Point Process}
 
@@
 
-Standard distance sampling methodology specifies $g(s)$ as a function that decays with increasing distance from the observer.
-For example, if $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated from the number of observed distances.
+Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance from the observer.
+For example, if $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated using the observed distances.
 The parameter $\sigma$ can only be estimated if some assumption is made about the intensity of the animal locations.
@@
 The thinning probability function $g(s)$ is then defined relative to the positions of the surveyed regions.
-For any area not surveyed the probability of observing a point is $0$, i.e.
-$g(s) = 0$ for any $s \notin \cup_{k=1}^K \Omega_k$.
-Let the probability of observing a point $s \in \Omega_k$ be denoted $g_k(s)$.
-Since the surveyed regions are non-overlapping each location $s$ is associated with only one thinning probability function $g_k$.
-For example, for an observer at location $s_k \in \Omega_k$, the half-normal detection function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
-The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation although we do not consider this here.
-The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for any $s \in \Omega_k$.
+The probability of observing a point outside the surveyed region is zero.
+We denote the probability of observing a point $s \in \Omega_k$ as $g_k(s)$.
+Since the surveyed regions are non-overlapping each location $s$ is associated with a single thinning probability function $g_k$.
+For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
+The assumption of non-overlapping survey regions can be relaxed by including extra information such as the time of each observation.
+For simplicity we do not consider this here.
+The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for $s \in \Omega_k$.
 
-The likelihood of observing points at locations $\bm{Y} = (s_1, \ldots, s_n)$ is
+The likelihood of observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
 \begin{equation}
-\pi(\bm{Y} | \theta, \phi) = \exp\left(-\int_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
+\label{lgcp-likelihood}
+\pi(\bm{Y} | \theta, \phi) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
 \end{equation}
 
+The integral component of the likelihood does not usually have analytical solutions.
+Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood.
+
+To evaluate the integral we introduce polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.
+In general the thinning function depends only on $r$ and not on $\theta$.
+We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$.
+It follows that the integral can be rewritten as a sum of weighted one-dimensional integrals $\sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$ using the assumption that $\lambda(s)$ is constant within each sampling unit.
+For each sampling unit we approximate the one-dimensional integral using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
+This gives
+\begin{equation*}
+	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj}) 
+\end{equation*}
+where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(s_{kj}) = \lambda(s_k) g_k(r_{kj})$.
+
+To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.
+Similarly let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$
+
+Then the approximate log-likelihood is
+
+\begin{equation}
+\label{approx-log-likelihood}
+	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}
+\end{equation}
+
+This approximation can be written as a modified Poisson likelihood.  To see this let $\exp \eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$, 
+$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $y = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.
+Then the approximate likelihood becomes
+\begin{equation}
+\pi(\bm{Y}) \approx \prod_{i=1}^{KM + n} \eta_i^{y_i}\exp(-\alpha_i\eta_i)
+\end{equation}
+
+This is similar to a Poisson likelihood with an offset term equal to $\log\alpha$.  
+[[Cite ``Going off Grid" and ``Berman Turner" papers here]]
+[[ I don't understand how INLA/inlabru handles this specific likelihood form.
+The data corresponding to $\exp(-\alpha_i\eta_i)$ is straightforward using the $E$ argument with ``family = poisson".
+The $\eta_i^{y_i}$ part sets $E = 0$ but I don't see anything that would set $\exp(-\alpha_i\eta_i) =1$ for this part of the stack which is what we need to get back to \eqref{approx-log-likelihood}.
+
+This all differs from Berman-Turner/Baddeley papers in one key respect - the observed points do not form part of the integration scheme.
+For Berman-Turner this is necessary so we can define pseudo-data $\tilde{y}_i = y_i / \alpha_i$.
+Since all points are in the integration scheme $\alpha_i \neq 0$.
+This is the trick they use to get it into a completely Poisson form (although now the pseudo-data is non-integer).]]
+\section*{Intensities for incomplete data}
+
+In the above we assume the data are complete records of animal location.
+However, in many distance sampling surveys the exact location of the animal is not recorded - instead only the location of the observer and the distance to the observer are recorded.
+Data of this type can be analysed within a point process framework by deriving the appropriate intensity function for the incomplete data.
+
+Using the polar coordinates notation given above, for true animal location $s_k(r, \theta) \in \Omega_k$, we consider the case where we observe $r$ but not $\theta$.
+It follows that the intensity for points at a distance $r$ observed within sampling unit $\Omega_k$ is
+\begin{align}
+\tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(s)g_k(s)\mathrm{d}s \nonumber \\
+&= 2\pi\lambda(s_k)rg_k(r)
+\end{align}
+ where $c_k(r)$ is a circle of radius $r$ centred at $s_k$ and the second line follows from changing to polar coordinates and the assumption of constant intensity within $\Omega_k$.
+This intensity differs from the full data case by including the $2\pi$  term to account for the fact that we do not observe $\theta$ and the additional $r$ term to account for the increasing area surveyed at larger distances.
+
+\section*{Modelling $\lambda(s)$ using the stochastic partial differential equation approach}
+
+To model the intensity of animal locations we use a combination of fixed effects of spatial covariates and a spatially structured random effect.
+This allows us to associate distance sampling data with environmental and ecological conditions that may be relevant to understanding the ecology of the species of interest.
+For cases where we have insufficient covariates to fully explain the location of animals the spatially structured random effect accounts for additional heterogeneity in the intensity and is required to assume conditional independence between points, given the intensity.
+
+The log-intensity of animal locations is given by
+\begin{equation*}
+\log \lambda(s) = \beta_0 + \sum_v \beta_v z_v(s) + \xi(s)
+\end{equation*}
+where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatially indexed covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean Gaussian random field with Mat\'ern covariance
+\begin{equation}
+C(s_1,s_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|s_1-s_2\|)^{\nu}K_\nu(\kappa \|s_1-s_2\|)
+\end{equation}
+where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.
+All three parameters are not simultaneously identifiable and it is conventional to assume a value for $\nu$ which controls the mean-square differentiability of the process.
+We set $\nu = 1$ which is the most common default value in the approximation approach outlined below.
+[[Would like to justify $\nu = 1$ somehow]]
+
+We use a Gaussian Markov random field approximation of $\xi(s)$ following the approach of [[Lindgren 2011]] based on specifying $\xi(s)$ via a stochastic partial differential equation.
+Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent $\xi$ as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
+The parameters $\xi_1, \ldots, \xi_L$ follow a multivariate normal distribution with sparse precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a stochastic partial differential equation.
+The precision parameter $\tau$ controls the marginal variance of the field and $\kappa$ is an inverse range parameter that controls the rate of decay of the covariance as distance between locations increases.
+Further details of this Gaussian Markov random field approximation are given in the supplementary materials.
+
+\section*{Thinned Intensities}
+
+The above specifies a model for all animal locations in the area of interest.
+The log-intensity for observed points is $\log\tl(s) = \log\lambda(s) + \log g(s)$.
+This presents a problem since $\log g(s)$ is not typically linear in its parameters.
+However, for any single set of parameter values, $g(s)$ is essentially an offset term.
+That is to say, if we could specify $g(s)$ as known then we could proceed with inference with a standard additive linear predictor.
+
+TO DO:
+
+\begin{itemize}
+ \item Description of iterated INLA here
+\end{itemize}
+
+\section*{Case-Study:  Hawaiian Akepa Survey}
+
+TO DO:
+
+\begin{itemize}
+\item Bird survey locations image
+\item This section to include code snippets from inlabru
+\item Also mentions principled choice of priors and our choice for this akepa example
+\item Image of posterior mean of intensity
+\item Image of CV of posterior intensity
+\item Talk to Rick about something interesting to do with excursions package?  Multiple-testing an issue with gridded data as well - how is this solved?
+\end{itemize}
+
+\section*{Discussion}
+
+\begin{enumerate}
+	\item Iterated INLA is a powerful and flexible way to include non-linear components to a predictor and has applications beyond this application to distance sampling
+	\item More to do in future to test how many parameters can be included in the non-linear component and what types of non-linear components are feasible
+
+\end{enumerate}
+
 \end{document}

### 2019-07-22 analyses/AS.Nposterior.R
add example analysis scripts

@@ -1,0 +1,28 @@
+#' Plot the posterior of abundance for inlabru point process model
+#' Does some rough things to get 
+#' 
+#' @param model  fitted inlabru model
+#' @param predpts the thing that ipoints() returns, or something similar, must have weight column
+#' @param ... other args passed on to predict()
+#' 
+
+AS.Nposterior <- function(model, predpts, ...){
+  
+  # First get posterior for 
+  Lambda <- predict(model, predpts, ~ sum(weight * exp(grf + Intercept))) 
+  Lambda
+  
+  Nlow <- round(Lambda$mean - 4*Lambda$sd)
+  Nhigh <- round(Lambda$mean + 4*Lambda$sd)
+  by <- round((Nhigh - Nlow)/500)   # should give roughly 500 
+  
+  Nrange <- seq(Nlow, Nhigh, by = by)
+  Npost = predict(model, predpts, 
+                 ~ data.frame(N = Nrange, 
+                              dpois(Nrange, lambda = sum(weight * exp(grf + Intercept)))))
+  
+  return(list(Npost = Npost, Nexp = Lambda))
+}
+
+
+

### 2019-07-22 analyses/data/create_mesh.R

@@ -1,0 +1,30 @@
+# Load required packages
+library(inlabru)
+library(INLA)  # need testing version of inla
+
+point_transects <- readRDS("samplers_no_crs.RDS")
+study_area <- readRDS("study_area_no_crs.RDS")
+
+# meshbuilder()
+
+## Generated by meshbuilder()
+
+## Build boundary information:
+## (fmesher supports SpatialPolygons, but this app is not (yet) intelligent enough for that.)
+boundary <- list(
+  as.inla.mesh.segment(study_area),
+  NULL)
+
+## Build the mesh:
+mesh <- inla.mesh.2d(boundary=boundary,
+                     max.edge=c(280, 960),
+                     min.angle=c(30, 21),
+                     max.n=c(48000, 16000), ## Safeguard against large meshes.
+                     max.n.strict=c(128000, 128000), ## Don't build a huge mesh!
+                     cutoff=2, ## Filter away adjacent points.
+                     offset=c(1100, 2000)) ## Offset for extra boundaries, if needed.
+
+ggplot() + gg(mesh) + gg(point_transects, colour = "red")
+
+# save the mesh
+saveRDS(object = mesh, file = "mesh_no_crs.RDS")

### 2019-07-22 analyses/eval_model.R

@@ -1,0 +1,89 @@
+# Evaluate fitted model
+library(INLA)
+library(inlabru)
+
+#### Fitted model  ####
+file_name = "akepa_2002_fitted_model.RDS"
+fit = readRDS(file_name)
+
+#### Specify detection functions  ####
+
+# Log half-normal
+log_hn = function(distance, lsig){ 
+  -0.5*(distance/exp(lsig))^2
+}
+
+# Half-normal
+hn <- function(distance, lsig) exp(log_hn(distance, lsig))
+
+# Include r for switching to polar coords
+dsamp = function(distance, lsig){ 
+  log(distance) + log_hn(distance, lsig)
+}
+
+
+#### Model evaluation ####
+
+summary(fit)
+
+# half-normal
+
+W = 60   # transect radius
+
+hnpred <- predict(fit, distdf, ~hn(distance,lsig))
+ggplot() + 
+  gg(hnpred)
+
+ggsave(filename = "../figures/halfnormal.png")
+
+# half-normal polar coords
+distdf <- data.frame(distance = seq(1,W,length=100))
+dfun <- predict(fit, distdf, ~ dsamp(distance,lsig))
+ggplot() + 
+  gg(dfun)
+
+# intensity
+pxl = pixels(mesh, nx = 100, ny = 100, mask = study_area)
+pr.int <- predict(fit, pxl, ~ exp(grf + Intercept))
+
+ggplot() + 
+  gg(study_area) + 
+  gg(pr.int) +
+  scale_fill_viridis_c()
+
+ggsave(filename = "../figures/intensity_mean.png")
+
+ggplot() +
+  gg(study_area)  +
+  gg(pr.int["cv"]) +
+  gg(samplers, colour = "red") +
+  scale_fill_viridis_c()
+
+ggsave(filename = "../figures/intensity_cv.png")
+
+# GMRF params
+spde.range = spde.posterior(fit, "grf", what = "range")
+spde.logvar = spde.posterior(fit, "grf", what = "log.variance")
+
+plot(spde.range)
+ggsave(filename = "../figures/range_posterior.png")
+
+plot(spde.logvar)
+ggsave(filename = "../figures/spde_logvar_posterior.png")
+
+# Abundance posterior
+source("AS.Nposterior.R") 
+predpts = ipoints(study_area, mesh)
+abnc = AS.Nposterior(fit, predpts)   # n.samples = 1000 default
+
[…]

### 2019-07-22 analyses/fit_model.R

@@ -1,0 +1,90 @@
+## Fit model and save model object
+
+# Load required packages
+library(inlabru)
+library(INLA)  # need testing version of inla
+# library(rgeos)
+# library(raster)  # raster::crs() easy way to remove CRS information
+# library(maptools)
+
+init.tutorial()
+
+#### Save Fitted Model Object? ####
+
+#### WARNING!!!! THIS WILL OVERWRITE EXISTING OBJECT!!!!
+to_save = TRUE
+file_name = "akepa_2002_fitted_model.RDS"
+
+
+#### Load Data ####
+
+realobs <- readRDS("data/obs_2002_no_crs.RDS")
+samplers <- readRDS("data/samplers_no_crs.RDS")
+study_area <- readRDS("data/study_area_no_crs.RDS")
+mesh <- readRDS("data/mesh_no_crs.RDS")
+
+#### Specify detection function  ####
+
+# Log half-normal
+log_hn = function(distance, lsig){ 
+  -0.5*(distance/exp(lsig))^2
+}
+
+# Half-normal
+hn <- function(distance, lsig) exp(log_hn(distance, lsig))
+
+# Include r for switching to polar coords
+dsamp = function(distance, lsig){ 
+  log(distance) + log_hn(distance, lsig)
+}
+
+
+# Plot the data
+g1 <- ggplot() +
+  gg(mesh) + 
+  gg(study_area) +
+  gg(samplers) +
+  gg(realobs, colour = "green")
+
+g1
+
+#### Specify and fit model ####
+
+# SPDE:
+# (xrange <- study_area@bbox["x",2] - study_area@bbox["x",1])
+# (yrange <- study_area@bbox["y",2] - study_area@bbox["y",1])
+# sort(gDistance(pts, byid = TRUE), decreasing = TRUE)
+matern <- inla.spde2.pcmatern(mesh, 
+                              prior.sigma = c(2, 0.01),    
+                              prior.range = c(130, 0.01))
+
+cmp <- ~ grf(map = coordinates, model = matern) + 
+  lsig + Intercept
+
+# Predictor formula:
+fml <- coordinates + distance ~ grf +
+  dsamp(distance, lsig) + 
+  log(2*pi) +   # 2*pi offset for not knowing angle theta
+  Intercept
+
+
+starting_values <- data.frame(grf = 0, lsig = 3.6, Intercept = 0)
+
+W <- 60   # transect radius
[…]

### 2019-07-22 analyses/AS.Nposterior.R
fix n.samples

@@ -12,14 +12,15 @@ AS.Nposterior <- function(model, predpts, ...){
   Lambda <- predict(model, predpts, ~ sum(weight * exp(grf + Intercept))) 
   Lambda
   
-  Nlow <- round(Lambda$mean - 4*Lambda$sd)
-  Nhigh <- round(Lambda$mean + 4*Lambda$sd)
+  Nlow <- round(Lambda$mean - 3*Lambda$sd)
+  Nhigh <- round(Lambda$mean + 3*Lambda$sd)
   by <- round((Nhigh - Nlow)/500)   # should give roughly 500 
   
   Nrange <- seq(Nlow, Nhigh, by = by)
   Npost = predict(model, predpts, 
                  ~ data.frame(N = Nrange, 
-                              dpois(Nrange, lambda = sum(weight * exp(grf + Intercept)))))
+                              dpois(Nrange, lambda = sum(weight * exp(grf + Intercept)))),
+                 ...)
   
   return(list(Npost = Npost, Nexp = Lambda))
 }

### 2019-07-22 analyses/sandpit/Nposterior.R
investigate N posterior

@@ -1,0 +1,27 @@
+# Evaluate fitted model
+library(INLA)
+library(inlabru)
+
+#### Fitted model  ####
+file_name = "../akepa_2002_fitted_model.RDS"
+fit = readRDS(file_name)
+
+source("AS.Nposterior.R") 
+predpts = ipoints(study_area, mesh)
+
+# system.time(AS.Nposterior(fit, predpts, n.samples = 4000))
+# abnc = AS.Nposterior(fit, predpts, n.samples = 1000)   # n.samples = 1000 default
+
+n.samples.list = c(1000, 2000, 4000, 8000, 16000)
+
+for (n in n.samples.list){
+  abnc = AS.Nposterior(fit, predpts, n.samples = n)   # n.samples = 1000 default
+  g = ggplot() + 
+    gg(abnc$Npost) +
+    ggtitle(paste("n = ", n)) +
+    xlim(3000, 9000)
+  print(g)
+}
+
+
+  

### 2019-08-01 analyses/AS.Nposterior.R
add arguments to Nposterior function

@@ -4,17 +4,18 @@
 #' @param model  fitted inlabru model
 #' @param predpts the thing that ipoints() returns, or something similar, must have weight column
 #' @param ... other args passed on to predict()
-#' 
+#' @param n.sd how many standard deviations above and below the mean to consider (for domain of N)
+#' @param n.grid roughly the number of discrete N values to consider
 
-AS.Nposterior <- function(model, predpts, ...){
+AS.Nposterior <- function(model, predpts, n.sd = 3, n.grid = 100, ...){
   
   # First get posterior for 
   Lambda <- predict(model, predpts, ~ sum(weight * exp(grf + Intercept))) 
   Lambda
   
-  Nlow <- round(Lambda$mean - 3*Lambda$sd)
-  Nhigh <- round(Lambda$mean + 3*Lambda$sd)
-  by <- round((Nhigh - Nlow)/500)   # should give roughly 500 
+  Nlow <- round(Lambda$mean - n.sd*Lambda$sd)
+  Nhigh <- round(Lambda$mean + n.sd*Lambda$sd)
+  by <- round((Nhigh - Nlow)/n.grid)   # should give roughly n.grid 
   
   Nrange <- seq(Nlow, Nhigh, by = by)
   Npost = predict(model, predpts, 

### 2019-08-01 analyses/eval_model.R
set plot dimensions

@@ -2,6 +2,7 @@
 library(INLA)
 library(inlabru)
 
+set.seed(9701071)
 
 #### WARNING - as it stands this script will overwrite figures for the paper
 
@@ -36,18 +37,21 @@ summary(fit)
 # half-normal
 
 W = 60   # transect radius
+distdf <- data.frame(distance = seq(1,W,length=100))
 
 hnpred <- predict(fit, distdf, ~hn(distance,lsig))
 ggplot() + 
-  gg(hnpred)
+  gg(hnpred) +
+  ggtitle("Posterior half-normal detection function")
 
-ggsave(filename = "../figures/halfnormal.png")
+ggsave(filename = "../figures/halfnormal.png",
+       width = 5, height = 5, units = "in")
 
-# half-normal polar coords
-distdf <- data.frame(distance = seq(1,W,length=100))
-dfun <- predict(fit, distdf, ~ dsamp(distance,lsig))
-ggplot() + 
-  gg(dfun)
+# half-normal polar coords?  Is that what this is?  
+# distdf <- data.frame(distance = seq(1,W,length=100))
+# dfun <- predict(fit, distdf, ~ dsamp(distance,lsig))
+# ggplot() + 
+#   gg(dfun)
 
 # intensity
 pxl = pixels(mesh, nx = 100, ny = 100, mask = study_area)
@@ -56,41 +60,131 @@ pr.int <- predict(fit, pxl, ~ exp(grf + Intercept))
 ggplot() + 
   gg(study_area) + 
   gg(pr.int) +
-  scale_fill_viridis_c()
+  scale_fill_viridis_c() +
+  ggtitle("Posterior mean intensity field")
 
-ggsave(filename = "../figures/intensity_mean.png")
+ggsave(filename = "../figures/intensity_mean.png",
+       width = 5, height = 6, units = "in")
 
 ggplot() +
   gg(study_area)  +
   gg(pr.int["cv"]) +
   gg(samplers, colour = "red") +
-  scale_fill_viridis_c()
+  scale_fill_viridis_c() +
+  ggtitle("Posterior CV of intensity field")
 
-ggsave(filename = "../figures/intensity_cv.png")
+ggsave(filename = "../figures/intensity_cv.png",
+       width = 5, height = 6, units = "in")
+
+# lower and upper quantiles
+
+v = c(pr.int["mean"]@data,
[…]

### 2019-08-01 analyses/excursions.R

@@ -3,6 +3,7 @@ library(INLA)
 library(inlabru)
 library(excursions)
 
+set.seed(1525)
 
 #### Fitted model  ####
 file_name = "akepa_2002_fitted_model.RDS"
@@ -19,7 +20,7 @@ mesh = readRDS("data/mesh_no_crs.RDS")
 # There must be a way to get the actual values from
 # predict with n.samples > 1 ?
 
-n.mc = 500
+n.mc = 500     # lower this when fiddling with things
 pxl = pixels(mesh, nx = 100, ny = 100, mask = study_area)
 X = matrix(NA, nrow = nrow(pxl@coords), ncol = n.mc)
 for (i in 1:n.mc){
@@ -37,7 +38,7 @@ Xkm = X*1000000
 Xhec = X*10000
 
 # per hectare threshold:
-u = 0.5
+u = 1
 alpha = 0.05    # Prob(intensity > u) >= 1 - alpha for all s in E
 
 ex = excursions.mc(Xhec, u = u, type = ">", alpha = alpha)
@@ -45,19 +46,36 @@ ex = excursions.mc(Xhec, u = u, type = ">", alpha = alpha)
 # plot
 Fpxl = pxl
 Fpxl$F = ex$F
+
+# png(filename = "../figures/excursion_function.png",
+#     width = 4.5, height = 5.5, units = "cm", res = 1000)
 ggplot() + 
   gg(Fpxl) +
-  scale_fill_viridis_c()
+  scale_fill_viridis_c() +
+  ggtitle("Excursion function for > 1 bird per hectare")
+
+ggsave(filename = "../figures/excursion_function.png",
+       width = 5, height = 6, units = "in")
 
 Epxl = pxl
 Epxl$E = ex$E
 ggplot() +
-  gg(Epxl)
+  gg(Epxl) +
+  ggtitle("Excursion set for > 1 bird per hectare,\nalpha = 0.05")
+
+ggsave(filename = "../figures/excursion_set.png",
+       width = 5, height = 6, units = "in")
 
 # contour map - how to visualise this?  ?
 cm = contourmap.mc(Xhec, levels = c(0.1, 0.5, 0.8, 1.5, 2, 4), alpha = alpha)
 str(cm)
 
+Gpxl = pxl
+Gpxl$G = cm$G
+ggplot() +
+  gg(Gpxl)
+
+cm = contourmap.mc(Xhec, levels = c(1), alpha = alpha)
 Gpxl = pxl
 Gpxl$G = cm$G
 ggplot() +

### 2019-08-01 analyses/eval_model.R
more plot tweaking

@@ -57,24 +57,23 @@ ggsave(filename = "../figures/halfnormal.png",
 pxl = pixels(mesh, nx = 100, ny = 100, mask = study_area)
 pr.int <- predict(fit, pxl, ~ exp(grf + Intercept))
 
-ggplot() + 
-  gg(study_area) + 
+p1 = ggplot() + 
+  gg(study_area) +
   gg(pr.int) +
   scale_fill_viridis_c() +
-  ggtitle("Posterior mean intensity field")
+  ggtitle("Posterior mean intensity")
 
-ggsave(filename = "../figures/intensity_mean.png",
-       width = 5, height = 6, units = "in")
-
-ggplot() +
+p2 = ggplot() +
   gg(study_area)  +
   gg(pr.int["cv"]) +
   gg(samplers, colour = "red") +
   scale_fill_viridis_c() +
-  ggtitle("Posterior CV of intensity field")
+  ggtitle("Posterior CV of intensity")
 
-ggsave(filename = "../figures/intensity_cv.png",
-       width = 5, height = 6, units = "in")
+png(filename = "../figures/intensity_mean_cv.png",
+    width = 10, height = 6, units = "in", res = 100)
+multiplot(p1, p2, cols = 2)
+dev.off()
 
 # lower and upper quantiles
 

### 2019-08-01 paper.tex
add images and update todo list
flesh out akepa case study

@@
 \doublespacing
+
+\usepackage{hyperref}
 
@@
 \usepackage{amsmath}
+\usepackage{caption}
 
@@
 \newcommand{\tL}{\tilde{\Lambda}}  % thinned big lambda
+
+\DeclareMathOperator*{\argmax}{arg\,max}  % * means _ puts thing beneath operator
 
@@
 This is similar to a Poisson likelihood with an offset term equal to $\log\alpha$.
-[[Cite ``Going off Grid" and ``Berman Turner" papers here]]
-[[ I don't understand how INLA/inlabru handles this specific likelihood form.
-The data corresponding to $\exp(-\alpha_i\eta_i)$ is straightforward using the $E$ argument with ``family = poisson".
-The $\eta_i^{y_i}$ part sets $E = 0$ but I don't see anything that would set $\exp(-\alpha_i\eta_i) =1$ for this part of the stack which is what we need to get back to \eqref{approx-log-likelihood}.
 
-This all differs from Berman-Turner/Baddeley papers in one key respect - the observed points do not form part of the integration scheme.
-For Berman-Turner this is necessary so we can define pseudo-data $\tilde{y}_i = y_i / \alpha_i$.
-Since all points are in the integration scheme $\alpha_i \neq 0$.
-This is the trick they use to get it into a completely Poisson form (although now the pseudo-data is non-integer).]]
 \section*{Intensities for incomplete data}
@@
 
-TO DO:
+Section outline:
+\begin{enumerate}
+	\item Introduce the Akepa data, survey method, conservation status, conservation approaches, questions managers have.
+	\item Example code snippets for how to specify the model in inlabru
+	\item This includes discussion of priors.
+Reference PC-priors, short description of how they work and why they are important for spatial modelling with Gaussian random fields.
+In this case:  the prior for the range parameter was chosen based on the smallest distance between sampled locations.
+The prior for sigma was based on a plausible scale for the maximum intensity.
+	\item Show main results:  (i) The posterior mean of intensity (ii) The CV of the posterior intensity (iii)  The expected abundance
+	\item Discussion of difference between \textit{expected} abundance and \textit{realized abundance}.
+Show how we create a posterior for realized abundance by monte-carlo sampling and averaging of Poisson probability mass functions.
+Highlight that this is not often considered and is a natural thing to do in Bayesian setting where it is possible to draw monte-carlo samples from the posterior.
+Plots of posterior for expected and realized abundance.
+Note extra variation in realized abundance over expected abundance.
+	\item Plot lower 0.025 quantile and upper 0.975 quantile of intensity.
+Note the problem of multiple hypothesis testing and that these maps may be misleading.
+Describe the excursion sets approach to solving this problem.
+	\item Show plots of estimated excursion sets for various thresholds and emphasise the usefulness from a management point of view.
+Emphasise the \textit{joint} probability of every location in the set exceeding a threshold.
+	\item A small comparison with the GAM approach (perhaps in an appendix?) to show the similarity for the mean intensity.  Compare the uncertainty.  Reassuring that the methods agree.    
+\end{enumerate}
 
-\begin{itemize}
-\item Bird survey locations image
-\item This section to include code snippets from inlabru
-\item Also mentions principled choice of priors and our choice for this akepa example
-\item Image of posterior mean of intensity
-\item Image of CV of posterior intensity
-\item Talk to Rick about something interesting to do with excursions package?  Multiple-testing an issue with gridded data as well - how is this solved?
-\end{itemize}
+\newpage
+
+TODO HERE:  Description of Akepa data, survey method, key questions, what do we care about from ecology/conservation point of view.
+Then, code snippets of how to specify the model in inlabru, brief discussion of PC priors and how they were chosen here
+
+\bigskip
+
+The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.
+This shows a region of high intensity in the south and much lower intensity in the north.
+The CV plot shows clearly that there is less variation in the posterior intensity field in areas with greater sampling effort.
+In general the south, with greater density of sampled locations, has a lower CV than the north, where sampling units are more spread out.
+The areas of highest CV are those regions furthest away from any sampled location, reflecting the intuitive fact that we are least certain in areas where we did not look.
+
+\begin{figure}
+	\includegraphics[scale=0.5]{figures/intensity_mean_cv.png}
+	\caption{(left) Posterior mean of the intensity (right) Posterior coefficient of variation of the intensity}
+	\label{fig:intensity-mean-cv}
+\end{figure}
+
+The CV plot shows a measure of uncertainty relative to the mean intensity in each prediction grid cell.
+However, often we are interested in the uncertainty around the mean intensity on the scale of the intensity function itself - perhaps seeking to identify areas where we can confidently predict particularly high or low intensities.
+To attempt to communicate this uncertainty around the mean it is common [[references?]] to present the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
+
+\begin{figure}
+	\includegraphics[scale=0.5]{figures/intensity_quantiles.png}
+	\caption{The predicted posterior intensity summarised in three plots by the mean (upper),  0.025 quantile (lower left) and 0.975 quantile (lower right) for each prediction grid cell.
+All three plots use a common colour scale}
+	\label{fig:intensity-quantiles}
+\end{figure}
+
+We believe quantile plots like those in \autoref{fig:intensity-quantiles} to be challenging to interpret and potentially misleading.
+The temptation is to interpet these plots as showing possible intensity surfaces that produced the observed data.
+We believe this misconception would be extremely common in particular for non-statistically trained audiences.
+The maps in \autoref{fig:intensity-quantiles} in fact only show \textit{summary statistics} of a posterior intensity which is specified (in our model) as a log-Gaussian random field.
+
+This difference becomes clear if one generates realizations of the posterior intensity field.
+Three such realizations are in \autoref{fig:intensity-realizations}.
+There are some notable differences when compared to the mean and quantile summary maps.
+First, each realization has finer-grained spatial structure than the mean of the field.
+What this means in practice is that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.
+The mean field plot hides the spatial structure the model has been able to estimate from the data.
+In general, the mean of a random field will contain less spatial structure than realisations of the field.
+
+\begin{figure}
+	\includegraphics[scale=0.35]{figures/intensity_realized.png}
+	\caption{Three realizations of the posterior intensity field}
+	\label{fig:intensity-realizations}
+\end{figure}
+
+The quantile plots are more concerning.
+As mentioned above, the temptation is to interpret these as possible intensities that produced our observed data.
+However, they in fact show the quantiles of the \textit{marginal} probability density for each prediction location.
+This essentially treats each prediction cell as independent of the others.
+Whilst any one prediction location may be well described by such quantiles, to present them side by side in one image obscures the fact that the probability that all (or some subset of) the prediction cells would \textit{simultaneously} achieve that value will be in fact much lower than the quantile value and in many cases would be essentially zero.
+To demonstrate this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 2500 for the 0.025 quantile and 12,300 for the 0.975.
+A naive (and tempting) interpretation of these numbers is as upper and lower limits of the 95\% credible interval for abundance.
+However, these abundance estimates are essentially impossible according to the estimated posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
+This emphasises that the temptation to treat these maps as showing possible intensity surfaces leads to outcomes (and possible management decisions) that are inconsistent with the very model that generated the maps.
+
+As an alternative we suggest using excursion sets and excursion functions.  This is a recently approach to representing uncertainty in stochastic processes that specifically considers
+joint probability of events across multiple locations.
+The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) < u \}$.
+For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
+
+\begin{equation*}
+E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \}
+\end{equation*}
+It is important to note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.
+This means that when considering random quantities such as $\lambda(s)$ the definition of the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is finding the largest such set for which such a condition holds simultaneously for all locations in a region.
+Because $\lambda(s)$ is stochastic we must use something akin to a confidence level by specifying $\alpha$.
+
+Excursion sets can be estimated using...  [[summarise Bolin paper here]]
+
+\autoref{fig:excursion-set} shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95.
+This figure can be interpreted in a natural way.
+It is the largest possible region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
+
+\begin{figure}
+	\includegraphics[scale=0.5]{figures/excursion_set.png}
+	\caption{The positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95}
+	\label{fig:excursion-set}
+\end{figure}
+
+To visualise multiple such maps we can use the excursion function $F_u^{+}(s) = \sup \{1 - \alpha ; s \in E_{u,\alpha}^+ \}$, which defines for each location something similar to a p-value.
+For each $s$ it is the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.
+The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursion-function}.
+
+\begin{figure}
+	\includegraphics[scale=0.5]{figures/excursion_function.png}
+	\caption{The positive excursion function with a level corresponding to 1 bird per hectare}
+	\label{fig:excursion-function}
+\end{figure}
+
+This figure can also be interpreted naturally, defining probabilities each location to be a member of a set for which the probability of exceeding a threshold simultaneously in all locations.
+A key feature of these plots is that they display either discrete sub-regions of the area of interest (for excursion sets) or the probability of membership of such discrete sub-regions (for excursion functions).
+We believe the risk of interpreting these maps as intensities is much lower than for the quantile plots.
+
+Another key aim of performing distance sampling surveys is to estimate the abundance of the population of interest.
+
+\begin{figure}
+	\includegraphics[scale=0.6]{figures/realized_abundance_posterior.png}
+	\caption{Posterior of realized abundance}
+	\label{fig:realized-abundance-posterior}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[scale=0.6]{figures/realized_abundance_vs_exp.png}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[scale=0.6]{figures/halfnormal.png}
+\end{figure}
+
+\newpage
 
@@
 \begin{enumerate}
+	\item Note the model-based Bayesian hierarchical approach provides a framework for working with distance sampling data.
+This allows simultaneous estimation of parameters associated with the detection process and the spatial distribution.
+It is in principle extendible...
+somehow...
+	\item Note that Bayesian distance sampling has been done before but that the INLA approach avoids data augmentation and MCMC.  
 	\item Iterated INLA is a powerful and flexible way to include non-linear components to a predictor and has applications beyond this application to distance sampling

### 2019-08-01 paper.tex
case study getting there

@@
 
-Excursion sets can be estimated using...  [[summarise Bolin paper here]]
+Excursion sets can be estimated by considering increasing candidate sets $D$ and a sequential integration scheme to estimate probabilities.
+An implementation is available in the \texttt{excursions} package available through the Comprehensive R Archive Network.
 
@@
 Another key aim of performing distance sampling surveys is to estimate the abundance of the population of interest.
+By specifying our spatial model as a log-Gaussian Cox process, the intensity is a $random field$.
+Let $n$ denote the abundance of the population within the region of interest and $N$ its corresponding random variable within our Bayesian framework.
+Integrating the mean of the random field provides a point estimate for the abundance, which we term the expected abundance (where the term expected comes from the fact this value is the rate parameter in the Poisson distribution which $N$ follows).
+Here we demonstrate the importance of considering the full posterior for $N$ which we call the \textit{realized} abundance.
+
+Let $\Lambda = \int_{\Omega}\lambda(s)\mathrm{d}s$.
+Note that $\Lambda$ is a random variable and the integral of $\mathbb{E}[\lambda(s)]$ is equivalent to $\mathbb{E}(\Lambda)$ by Fubini's theorem.
+As we mentioned above, this is the most sensible point estimate for the abundance.
+We estimate the posterior $\pi(N | \Lambda)$ by a monte-carlo method.
+Taking $m$ monte-carlo samples of the posterior intensity field $\lambda_1, \ldots, \lambda_m$ with corresponding expected abundance estimates $\Lambda_i = \int \lambda_i(s)\mathrm{d}s$ we estimate the posterior for the abundance as $\pi(N | \Lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \Lambda = \Lambda_i)$.
+This is well defined since $\pi(N | \Lambda = \Lambda_i) = \text{Poisson}(N, \Lambda_i)$.
+\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
+
+For reference \autoref{fig:realized-abundance-comparison} shows a single Poisson PMF with rate parameter equal to the expected abundance.
 
@@
 	\includegraphics[scale=0.6]{figures/realized_abundance_posterior.png}
-	\caption{Posterior of realized abundance}
+	\caption{Posterior of realized abundance.  The blue line marks the expected abundance}
 	\label{fig:realized-abundance-posterior}
@@
 	\includegraphics[scale=0.6]{figures/realized_abundance_vs_exp.png}
+	\caption{Solid:  Posterior of realized abundance.  Dashed:  Poisson PMF with rate parameter $\mathbb{E}[\Lambda]$}
+	\label{fig:realized-abundance-comparison}
 \end{figure}
@@
 \end{figure}
+
+\bigskip
+
+TODO HERE:  Something about how this is all a one-stage model and all these plots are based on sampling from the joint posterior of all parameters.
+As such all plots can be viewed as incorporating the uncertainty from the detection model.
+This is a lovely and natural way to present results and an advantage of having a joint model.
+I'm not sure varprop can do this?
 

### 2019-08-01 paper.tex
tidy up a bit to send to supervisors

@@
 
+\newpage
+
 \section*{Case-Study:  Hawaiian Akepa Survey}
-
-Section outline:
-\begin{enumerate}
-	\item Introduce the Akepa data, survey method, conservation status, conservation approaches, questions managers have.
-	\item Example code snippets for how to specify the model in inlabru
-	\item This includes discussion of priors.
-Reference PC-priors, short description of how they work and why they are important for spatial modelling with Gaussian random fields.
-In this case:  the prior for the range parameter was chosen based on the smallest distance between sampled locations.
-The prior for sigma was based on a plausible scale for the maximum intensity.
-	\item Show main results:  (i) The posterior mean of intensity (ii) The CV of the posterior intensity (iii)  The expected abundance
-	\item Discussion of difference between \textit{expected} abundance and \textit{realized abundance}.
-Show how we create a posterior for realized abundance by monte-carlo sampling and averaging of Poisson probability mass functions.
-Highlight that this is not often considered and is a natural thing to do in Bayesian setting where it is possible to draw monte-carlo samples from the posterior.
-Plots of posterior for expected and realized abundance.
-Note extra variation in realized abundance over expected abundance.
-	\item Plot lower 0.025 quantile and upper 0.975 quantile of intensity.
-Note the problem of multiple hypothesis testing and that these maps may be misleading.
-Describe the excursion sets approach to solving this problem.
-	\item Show plots of estimated excursion sets for various thresholds and emphasise the usefulness from a management point of view.
-Emphasise the \textit{joint} probability of every location in the set exceeding a threshold.
-	\item A small comparison with the GAM approach (perhaps in an appendix?) to show the similarity for the mean intensity.  Compare the uncertainty.  Reassuring that the methods agree.    
-\end{enumerate}
-
-\newpage
 
@@
 It is important to note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.
-This means that when considering random quantities such as $\lambda(s)$ the definition of the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is finding the largest such set for which such a condition holds simultaneously for all locations in a region.
-Because $\lambda(s)$ is stochastic we must use something akin to a confidence level by specifying $\alpha$.
+This means that when considering random quantities such as $\lambda(s)$ the definition of the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.
+Because $\lambda(s)$ is stochastic we can only ever say this is true with some probability and use something akin to a confidence level by specifying $\alpha$.
 
@@
 
-This figure can also be interpreted naturally, defining probabilities each location to be a member of a set for which the probability of exceeding a threshold simultaneously in all locations.
+This figure can also be interpreted naturally.
+It shows the largest possible probability for each location to be a member of a set for in which the intensity exceeds a threshold simultaneously across all locations in the set.
+It can be viewed intuitively as a kind of fuzzier version of \autoref{fig:excursion-set}.
+Regions on the edge of the excursion set would be included if the $\alpha$ value were allowed to increase slightly.
+However, for regions in the north there is essentially no probability level for which those locations would be included in $A_u^{+}(\lambda)$.
+
   A key feature of these plots is that they display either discrete sub-regions of the area of interest (for excursion sets) or the probability of membership of such discrete sub-regions (for excursion functions).
 We believe the risk of interpreting these maps as intensities is much lower than for the quantile plots.
+The fact that it is possible to estimate discrete regions may also be useful to communicate results to conservation managers and decision-makers.
+We anticipate the excursion function will be most useful in communicating to audiences with statistical training.
 
 Another key aim of performing distance sampling surveys is to estimate the abundance of the population of interest.
-By specifying our spatial model as a log-Gaussian Cox process, the intensity is a $random field$.
+By specifying our spatial model as a log-Gaussian Cox process, the intensity is a random field rather than a fixed function of space.
 Let $n$ denote the abundance of the population within the region of interest and $N$ its corresponding random variable within our Bayesian framework.
-Integrating the mean of the random field provides a point estimate for the abundance, which we term the expected abundance (where the term expected comes from the fact this value is the rate parameter in the Poisson distribution which $N$ follows).
+Integrating the mean of the random field provides a point estimate for the abundance, which we term the expected abundance.
 Here we demonstrate the importance of considering the full posterior for $N$ which we call the \textit{realized} abundance.
 
-Let $\Lambda = \int_{\Omega}\lambda(s)\mathrm{d}s$.
-Note that $\Lambda$ is a random variable and the integral of $\mathbb{E}[\lambda(s)]$ is equivalent to $\mathbb{E}(\Lambda)$ by Fubini's theorem.
+Let $N = \int_{\Omega}\lambda(s)\mathrm{d}s$.
+Note that $N$ is a random variable and the integral of $\mathbb{E}[\lambda(s)]$ is equivalent to $\mathbb{E}(N)$ by Fubini's theorem, hence the term expected abundance.
 As we mentioned above, this is the most sensible point estimate for the abundance.
-We estimate the posterior $\pi(N | \Lambda)$ by a monte-carlo method.
-Taking $m$ monte-carlo samples of the posterior intensity field $\lambda_1, \ldots, \lambda_m$ with corresponding expected abundance estimates $\Lambda_i = \int \lambda_i(s)\mathrm{d}s$ we estimate the posterior for the abundance as $\pi(N | \Lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \Lambda = \Lambda_i)$.
-This is well defined since $\pi(N | \Lambda = \Lambda_i) = \text{Poisson}(N, \Lambda_i)$.
+We estimate the posterior $\pi(N | \lambda)$ by a monte-carlo method.
+Taking $m$ monte-carlo samples of the posterior intensity field $\lambda_1, \ldots, \lambda_m$ we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
+This is well defined since $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
 \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
-
-For reference \autoref{fig:realized-abundance-comparison} shows a single Poisson PMF with rate parameter equal to the expected abundance.
+This allows us to estimate the probability for any specific value of \textit{realized} abundance $n$ by calculating $\pi(N = n | \lambda)$.
+For reference, \autoref{fig:realized-abundance-comparison} shows a single Poisson PMF with rate parameter equal to the expected abundance.
 

### 2019-08-09 paper.tex
start working on introduction

@@
 
-\usepackage{hyperref}
+\usepackage{hyperref} % auto detect ref type
+\usepackage{natbib}
+\setcitestyle{authoryear}
 
-% remove preprent footnote
+% remove preprint footnote
 \makeatletter
@@
 \end{enumerate}
+
+The estimation of the size of wild populations of animals is a critical objective within ecology and conservation [[citation]].
+Distance sampling methods aim to achieve this by using a spatially explicit sampling design and estimating the detectability of animals as a function of distance from observer \citep{buckland_advanced_2008, buckland_distance_2015}.
+Standard distance sampling approaches use a hybrid of design and model-based inference to estimate population size.
+Probability of detection is modelled as a function of distance from observer and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.
+
+Over time interest grew in fully model-based approaches which allow the use of non-randomized survey designs by considering a spatial model for animal density so observations can be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, buckland_model-based_2016, miller_spatial_2013}.
+Model-based distance sampling has typically been implemented in a two-stage modelling process.
+In the first stage, detectability is estimated within each sampling unit.
+In the second stage, detections within sampling units are binned into counts and used as the response variable in generalized additive model framework where the detectability estimates from the first stage are used as an offset.
+Due to often sparse nature of wildlife survey data this may require consideration of over or under-dispersion and zero-inflation.
+Negative-binomial and Tweedie distributions are common choices to model the distribution of counts [[citation?]].
+A key concern with the two-stage approach is the propogation of the uncertainty in the detectability estimates to the second-stage spatial model.
+Early attempts to address this focused on bootstrapping [[citations]], but more recent work has pointed to potentially irresolvable difficulties  [[flesh out]] with the bootstrapping approach and instead proposed propagating error based on a second-order Taylor approximation of detectability around the maximum-likelihood estimate from the first stage \citep{bravington_reliable_2018-1}.
+
+\citet{yuan_point_2017} presented a one-stage approach for analysing line-transect distance sampling within a Bayesian framework that simulateously estimates the detectability 
+and spatial distribution of animals.
+Key to their approach is the use of sparse Gaussian Markov random field (GMRF) approximations to continously indexed Gaussian random fields (GRF) and a log-Gaussian Cox point process model that avoids the need to bin data into counts and naturally sidesteps issues of zero-inflation and dispersion.
+The observation process is formulated as a thinning of the log-Gaussian Cox process.
+Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.
+The model was specified as a latent Gaussian model and fitted using integrated nested Laplace approximations using the \texttt{R-INLA} package \citep{rue_approximate_2009}.
+
+Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.
+Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
+Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data as a thinned point process.
+However, the point process viewpoint is not new and has been applied numerous times to line transect data [[citations here]].
 
@@
 
+\bibliographystyle{chicago}
+\bibliography{paper}
+
 \end{document}

### 2019-08-09 paper.tex
Update paper.tex
Added shortcuts, Case Study text and links to figures.

@@
 \newcommand{\tL}{\tilde{\Lambda}}  % thinned big lambda
+% RJC 09/08/2019 Added shortcuts for Hawaiian words
+\newcommand{\akepa}{\textquotesingle\={a}kepa}  % adds Hawaiian diacritical marks
+\newcommand{\Akepa}{\textquotesingle\={A}kepa}  % adds Hawaiian diacritical marks
+\newcommand{\hawaii}{Hawai\textquotesingle i}   % adds Hawaiian diacritical marks
 
@@
 %% \address[label1]{<address>}
-%% \address[label2]{<address>}
+%% \address[label,2]{<address>}
 
-\author{Andrew E Seaton}
+% RJC 09/08/2019 Added co-authors and affiliations. Note that I used \affil[]{} instead of \address[]{}
+\author[1,*]{Andrew E Seaton}
+\author[1,2]{Richard J Camp}
+\author[1]{Janine B Illian}
+\author[1]{David L Borchers}
+\author[1]{David L Miller}      % Ask about co-authoring the manuscript
+\author[1]{Len Thomas}          % Ask about co-authoring the manuscript
+\author[1]{Stephen T Buckland}  % Ask about co-authoring the manuscript
+\author[3]{Steve J Kendall}     % It is Steve's data we are using.
 
-\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}
+%\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}
+\affil[1]{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}
+\affil[2]{U. S. Geological Survey, Pacific Island Ecosystems Research Center, P.O. Box 44, \hawaii{} National Park, HI 96718, U.S.A.}
+\affil[3]{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}
+\affil[*]{Correspondence: Andrew E Seaton, Email: aes22@st-andrews.ac.uk}
 
@@
 
+% RJC 09/08/2019 Added text describing the akepa data, survey method, key questions and what we care about from ecology/conservation point of view.
+This text provides a background on Hawaiian forest birds, why it is important to monitor their populations and the need for unbiased and precise estimates.
+I am wondering if we need to strengthen the 'key questions' text.
+
 \bigskip
+
+% RJC 09/08/2019 Added case study text between % addtextstart and % addtextend markers.
+% addtextstart
+In this section, we present a description of Hawaiian forest birds, forest bird monitoring, the study species, study area, bird sampling, and description of the data.
+\hawaii{} is home to the largest adaptive radiation of oceanic archipelago birds worldwide.
+The Hawaiian Honeycreepers (Drepanidinae) is comprised of 21 genera and more than 50 species generally grouped according to feeding niches with many species and subspecies being island endemics.
+The honeycreepers, like most insular bird species, evolved life history strategies that include greater longevity with high adult survival, delayed maturation with an extended post-fledging dependency period, decreased clutch size and fecundity, and reduced dispersal abilities with high site fidelity \citep{woodworth_demography_2009}.
+While the honeycreepers diversified into the plethora of ecological niches they also had a loss or reduction of defences against competitors, predators and diseases.
+The loss of defensive adaptations made the honeycreepers particularly vulnerable to humans and the changes and threats they brought.
+As a result, more than half of the honeycreepers are extinct \citep{banko_extinctions_2009}.
+A mere 23 honeycreeper species persist with most species restricted to minute portions of their original ranges and in small numbers.
+The IUCN lists 21 honeycreeper species as critically endangered, endangered or vulnerable, and only two species are categorized as least concern \citep{iucn_redlist_2018}.
+
+In the late $19^{\textnormal{th}}$ century naturalists to \hawaii{} described the Hawaiian forest birds and documented species distributions and abundances.
+Subsequent naturalists to the islands documented range contractions and population declines of native species while noting expansion in non-native bird species ranges and abundances.
+It was not until standardized and statistically robust sampling techniques to estimate population sizes and associated standard errors were initiated that the extent to which honeycreepers had declined was appreciated.
+Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}.
+Information from the HFBS was used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge that was the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
+Due to the broad-scale coverage and robust design HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.
+
+\subsection*{Study species}
+The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.
+\Akepa{} is a foraging specialist whose diet consists of spiders, insects and psyllids, as well as nectar.
+\Akepa{} uses its cross-bill to probe and open terminal leaf clusters and extracts prey with its tongue.
+During the 20th century \akepa{} declined dramatically due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994},  mosquito-transmitted avian diseases \citep{pratt_avifaunal_1994, atkinson_wildlife_1995}, introduced predators \citep{lepson_akepa_1997}, and food resources competitors \citep{lepson_akepa_1997}.
+\Akepa{} has a global abundance of 16,248 (95\%CI 10,074\textendash25,198) birds that has been restricted to five spatially distinct populations \citep{judge_akepa_2018}.
+Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau; 19$^{\circ}$ 51\textquotesingle N, 155$^{\circ}$ 18\textquotesingle W), supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}.
+Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern.
+Moreover, having unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.
+
+\subsection*{Study area}
+Hakalau was established in 1985 to conserve 15,390-ha of montane forest habitat for native forest birds and rainforest plants.
+Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance.
+Survey points were established along 15 transects following a systematic, random design with points approximately 150-m apart on transects located either 500 or 1,000 m apart.
+Following \cite{camp_population_2010, camp_statespace_2016}, we restrict our study area to the 3,061 ha open-forest stratum of Hakalau.
+The study area is between 1,400 and 1,920 m elevation of montane forest dominated by native \textquotesingle \={o}hi\textquotesingle a (\textit{Metrosideros polymorpha}) and koa (\textit{Acacia koa}) with a mixture of native and non-native understory plants.
+The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.
+To the north the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig.
+\ref{fig:2002studyareapointspt}).
+The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of their study area.
+The west side of the study area is bounded by pasture that is dominated by grass and is unsuitable habitat for \akepa.
+The substrate in the southern portion of the study area is nutrient-rich, old lava flows (11,000 to 64,000 yrs old) whereas the northern portion occurs on new lava flows (5,000 to 10,000 yrs old) that are nutrient depleted \citep{ccp_hakalau_2010}.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\linewidth]{figures/2002studyareapoints_pt}
+	\caption{Study area (light blue) showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.}
+	\label{fig:2002studyareapointspt}
+\end{figure}
+
+\subsection*{Bird sampling}
+Surveys used point-transect methods, a form of distance sampling, where horizontal distances from survey points to individually detected birds are used to model a species-specific detection probability that is used to estimate absolute abundance and variance \citep{buckland_distance_2015}.
+Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength \textgreater Baufort scale 3).
+During 8-min counts trained observers recorded the species, exact distance to the nearest meter and detection type for each bird detected, along with the sampling conditions cloud cover, rain, wind strength, gust strength, and time of day each point was surveyed.
+
+\subsection*{Data description}
+For purposes of our analysis, we selected a single survey from the \akepa{} time series based on broad sampling of the study area and sufficient numbers of detections.
+In 2002, 195 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig.
+\ref{fig:2002studyareapointspt}).
+On 67 points 152 \akepa{} were detected.
+The number of detections ranged from zero to 6 (Fig.
+\ref{fig:2002countspt}).
+These data are used for both the two-stage and one-stage model-based analyses.
+\cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the open-forest study area and the bird surveys.
+
+\begin{figure}
+	\centering
+	\includegraphics[width=0.7\linewidth]{figures/2002counts_pt}
+	\caption{Counts of \akepa{} by point during the 2002 survey.}
+	\label{fig:2002countspt}
+\end{figure}
+
+% addtextend
 

### 2019-08-10 paper.tex
Update paper.tex
Many edits, see email.

@@
 \author[1,2]{Richard J Camp}
+\author[3]{Finn Lindgren}
 \author[1]{Janine B Illian}
@@
 \author[1]{Stephen T Buckland}  % Ask about co-authoring the manuscript
-\author[3]{Steve J Kendall}     % It is Steve's data we are using.
+\author[4]{Steve J Kendall}     % It is Steve's data we are using.
 
@@
 \affil[2]{U. S. Geological Survey, Pacific Island Ecosystems Research Center, P.O. Box 44, \hawaii{} National Park, HI 96718, U.S.A.}
-\affil[3]{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}
+\affil[3]{School of Mathematics, University of Edinburgh, Edinburgh, Scotland}
+\affil[4]{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}
 \affil[*]{Correspondence: Andrew E Seaton, Email: aes22@st-andrews.ac.uk}
@@
 Firstly, to justify the assumption that the true distribution of animals is uniform with respect to distance from the observer.
-Secondly, to construct Horvit-Thompson-like estimators of animal density, which are then converted to abundance estimates for regions of interest
+Secondly, to construct Horvitz-Thompson-like estimators of animal density, which are then converted to abundance estimates for regions of interest
 	\item Interest in the spatial distribution of animals has led to the development of approaches that combine traditional distance sampling methods and spatially explicit generalized additive models
 	\item In this approach the estimated detection probabilities for discrete spatial units are used as an offset in a generalized additive model where the response variable is a count of detections within each unit
-	\item This two-stage approach needs a way to propogate uncertainty from the detection model to the spatial model
+	\item This two-stage approach needs a way to propagate uncertainty from the detection model to the spatial model
 	\item Recently [[JOYCE PAPER]] presented a one-stage approach that simultaneously estimates the detection and spatial processes, avoids binning data into counts and removes the need to consider uncertainty propagation
@@
 	\item Point transect surveys require a different approach than line transects because the area surveyed increases with distance from observer
-	\item In addition to accounting for this we present an alternative way to specify the detection function.
-We replace the spline approach with a parametric family of detection functions such as half-normal and hazard rate functions
+	\item In addition to accounting for this [RJC - what does 'this' refer to?] we present an alternative way to specify the detection function.
+We replace the spline approach with a parametric family of detection functions such as half-normal and hazard rate functions [RJC - I thought we couldn't use the hazard rate (hr) function?
+If the updated inlabru package includes the hr detection function model I need to rerun dsm as hr better fits the data.]
 	\item  These detection functions are typically not linear in their parameters and so cannot be straightforwardly estimated using methods for additive linear predictors
@@
 
-The estimation of the size of wild populations of animals is a critical objective within ecology and conservation [[citation]].
+The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
 Distance sampling methods aim to achieve this by using a spatially explicit sampling design and estimating the detectability of animals as a function of distance from observer \citep{buckland_advanced_2008, buckland_distance_2015}.
-Standard distance sampling approaches use a hybrid of design and model-based inference to estimate population size.
+Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.
 Probability of detection is modelled as a function of distance from observer and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.
 
-Over time interest grew in fully model-based approaches which allow the use of non-randomized survey designs by considering a spatial model for animal density so observations can be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, buckland_model-based_2016, miller_spatial_2013}.
+Over time interest grew in fully model-based approaches which allow the use of non-randomized survey designs by considering a spatial model for animal density so observations can be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
 Model-based distance sampling has typically been implemented in a two-stage modelling process.
 In the first stage, detectability is estimated within each sampling unit.
-In the second stage, detections within sampling units are binned into counts and used as the response variable in generalized additive model framework where the detectability estimates from the first stage are used as an offset.
-Due to often sparse nature of wildlife survey data this may require consideration of over or under-dispersion and zero-inflation.
-Negative-binomial and Tweedie distributions are common choices to model the distribution of counts [[citation?]].
-A key concern with the two-stage approach is the propogation of the uncertainty in the detectability estimates to the second-stage spatial model.
-Early attempts to address this focused on bootstrapping [[citations]], but more recent work has pointed to potentially irresolvable difficulties  [[flesh out]] with the bootstrapping approach and instead proposed propagating error based on a second-order Taylor approximation of detectability around the maximum-likelihood estimate from the first stage \citep{bravington_reliable_2018-1}.
+In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset in the second stage.
+Due to often sparse nature of wildlife survey data this may require consideration of over or under-dispersion and zero-inflated distributions.
+Negative-binomial and Tweedie distributions are common choices to model the distribution of counts \citep{wood_gam_2017}.
+A key concern with the two-stage approach is the propagation of the uncertainty in the detectability estimates to the second-stage spatial model.
+Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003}, but more recent work with the bootstrapping approach has pointed to potentially irresolvable difficulties that require resampling blocks at multiple spatial scales.
+Instead \cite{bravington_reliable_2018-1} proposed propagating error based on a second-order Taylor approximation of detectability around the maximum-likelihood estimate from the first stage.
 
-\citet{yuan_point_2017} presented a one-stage approach for analysing line-transect distance sampling within a Bayesian framework that simulateously estimates the detectability
-and spatial distribution of animals.
-Key to their approach is the use of sparse Gaussian Markov random field (GMRF) approximations to continously indexed Gaussian random fields (GRF) and a log-Gaussian Cox point process model that avoids the need to bin data into counts and naturally sidesteps issues of zero-inflation and dispersion.
+\citet{yuan_point_2017} presented a one-stage approach for analysing line-transect distance sampling within a Bayesian framework that simultaneously estimates the detectability and spatial distribution of animals.
+Key to their approach is the use of sparse Gaussian Markov random field (GMRF) approximations to continuously indexed Gaussian random fields (GRF) and a log-Gaussian Cox point process model that avoids the need to bin data into counts and naturally sidesteps issues of zero-inflation and dispersion.
 The observation process is formulated as a thinning of the log-Gaussian Cox process.
@@
 
-Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.
+Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point-transect distance sampling data.
 Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
-Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data as a thinned point process.
+Our case study is, to the best of our knowledge, the first analysis of point-transect distance sampling data as a thinned point process.
 However, the point process viewpoint is not new and has been applied numerous times to line transect data [[citations here]].
@@
 For example, if $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated using the observed distances.
-The parameter $\sigma$ can only be estimated if some assumption is made about the intensity of the animal locations.
+The parameter $\sigma$ can only be estimated if an assumption is made about the intensity of the animal locations.
 Without such an assumption detectability and intensity are confounded.
@@
 The thinning probability function $g(s)$ is then defined relative to the positions of the surveyed regions.
+We denote the probability of observing a point $s \in \Omega_k$ as $g_k(s)$.
 The probability of observing a point outside the surveyed region is zero.
-We denote the probability of observing a point $s \in \Omega_k$ as $g_k(s)$.
-Since the surveyed regions are non-overlapping each location $s$ is associated with a single thinning probability function $g_k$.
+[RJC - I switched the order of the two previous sentences.
+That way we start with the relevant probability, followed by defining the probability elsewhere is zero]  Since the surveyed regions are non-overlapping each location $s$ is associated with a single thinning probability function $g_k$.
 For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
@@
 However, for any single set of parameter values, $g(s)$ is essentially an offset term.
-That is to say, if we could specify $g(s)$ as known then we could proceed with inference with a standard additive linear predictor.
+That is, if we specify $g(s)$ as known then we can proceed with inference within a standard additive linear predictor framework.
 
@@
 \end{itemize}
-
-\newpage
 
@@
 	\centering
-	\includegraphics[width=0.7\linewidth]{figures/2002studyareapoints_pt}
+	\includegraphics[scale=0.5]{figures/2002studyareapoints_pt}
 	\caption{Study area (light blue) showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.}
@@
 	\centering
-	\includegraphics[width=0.7\linewidth]{figures/2002counts_pt}
+	\includegraphics[scale=0.5]{figures/2002counts_pt}
 	\caption{Counts of \akepa{} by point during the 2002 survey.}
@@
 To attempt to communicate this uncertainty around the mean it is common [[references?]] to present the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
+[RJC - I don't think a reference is needed to substantiate presenting credible intervals or computing them using quantiles, particularly in a stats journal.]
 
@@
 First, each realization has finer-grained spatial structure than the mean of the field.
-What this means in practice is that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.
+In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.
 The mean field plot hides the spatial structure the model has been able to estimate from the data.
 In general, the mean of a random field will contain less spatial structure than realisations of the field.
+[RJC - 'First' is not followed up with subsequent ordinal numbers; I suggest rewriting the above sentence to remove 'first' or add ordinal numbers below.]
 
@@
 Whilst any one prediction location may be well described by such quantiles, to present them side by side in one image obscures the fact that the probability that all (or some subset of) the prediction cells would \textit{simultaneously} achieve that value will be in fact much lower than the quantile value and in many cases would be essentially zero.
-To demonstrate this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 2500 for the 0.025 quantile and 12,300 for the 0.975.
+[RJC - I stumbled over this sentence several times and I think it needs to be rewritten.]  To demonstrate this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 2500 for the 0.025 quantile and 12,300 for the 0.975.
 A naive (and tempting) interpretation of these numbers is as upper and lower limits of the 95\% credible interval for abundance.
@@
 
-As an alternative we suggest using excursion sets and excursion functions.  This is a recently approach to representing uncertainty in stochastic processes that specifically considers
-joint probability of events across multiple locations.
+As an alternative, we suggest using excursion sets and excursion functions.
+This is a recently developed approach to representing uncertainty in stochastic processes that specifically considers joint probability of events across multiple locations.
 The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) < u \}$.
@@
 Excursion sets can be estimated by considering increasing candidate sets $D$ and a sequential integration scheme to estimate probabilities.
-An implementation is available in the \texttt{excursions} package available through the Comprehensive R Archive Network.
+An implementation is available in the \texttt{excursions} package \citep{bolin_calculating_2018} available through the Comprehensive R Archive Network \citep{r_2017}.
 
@@
 We believe the risk of interpreting these maps as intensities is much lower than for the quantile plots.
-The fact that it is possible to estimate discrete regions may also be useful to communicate results to conservation managers and decision-makers.
+The fact that it is possible to estimate discrete regions may also be useful to communicate results to conservation practitioners, land and resource managers and policy decision-makers.
 We anticipate the excursion function will be most useful in communicating to audiences with statistical training.
+[RJC - I think statistically savvy audiences will readily grasp the excursion functions.
+Do you mean "without" statistical training?]
 
@@
 I'm not sure varprop can do this?
+[RJC - Varprop does do this through the posterior covariance V_beta matrix of the intercept, smoother and detection probability terms.
+But it doesn't do so in such an elegant way.
+Varprop is still reliant on a two-stage model.]
 
@@
 somehow...
-	\item Note that Bayesian distance sampling has been done before but that the INLA approach avoids data augmentation and MCMC.
+	\item Note that Bayesian distance sampling [RJC - References for this include Oedekoven et al (2014, JABES 19:219-239) and Buckland et al.
+(2015).
+We could also include Buckland et al.
+(2016, JABES 21:58-75) but it deals more with model-based design distance sampling than Bayesian approaches specifically.
+The more recent papers that reference Cornelia's paper are applied.
+It is also worth looking at Yuan's paper on point process models of blue whales.] has been done before but that the INLA approach avoids data augmentation and MCMC.
 	\item Iterated INLA is a powerful and flexible way to include non-linear components to a predictor and has applications beyond this application to distance sampling

### 2019-08-12 paper.tex
slight tweak

--- 2019-08-09 paper.tex
@@
 The observation process is formulated as a thinning of the log-Gaussian Cox process.
-Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.
+Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.
 The model was specified as a latent Gaussian model and fitted using integrated nested Laplace approximations using the \texttt{R-INLA} package \citep{rue_approximate_2009}.
@@
 Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
-Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data as a thinned point process.
+Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
 However, the point process viewpoint is not new and has been applied numerous times to line transect data [[citations here]].
+We also present the first use of sparse GMRF with point transect distance sampling data.
+
+We demonstrate an application of a new iterative model fitting approach that allows the inclusion of model components that are non-linear in their parameters.
+This allows us to model detectability with a parametric family of detection functions, an approach that more closely resembles standard distance sampling methods than the SPDE formulation presented in \citet{yuan_point_2017}.
 

### 2019-08-12 paper.tex

--- 2019-08-10 paper.tex
@@
 
-\citet{yuan_point_2017} presented a one-stage approach for analysing line-transect distance sampling within a Bayesian framework that simultaneously estimates the detectability and spatial distribution of animals.
-Key to their approach is the use of sparse Gaussian Markov random field (GMRF) approximations to continuously indexed Gaussian random fields (GRF) and a log-Gaussian Cox point process model that avoids the need to bin data into counts and naturally sidesteps issues of zero-inflation and dispersion.
+\citet{yuan_point_2017} presented a one-stage approach for analysing line-transect distance sampling within a Bayesian framework that simulateously estimates the detectability 
+and spatial distribution of animals.
+Key to their approach is the use of sparse Gaussian Markov random field (GMRF) approximations to continously indexed Gaussian random fields (GRF) and a log-Gaussian Cox point process model that avoids the need to bin data into counts and naturally sidesteps issues of zero-inflation and dispersion.
 The observation process is formulated as a thinning of the log-Gaussian Cox process.
-Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.
+Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.
 The model was specified as a latent Gaussian model and fitted using integrated nested Laplace approximations using the \texttt{R-INLA} package \citep{rue_approximate_2009}.
 
-Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point-transect distance sampling data.
+Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.
 Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
-Our case study is, to the best of our knowledge, the first analysis of point-transect distance sampling data as a thinned point process.
+Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
 However, the point process viewpoint is not new and has been applied numerous times to line transect data [[citations here]].
+We also present the first use of sparse GMRF with point transect distance sampling data.
+
+We demonstrate an application of a new iterative model fitting approach that allows the inclusion of model components that are non-linear in their parameters.
+This allows us to model detectability with a parametric family of detection functions, an approach that more closely resembles standard distance sampling methods than the SPDE formulation presented in \citet{yuan_point_2017}.
 

### 2019-08-14 paper.tex
small changes
I made small changes to both paper .bib and .tex so that I can commit and push the files upward.
fix compile issues

@@
 \documentclass[preprint,12pt]{elsarticle}
-
-\usepackage{setspace}
-\doublespacing
-
-\usepackage{hyperref} % auto detect ref type
-\usepackage{natbib}
-\setcitestyle{authoryear}
 
@@
 \makeatother
+
+\usepackage{setspace}
+\doublespacing
+
+\usepackage{hyperref} % auto detect ref type
+\usepackage{natbib}
+\setcitestyle{authoryear}
 
@@
 \usepackage{caption}
+\usepackage{textcomp}    % for Hawaii characters
 
@@
 \newcommand{\tL}{\tilde{\Lambda}}  % thinned big lambda
+
 % RJC 09/08/2019 Added shortcuts for Hawaiian words
@@
 \newcommand{\hawaii}{Hawai\textquotesingle i}   % adds Hawaiian diacritical marks
-
 \DeclareMathOperator*{\argmax}{arg\,max}  % * means _ puts thing beneath operator
@@
 \author[1]{Stephen T Buckland}  % Ask about co-authoring the manuscript
-\author[4]{Steve J Kendall}     % It is Steve's data we are using.
+\author[4]{Steve J Kendall}     % It is Steve's data we are using and I have already mentioned this manuscript to him.
 
 %\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}
-\affil[1]{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}
-\affil[2]{U. S. Geological Survey, Pacific Island Ecosystems Research Center, P.O. Box 44, \hawaii{} National Park, HI 96718, U.S.A.}
-\affil[3]{School of Mathematics, University of Edinburgh, Edinburgh, Scotland}
-\affil[4]{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}
-\affil[*]{Correspondence: Andrew E Seaton, Email: aes22@st-andrews.ac.uk}
+\address[1]{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}
+\address[2]{U. S. Geological Survey, Pacific Island Ecosystems Research Center, P.O. Box 44, \hawaii{} National Park, HI 96718, U.S.A.}
+\address[3]{School of Mathematics, University of Edinburgh, Edinburgh, Scotland}
+\address[4]{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}
+\address[*]{Correspondence: Andrew E Seaton, Email: aes22@st-andrews.ac.uk}
 
@@
 I'm not sure varprop can do this?
-[RJC - Varprop does do this through the posterior covariance V_beta matrix of the intercept, smoother and detection probability terms.
+[RJC - Varprop does do this through the posterior covariance $V_beta$ matrix of the intercept, smoother and detection probability terms.
 But it doesn't do so in such an elegant way.

### 2019-08-14 paper.tex
1st draft of introduction complete

@@
 
-Introduction outline:
-\begin{enumerate}
-	\item Estimation of abundance and spatial distribution of animals is an important topic
-	\item Distance sampling is a family of methods that aim to do this whilst accounting for imperfect detectability of animals
-	\item Distance sampling began life under a design-based inference paradigm
-	\item Randomised survey design is used in two ways.
-Firstly, to justify the assumption that the true distribution of animals is uniform with respect to distance from the observer.
-Secondly, to construct Horvitz-Thompson-like estimators of animal density, which are then converted to abundance estimates for regions of interest
-	\item Interest in the spatial distribution of animals has led to the development of approaches that combine traditional distance sampling methods and spatially explicit generalized additive models
-	\item In this approach the estimated detection probabilities for discrete spatial units are used as an offset in a generalized additive model where the response variable is a count of detections within each unit
-	\item This two-stage approach needs a way to propagate uncertainty from the detection model to the spatial model
-	\item Recently [[JOYCE PAPER]] presented a one-stage approach that simultaneously estimates the detection and spatial processes, avoids binning data into counts and removes the need to consider uncertainty propagation
-	\item This approach considered line-transect distance sampling as thinned log-Gaussian Cox process
-	\item To estimate the detection function [[JOYCE et al]] constructed a spline-like function using an SPDE approach and fit the model using \texttt{R-INLA}
-	\item In this paper we show how to achieve a similar one-stage model for point transect distance sampling
-	\item Point transect surveys require a different approach than line transects because the area surveyed increases with distance from observer
-	\item In addition to accounting for this [RJC - what does 'this' refer to?] we present an alternative way to specify the detection function.
-We replace the spline approach with a parametric family of detection functions such as half-normal and hazard rate functions [RJC - I thought we couldn't use the hazard rate (hr) function?
-If the updated inlabru package includes the hr detection function model I need to rerun dsm as hr better fits the data.]
-	\item  These detection functions are typically not linear in their parameters and so cannot be straightforwardly estimated using methods for additive linear predictors
-	\item To solve this we implement an iterative approach based on a Taylor approximation of the detection function and fit the model using the \texttt{inlabru} package which extends the functionality of \texttt{R-INLA}
-	\item This represents a novel development in distance sampling methods that allows simultaneous estimation of parametric detection functions and the spatial distribution of animals from point transect data
-\end{enumerate}
-
 The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
@@
 
-Over time interest grew in fully model-based approaches which allow the use of non-randomized survey designs by considering a spatial model for animal density so observations can be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
+Over time interest grew in fully model-based approaches which allow the use of non-randomized survey designs by considering a spatial model for animal density.
+This allows observations to be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
 Model-based distance sampling has typically been implemented in a two-stage modelling process.
 In the first stage, detectability is estimated within each sampling unit.
-In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset in the second stage.
+In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset.
 Due to often sparse nature of wildlife survey data this may require consideration of over or under-dispersion and zero-inflated distributions.
 Negative-binomial and Tweedie distributions are common choices to model the distribution of counts \citep{wood_gam_2017}.
+
 A key concern with the two-stage approach is the propagation of the uncertainty in the detectability estimates to the second-stage spatial model.
@@
 and spatial distribution of animals.
-Key to their approach is the use of sparse Gaussian Markov random field (GMRF) approximations to continously indexed Gaussian random fields (GRF) and a log-Gaussian Cox point process model that avoids the need to bin data into counts and naturally sidesteps issues of zero-inflation and dispersion.
+Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF) and a log-Gaussian Cox point process model that avoids the need to bin data into counts and naturally sidesteps issues of zero-inflation and dispersion.
 The observation process is formulated as a thinning of the log-Gaussian Cox process.
@@
 
-Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.
+Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data and replace the SPDE specification of the detection function with a parametric family of detection functions which are familiar to users of standard distance sampling methods.
 Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
 Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
-However, the point process viewpoint is not new and has been applied numerous times to line transect data [[citations here]].
-We also present the first use of sparse GMRF with point transect distance sampling data.
+However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004, stoyan_remark_1982, hogmander_random_1991}.
+We also present the first use of a sparse GMRF spatial random effect with point transect distance sampling data.
+The sparsity of the GMRF is useful for computational efficiency and the finite element mesh motivated by the SPDE allows efficient handling of complex spatial domains such as coastlines \citep{simpson_going_2016}.
 
-We demonstrate an application of a new iterative model fitting approach that allows the inclusion of model components that are non-linear in their parameters.
-This allows us to model detectability with a parametric family of detection functions, an approach that more closely resembles standard distance sampling methods than the SPDE formulation presented in \citet{yuan_point_2017}.
+To demonstrate the benefits of this one-stage approach we highlight potential issues regarding model evaluation and communication of results that are common in many formulations of species distribution modelling.
+We discuss potentially misleading summary maps of the posterior intensity process and propose an alternative based on excursion sets \citep{bolin_excursion_2015, bolin_calculating_2018}.
+A key theme in our case-study is to highlight the importance of considering \textit{realizations} of random quantities rather than rely too much as summaries such as expectations and quantiles.
+The one-stage approach is invaluable since model summaries are based on monte-carlo sampling from the joint posterior of all model parameters.
+Since the spatial process and observation process are estimated jointly this means figures and summaries produced naturally average over the estimated uncertainty in the observation processes.
+This is an advantage over the two-stage approach even if uncertainty has been effectively propagated in some way.
+
+The article proceeds as follows.
+First we present the perspective of point transect distance sampling as a thinned poisson process, describing in detail how this can be formulated as a latent Gaussian model with a latent GMRF random effect.
+We also show how the point process viewpoint can still be used in the case of incomplete location information where only the distance from the observer is recorded, not the exact location of the animal.
+Secondly we describe how to specify the joint model using an iterative approach to fitting models with components that are non-linear in their parameters.
+Thirdly, we introduce our case-study using a point transect distance sampling survey of the Hawaiian \akepa{} and show how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package.
+We focus our model evaluation on the effective understanding and communication of uncertainty in the posterior intensity and abundance estimates, which are typically of greatest interest to stakeholders.
 
@@
 \sloppy For the case with imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected})$.
-A key property of the log-Gaussian Cox process is that a realisation thinned with thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.
+A key property of the log-Gaussian Cox process is that a realisation of a point process with intensity process $\lambda(s)$ that is thinned by thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.
 
@@
 
-TODO HERE:  Description of Akepa data, survey method, key questions, what do we care about from ecology/conservation point of view.
-Then, code snippets of how to specify the model in inlabru, brief discussion of PC priors and how they were chosen here
+TODO HERE:  Add code snippets of how to specify the model in inlabru, brief discussion of PC priors and how they were chosen here
 
@@
 I'm not sure varprop can do this?
-[RJC - Varprop does do this through the posterior covariance $V_beta$ matrix of the intercept, smoother and detection probability terms.
+[RJC - Varprop does do this through the posterior covariance $V_{\beta}$ matrix of the intercept, smoother and detection probability terms.
 But it doesn't do so in such an elegant way.

### 2019-08-28 analyses/data/create_extended_mesh.R
add + preprocess extended dataset

@@ -1,0 +1,30 @@
+# Load required packages
+library(inlabru)
+library(INLA)  # need testing version of inla
+
+point_transects <- readRDS("samplers_extended_no_crs.RDS")
+study_area <- readRDS("study_area_extended_no_crs.RDS")
+
+# meshbuilder()
+
+## Generated by meshbuilder()
+
+## Build boundary information:
+## (fmesher supports SpatialPolygons, but this app is not (yet) intelligent enough for that.)
+boundary <- list(
+  as.inla.mesh.segment(study_area),
+  NULL)
+
+## Build the mesh:
+mesh <- inla.mesh.2d(boundary=boundary,
+                     max.edge=c(280, 960),
+                     min.angle=c(30, 21),
+                     max.n=c(48000, 16000), ## Safeguard against large meshes.
+                     max.n.strict=c(128000, 128000), ## Don't build a huge mesh!
+                     cutoff=2, ## Filter away adjacent points.
+                     offset=c(1100, 2000)) ## Offset for extra boundaries, if needed.
+
+ggplot() + gg(mesh) + gg(point_transects, colour = "red")
+
+# save the mesh
+saveRDS(object = mesh, file = "mesh_extended_no_crs.RDS")

### 2019-08-28 analyses/data/create_study_area.R

@@ -1,0 +1,29 @@
+# create polygon defining study area
+# use simple convex hull and combindation with open forest polygon for now
+
+library(sp)
+library(rgeos)
+library(inlabru)
+
+samplers <- readRDS("samplers_extended_no_crs.RDS")
+samplersnew = samplers[samplers$Stratum == "CF",]
+old_study_area = readRDS("study_area_no_crs.RDS")
+chull = gConvexHull(samplersnew)
+chull = gBuffer(chull, width = 500, capStyle = "SQUARE")
+
+ggplot() +
+  gg(old_study_area) +
+  gg(chull) +
+  gg(samplers) +
+  gg(samplersnew, colour = "red") +
+  coord_equal()
+
+study_area_new = gUnion(old_study_area, chull)
+
+ggplot() +
+  gg(study_area_new) +
+  gg(samplers) +
+  gg(samplersnew, colour = "red") +
+  coord_equal()
+
+saveRDS(study_area_new, file = "study_area_extended_no_crs.RDS")

### 2019-08-28 analyses/data/process_AnalysisDataset2002.R

@@ -1,0 +1,38 @@
+# convert AnalysisDataset2002.csv to appropriate SpatialPointDataFrame objects
+
+library(dplyr)
+library(sp)
+library(inlabru)   # for quick plotting
+
+newdata = read.csv("AnalysisDataset2002.csv", stringsAsFactors = FALSE)
+
+old_obs = readRDS("obs_2002_no_crs.RDS")
+old_samplers = readRDS("samplers_no_crs.RDS")
+
+# Stratum codes:
+# OF = open forest
+# CF = closed forest
+# RF = reforested    - no akepa here
+
+newdata %>%
+  select(SampleLabel, Easting, Northing, Stratum) %>% 
+  filter(Stratum %in% c("OF", "CF")) %>%
+  distinct() -> samplers
+
+samplers$weight = 1
+coordinates(samplers) = ~ Easting + Northing
+
+obs = newdata %>% 
+  select(SampleLabel, Distance, Easting, Northing, Stratum) %>% 
+  filter(Stratum %in% c("OF", "CF"), !is.na(Distance)) -> obs
+
+coordinates(obs) = ~ Easting + Northing
+
+# check it looks okay
+
+ggplot() +
+  gg(samplers) +
+  gg(obs, col = "green")
+
+saveRDS(samplers, file = "samplers_extended_no_crs.RDS")
+saveRDS(samplers, file = "obs_extended_no_crs.RDS")

### 2019-08-29 analyses/data/process_AnalysisDataset2002.R
more extended data work

@@ -24,7 +24,8 @@ coordinates(samplers) = ~ Easting + Northing
 
 obs = newdata %>% 
   select(SampleLabel, Distance, Easting, Northing, Stratum) %>% 
-  filter(Stratum %in% c("OF", "CF"), !is.na(Distance)) -> obs
+  filter(Stratum %in% c("OF", "CF"), !is.na(Distance)) %>% 
+  rename(distance = Distance) -> obs
 
 coordinates(obs) = ~ Easting + Northing
 
@@ -35,4 +36,4 @@ ggplot() +
   gg(obs, col = "green")
 
 saveRDS(samplers, file = "samplers_extended_no_crs.RDS")
-saveRDS(samplers, file = "obs_extended_no_crs.RDS")
+saveRDS(obs, file = "obs_extended_no_crs.RDS")

### 2019-08-29 analyses/eval_model.R

@@ -10,9 +10,9 @@ set.seed(9701071)
 file_name = "akepa_2002_fitted_model.RDS"
 fit = readRDS(file_name)
 
-study_area = readRDS("data/study_area_no_crs.RDS")
-samplers = readRDS("data/samplers_no_crs.RDS")
-mesh = readRDS("data/mesh_no_crs.RDS")
+study_area = readRDS("data/study_area_extended_no_crs.RDS")
+samplers = readRDS("data/samplers_extended_no_crs.RDS")
+mesh = readRDS("data/mesh_extended_no_crs.RDS")
 
 #### Specify detection functions  ####
 

### 2019-08-29 analyses/excursions.R

@@ -9,9 +9,9 @@ set.seed(1525)
 file_name = "akepa_2002_fitted_model.RDS"
 fit = readRDS(file_name)
 
-study_area = readRDS("data/study_area_no_crs.RDS")
-samplers = readRDS("data/samplers_no_crs.RDS")
-mesh = readRDS("data/mesh_no_crs.RDS")
+study_area = readRDS("data/study_area_extended_no_crs.RDS")
+samplers = readRDS("data/samplers_extended_no_crs.RDS")
+mesh = readRDS("data/mesh_extended_no_crs.RDS")
 
 #### monte-carlo samples of intensity ####
 

### 2019-08-29 analyses/fit_model.R

@@ -18,10 +18,10 @@ file_name = "akepa_2002_fitted_model.RDS"
 
 #### Load Data ####
 
-realobs <- readRDS("data/obs_2002_no_crs.RDS")
-samplers <- readRDS("data/samplers_no_crs.RDS")
-study_area <- readRDS("data/study_area_no_crs.RDS")
-mesh <- readRDS("data/mesh_no_crs.RDS")
+realobs <- readRDS("data/obs_extended_no_crs.RDS")
+samplers <- readRDS("data/samplers_extended_no_crs.RDS")
+study_area <- readRDS("data/study_area_extended_no_crs.RDS")
+mesh <- readRDS("data/mesh_extended_no_crs.RDS")
 
 #### Specify detection function  ####
 
@@ -68,7 +68,7 @@ fml <- coordinates + distance ~ grf +
   Intercept
 
 
-starting_values <- data.frame(grf = 0, lsig = 3.6, Intercept = 0)
+starting_values <- data.frame(grf = 0, lsig = 3.65, Intercept = 0)
 
 W <- 60   # transect radius
 distance_domain <- seq(.Machine$double.eps, W, length.out = 30)
@@ -78,7 +78,7 @@ fit <-  lgcp(components = cmp,
              samplers = samplers,
              domain = list(coordinates = mesh, distance = distance_domain),
              formula = fml,
-             options = list(result = starting_values, max.iter = 20))
+             options = list(result = starting_values, max.iter = 30))
 
 summary(fit)
 

### 2019-08-29 paper.tex
update figures for extended data

@@
 Secondly we describe how to specify the joint model using an iterative approach to fitting models with components that are non-linear in their parameters.
-Thirdly, we introduce our case-study using a point transect distance sampling survey of the Hawaiian \akepa{} and show how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package.
+Thirdly, we introduce our case-study using a point transect distance sampling survey of the Hawaiian \akepa{}, an endangered bird of the honeycreeper tribe.
+We show how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package.
 We focus our model evaluation on the effective understanding and communication of uncertainty in the posterior intensity and abundance estimates, which are typically of greatest interest to stakeholders.
@@
 
-This is similar to a Poisson likelihood with an offset term equal to $\log\alpha$.
+This is similar to a Poisson likelihood with an offset term equal to $\log\alpha$ and can be specified using the Poisson likelihood implementation available in \texttt{R-INLA}.
 
@@
 In the above we assume the data are complete records of animal location.
-However, in many distance sampling surveys the exact location of the animal is not recorded - instead only the location of the observer and the distance to the observer are recorded.
+However, in many distance sampling surveys only the location of the observer and the distance to the observer are recorded.
 Data of this type can be analysed within a point process framework by deriving the appropriate intensity function for the incomplete data.
 
-Using the polar coordinates notation given above, for true animal location $s_k(r, \theta) \in \Omega_k$, we consider the case where we observe $r$ but not $\theta$.
+Using the polar coordinates notation given above, for true animal location $s_k(r, \theta) \in \Omega_k$ we consider the case where we observe $r$ but not $\theta$.
 It follows that the intensity for points at a distance $r$ observed within sampling unit $\Omega_k$ is
@@
 where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.
-All three parameters are not simultaneously identifiable and it is conventional to assume a value for $\nu$ which controls the mean-square differentiability of the process.
-We set $\nu = 1$ which is the most common default value in the approximation approach outlined below.
-[[Would like to justify $\nu = 1$ somehow]]
+All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which controls the mean-square differentiability of the process.
+We set $\nu = 1$ which is the most default value in \texttt{R-INLA}.
+[[Would like to justify $\nu = 1$ somehow?]]
 
-We use a Gaussian Markov random field approximation of $\xi(s)$ following the approach of [[Lindgren 2011]] based on specifying $\xi(s)$ via a stochastic partial differential equation.
+We use a Gaussian Markov random field approximation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a stochastic partial differential equation.
 Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent $\xi$ as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
-The parameters $\xi_1, \ldots, \xi_L$ follow a multivariate normal distribution with sparse precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a stochastic partial differential equation.
+The parameters $\xi_1, \ldots, \xi_L$ are a Gaussian Markov random field with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a stochastic partial differential equation.
 The precision parameter $\tau$ controls the marginal variance of the field and $\kappa$ is an inverse range parameter that controls the rate of decay of the covariance as distance between locations increases.
@@
 This presents a problem since $\log g(s)$ is not typically linear in its parameters.
-However, for any single set of parameter values, $g(s)$ is essentially an offset term.
-That is, if we specify $g(s)$ as known then we can proceed with inference within a standard additive linear predictor framework.
-
-TO DO:
+However, for any single set of parameter values $g(s)$ can be viewed as acting like an offset.
+If we could specify $g(s)$ as known then we can proceed with an inference approach that applies to additive linear predictors.
 
@@
 \section*{Case-Study:  Hawaiian Akepa Survey}
-
-TODO HERE:  Add code snippets of how to specify the model in inlabru, brief discussion of PC priors and how they were chosen here
 
@@
 
+\subsection*{Model specification}
+
+TODO HERE:  Add code snippets of how to specify the model in inlabru, brief discussion of PC priors and how they were chosen here
+
+\subsection*{Results}
+
 The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.
@@
 In general the south, with greater density of sampled locations, has a lower CV than the north, where sampling units are more spread out.
-The areas of highest CV are those regions furthest away from any sampled location, reflecting the intuitive fact that we are least certain in areas where we did not look.
+The areas of highest CV are those regions furthest away from any sampled location, reflecting the intuitive fact that we are least certain in areas further away from the sampled locations.
 
@@
 To attempt to communicate this uncertainty around the mean it is common [[references?]] to present the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
-[RJC - I don't think a reference is needed to substantiate presenting credible intervals or computing them using quantiles, particularly in a stats journal.]
+[RJC - I don't think a reference is needed to substantiate presenting credible intervals or computing them using quantiles, particularly in a stats journal.
+AES - I was trying to find an example of someone presenting maps like these in a paper, yet to find one.
+If you know of any please let me know!]
 
@@
 
-As an alternative, we suggest using excursion sets and excursion functions.
+As an alternative, we suggest using excursion sets and excursion functions \citep{bolin_excursion_2015}.
 This is a recently developed approach to representing uncertainty in stochastic processes that specifically considers joint probability of events across multiple locations.
-The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) < u \}$.
+The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e.
+the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$.
 For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
@@
 
-Excursion sets can be estimated by considering increasing candidate sets $D$ and a sequential integration scheme to estimate probabilities.
+Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.
 An implementation is available in the \texttt{excursions} package \citep{bolin_calculating_2018} available through the Comprehensive R Archive Network \citep{r_2017}.
 
-\autoref{fig:excursion-set} shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95.
+\autoref{fig:excursions} (left panel) shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95.
 This figure can be interpreted in a natural way.
@@
 \begin{figure}
-	\includegraphics[scale=0.5]{figures/excursion_set.png}
-	\caption{The positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95}
-	\label{fig:excursion-set}
+	\includegraphics[scale=0.5]{figures/excursions.png}
+	\caption{Left:  The positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.
+Right: The positive excursion function with a level corresponding to 1 bird per hectare}
+	\label{fig:excursions}
 \end{figure}
@@
 For each $s$ it is the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.
-The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursion-function}.
-
-\begin{figure}
-	\includegraphics[scale=0.5]{figures/excursion_function.png}
-	\caption{The positive excursion function with a level corresponding to 1 bird per hectare}
-	\label{fig:excursion-function}
-\end{figure}
+The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursions} (right panel).
 
@@
 It shows the largest possible probability for each location to be a member of a set for in which the intensity exceeds a threshold simultaneously across all locations in the set.
-It can be viewed intuitively as a kind of fuzzier version of \autoref{fig:excursion-set}.
 Regions on the edge of the excursion set would be included if the $\alpha$ value were allowed to increase slightly.
@@
 Another key aim of performing distance sampling surveys is to estimate the abundance of the population of interest.
-By specifying our spatial model as a log-Gaussian Cox process, the intensity is a random field rather than a fixed function of space.
+By specifying our spatial model as a log-Gaussian Cox process, the posterior intensity is a random field rather than a deterministic function.
 Let $n$ denote the abundance of the population within the region of interest and $N$ its corresponding random variable within our Bayesian framework.
@@
 We estimate the posterior $\pi(N | \lambda)$ by a monte-carlo method.
-Taking $m$ monte-carlo samples of the posterior intensity field $\lambda_1, \ldots, \lambda_m$ we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
-This is well defined since $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
+Taking $m$ monte-carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
+This is well-defined since $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
 \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
@@
 somehow...
+
 	\item Note that Bayesian distance sampling [RJC - References for this include Oedekoven et al (2014, JABES 19:219-239) and Buckland et al.
@@
 	\item More to do in future to test how many parameters can be included in the non-linear component and what types of non-linear components are feasible
-
+	\item Note similarities and differences with two-stage GAM approach
+	\item Note that presenting uncertainty in maps is difficult and excursion sets are a promising contribution relevant to many areas of spatial statistics.
+	\item Links to species "range" estimates - excursions possibly useful here.
 \end{enumerate}

### 2019-08-29 analyses/eval_model.R

@@ -2,6 +2,9 @@
 library(INLA)
 library(inlabru)
 
+# set ggplot theme
+theme_set(theme_minimal())
+
 set.seed(9701071)
 
 #### WARNING - as it stands this script will overwrite figures for the paper
@@ -61,14 +64,18 @@ p1 = ggplot() +
   gg(study_area) +
   gg(pr.int) +
   scale_fill_viridis_c() +
-  ggtitle("Posterior mean intensity")
+  ggtitle("Posterior mean intensity") +  
+  xlab("Easting") +
+  ylab("Northing")
 
 p2 = ggplot() +
   gg(study_area)  +
   gg(pr.int["cv"]) +
   gg(samplers, colour = "red") +
   scale_fill_viridis_c() +
-  ggtitle("Posterior CV of intensity")
+  ggtitle("Posterior CV of intensity") +  
+  xlab("Easting") +
+  ylab("Northing")
 
 png(filename = "../figures/intensity_mean_cv.png",
     width = 10, height = 6, units = "in", res = 100)
@@ -86,17 +93,23 @@ viridisscale =   scale_fill_viridis_c(limits = range(v))
 p1 = ggplot() + 
   gg(study_area) + 
   gg(pr.int) +
-  viridisscale
+  viridisscale +
+  xlab("Easting") +
+  ylab("Northing")
 
 p2 = ggplot() +
   gg(study_area) +
   gg(pr.int["q0.025"]) +
-  viridisscale
+  viridisscale +
+  xlab("Easting") +
+  ylab("Northing")
 
 p3 = ggplot() +
   gg(study_area) +
   gg(pr.int["q0.975"]) +
-  viridisscale
+  viridisscale +
+  xlab("Easting") +
+  ylab("Northing")
 
 layout_matrix <- matrix(c(4, 1, 1, 4, 2, 2, 3, 3), nrow = 2, byrow = TRUE)  # triangle arrangement
 
@@ -126,17 +139,23 @@ viridisscale =   scale_fill_viridis_c(limits = range(draw_all))
 p1 = ggplot() + 
   gg(study_area) + 
   gg(draw1["mean"]) +
-  viridisscale
+  viridisscale +
+  xlab("Easting") +
+  ylab("Northing")
 
 p2 = ggplot() + 
   gg(study_area) + 
   gg(draw2["mean"]) +
-  viridisscale
+  viridisscale +
+  xlab("Easting") +
+  ylab("Northing")
 
 p3 = ggplot() + 
[…]

### 2019-08-29 analyses/excursions.R

@@ -2,6 +2,12 @@
 library(INLA)
 library(inlabru)
 library(excursions)
+source("gg.R")   # small edit of gg.SpatialPixelsDataFrame
+
+## todo - replace predict loop with generate() ?
+
+# set ggplot theme
+theme_set(theme_minimal())
 
 set.seed(1525)
 
@@ -49,22 +55,35 @@ Fpxl$F = ex$F
 
 # png(filename = "../figures/excursion_function.png",
 #     width = 4.5, height = 5.5, units = "cm", res = 1000)
-ggplot() + 
+p2 = ggplot() + 
   gg(Fpxl) +
   scale_fill_viridis_c() +
-  ggtitle("Excursion function for > 1 bird per hectare")
+  ggtitle("Excursion function for > 1 bird per hectare") +
+  xlab("Easting") +
+  ylab("Northing")
 
-ggsave(filename = "../figures/excursion_function.png",
-       width = 5, height = 6, units = "in")
+p2
+# ggsave(filename = "../figures/excursion_function.png",
+#        width = 5, height = 6, units = "in")
 
 Epxl = pxl
-Epxl$E = ex$E
-ggplot() +
-  gg(Epxl) +
-  ggtitle("Excursion set for > 1 bird per hectare,\nalpha = 0.05")
+Epxl$E = as.factor(ex$E)
+p1 = ggplot() +
+  gg(Epxl, aes(group = E)) +
+  ggtitle("Excursion set for > 1 bird per hectare,\nalpha = 0.05") +
+  scale_fill_manual(values = c("gray", "blue")) +
+  xlab("Easting") +
+  ylab("Northing")
 
-ggsave(filename = "../figures/excursion_set.png",
-       width = 5, height = 6, units = "in")
+p1
+
+# ggsave(filename = "../figures/excursion_set.png",
+#        width = 5, height = 6, units = "in")
+
+png(filename = "../figures/excursions.png",
+    width = 10, height = 6, units = "in", res = 100)
+multiplot(p1, p2, cols = 2)
+dev.off()
 
 # contour map - how to visualise this?  ?
 cm = contourmap.mc(Xhec, levels = c(0.1, 0.5, 0.8, 1.5, 2, 4), alpha = alpha)

### 2019-08-29 analyses/gg.R

@@ -1,0 +1,50 @@
+# update gg.SpatialPixelsDataFrame to play well with factor
+# this could be improved by allowing pixels to have multiple columns, only do default
+# behaviour if length(names(data)) == 1, otherwise make user specify
+
+# the long lat default name behaviour comes from the fortify.sp function
+# by default this labels coords long lat, regardless of crs
+# I assume this is why Fabian just kept long lat as default for sp objects
+# That this behaviour comes from fortify means that we cannot use the aes(...) argument
+# to provide the names of the coordinates.  The created dataframe will have columns
+# long and lat.  
+# So seems like easiest thing is to just just xlab and ylab to get labels you want
+
+gg.SpatialPixelsDataFrame <- function(data,
+                                      mapping = NULL,
+                                      alpha = NULL,
+                                      crs = NULL,
+                                      mask = NULL, ...) {
+  if (!is.null(crs)) {
+    data <- spTransform(data, crs)
+  }
+  if (!is.null(mask)) {
+    data <- data[as.vector(!is.na(over(data, mask))), ]
+  }
+  
+  df <- as.data.frame(data)
+  
+  if (!is.factor(data[,c(names(data)[[1]])])){
+  dmap <- aes_string(
+    x = coordnames(data)[1],
+    y = coordnames(data)[2],
+    fill = names(data)[[1]]
+  )} else
+    dmap <- aes_string(
+      x = coordnames(data)[1],
+      y = coordnames(data)[2],
+      group = names(data)[[1]]
+    )
+  
+  if (!is.null(mapping)) {
+    dmap <- modifyList(dmap, mapping)
+  }
+  if (!is.null(alpha)) dmap <- modifyList(dmap, aes_string(alpha = alpha))
+  gm <- geom_tile(data = df, mapping = dmap, ...)
+  
+  # If data is not discrete (factor), add default color palette
+  # if (!inherits(data[[deparse(dmap$fill)]], "factor")) {
+  #  gm = c(gm, scale_fill_gradientn(colours = bru.pal()))
+  # }
+  gm
+}

### 2019-09-02 paper.tex
redo introduction

@@
 \usepackage{textcomp}    % for Hawaii characters
+%\usepackage{listings}
+%\lstset{language=R}
+\usepackage[T1]{fontenc}  % tilde in middle in lstlisting
+\usepackage[formats]{listings}
+\lstset{language=R,
+		basicstyle=\ttfamily,
+		columns=fixed,
+		basewidth=0.5em,
+		literate={~}{{$\sim$}}1
+		}
 
@@
 The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
-Distance sampling methods aim to achieve this by using a spatially explicit sampling design and estimating the detectability of animals as a function of distance from observer \citep{buckland_advanced_2008, buckland_distance_2015}.
+Distance sampling methods aim to achieve this by using a spatially explicit sampling design and estimating the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2008, buckland_distance_2015}.
 Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.
-Probability of detection is modelled as a function of distance from observer and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.
+Probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.
 
-Over time interest grew in fully model-based approaches which allow the use of non-randomized survey designs by considering a spatial model for animal density.
-This allows observations to be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
-Model-based distance sampling has typically been implemented in a two-stage modelling process.
+Over time interest grew in fully model-based approaches with a spatially explicit model for animal density that allow the use of non-randomized survey designs.
+This also allows observations to be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
+Model-based distance sampling has been implemented in a two-stage modelling framework.
 In the first stage, detectability is estimated within each sampling unit.
 In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset.
-Due to often sparse nature of wildlife survey data this may require consideration of over or under-dispersion and zero-inflated distributions.
+Due to often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.
 Negative-binomial and Tweedie distributions are common choices to model the distribution of counts \citep{wood_gam_2017}.
@@
 A key concern with the two-stage approach is the propagation of the uncertainty in the detectability estimates to the second-stage spatial model.
-Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003}, but more recent work with the bootstrapping approach has pointed to potentially irresolvable difficulties that require resampling blocks at multiple spatial scales.
-Instead \cite{bravington_reliable_2018-1} proposed propagating error based on a second-order Taylor approximation of detectability around the maximum-likelihood estimate from the first stage.
+Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units as well as the challenge of combining smoothers (a Bayesian/random effects concept) and boostraps (a frequentist concept) \citep{bravington_reliable_2018-1}.
+Instead, \cite{bravington_reliable_2018-1}  proposed avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the maximum-likelihood estimate from the first stage.
 
-\citet{yuan_point_2017} presented a one-stage approach for analysing line-transect distance sampling within a Bayesian framework that simulateously estimates the detectability 
-and spatial distribution of animals.
-Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF) and a log-Gaussian Cox point process model that avoids the need to bin data into counts and naturally sidesteps issues of zero-inflation and dispersion.
+One-stage modelling approaches to distance sampling have also been developed within a Bayesian framework.
+The first efforts in this area used data augmentation approaches for unobserved individuals or groups and were fit using Markov-chain Monte Carlo (MCMC) methods \citep{royle_hierarchical_2008, schmidt_using_2012}.
+More recently  \cite{oedekoven_bayesian_2014} provided a one-stage model that avoids data augmention by specifying a combined likelihood of the detection and spatially explicit count models and incorporated model uncertainty by applying a reversible jump MCMC algorithm.
+
+The reliance on MCMC for one-stage modelling presents a challenge for using spatially structured random effects, which can be computationally expensive.
+\citet{yuan_point_2017} presented a one-stage approach fitted using integrated nested Laplace approximations (INLA).
+They considered line-transect distance sampling and simulateously estimated the detectability and spatial distribution of animals.
+Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF).
+They consider the data as a realisation of a spatio-temporal log-Gaussian Cox point process, an approach that avoids the need to bin data into counts and sidesteps the issues of zero-inflation and dispersion.
 The observation process is formulated as a thinning of the log-Gaussian Cox process.
 Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.
-The model was specified as a latent Gaussian model and fitted using integrated nested Laplace approximations using the \texttt{R-INLA} package \citep{rue_approximate_2009}.
 
-Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data and replace the SPDE specification of the detection function with a parametric family of detection functions which are familiar to users of standard distance sampling methods.
+Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.
+We replace the SPDE specification of the detection function with a parametric family of detection functions that are more familiar to users of standard distance sampling methods.
 Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
@@
 We also present the first use of a sparse GMRF spatial random effect with point transect distance sampling data.
-The sparsity of the GMRF is useful for computational efficiency and the finite element mesh motivated by the SPDE allows efficient handling of complex spatial domains such as coastlines \citep{simpson_going_2016}.
+The sparsity of the GMRF is useful for computational efficiency and the finite element mesh motivated by the SPDE allows for efficient handling of complex spatial domains such as coastlines \citep{simpson_going_2016}.
 
-To demonstrate the benefits of this one-stage approach we highlight potential issues regarding model evaluation and communication of results that are common in many formulations of species distribution modelling.
+To demonstrate the benefits of our approach we discuss issues regarding model evaluation and communication of results that are common across the field of species distribution modelling.
 We discuss potentially misleading summary maps of the posterior intensity process and propose an alternative based on excursion sets \citep{bolin_excursion_2015, bolin_calculating_2018}.
-A key theme in our case-study is to highlight the importance of considering \textit{realizations} of random quantities rather than rely too much as summaries such as expectations and quantiles.
-The one-stage approach is invaluable since model summaries are based on monte-carlo sampling from the joint posterior of all model parameters.
-Since the spatial process and observation process are estimated jointly this means figures and summaries produced naturally average over the estimated uncertainty in the observation processes.
+A key theme in our case-study is to highlight the importance of considering \textit{realizations} of random quantities rather than rely too much on summaries such as expectations and quantiles.
+The one-stage Bayesian approach is invaluable since model summaries are based on monte-carlo sampling from the joint posterior of all model parameters.
+The spatial process and observation process are estimated jointly and so figures and summaries produced using posterior samples will naturally average over the uncertainty in the observation processes.
 This is an advantage over the two-stage approach even if uncertainty has been effectively propagated in some way.
@@
 The article proceeds as follows.
-First we present the perspective of point transect distance sampling as a thinned poisson process, describing in detail how this can be formulated as a latent Gaussian model with a latent GMRF random effect.
-We also show how the point process viewpoint can still be used in the case of incomplete location information where only the distance from the observer is recorded, not the exact location of the animal.
-Secondly we describe how to specify the joint model using an iterative approach to fitting models with components that are non-linear in their parameters.
-Thirdly, we introduce our case-study using a point transect distance sampling survey of the Hawaiian \akepa{}, an endangered bird of the honeycreeper tribe.
-We show how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package.
-We focus our model evaluation on the effective understanding and communication of uncertainty in the posterior intensity and abundance estimates, which are typically of greatest interest to stakeholders.
+First, we present point transect distance sampling as a thinned log-Gaussian Cox process, describing how this can be formulated as a latent Gaussian model with a spatial GMRF random effect that can be fitted using INLA.
+We also show how the point process model can be applied in the case of incomplete location information where only the distance from the observer is recorded, not the exact location of the animal.
+This is a common data format for distance sampling data.
+Second, we describe how to specify and fit the joint model using an iterative approach to fitting models with components that are non-linear in their parameters - we call this the "iterated INLA" approach.
+This allows us to avoid the SPDE approach to the detection function, instead using the parametric families common in the literature.
+Third, we introduce our case-study of a point transect distance sampling survey of the Hawaiian \akepa{}, an endangered bird of the honeycreeper tribe.
+We show how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package with additional functionality for the iterated INLA approach.
+The paper includes example code to specify the model as well as full scripts for the analysis in the supplementary materials.
+We focus model evaluation and discussion on the effective understanding and communication of uncertainty in the posterior intensity and abundance estimates, which are typically of greatest interest to stakeholders.
 
@@
 \begin{align}
+\label{intensity-incomplete}
 \tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(s)g_k(s)\mathrm{d}s \nonumber \\
@@
 
-\subsection*{Model specification}
+\subsection*{Model specification in \texttt{inlabru}}
 
-TODO HERE:  Add code snippets of how to specify the model in inlabru, brief discussion of PC priors and how they were chosen here
+In this section we introduce \texttt{inlabru} syntax and provide example code to demonstrate to specify the model.  First we specify the SPDE model as
+\begin{lstlisting}
+matern <- inla.spde2.pcmatern(mesh, 
+                              prior.sigma = c(2, 0.01),    
+                              prior.range = c(130, 0.01))
+\end{lstlisting}
+which uses the penalised complexity priors for the hyperparameters of the GMRF as introduced by \cite{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$ where $\rho$ is the range of the Mat\'ern covariance.
+
+The data consists only of distances from the observer so we use equation \eqref{intensity-incomplete} to specify the log thinning probability function as
+\begin{lstlisting}
+thinning_function <- function(distance, lsig){
+    log(2*pi) + log(distance) + log_halfnormal(distance, lsig)
+    }
+\end{lstlisting}
+where \texttt{log\_halfnormal} is the log half-normal detection function $\log g(r | \sigma) = -1/2(r/\sigma)^2$ and the argument \texttt{lsig} is $\log(\sigma)$.
+The column in the dataset containing the observed distances is named \texttt{distance} amd the software will search for the appropriate data using the names of the arguments specified in this function.
+To allow the software to distinguish between names of covariates and parameters to be estimated we must also specify a model components object
+\begin{lstlisting}
+cmp <- ~ Intercept + grf(map = coordinates, model = matern) + lsig 
+\end{lstlisting} 
+This object names all model components to be estimated.
+Note that the \texttt{grf} component is a name we have chosen and not a function.
+The software allows user-specified names for the \texttt{f()} function known to users of \texttt{R-INLA}.
+\texttt{coordinates} is a keyword that allows users to supply their data in the form of a spatial object using the \texttt{sp} package \citep{pebesma_spatial_2005}.
+
+The model formula object is
+\begin{lstlisting}
+fml <- coordinates + distance ~ Intercept + grf +
+	thinning_function(distance, lsig)
+\end{lstlisting}
+To fit the model we specify starting values for the parameters, the domain of possible observed distances and fit the model using the \texttt{lgcp} function
+\begin{lstlisting}
+starting_values <- data.frame(grf = 0, lsig = 3.65, Intercept = 0)
+W <- 60   # transect radius
+distance_domain <- seq(.Machine$double.eps, W, length.out = 30)
+lgcp(components = cmp, 
+     data = obs,
+     samplers = samplers,
+     domain = list(distance = distance_domain),
+     formula = fml,
+     options = list(result = starting_values, max.iter = 30))
+\end{lstlisting}
+where our observed data is contained in a \texttt{SpatialPointsDataFrame} object named \texttt{obs}.
+Since not all sampling units observed birds we also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sample units%.
 
@@
 Whilst any one prediction location may be well described by such quantiles, to present them side by side in one image obscures the fact that the probability that all (or some subset of) the prediction cells would \textit{simultaneously} achieve that value will be in fact much lower than the quantile value and in many cases would be essentially zero.
-[RJC - I stumbled over this sentence several times and I think it needs to be rewritten.]  To demonstrate this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 2500 for the 0.025 quantile and 12,300 for the 0.975.
+[RJC - I stumbled over this sentence several times and I think it needs to be rewritten.]  To demonstrate this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile and 12,400 for the 0.975 quantile.
 A naive (and tempting) interpretation of these numbers is as upper and lower limits of the 95\% credible interval for abundance.
@@
 	\item Links to species "range" estimates - excursions possibly useful here.
+	\item model selection is difficult.  Oedekoven 2014 addresses using RJMCMC.  Unclear what to do with this approach.
 \end{enumerate}

### 2019-09-02 analyses/fit_model.R
remove unnecessary spatial domain specification

@@ -76,7 +76,8 @@ distance_domain <- seq(.Machine$double.eps, W, length.out = 30)
 fit <-  lgcp(components = cmp, 
              data = realobs,
              samplers = samplers,
-             domain = list(coordinates = mesh, distance = distance_domain),
+             # domain = list(coordinates = mesh, distance = distance_domain),
+             domain = list(distance = distance_domain),
              formula = fml,
              options = list(result = starting_values, max.iter = 30))
 

### 2019-09-02 paper.tex
edit model description sections

@@
 
-Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance from the observer.
-For example, if $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated using the observed distances.
+Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance.
+The relevant distance depends on the survey type.
+For line transects the perpendicular distance to the transect line is used.
+For point transects the distance to the observer is used.
+For the remainder of the paper we assume a point transect survey design.
+
+The thinning probability function $g(s)$ is specified as a member of a parametric family of functions.
+In this paper we demonstrate our approach using the half-normal formulation, although others such as hazard rate, are also common.
+If $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated.
 The parameter $\sigma$ can only be estimated if an assumption is made about the intensity of the animal locations.
@@
 The standard assumption of distance sampling is that the intensity is constant with respect to changes in $r(s)$.
-Thus any observed differences from uniformity can be attributed to detectability, and not variation in the intensity.
+Thus any observed differences from uniformity can be attributed to detectability and not variation in the intensity.
 
-A point transect distance sampling survey consists of a set of $K$ sampled regions.
-The $k$-th sampled region we denote $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region $\Omega = \cup_{k=1}^K \Omega_k$.
-For simplicity we assume that all sampled regions are discs with radius $W$ and $\Omega_j \cap \Omega_k = \emptyset$ for all $j \neq k$.
-The thinning probability function $g(s)$ is then defined relative to the positions of the surveyed regions.
-We denote the probability of observing a point $s \in \Omega_k$ as $g_k(s)$.
+A point transect distance sampling survey consists of a set of $K$ sampling units.
+The $k$-th sample unit we denote $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region $\Omega = \cup_{k=1}^K \Omega_k$.
+For simplicity we assume that all sampling units are discs with radius $W$ and $\Omega_j \cap \Omega_k = \emptyset$ for all $j \neq k$.
+The thinning probability function $g(s)$ is defined relative to the positions of the sample units.
+The probability of observing a point at location $s$ in the $k$-th sampling unit $\Omega_k$ we denote $g_k(s)$.
 The probability of observing a point outside the surveyed region is zero.
-[RJC - I switched the order of the two previous sentences.
-That way we start with the relevant probability, followed by defining the probability elsewhere is zero]  Since the surveyed regions are non-overlapping each location $s$ is associated with a single thinning probability function $g_k$.
+Since the surveyed regions are non-overlapping each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.
 For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
@@
 
-The likelihood of observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
+The thinne log-Gaussian Cox process likelihood of observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
 \begin{equation}
@@
 \end{equation}
-
 The integral component of the likelihood does not usually have analytical solutions.
-Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood.
+Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach of \cite{simpson_going_2016} for a fully observed point pattern.
 
@@
 
-Using the polar coordinates notation given above, for true animal location $s_k(r, \theta) \in \Omega_k$ we consider the case where we observe $r$ but not $\theta$.
+Using the polar coordinates notation given above, for a detected point at location $s_k(r, \theta) \in \Omega_k$ we consider the case where we observe $r$ but not $\theta$.
 It follows that the intensity for points at a distance $r$ observed within sampling unit $\Omega_k$ is
@@
 All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which controls the mean-square differentiability of the process.
-We set $\nu = 1$ which is the most default value in \texttt{R-INLA}.
+We set $\nu = 1$ which is the default value in \texttt{R-INLA}.
 [[Would like to justify $\nu = 1$ somehow?]]
@@
 We use a Gaussian Markov random field approximation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a stochastic partial differential equation.
-Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent $\xi$ as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
-The parameters $\xi_1, \ldots, \xi_L$ are a Gaussian Markov random field with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a stochastic partial differential equation.
+Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
+The parameters $\xi_1, \ldots, \xi_L$ form a Gaussian Markov random field with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a stochastic partial differential equation.
 The precision parameter $\tau$ controls the marginal variance of the field and $\kappa$ is an inverse range parameter that controls the rate of decay of the covariance as distance between locations increases.
@@
 However, for any single set of parameter values $g(s)$ can be viewed as acting like an offset.
-If we could specify $g(s)$ as known then we can proceed with an inference approach that applies to additive linear predictors.
+If we could specify $g(s)$ as known then we could proceed with an inference approach suitable for additive linear predictors.
 

### 2019-09-02 paper.tex
edit case study

@@
 % addtextstart
-In this section, we present a description of Hawaiian forest birds, forest bird monitoring, the study species, study area, bird sampling, and description of the data.
-\hawaii{} is home to the largest adaptive radiation of oceanic archipelago birds worldwide.
-The Hawaiian Honeycreepers (Drepanidinae) is comprised of 21 genera and more than 50 species generally grouped according to feeding niches with many species and subspecies being island endemics.
-The honeycreepers, like most insular bird species, evolved life history strategies that include greater longevity with high adult survival, delayed maturation with an extended post-fledging dependency period, decreased clutch size and fecundity, and reduced dispersal abilities with high site fidelity \citep{woodworth_demography_2009}.
-While the honeycreepers diversified into the plethora of ecological niches they also had a loss or reduction of defences against competitors, predators and diseases.
-The loss of defensive adaptations made the honeycreepers particularly vulnerable to humans and the changes and threats they brought.
-As a result, more than half of the honeycreepers are extinct \citep{banko_extinctions_2009}.
-A mere 23 honeycreeper species persist with most species restricted to minute portions of their original ranges and in small numbers.
-The IUCN lists 21 honeycreeper species as critically endangered, endangered or vulnerable, and only two species are categorized as least concern \citep{iucn_redlist_2018}.
-
-In the late $19^{\textnormal{th}}$ century naturalists to \hawaii{} described the Hawaiian forest birds and documented species distributions and abundances.
-Subsequent naturalists to the islands documented range contractions and population declines of native species while noting expansion in non-native bird species ranges and abundances.
-It was not until standardized and statistically robust sampling techniques to estimate population sizes and associated standard errors were initiated that the extent to which honeycreepers had declined was appreciated.
+In this section, we apply the above methods to point transect data on the \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}), an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.
 Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}.
@@
 
-\subsection*{Study species}
-The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.
-\Akepa{} is a foraging specialist whose diet consists of spiders, insects and psyllids, as well as nectar.
-\Akepa{} uses its cross-bill to probe and open terminal leaf clusters and extracts prey with its tongue.
 During the 20th century \akepa{} declined dramatically due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994},  mosquito-transmitted avian diseases \citep{pratt_avifaunal_1994, atkinson_wildlife_1995}, introduced predators \citep{lepson_akepa_1997}, and food resources competitors \citep{lepson_akepa_1997}.
 \Akepa{} has a global abundance of 16,248 (95\%CI 10,074\textendash25,198) birds that has been restricted to five spatially distinct populations \citep{judge_akepa_2018}.
-Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau; 19$^{\circ}$ 51\textquotesingle N, 155$^{\circ}$ 18\textquotesingle W), supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}.
+Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau), supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}.
 Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern.
@@
 
-\subsection*{Study area}
+\subsection*{Study area and sampling method}
+
+[[Need to update this and the figure with the extended data area]]
+
 Hakalau was established in 1985 to conserve 15,390-ha of montane forest habitat for native forest birds and rainforest plants.
 Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance.
-Survey points were established along 15 transects following a systematic, random design with points approximately 150-m apart on transects located either 500 or 1,000 m apart.
+Survey points were established along 15 transects following a systematic, random design with points approximately 150 m apart on transects located either 500 or 1,000 m apart.
 Following \cite{camp_population_2010, camp_statespace_2016}, we restrict our study area to the 3,061 ha open-forest stratum of Hakalau.
-The study area is between 1,400 and 1,920 m elevation of montane forest dominated by native \textquotesingle \={o}hi\textquotesingle a (\textit{Metrosideros polymorpha}) and koa (\textit{Acacia koa}) with a mixture of native and non-native understory plants.
 The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.
@@
 The west side of the study area is bounded by pasture that is dominated by grass and is unsuitable habitat for \akepa.
-The substrate in the southern portion of the study area is nutrient-rich, old lava flows (11,000 to 64,000 yrs old) whereas the northern portion occurs on new lava flows (5,000 to 10,000 yrs old) that are nutrient depleted \citep{ccp_hakalau_2010}.
 
@@
 
-\subsection*{Bird sampling}
-Surveys used point-transect methods, a form of distance sampling, where horizontal distances from survey points to individually detected birds are used to model a species-specific detection probability that is used to estimate absolute abundance and variance \citep{buckland_distance_2015}.
+Surveys used point-transect methods, recording horizontal distances from survey points to individually detected birds.
 Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength \textgreater Baufort scale 3).
 During 8-min counts trained observers recorded the species, exact distance to the nearest meter and detection type for each bird detected, along with the sampling conditions cloud cover, rain, wind strength, gust strength, and time of day each point was surveyed.
+\cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the open-forest study area and the bird surveys.
 
 \subsection*{Data description}
-For purposes of our analysis, we selected a single survey from the \akepa{} time series based on broad sampling of the study area and sufficient numbers of detections.
+[[ Need to update this with extended data ]]
+
+For the purposes of our analysis we selected a single survey from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability.
 In 2002, 195 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig.
@@
 On 67 points 152 \akepa{} were detected.
-The number of detections ranged from zero to 6 (Fig.
-\ref{fig:2002countspt}).
-These data are used for both the two-stage and one-stage model-based analyses.
-\cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the open-forest study area and the bird surveys.
+The number of detections within each sampling unit ranged from zero to 6.
 
-\begin{figure}
-	\centering
-	\includegraphics[scale=0.5]{figures/2002counts_pt}
-	\caption{Counts of \akepa{} by point during the 2002 survey.}
-	\label{fig:2002countspt}
-\end{figure}
+%\begin{figure}
+%	\centering
+%	\includegraphics[scale=0.5]{figures/2002counts_pt}
+%	\caption{Counts of \akepa{} by point during the 2002 survey.}
+%	\label{fig:2002countspt}
+%\end{figure}
 
@@
 
-In this section we introduce \texttt{inlabru} syntax and provide example code to demonstrate to specify the model.  First we specify the SPDE model as
+In this section we introduce \texttt{inlabru} syntax and provide example code to demonstrate to specify the model.  First, we specify the SPDE model as
 \begin{lstlisting}
@@
 
-The data consists only of distances from the observer so we use equation \eqref{intensity-incomplete} to specify the log thinning probability function as
+Second, we specify the thinning probability function.
+The data contains only distances from the observer so we use equation \eqref{intensity-incomplete} to specify the log thinning probability function as
 \begin{lstlisting}
@@
 where \texttt{log\_halfnormal} is the log half-normal detection function $\log g(r | \sigma) = -1/2(r/\sigma)^2$ and the argument \texttt{lsig} is $\log(\sigma)$.
-The column in the dataset containing the observed distances is named \texttt{distance} amd the software will search for the appropriate data using the names of the arguments specified in this function.
+The dataset contains a column named \texttt{distance} that contains the observed distances.
+The software will search for the appropriate columns using the names of the arguments specified in this function.
 To allow the software to distinguish between names of covariates and parameters to be estimated we must also specify a model components object
@@
 The software allows user-specified names for the \texttt{f()} function known to users of \texttt{R-INLA}.
-\texttt{coordinates} is a keyword that allows users to supply their data in the form of a spatial object using the \texttt{sp} package \citep{pebesma_spatial_2005}.
+Note that \texttt{coordinates} is a keyword that allows users to supply their data in the form of a spatial object using the \texttt{sp} package \citep{pebesma_spatial_2005}.
 
@@
 where our observed data is contained in a \texttt{SpatialPointsDataFrame} object named \texttt{obs}.
-Since not all sampling units observed birds we also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sample units%.
+Since not all sampling units observed birds we also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sampling units.
 
@@
 This shows a region of high intensity in the south and much lower intensity in the north.
-The CV plot shows clearly that there is less variation in the posterior intensity field in areas with greater sampling effort.
-In general the south, with greater density of sampled locations, has a lower CV than the north, where sampling units are more spread out.
-The areas of highest CV are those regions furthest away from any sampled location, reflecting the intuitive fact that we are least certain in areas further away from the sampled locations.
+The CV plot shows that the CV in the posterior intensity field is lower areas with greater sampling effort.
+There is clear preferential sampling and in the north estimated intensities are lower which will also inflate the posterior CV.
+It is difficult to know whether the high CV in the north is primarily caused by lower sampling effort or lower estimated abundance, or some combination of the two.
 
@@
 The CV plot shows a measure of uncertainty relative to the mean intensity in each prediction grid cell.
-However, often we are interested in the uncertainty around the mean intensity on the scale of the intensity function itself - perhaps seeking to identify areas where we can confidently predict particularly high or low intensities.
-To attempt to communicate this uncertainty around the mean it is common [[references?]] to present the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
-[RJC - I don't think a reference is needed to substantiate presenting credible intervals or computing them using quantiles, particularly in a stats journal.
-AES - I was trying to find an example of someone presenting maps like these in a paper, yet to find one.
-If you know of any please let me know!]
+However, often we are interested in the uncertainty around the mean intensity on the scale of the estimated intensity itself - perhaps seeking to identify areas where we can confidently predict particularly high or low intensities.
+One approach attempt to communicate this uncertainty around the mean is to present the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
 
@@
 
-This difference becomes clear if one generates realizations of the posterior intensity field.
+This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.
 Three such realizations are in \autoref{fig:intensity-realizations}.
 There are some notable differences when compared to the mean and quantile summary maps.
-First, each realization has finer-grained spatial structure than the mean of the field.
+Each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.
 In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.
-The mean field plot hides the spatial structure the model has been able to estimate from the data.
+The summary plots hide the spatial structure the model has been able to estimate from the data.
 In general, the mean of a random field will contain less spatial structure than realisations of the field.
-[RJC - 'First' is not followed up with subsequent ordinal numbers; I suggest rewriting the above sentence to remove 'first' or add ordinal numbers below.]
 
@@
 The quantile plots are more concerning.
-As mentioned above, the temptation is to interpret these as possible intensities that produced our observed data.
+As mentioned above, the temptation is to interpret these as possible intensities that produced the observed data.
 However, they in fact show the quantiles of the \textit{marginal} probability density for each prediction location.
-This essentially treats each prediction cell as independent of the others.
-Whilst any one prediction location may be well described by such quantiles, to present them side by side in one image obscures the fact that the probability that all (or some subset of) the prediction cells would \textit{simultaneously} achieve that value will be in fact much lower than the quantile value and in many cases would be essentially zero.
-[RJC - I stumbled over this sentence several times and I think it needs to be rewritten.]  To demonstrate this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile and 12,400 for the 0.975 quantile.
+Presenting these quantiles side by side in a single map obscures the problem of multiple hypothesis testing.
+To demonstrate this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile and 12,400 for the 0.975 quantile.
 A naive (and tempting) interpretation of these numbers is as upper and lower limits of the 95\% credible interval for abundance.
-However, these abundance estimates are essentially impossible according to the estimated posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
-This emphasises that the temptation to treat these maps as showing possible intensity surfaces leads to outcomes (and possible management decisions) that are inconsistent with the very model that generated the maps.
+However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
+This emphasises that the temptation to treat these maps as showing possible intensity surfaces leads to interpretations (and possible management decisions) that are inconsistent with the very model that generated the maps.
 
@@
 For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
-
 \begin{equation*}
@@
 Because $\lambda(s)$ is stochastic we can only ever say this is true with some probability and use something akin to a confidence level by specifying $\alpha$.
-
 Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.
@@
 It is the largest possible region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
-
 \begin{figure}
@@
 \end{figure}
-
 To visualise multiple such maps we can use the excursion function $F_u^{+}(s) = \sup \{1 - \alpha ; s \in E_{u,\alpha}^+ \}$, which defines for each location something similar to a p-value.
@@
 We anticipate the excursion function will be most useful in communicating to audiences with statistical training.
-[RJC - I think statistically savvy audiences will readily grasp the excursion functions.
+% [RJC - I think statistically savvy audiences will readily grasp the excursion functions.
 Do you mean "without" statistical training?]
@@
 Another key aim of performing distance sampling surveys is to estimate the abundance of the population of interest.
-By specifying our spatial model as a log-Gaussian Cox process, the posterior intensity is a random field rather than a deterministic function.
+We describe here how to estimate the posterior for the abundance.
 Let $n$ denote the abundance of the population within the region of interest and $N$ its corresponding random variable within our Bayesian framework.
-Integrating the mean of the random field provides a point estimate for the abundance, which we term the expected abundance.
-Here we demonstrate the importance of considering the full posterior for $N$ which we call the \textit{realized} abundance.
+Integrating the mean of the random field provides a point estimate for the abundance, which we call the \textit{expected} abundance.
+Here we demonstrate the importance of considering the full posterior for $N$ which we call the \textit{realised} abundance.
 
@@
 As we mentioned above, this is the most sensible point estimate for the abundance.
-We estimate the posterior $\pi(N | \lambda)$ by a monte-carlo method.
+To communicate uncertainty around the expected abundance we estimate the posterior $\pi(N | \lambda)$ by a monte-carlo method.
 Taking $m$ monte-carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
@@
 \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
-This allows us to estimate the probability for any specific value of \textit{realized} abundance $n$ by calculating $\pi(N = n | \lambda)$.
+This allows us to estimate the probability for any specific value of realised abundance $n$ by calculating $\pi(N = n | \lambda)$.
 For reference, \autoref{fig:realized-abundance-comparison} shows a single Poisson PMF with rate parameter equal to the expected abundance.
+
+\autoref{fig:half-normal} shows a summary of the posterior for the detection function, plotting the posterior mean along with 95\% credible intervals.
+We note again that all model summaries shown here are based on monte-carlo sampling of the joint posterior of all model parameters.
+As such all the plots incorporate the uncertainty in the detection process.
 
@@
 	\includegraphics[scale=0.6]{figures/halfnormal.png}
+	\caption{Posterior detection function}
+	\label{fig:half-normal}
 \end{figure}
@@
 \bigskip
-
-TODO HERE:  Something about how this is all a one-stage model and all these plots are based on sampling from the joint posterior of all parameters.
-As such all plots can be viewed as incorporating the uncertainty from the detection model.
-This is a lovely and natural way to present results and an advantage of having a joint model.
-I'm not sure varprop can do this?
-[RJC - Varprop does do this through the posterior covariance $V_{\beta}$ matrix of the intercept, smoother and detection probability terms.
-But it doesn't do so in such an elegant way.
-Varprop is still reliant on a two-stage model.]
 

### 2019-09-02 paper.tex
add discussion

@@
 
-\begin{enumerate}
-	\item Note the model-based Bayesian hierarchical approach provides a framework for working with distance sampling data.
-This allows simultaneous estimation of parameters associated with the detection process and the spatial distribution.
-It is in principle extendible...
-somehow...
+\subsection*{General remarks}
 
-	\item Note that Bayesian distance sampling [RJC - References for this include Oedekoven et al (2014, JABES 19:219-239) and Buckland et al.
-(2015).
-We could also include Buckland et al.
-(2016, JABES 21:58-75) but it deals more with model-based design distance sampling than Bayesian approaches specifically.
-The more recent papers that reference Cornelia's paper are applied.
-It is also worth looking at Yuan's paper on point process models of blue whales.] has been done before but that the INLA approach avoids data augmentation and MCMC.
-	\item Iterated INLA is a powerful and flexible way to include non-linear components to a predictor and has applications beyond this application to distance sampling
-	\item More to do in future to test how many parameters can be included in the non-linear component and what types of non-linear components are feasible
-	\item Note similarities and differences with two-stage GAM approach
-	\item Note that presenting uncertainty in maps is difficult and excursion sets are a promising contribution relevant to many areas of spatial statistics.
-	\item Links to species "range" estimates - excursions possibly useful here.
-	\item model selection is difficult.  Oedekoven 2014 addresses using RJMCMC.  Unclear what to do with this approach.
-\end{enumerate}
+The most common method to analyse distance sampling data is a two-stage approach with a generalised additive model that includes a spatial smoothing spline.
+The approach here of specifying a point process model and including an SPDE effect can seem fundamentally different.
+However, these two approaches are related and an analysis of the \akepa{} data using the GAM approach provides similar point estimates.
+There is a correspondence between SPDEs and smoothing splines: the smoothing spline is an estimator of the posterior mean of the SPDE \citep{kimeldorf_spline_2002}.
+The SPDE and the smoothing spline play a similar role within the model and the resulting mean field map looks similar in both approaches.
+In addition, via the point process model, we have avoided the need to bin data into counts, which may at first appear as though we have avoided some loss of information.
+This is not the case since we also applied the standard distance sampling assumption of constant density within each point transect.
+This can be thought of as the point process version of binning data into counts since the location of points within transects does not inform the spatial distribution of points.
+The main advantage of the log-Gaussian Cox process is that the discretisation of space occurs at the level of approximating the GRF.
+This is in contrast to motivating the discretisation by binning the observed data which requires consideration of zero-inflation and dispersion.
+
+We also note that \texttt{inlabru} provides functionality for fitting models with multiple likelihoods \citep{bachl_inlabru_2019}.
+As such our approach is extendible to more complicated contexts such as marked point processes.
+
+We believe the excursions approach is a valuable addition to the literature on communicating uncertainty in estimated surfaces.
+For species distribution modelling this has the potential to be a useful tool for model evaluation and communication with stakeholders.
+In particular the approach requires users to decide (i) a threshold or set of threshold values of interest and (ii) an acceptable probability of error.
+We believe that, in most situations, these decisions should be made by domain experts who have an understanding of the phenomena of interest, the set of possible decisions that may be taken on the basis of the analysis and the consequences of error.
+Once these decisions have been made the statistician can then proceed accordingly, providing outputs  that inform on quantities of interest with a level of uncertainty that stakeholders have agreed is acceptable.
+We note that the excursions approach applied here has the potential to be useful in other methods of species distribution mapping, environmental statistics and spatial statistics in general.
+There is the potential to use excursions to motivate both the \textit{definition} and \textit{estimation} of concepts of interest to ecologists, such as a species' home range or the notion of occupancy.
+
+\subsection*{Possible future directions}
+
+Here we applied the iterated INLA approach in order to estimate the half-normal detection function which is non-linear in its $\log \sigma$ parameter.
+We note that this approach provides a general framework for estimating parametric functions from data and is not restricted to detection functions within a distance sampling model.
+The \texttt{inlabru} software allows users to specify the model in an intuive and modular way.
+Users write individual functions to define (possibly non-linear) model components and specify how these are combined to form the additive predictor.
+By additive we mean a predictor that is a sum of model components, and not necessarily additive in terms of parameters to be estimated.
+
+We presented a simple choice for the detection model.
+The half-normal function has only one parameter to be estimated but other, more flexible, detection functions are common in applications.
+The potential to include further flexibility in the detection model through covariates or random-effects is an open research question and would dramatically increase the usability of our approach presented here.
+Increasing the number of parameters to estimate in the non-linear model components may run into computational challenges.
+This has yet to be explored.
+
+Model selection is a general challenge for models with spatially structured random effects.
+Traditional tools of model assessment and selection do not apply in many contexts.
+For model assessment we advise a pragmatic approach based on comparing summaries of posterior samples to the observed data.
+A moodel selection procedure is useful to users to justify decisions regarding which covariates to include, whether or not to include a spatially structured random effect, and if so, what type of effect should be used.
 

### 2019-09-02 paper.tex
update abstract and title

@@
 \begin{frontmatter}
-\title{Point transect distance sampling in inlabru}
+\title{One-stage point transect distance sampling using iterated integrated nested Laplace approximations}
 
@@
 \begin{abstract}
-Point transect distance sampling in \texttt{inlabru}
+Distance sampling methods aim to estimate the size and distribution of animal populations.
+Traditionally this has been achieved using a two-stage approach, using estimates of detectability as an offset in a generalised additive model.
+One-stage approaches have typically been implemented using data augmentation with an MCMC algorithm.
+Here we present a new approach to one-stage distance sampling models, using iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations (INLA) to simultaneously estimate the detectability of animals and their spatial distribution.
+We model the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields.
+We demonstrate the benefits of the approach by applying the methods to point count survey data of Hawaiian \akepa{}.
+The Bayesian one-stage approach allows sampling from the joint posterior of all model parameters and so resulting summaries and plots naturally incorporate uncertainty in the observation process.
 \end{abstract}

### 2019-09-03 analyses/data/create_study_area.R
create bnd_forest2.csv boundary and mesh

@@ -7,18 +7,21 @@ library(inlabru)
 
 samplers <- readRDS("samplers_extended_no_crs.RDS")
 samplersnew = samplers[samplers$Stratum == "CF",]
-old_study_area = readRDS("study_area_no_crs.RDS")
-chull = gConvexHull(samplersnew)
-chull = gBuffer(chull, width = 500, capStyle = "SQUARE")
+# old_study_area = readRDS("study_area_no_crs.RDS")
+# chull = gConvexHull(samplersnew)
+# chull = gBuffer(chull, width = 500, capStyle = "SQUARE")
+# 
+# ggplot() +
+#   gg(old_study_area) +
+#   gg(chull) +
+#   gg(samplers) +
+#   gg(samplersnew, colour = "red") +
+#   coord_equal()
+# 
+# study_area_new = gUnion(old_study_area, chull)
 
-ggplot() +
-  gg(old_study_area) +
-  gg(chull) +
-  gg(samplers) +
-  gg(samplersnew, colour = "red") +
-  coord_equal()
-
-study_area_new = gUnion(old_study_area, chull)
+bnd.csv = read.csv("bnd_forest2.csv")
+study_area_new <- SpatialPolygons(list(Polygons(list(Polygon(bnd.csv)), ID=1)))
 
 ggplot() +
   gg(study_area_new) +

### 2019-09-03 analyses/fit_model.R
update fit and figures for extended boundary

@@ -68,7 +68,7 @@ fml <- coordinates + distance ~ grf +
   Intercept
 
 
-starting_values <- data.frame(grf = 0, lsig = 3.65, Intercept = 0)
+starting_values <- data.frame(grf = 0, lsig = 3.4, Intercept = 0)
 
 W <- 60   # transect radius
 distance_domain <- seq(.Machine$double.eps, W, length.out = 30)
@@ -79,7 +79,7 @@ fit <-  lgcp(components = cmp,
              # domain = list(coordinates = mesh, distance = distance_domain),
              domain = list(distance = distance_domain),
              formula = fml,
-             options = list(result = starting_values, max.iter = 30))
+             options = list(result = starting_values, max.iter = 40))
 
 summary(fit)
 

### 2019-10-10 paper.tex
cutting out excursions waffle

--- 2019-09-02 paper.tex
@@
 \begin{abstract}
-Point transect distance sampling in \texttt{inlabru}
+Distance sampling methods aim to estimate the size and distribution of animal populations.
+Traditionally this has been achieved using a two-stage approach by using first-stage estimates of detectability as an offset in a second-stage generalised additive model.
+One-stage approaches have typically been implemented using MCMC and data augmentation methods.
+Here we present a new approach to one-stage distance sampling models, using iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations (INLA) to simultaneously estimate the detectability and spatial distribution.
+We model the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields.
+We demonstrate the approach by applying the methods to point count survey data of Hawaiian \akepa{}.
+Our Bayesian one-stage approach allows sampling from the joint posterior of all model parameters and so the resulting summaries and plots naturally incorporate uncertainty in the observation model estimates.
 \end{abstract}
@@
 The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
-Distance sampling methods aim to achieve this by using a spatially explicit sampling design and estimating the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2008, buckland_distance_2015}.
+Distance sampling methods aim to achieve this by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2008, buckland_distance_2015}.
 Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.
-Probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.
+The probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.
 
-Over time interest grew in fully model-based approaches with a spatially explicit model for animal density that allow the use of non-randomized survey designs.
-This also allows observations to be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
+More recently, interest has focused on fully model-based approaches that include a spatially explicit model for animal density and allow the use of non-randomized survey designs.
+These methods also allow density to be associated with spatially-indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
+
 Model-based distance sampling has been implemented in a two-stage modelling framework.
@@
 Due to often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.
-Negative-binomial and Tweedie distributions are common choices to model the distribution of counts \citep{wood_gam_2017}.
+Negative-binomial and Tweedie distributions are common choices to model the distribution of counts.
+This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review and software package \texttt{dsm} to fit the models.
 
-A key concern with the two-stage approach is the propagation of the uncertainty in the detectability estimates to the second-stage spatial model.
+A key concern with the two-stage approach is the propagation of the uncertainty from the detectability estimates to the second-stage spatial model.
 Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units as well as the challenge of combining smoothers (a Bayesian/random effects concept) and boostraps (a frequentist concept) \citep{bravington_reliable_2018-1}.
-Instead, \cite{bravington_reliable_2018-1}  proposed avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the maximum-likelihood estimate from the first stage.
+Instead, \cite{bravington_reliable_2018-1}  proposed avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the first-stage maximum-likelihood estimate.
 
 One-stage modelling approaches to distance sampling have also been developed within a Bayesian framework.
-The first efforts in this area used data augmentation approaches for unobserved individuals or groups and were fit using Markov-chain Monte Carlo (MCMC) methods \citep{royle_hierarchical_2008, schmidt_using_2012}.
-More recently  \cite{oedekoven_bayesian_2014} provided a one-stage model that avoids data augmention by specifying a combined likelihood of the detection and spatially explicit count models and incorporated model uncertainty by applying a reversible jump MCMC algorithm.
+The first efforts in this area used data augmentation approaches to account for unobserved individuals or groups and were fit using Markov-chain Monte Carlo (MCMC) methods \citep{royle_hierarchical_2008, schmidt_using_2012}.
+More recently  \cite{oedekoven_bayesian_2014} provided a one-stage model that avoids data augmention by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty by applying a reversible jump MCMC algorithm.
 
-The reliance on MCMC for one-stage modelling presents a challenge for using spatially structured random effects, which can be computationally expensive.
+The reliance on MCMC for one-stage modelling presents a challenge for using spatially-structured random effects, which can be computationally expensive.
 \citet{yuan_point_2017} presented a one-stage approach fitted using integrated nested Laplace approximations (INLA).
-They considered line-transect distance sampling and simulateously estimated the detectability and spatial distribution of animals.
+Applying their method to line-transect distance sampling, they simulateously estimated the detectability and spatial distribution of animals.
 Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF).
@@
 Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.
-We replace the SPDE specification of the detection function with a parametric family of detection functions that are more familiar to users of standard distance sampling methods.
 Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
+
 Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
 However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004, stoyan_remark_1982, hogmander_random_1991}.
-We also present the first use of a sparse GMRF spatial random effect with point transect distance sampling data.
+We also further develop the point process viewpoint to allow point process models to be applied to data with incomplete location information.
+It is common for distance sampling surveys to only record the distance to the observer and the observer location.
+We show how point process methods can handle this type of data even though the exact location of the point is not known.
+
+We replace the SPDE specification of the detection function in \cite{yuan_point_2017} with a parametric family of detection functions that are more familiar to users of standard distance sampling methods.
+This results in components of the additive predictor that are non-linear in their parameters.
+To deal with this we use a method of iterated model fits based on a Taylor expansion of the non-linear model components.
+We name this approach \textit{iterated INLA}.
+
+To the best of our knowledge we also demonstrate the first use of a sparse GMRF spatial random effect with point transect distance sampling data.
 The sparsity of the GMRF is useful for computational efficiency and the finite element mesh motivated by the SPDE allows for efficient handling of complex spatial domains such as coastlines \citep{simpson_going_2016}.
+The flexibility of the mesh allows a higher-resolution discretisation of space where complexity requires it but also lower resolutions in areas where this is not required.
+As collecting high-resolution data and incorporating complex spatial domains (broadly defined here as domains where regular lattice discretisations of space lead to a computational bottleneck) becomes more common we believe this approximate-Bayesian GMRF approach will become a valuable addition to the spatial ecology toolbox.
 
-To demonstrate the benefits of our approach we discuss issues regarding model evaluation and communication of results that are common across the field of species distribution modelling.
-We discuss potentially misleading summary maps of the posterior intensity process and propose an alternative based on excursion sets \citep{bolin_excursion_2015, bolin_calculating_2018}.
-A key theme in our case-study is to highlight the importance of considering \textit{realizations} of random quantities rather than rely too much on summaries such as expectations and quantiles.
-The one-stage Bayesian approach is invaluable since model summaries are based on monte-carlo sampling from the joint posterior of all model parameters.
-The spatial process and observation process are estimated jointly and so figures and summaries produced using posterior samples will naturally average over the uncertainty in the observation processes.
-This is an advantage over the two-stage approach even if uncertainty has been effectively propagated in some way.
+% DRASTICALLY REWORK NEXT TWO PARAGRAPHS
 
-The article proceeds as follows.
-First, we present point transect distance sampling as a thinned log-Gaussian Cox process, describing how this can be formulated as a latent Gaussian model with a spatial GMRF random effect that can be fitted using INLA.
-We also show how the point process model can be applied in the case of incomplete location information where only the distance from the observer is recorded, not the exact location of the animal.
-This is a common data format for distance sampling data.
-Second, we describe how to specify and fit the joint model using an iterative approach to fitting models with components that are non-linear in their parameters - we call this the "iterated INLA" approach.
-This allows us to avoid the SPDE approach to the detection function, instead using the parametric families common in the literature.
-Third, we introduce our case-study of a point transect distance sampling survey of the Hawaiian \akepa{}, an endangered bird of the honeycreeper tribe.
-We show how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package with additional functionality for the iterated INLA approach.
-The paper includes example code to specify the model as well as full scripts for the analysis in the supplementary materials.
-We focus model evaluation and discussion on the effective understanding and communication of uncertainty in the posterior intensity and abundance estimates, which are typically of greatest interest to stakeholders.
+We apply the methods to point transect survey data of Hawaiian \akepa{}, an endangered bird that is the focus of targeted conservation efforts.
+We simultaneously estimate the detectability and spatial distribution of \akepa{} in a region of Hawaii island.
+Model evaluation focuses on uncertainty in model outputs that are based on sampling from the joint posterior of all model parameters.
+This provides a natural way to produce useful outputs from the analysis that average over detection model uncertainty.
+The posterior intensity field is investigated in detail and its uncertainty visualised using excursion sets and excursion functions \citep{bolin_excursion_2015, bolin_calculating_2018}, approaches which deal with the problem of multiple hypothesis testing when mapping prediction quantiles.
+
+The article proceeds as follows:
+\begin{enumerate}[(i)]
+\item We present point transect distance sampling as a thinned log-Gaussian Cox process, describing how this can be formulated as a latent Gaussian model with a spatial GMRF random effect that can be fitted using INLA.
+We also show how the point process model can be applied in the case of incomplete location information.
+\item We describe how the model can be specified in a modified Poisson log-likelihood form and fitted using the iterated INLA approach.
+\item A case-study using point transect distance sampling survey of the Hawaiian \akepa{}.
+We give example code showing how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package with additional functionality for iterated INLA.
+\end{enumerate}
+Full scripts for the analysis can be found in the supplementary materials.
 
@@
 We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(s)$.
-The log-Gaussian Cox process is a flexible approach that can include spatially structured random effects on the intensity process to account for unexplained heterogeneity not captured by fixed effect covariates.
+The log-Gaussian Cox process is a flexible approach that can include spatially structured random effects on the intensity process to account for unexplained heterogeneity not captured by fixed-effect covariates.
 
@@
 Without such an assumption detectability and intensity are confounded.
-The standard assumption of distance sampling is that the intensity is constant with respect to changes in $r(s)$.
+The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(s)$.
 Thus any observed differences from uniformity can be attributed to detectability and not variation in the intensity.
@@
 For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
-The assumption of non-overlapping survey regions can be relaxed by including extra information such as the time of each observation.
-For simplicity we do not consider this here.
+The assumption of non-overlapping survey regions can be relaxed by including extra information such as the time of each observation but, for simplicity, we do not consider this here.
 The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for $s \in \Omega_k$.
 
-The thinne log-Gaussian Cox process likelihood of observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
+The thinned log-Gaussian Cox process likelihood with observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
 \begin{equation}
@@
 
-In this section we introduce \texttt{inlabru} syntax and provide example code to demonstrate to specify the model.  First, we specify the SPDE model as
+In this section we introduce \texttt{inlabru} syntax and provide example code to demonstrate how to specify the model.  First, we specify the SPDE model component as
 \begin{lstlisting}
@@
 \end{lstlisting}
-which uses the penalised complexity priors for the hyperparameters of the GMRF as introduced by \cite{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$ where $\rho$ is the range of the Mat\'ern covariance.
+which uses penalised complexity priors for the hyperparameters of the GMRF \citep{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$ where $\rho$ is the range of the Mat\'ern covariance.
+The value of 130 was selected based on the minimum distance between sampling locations.
 
-Second, we specify the thinning probability function.
+Second, we specify the thinning probability function model component.
 The data contains only distances from the observer so we use equation \eqref{intensity-incomplete} to specify the log thinning probability function as
@@
 The software will search for the appropriate columns using the names of the arguments specified in this function.
-To allow the software to distinguish between names of covariates and parameters to be estimated we must also specify a model components object
+To allow the software to distinguish between named covariates and parameters - which both appear as arguments of the function to be estimated - we must also specify a model components object
 \begin{lstlisting}
@@
 
-The model formula object is
+We can now specify the predictor formula using the components defined above.  
 \begin{lstlisting}
@@
 where our observed data is contained in a \texttt{SpatialPointsDataFrame} object named \texttt{obs}.
-Since not all sampling units observed birds we also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sampling units.
+We also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sampling units.
+Full scripts for the analysis are available in the supplementary materials.
 
@@
 The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.
-This shows a region of high intensity in the south and much lower intensity in the north.
+This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary materials).
 The CV plot shows that the CV in the posterior intensity field is lower areas with greater sampling effort.
-There is clear preferential sampling and in the north estimated intensities are lower which will also inflate the posterior CV.
-It is difficult to know whether the high CV in the north is primarily caused by lower sampling effort or lower estimated abundance, or some combination of the two.
-
+There is clear preferential sampling and in the north estimated intensities are lower which also contributes to larger CV values.
 \begin{figure}
@@
 
-The CV plot shows a measure of uncertainty relative to the mean intensity in each prediction grid cell.
-However, often we are interested in the uncertainty around the mean intensity on the scale of the estimated intensity itself - perhaps seeking to identify areas where we can confidently predict particularly high or low intensities.
-One approach attempt to communicate this uncertainty around the mean is to present the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
-
+The fact that CV values are influenced both by sampling effort and the estimated intensity means other methods to visualise uncertainty in the posterior intensity field are also often considered.
+A common approach is to map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
 \begin{figure}
@@
 \end{figure}
+Although commonly used, we believe quantile plots like those in \autoref{fig:intensity-quantiles} to be challenging to interpret and potentially misleading.
+The temptation is to interpet these plots as showing possible intensity surfaces that produced the observed data.
+Presenting these quantiles side by side in a single map obscures the problem of multiple hypothesis testing.
+To demonstrate the fallacy of this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile and 12,400 for the 0.975 quantile.
+A naive (and tempting) interpretation of these numbers is as upper and lower limits of the 95\% credible interval for abundance.
+However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
 
-We believe quantile plots like those in \autoref{fig:intensity-quantiles} to be challenging to interpret and potentially misleading.
-The temptation is to interpet these plots as showing possible intensity surfaces that produced the observed data.
-We believe this misconception would be extremely common in particular for non-statistically trained audiences.
-The maps in \autoref{fig:intensity-quantiles} in fact only show \textit{summary statistics} of a posterior intensity which is specified (in our model) as a log-Gaussian random field.
-
-This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.
+As an alternative to the quantile maps we use excursion sets and excursion functions \citep{bolin_excursion_2015}.
+This is a recent addition to the literature so we take some time here to define and elaborate on the concepts used.
+This approach to representing variability in stochastic processes specifically considers the joint probability of events across multiple locations.
+The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e.
+the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$.
+For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
+% Somehow keep this information somewhere in the paper:
+%This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.
 Three such realizations are in \autoref{fig:intensity-realizations}.
@@
 In general, the mean of a random field will contain less spatial structure than realisations of the field.
-
-\begin{figure}
-	\includegraphics[scale=0.35]{figures/intensity_realized.png}
-	\caption{Three realizations of the posterior intensity field}
-	\label{fig:intensity-realizations}
-\end{figure}
-
-The quantile plots are more concerning.
-As mentioned above, the temptation is to interpret these as possible intensities that produced the observed data.
-However, they in fact show the quantiles of the \textit{marginal} probability density for each prediction location.
-Presenting these quantiles side by side in a single map obscures the problem of multiple hypothesis testing.
-To demonstrate this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile and 12,400 for the 0.975 quantile.
-A naive (and tempting) interpretation of these numbers is as upper and lower limits of the 95\% credible interval for abundance.
-However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
-This emphasises that the temptation to treat these maps as showing possible intensity surfaces leads to interpretations (and possible management decisions) that are inconsistent with the very model that generated the maps.
-
-As an alternative, we suggest using excursion sets and excursion functions \citep{bolin_excursion_2015}.
-This is a recently developed approach to representing uncertainty in stochastic processes that specifically considers joint probability of events across multiple locations.
-The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e.
-the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$.
-For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
+%
+%\begin{figure}
+%	\includegraphics[scale=0.35]{figures/intensity_realized.png}
+%	\caption{Three realizations of the posterior intensity field}
+%	\label{fig:intensity-realizations}
+%\end{figure} 
 \begin{equation*}
@@
 \end{equation*}
-It is important to note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.
-This means that when considering random quantities such as $\lambda(s)$ the definition of the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.
-Because $\lambda(s)$ is stochastic we can only ever say this is true with some probability and use something akin to a confidence level by specifying $\alpha$.
+Note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.
+As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.
+Negative excursion sets can be similarly defined.
 Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.
@@
 This figure can be interpreted in a natural way.
-It is the largest possible region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
+It is the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
 \begin{figure}
@@
 The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursions} (right panel).
+This figure can also be interpreted naturally.
+It shows the largest possible probability for each location to be a member of a set in which the intensity exceeds a threshold simultaneously across all locations in the set.
+It is clear from the figure that regions on the edge of the excursion set would be included if the $\alpha$ value were allowed to increase slightly.
+However, for regions in the north there is essentially no probability level for which those locations would be included in $A_u^{+}(\lambda)$.
+In the above we used 1 bird per hectare as an illustrative threshold.
+In applications multiple thresholds can be considered - based on the research question at hand.
+An attractive feature of the excursions approach is that it forces researchers to identify values that are of interest and choose an acceptable probability of error given the wider context of the research.
 
-This figure can also be interpreted naturally.
-It shows the largest possible probability for each location to be a member of a set for in which the intensity exceeds a threshold simultaneously across all locations in the set.
-Regions on the edge of the excursion set would be included if the $\alpha$ value were allowed to increase slightly.
-However, for regions in the north there is essentially no probability level for which those locations would be included in $A_u^{+}(\lambda)$.
-
-A key feature of these plots is that they display either discrete sub-regions of the area of interest (for excursion sets) or the probability of membership of such discrete sub-regions (for excursion functions).
+%A key feature of these plots is that they display either discrete sub-regions of the area of interest (for excursion sets) or the probability of membership of such discrete sub-regions (for excursion functions).
 We believe the risk of interpreting these maps as intensities is much lower than for the quantile plots.
@@
 We describe here how to estimate the posterior for the abundance.
-Let $n$ denote the abundance of the population within the region of interest and $N$ its corresponding random variable within our Bayesian framework.
-Integrating the mean of the random field provides a point estimate for the abundance, which we call the \textit{expected} abundance.
-Here we demonstrate the importance of considering the full posterior for $N$ which we call the \textit{realised} abundance.
-
-Let $N = \int_{\Omega}\lambda(s)\mathrm{d}s$.
-Note that $N$ is a random variable and the integral of $\mathbb{E}[\lambda(s)]$ is equivalent to $\mathbb{E}(N)$ by Fubini's theorem, hence the term expected abundance.
-As we mentioned above, this is the most sensible point estimate for the abundance.
-To communicate uncertainty around the expected abundance we estimate the posterior $\pi(N | \lambda)$ by a monte-carlo method.
+Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.
+Integrating the mean of the posterior intensity field provides a point estimate for the abundance, the \textit{expected} abundance.
+We approximate the posterior $\pi(N | \lambda)$ by a monte-carlo method.
 Taking $m$ monte-carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
-This is well-defined since $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
+Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
 \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
@@
 For reference, \autoref{fig:realized-abundance-comparison} shows a single Poisson PMF with rate parameter equal to the expected abundance.
+We note here that in a GAM approach it is common to only report the expected abundance and the error in expected abundance.
+This does not include the additional variability that comes from the Poisson distribution with a rate parameter equal to the expected abundance.
 
+Finally, we show that the method has simultaneusly identified a plausible detection function.
 \autoref{fig:half-normal} shows a summary of the posterior for the detection function, plotting the posterior mean along with 95\% credible intervals.
-We note again that all model summaries shown here are based on monte-carlo sampling of the joint posterior of all model parameters.
+We note again that all model summaries shown in this section are based on monte-carlo sampling of the joint posterior of all model parameters.
 As such all the plots incorporate the uncertainty in the detection process.
-
 \begin{figure}

### 2019-10-10 paper.tex
more waffle cut

@@
 %
-%\begin{figure}
-%	\includegraphics[scale=0.35]{figures/intensity_realized.png}
-%	\caption{Three realizations of the posterior intensity field}
-%	\label{fig:intensity-realizations}
-%\end{figure} 
+
 \begin{equation*}
@@
 
-%A key feature of these plots is that they display either discrete sub-regions of the area of interest (for excursion sets) or the probability of membership of such discrete sub-regions (for excursion functions).
-We believe the risk of interpreting these maps as intensities is much lower than for the quantile plots.
-The fact that it is possible to estimate discrete regions may also be useful to communicate results to conservation practitioners, land and resource managers and policy decision-makers.
-We anticipate the excursion function will be most useful in communicating to audiences with statistical training.
-% [RJC - I think statistically savvy audiences will readily grasp the excursion functions.
-Do you mean "without" statistical training?]
+Given the issues of interpreting maps that simultaneously show summaries of the posterior random field it can often be informative to look directly at realisations themselves.
+Three such realizations are in \autoref{fig:intensity-realizations}.
+\begin{figure}
+	\includegraphics[scale=0.35]{figures/intensity_realized.png}
+	\caption{Three realizations of the posterior intensity field}
+	\label{fig:intensity-realizations}
+\end{figure}
+There are some notable differences when compared to the mean and quantile summary maps.
+Note each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.
+In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only considered the posterior mean of the field.
+The summary maps hide the spatial structure the model has been able to estimate from the data.
 
@@
 Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.
-Integrating the mean of the posterior intensity field provides a point estimate for the abundance, the \textit{expected} abundance.
+Integrating the mean of the posterior intensity field provides a point estimate for the abundance - the expected abundance.
 We approximate the posterior $\pi(N | \lambda)$ by a monte-carlo method.
@@
 For reference, \autoref{fig:realized-abundance-comparison} shows a single Poisson PMF with rate parameter equal to the expected abundance.
-We note here that in a GAM approach it is common to only report the expected abundance and the error in expected abundance.
-This does not include the additional variability that comes from the Poisson distribution with a rate parameter equal to the expected abundance.
-
-Finally, we show that the method has simultaneusly identified a plausible detection function.
-\autoref{fig:half-normal} shows a summary of the posterior for the detection function, plotting the posterior mean along with 95\% credible intervals.
-We note again that all model summaries shown in this section are based on monte-carlo sampling of the joint posterior of all model parameters.
-As such all the plots incorporate the uncertainty in the detection process.
 \begin{figure}
@@
 \end{figure}
-
 \begin{figure}
@@
 \end{figure}
+We note here that in a GAM approach it is common to only report the estimated expected abundance and the uncertainty in this estimate.
+This does not include the additional variability that comes from the Poisson distribution with a rate parameter equal to the expected abundance.
 
+Finally, we show that the one-stage model has estimated a plausible detection function.
+\autoref{fig:half-normal} shows a summary of the posterior for the detection function, plotting the posterior mean along with 95\% credible intervals.
+We note again that all model summaries shown in this section are based on monte-carlo sampling of the joint posterior of all model parameters.
+As such all the plots incorporate the uncertainty in the detection process.
 \begin{figure}
@@
 The SPDE and the smoothing spline play a similar role within the model and the resulting mean field map looks similar in both approaches.
-In addition, via the point process model, we have avoided the need to bin data into counts, which may at first appear as though we have avoided some loss of information.
-This is not the case since we also applied the standard distance sampling assumption of constant density within each point transect.
-This can be thought of as the point process version of binning data into counts since the location of points within transects does not inform the spatial distribution of points.
-The main advantage of the log-Gaussian Cox process is that the discretisation of space occurs at the level of approximating the GRF.
-This is in contrast to motivating the discretisation by binning the observed data which requires consideration of zero-inflation and dispersion.
-
+In addition, it can appear as though the point process model avoids the need to bin data into counts and so avoids some loss of information.
+This is not the case since we also applied the standard distance sampling assumption of constant intensity within each point transect.
+This can be thought of as the point process version of binning data into counts since the location of points within transects does not inform the spatial distribution of points across the whole study region.
+The main advantage of the log-Gaussian Cox process is that the discretisation of space occurs during the step of approximating the GRF.
+This is in contrast to motivating the discretisation by the act of binning the data into counts so as to create a response variable that can be analysed using a GAM model.
 We also note that \texttt{inlabru} provides functionality for fitting models with multiple likelihoods \citep{bachl_inlabru_2019}.
@@
 
-\subsection*{Possible future directions}
+\subsection*{Possible future research}
 
 Here we applied the iterated INLA approach in order to estimate the half-normal detection function which is non-linear in its $\log \sigma$ parameter.
-We note that this approach provides a general framework for estimating parametric functions from data and is not restricted to detection functions within a distance sampling model.
+We note that this approach provides a general framework for estimating parametric functions from data and is not restricted to detection functions in a distance sampling model.
 The \texttt{inlabru} software allows users to specify the model in an intuive and modular way.

### 2019-11-19 paper-v2.tex
new paper structure outline

--- 2019-09-02 paper.tex
@@
 \usepackage{amsmath}
+\usepackage{enumerate}
 \usepackage{caption}
@@
 \begin{abstract}
-Distance sampling methods aim to estimate the size and distribution of animal populations.
-Traditionally this has been achieved using a two-stage approach, using estimates of detectability as an offset in a generalised additive model.
-One-stage approaches have typically been implemented using data augmentation with an MCMC algorithm.
-Here we present a new approach to one-stage distance sampling models, using iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations (INLA) to simultaneously estimate the detectability of animals and their spatial distribution.
-We model the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields.
-We demonstrate the benefits of the approach by applying the methods to point count survey data of Hawaiian \akepa{}.
-The Bayesian one-stage approach allows sampling from the joint posterior of all model parameters and so resulting summaries and plots naturally incorporate uncertainty in the observation process.
+A New Abstract Goes Here   
 \end{abstract}
@@
 
-The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
-Distance sampling methods aim to achieve this by using a spatially explicit sampling design and estimating the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2008, buckland_distance_2015}.
-Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.
-Probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.
+\begin{itemize}
+	\item Estimating population size and spatial distribution is an important question in ecology and conservation
+	\item Here we present the Akepa data:
+		\begin{itemize}
+			\item endangered
+			\item subject of conservation efforts
+			\item require monitoring to understand changes in overall abundance and the spatial distribution to target conservation efforts
+		\end{itemize}
+	\item Estimating the population size and spatial distribution of Akepa is challenging because
+		\begin{enumerate}[(i)]
+			\item cannot do a census (large numbers, dense forest, logistical constraints) and so must subsample in space and time (describe the HFBS design)
+			\item imperfect detection when we do sample - a complex observation process (describe point transect distance sampling approach)
+			\item interest in generating spatial predictions but we usually have missing covariates and use spatial random effects (No covariate for north-south in Akepa data, other studies ambiguous)
+			\item complex models that are difficult to communicate with stakeholders such as conservation managers and policy officers (communicating uncertainty in abundance estimation and prediction maps)
+		\end{enumerate}
+\end{itemize}
 
-Over time interest grew in fully model-based approaches with a spatially explicit model for animal density that allow the use of non-randomized survey designs.
-This also allows observations to be associated with spatially indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
-Model-based distance sampling has been implemented in a two-stage modelling framework.
-In the first stage, detectability is estimated within each sampling unit.
-In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset.
-Due to often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.
-Negative-binomial and Tweedie distributions are common choices to model the distribution of counts \citep{wood_gam_2017}.
+In this paper we present an analysis of the Akepa data that seeks to address these issues
 
-A key concern with the two-stage approach is the propagation of the uncertainty in the detectability estimates to the second-stage spatial model.
-Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units as well as the challenge of combining smoothers (a Bayesian/random effects concept) and boostraps (a frequentist concept) \citep{bravington_reliable_2018-1}.
-Instead, \cite{bravington_reliable_2018-1}  proposed avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the maximum-likelihood estimate from the first stage.
+\begin{enumerate}[(i)]
+	\item a spatial point process formulation of distance sampling
+	\item a one-stage estimation process to simultaneously estimate the spatial model and the observation process
+	\item a computationally efficient Gaussian Markov random field effect based on a stochastic partial differential equation approach
+	\item model evaluation and communication based on join posterior sampling and excursions package
+\end{enumerate}
 
-One-stage modelling approaches to distance sampling have also been developed within a Bayesian framework.
-The first efforts in this area used data augmentation approaches for unobserved individuals or groups and were fit using Markov-chain Monte Carlo (MCMC) methods \citep{royle_hierarchical_2008, schmidt_using_2012}.
-More recently  \cite{oedekoven_bayesian_2014} provided a one-stage model that avoids data augmention by specifying a combined likelihood of the detection and spatially explicit count models and incorporated model uncertainty by applying a reversible jump MCMC algorithm.
+We work in a Bayesian context here but note that maximum-likelihood frameworks could also be used.
 
-The reliance on MCMC for one-stage modelling presents a challenge for using spatially structured random effects, which can be computationally expensive.
-\citet{yuan_point_2017} presented a one-stage approach fitted using integrated nested Laplace approximations (INLA).
-They considered line-transect distance sampling and simulateously estimated the detectability and spatial distribution of animals.
-Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF).
-They consider the data as a realisation of a spatio-temporal log-Gaussian Cox point process, an approach that avoids the need to bin data into counts and sidesteps the issues of zero-inflation and dispersion.
-The observation process is formulated as a thinning of the log-Gaussian Cox process.
-Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.
-
-Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.
-We replace the SPDE specification of the detection function with a parametric family of detection functions that are more familiar to users of standard distance sampling methods.
-Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
-Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
-However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004, stoyan_remark_1982, hogmander_random_1991}.
-We also present the first use of a sparse GMRF spatial random effect with point transect distance sampling data.
-The sparsity of the GMRF is useful for computational efficiency and the finite element mesh motivated by the SPDE allows for efficient handling of complex spatial domains such as coastlines \citep{simpson_going_2016}.
-
-To demonstrate the benefits of our approach we discuss issues regarding model evaluation and communication of results that are common across the field of species distribution modelling.
-We discuss potentially misleading summary maps of the posterior intensity process and propose an alternative based on excursion sets \citep{bolin_excursion_2015, bolin_calculating_2018}.
-A key theme in our case-study is to highlight the importance of considering \textit{realizations} of random quantities rather than rely too much on summaries such as expectations and quantiles.
-The one-stage Bayesian approach is invaluable since model summaries are based on monte-carlo sampling from the joint posterior of all model parameters.
-The spatial process and observation process are estimated jointly and so figures and summaries produced using posterior samples will naturally average over the uncertainty in the observation processes.
-This is an advantage over the two-stage approach even if uncertainty has been effectively propagated in some way.
-
-The article proceeds as follows.
-First, we present point transect distance sampling as a thinned log-Gaussian Cox process, describing how this can be formulated as a latent Gaussian model with a spatial GMRF random effect that can be fitted using INLA.
-We also show how the point process model can be applied in the case of incomplete location information where only the distance from the observer is recorded, not the exact location of the animal.
-This is a common data format for distance sampling data.
-Second, we describe how to specify and fit the joint model using an iterative approach to fitting models with components that are non-linear in their parameters - we call this the "iterated INLA" approach.
-This allows us to avoid the SPDE approach to the detection function, instead using the parametric families common in the literature.
-Third, we introduce our case-study of a point transect distance sampling survey of the Hawaiian \akepa{}, an endangered bird of the honeycreeper tribe.
-We show how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package with additional functionality for the iterated INLA approach.
-The paper includes example code to specify the model as well as full scripts for the analysis in the supplementary materials.
-We focus model evaluation and discussion on the effective understanding and communication of uncertainty in the posterior intensity and abundance estimates, which are typically of greatest interest to stakeholders.
-
-\section*{Distance Sampling as a Thinned Point Process}
-
-We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(s)$.
-The log-Gaussian Cox process is a flexible approach that can include spatially structured random effects on the intensity process to account for unexplained heterogeneity not captured by fixed effect covariates.
-
-\sloppy For the case with imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected})$.
-A key property of the log-Gaussian Cox process is that a realisation of a point process with intensity process $\lambda(s)$ that is thinned by thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.
-
-Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance.
-The relevant distance depends on the survey type.
-For line transects the perpendicular distance to the transect line is used.
-For point transects the distance to the observer is used.
-For the remainder of the paper we assume a point transect survey design.
-
-The thinning probability function $g(s)$ is specified as a member of a parametric family of functions.
-In this paper we demonstrate our approach using the half-normal formulation, although others such as hazard rate, are also common.
-If $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated.
-The parameter $\sigma$ can only be estimated if an assumption is made about the intensity of the animal locations.
-Without such an assumption detectability and intensity are confounded.
-The standard assumption of distance sampling is that the intensity is constant with respect to changes in $r(s)$.
-Thus any observed differences from uniformity can be attributed to detectability and not variation in the intensity.
-
-A point transect distance sampling survey consists of a set of $K$ sampling units.
-The $k$-th sample unit we denote $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region $\Omega = \cup_{k=1}^K \Omega_k$.
-For simplicity we assume that all sampling units are discs with radius $W$ and $\Omega_j \cap \Omega_k = \emptyset$ for all $j \neq k$.
-The thinning probability function $g(s)$ is defined relative to the positions of the sample units.
-The probability of observing a point at location $s$ in the $k$-th sampling unit $\Omega_k$ we denote $g_k(s)$.
-The probability of observing a point outside the surveyed region is zero.
-Since the surveyed regions are non-overlapping each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.
-For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
-The assumption of non-overlapping survey regions can be relaxed by including extra information such as the time of each observation.
-For simplicity we do not consider this here.
-The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for $s \in \Omega_k$.
-
-The thinne log-Gaussian Cox process likelihood of observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
-\begin{equation}
-\label{lgcp-likelihood}
-\pi(\bm{Y} | \theta, \phi) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
-\end{equation}
-The integral component of the likelihood does not usually have analytical solutions.
-Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach of \cite{simpson_going_2016} for a fully observed point pattern.
-
-To evaluate the integral we introduce polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.
-In general the thinning function depends only on $r$ and not on $\theta$.
-We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$.
-It follows that the integral can be rewritten as a sum of weighted one-dimensional integrals $\sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$ using the assumption that $\lambda(s)$ is constant within each sampling unit.
-For each sampling unit we approximate the one-dimensional integral using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
-This gives
-\begin{equation*}
-	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj})
-\end{equation*}
-where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(s_{kj}) = \lambda(s_k) g_k(r_{kj})$.
-
-To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.
-Similarly let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$
-
-Then the approximate log-likelihood is
-
-\begin{equation}
-\label{approx-log-likelihood}
-	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}
-\end{equation}
-
-This approximation can be written as a modified Poisson likelihood.  To see this let $\exp \eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
-$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $y = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.
-Then the approximate likelihood becomes
-\begin{equation}
-\pi(\bm{Y}) \approx \prod_{i=1}^{KM + n} \eta_i^{y_i}\exp(-\alpha_i\eta_i)
-\end{equation}
-
-This is similar to a Poisson likelihood with an offset term equal to $\log\alpha$ and can be specified using the Poisson likelihood implementation available in \texttt{R-INLA}.
-
-\section*{Intensities for incomplete data}
-
-In the above we assume the data are complete records of animal location.
-However, in many distance sampling surveys only the location of the observer and the distance to the observer are recorded.
-Data of this type can be analysed within a point process framework by deriving the appropriate intensity function for the incomplete data.
-
-Using the polar coordinates notation given above, for a detected point at location $s_k(r, \theta) \in \Omega_k$ we consider the case where we observe $r$ but not $\theta$.
-It follows that the intensity for points at a distance $r$ observed within sampling unit $\Omega_k$ is
-\begin{align}
-\label{intensity-incomplete}
-\tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(s)g_k(s)\mathrm{d}s \nonumber \\
-&= 2\pi\lambda(s_k)rg_k(r)
-\end{align}
- where $c_k(r)$ is a circle of radius $r$ centred at $s_k$ and the second line follows from changing to polar coordinates and the assumption of constant intensity within $\Omega_k$.
-This intensity differs from the full data case by including the $2\pi$  term to account for the fact that we do not observe $\theta$ and the additional $r$ term to account for the increasing area surveyed at larger distances.
-
-\section*{Modelling $\lambda(s)$ using the stochastic partial differential equation approach}
-
-To model the intensity of animal locations we use a combination of fixed effects of spatial covariates and a spatially structured random effect.
-This allows us to associate distance sampling data with environmental and ecological conditions that may be relevant to understanding the ecology of the species of interest.
-For cases where we have insufficient covariates to fully explain the location of animals the spatially structured random effect accounts for additional heterogeneity in the intensity and is required to assume conditional independence between points, given the intensity.
-
-The log-intensity of animal locations is given by
-\begin{equation*}
-\log \lambda(s) = \beta_0 + \sum_v \beta_v z_v(s) + \xi(s)
-\end{equation*}
-where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatially indexed covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean Gaussian random field with Mat\'ern covariance
-\begin{equation}
-C(s_1,s_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|s_1-s_2\|)^{\nu}K_\nu(\kappa \|s_1-s_2\|)
-\end{equation}
-where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.
-All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which controls the mean-square differentiability of the process.
-We set $\nu = 1$ which is the default value in \texttt{R-INLA}.
-[[Would like to justify $\nu = 1$ somehow?]]
-
-We use a Gaussian Markov random field approximation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a stochastic partial differential equation.
-Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
-The parameters $\xi_1, \ldots, \xi_L$ form a Gaussian Markov random field with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a stochastic partial differential equation.
-The precision parameter $\tau$ controls the marginal variance of the field and $\kappa$ is an inverse range parameter that controls the rate of decay of the covariance as distance between locations increases.
-Further details of this Gaussian Markov random field approximation are given in the supplementary materials.
-
-\section*{Thinned Intensities}
-
-The above specifies a model for all animal locations in the area of interest.
-The log-intensity for observed points is $\log\tl(s) = \log\lambda(s) + \log g(s)$.
-This presents a problem since $\log g(s)$ is not typically linear in its parameters.
-However, for any single set of parameter values $g(s)$ can be viewed as acting like an offset.
-If we could specify $g(s)$ as known then we could proceed with an inference approach suitable for additive linear predictors.
+\section{Study design}
 
 \begin{itemize}
- \item Description of iterated INLA here
+	\item describe survey protocol in detail
+	\item mini lit review - DS, 2-stage, Joyce
 \end{itemize}
 
-\section*{Case-Study:  Hawaiian Akepa Survey}
+\section{Distance Sampling as a Thinned Point Process}
 
-% RJC 09/08/2019 Added text describing the akepa data, survey method, key questions and what we care about from ecology/conservation point of view.
-This text provides a background on Hawaiian forest birds, why it is important to monitor their populations and the need for unbiased and precise estimates.
-I am wondering if we need to strengthen the 'key questions' text.
+\section{Iterated INLA}
 
-\bigskip
+\begin{itemize}
+	\item mini lit review - communicating uncertainty in spatial models w random effects
+	(should this really be here??  Think about moving)
+\end{itemize}
 
-% RJC 09/08/2019 Added case study text between % addtextstart and % addtextend markers.
-% addtextstart
-In this section, we apply the above methods to point transect data on the \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}), an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.
-Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}.
-Information from the HFBS was used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge that was the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
-Due to the broad-scale coverage and robust design HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.
+\section{GMRF + SPDE}
 
-During the 20th century \akepa{} declined dramatically due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994},  mosquito-transmitted avian diseases \citep{pratt_avifaunal_1994, atkinson_wildlife_1995}, introduced predators \citep{lepson_akepa_1997}, and food resources competitors \citep{lepson_akepa_1997}.
-\Akepa{} has a global abundance of 16,248 (95\%CI 10,074\textendash25,198) birds that has been restricted to five spatially distinct populations \citep{judge_akepa_2018}.
-Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau), supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}.
-Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern.
-Moreover, having unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.
+\section{Results}
 
-\subsection*{Study area and sampling method}
+\section{Model Evaluation}
 
-[[Need to update this and the figure with the extended data area]]
+\begin{itemize}
+	\item Posterior sampling
+	\item Barplot for random effect
+	\item Excursions
+\end{itemize}
 
-Hakalau was established in 1985 to conserve 15,390-ha of montane forest habitat for native forest birds and rainforest plants.
-Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance.
-Survey points were established along 15 transects following a systematic, random design with points approximately 150 m apart on transects located either 500 or 1,000 m apart.
-Following \cite{camp_population_2010, camp_statespace_2016}, we restrict our study area to the 3,061 ha open-forest stratum of Hakalau.
-The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.
-To the north the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig.
-\ref{fig:2002studyareapointspt}).
-The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of their study area.
-The west side of the study area is bounded by pasture that is dominated by grass and is unsuitable habitat for \akepa.
+\section{Discussion}
 
-\begin{figure}
-	\centering
-	\includegraphics[scale=0.5]{figures/2002studyareapoints_pt}
-	\caption{Study area (light blue) showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.}
-	\label{fig:2002studyareapointspt}
-\end{figure}
-
-Surveys used point-transect methods, recording horizontal distances from survey points to individually detected birds.
-Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength \textgreater Baufort scale 3).
-During 8-min counts trained observers recorded the species, exact distance to the nearest meter and detection type for each bird detected, along with the sampling conditions cloud cover, rain, wind strength, gust strength, and time of day each point was surveyed.
-\cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the open-forest study area and the bird surveys.
-
-\subsection*{Data description}
-[[ Need to update this with extended data ]]
-
-For the purposes of our analysis we selected a single survey from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability.
-In 2002, 195 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig.
-\ref{fig:2002studyareapointspt}).
-On 67 points 152 \akepa{} were detected.
-The number of detections within each sampling unit ranged from zero to 6.
-
-%\begin{figure}
-%	\centering
-%	\includegraphics[scale=0.5]{figures/2002counts_pt}
-%	\caption{Counts of \akepa{} by point during the 2002 survey.}
-%	\label{fig:2002countspt}
-%\end{figure}
-
-% addtextend
-
-\subsection*{Model specification in \texttt{inlabru}}
-
-In this section we introduce \texttt{inlabru} syntax and provide example code to demonstrate to specify the model.  First, we specify the SPDE model as
-\begin{lstlisting}
-matern <- inla.spde2.pcmatern(mesh, 
-                              prior.sigma = c(2, 0.01),    
-                              prior.range = c(130, 0.01))
-\end{lstlisting}
-which uses the penalised complexity priors for the hyperparameters of the GMRF as introduced by \cite{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$ where $\rho$ is the range of the Mat\'ern covariance.
-
-Second, we specify the thinning probability function.
-The data contains only distances from the observer so we use equation \eqref{intensity-incomplete} to specify the log thinning probability function as
-\begin{lstlisting}
-thinning_function <- function(distance, lsig){
-    log(2*pi) + log(distance) + log_halfnormal(distance, lsig)
-    }
-\end{lstlisting}
-where \texttt{log\_halfnormal} is the log half-normal detection function $\log g(r | \sigma) = -1/2(r/\sigma)^2$ and the argument \texttt{lsig} is $\log(\sigma)$.
-The dataset contains a column named \texttt{distance} that contains the observed distances.
-The software will search for the appropriate columns using the names of the arguments specified in this function.
-To allow the software to distinguish between names of covariates and parameters to be estimated we must also specify a model components object
-\begin{lstlisting}
-cmp <- ~ Intercept + grf(map = coordinates, model = matern) + lsig 
-\end{lstlisting} 
-This object names all model components to be estimated.
-Note that the \texttt{grf} component is a name we have chosen and not a function.
-The software allows user-specified names for the \texttt{f()} function known to users of \texttt{R-INLA}.
-Note that \texttt{coordinates} is a keyword that allows users to supply their data in the form of a spatial object using the \texttt{sp} package \citep{pebesma_spatial_2005}.
-
-The model formula object is
-\begin{lstlisting}
-fml <- coordinates + distance ~ Intercept + grf +
-	thinning_function(distance, lsig)
-\end{lstlisting}
-To fit the model we specify starting values for the parameters, the domain of possible observed distances and fit the model using the \texttt{lgcp} function
-\begin{lstlisting}
-starting_values <- data.frame(grf = 0, lsig = 3.65, Intercept = 0)
-W <- 60   # transect radius
-distance_domain <- seq(.Machine$double.eps, W, length.out = 30)
-lgcp(components = cmp, 
-     data = obs,
-     samplers = samplers,
-     domain = list(distance = distance_domain),
-     formula = fml,
-     options = list(result = starting_values, max.iter = 30))
-\end{lstlisting}
-where our observed data is contained in a \texttt{SpatialPointsDataFrame} object named \texttt{obs}.
-Since not all sampling units observed birds we also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sampling units.
-
-\subsection*{Results}
-
-The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.
-This shows a region of high intensity in the south and much lower intensity in the north.
-The CV plot shows that the CV in the posterior intensity field is lower areas with greater sampling effort.
-There is clear preferential sampling and in the north estimated intensities are lower which will also inflate the posterior CV.
-It is difficult to know whether the high CV in the north is primarily caused by lower sampling effort or lower estimated abundance, or some combination of the two.
-
-\begin{figure}
-	\includegraphics[scale=0.5]{figures/intensity_mean_cv.png}
-	\caption{(left) Posterior mean of the intensity (right) Posterior coefficient of variation of the intensity}
-	\label{fig:intensity-mean-cv}
-\end{figure}
-
-The CV plot shows a measure of uncertainty relative to the mean intensity in each prediction grid cell.
-However, often we are interested in the uncertainty around the mean intensity on the scale of the estimated intensity itself - perhaps seeking to identify areas where we can confidently predict particularly high or low intensities.
-One approach attempt to communicate this uncertainty around the mean is to present the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
-
-\begin{figure}
-	\includegraphics[scale=0.5]{figures/intensity_quantiles.png}
-	\caption{The predicted posterior intensity summarised in three plots by the mean (upper),  0.025 quantile (lower left) and 0.975 quantile (lower right) for each prediction grid cell.
-All three plots use a common colour scale}
-	\label{fig:intensity-quantiles}
-\end{figure}
-
-We believe quantile plots like those in \autoref{fig:intensity-quantiles} to be challenging to interpret and potentially misleading.
-The temptation is to interpet these plots as showing possible intensity surfaces that produced the observed data.
-We believe this misconception would be extremely common in particular for non-statistically trained audiences.
-The maps in \autoref{fig:intensity-quantiles} in fact only show \textit{summary statistics} of a posterior intensity which is specified (in our model) as a log-Gaussian random field.
-
-This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.
-Three such realizations are in \autoref{fig:intensity-realizations}.
-There are some notable differences when compared to the mean and quantile summary maps.
-Each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.
-In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.
-The summary plots hide the spatial structure the model has been able to estimate from the data.
-In general, the mean of a random field will contain less spatial structure than realisations of the field.
-
-\begin{figure}
-	\includegraphics[scale=0.35]{figures/intensity_realized.png}
-	\caption{Three realizations of the posterior intensity field}
-	\label{fig:intensity-realizations}
-\end{figure}
-
-The quantile plots are more concerning.
-As mentioned above, the temptation is to interpret these as possible intensities that produced the observed data.
-However, they in fact show the quantiles of the \textit{marginal} probability density for each prediction location.
-Presenting these quantiles side by side in a single map obscures the problem of multiple hypothesis testing.
-To demonstrate this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile and 12,400 for the 0.975 quantile.
-A naive (and tempting) interpretation of these numbers is as upper and lower limits of the 95\% credible interval for abundance.
-However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
-This emphasises that the temptation to treat these maps as showing possible intensity surfaces leads to interpretations (and possible management decisions) that are inconsistent with the very model that generated the maps.
-
-As an alternative, we suggest using excursion sets and excursion functions \citep{bolin_excursion_2015}.
-This is a recently developed approach to representing uncertainty in stochastic processes that specifically considers joint probability of events across multiple locations.
-The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e.
-the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$.
-For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
-\begin{equation*}
-E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \}
-\end{equation*}
-It is important to note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.
-This means that when considering random quantities such as $\lambda(s)$ the definition of the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.
-Because $\lambda(s)$ is stochastic we can only ever say this is true with some probability and use something akin to a confidence level by specifying $\alpha$.
-Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.
-An implementation is available in the \texttt{excursions} package \citep{bolin_calculating_2018} available through the Comprehensive R Archive Network \citep{r_2017}.
-
-\autoref{fig:excursions} (left panel) shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95.
-This figure can be interpreted in a natural way.
-It is the largest possible region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
-\begin{figure}
-	\includegraphics[scale=0.5]{figures/excursions.png}
-	\caption{Left:  The positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.
-Right: The positive excursion function with a level corresponding to 1 bird per hectare}
-	\label{fig:excursions}
-\end{figure}
-To visualise multiple such maps we can use the excursion function $F_u^{+}(s) = \sup \{1 - \alpha ; s \in E_{u,\alpha}^+ \}$, which defines for each location something similar to a p-value.
-For each $s$ it is the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.
-The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursions} (right panel).
-
-This figure can also be interpreted naturally.
-It shows the largest possible probability for each location to be a member of a set for in which the intensity exceeds a threshold simultaneously across all locations in the set.
-Regions on the edge of the excursion set would be included if the $\alpha$ value were allowed to increase slightly.
-However, for regions in the north there is essentially no probability level for which those locations would be included in $A_u^{+}(\lambda)$.
-
-A key feature of these plots is that they display either discrete sub-regions of the area of interest (for excursion sets) or the probability of membership of such discrete sub-regions (for excursion functions).
-We believe the risk of interpreting these maps as intensities is much lower than for the quantile plots.
-The fact that it is possible to estimate discrete regions may also be useful to communicate results to conservation practitioners, land and resource managers and policy decision-makers.
-We anticipate the excursion function will be most useful in communicating to audiences with statistical training.
-% [RJC - I think statistically savvy audiences will readily grasp the excursion functions.
-Do you mean "without" statistical training?]
-
-Another key aim of performing distance sampling surveys is to estimate the abundance of the population of interest.
-We describe here how to estimate the posterior for the abundance.
-Let $n$ denote the abundance of the population within the region of interest and $N$ its corresponding random variable within our Bayesian framework.
-Integrating the mean of the random field provides a point estimate for the abundance, which we call the \textit{expected} abundance.
-Here we demonstrate the importance of considering the full posterior for $N$ which we call the \textit{realised} abundance.
-
-Let $N = \int_{\Omega}\lambda(s)\mathrm{d}s$.
-Note that $N$ is a random variable and the integral of $\mathbb{E}[\lambda(s)]$ is equivalent to $\mathbb{E}(N)$ by Fubini's theorem, hence the term expected abundance.
-As we mentioned above, this is the most sensible point estimate for the abundance.
-To communicate uncertainty around the expected abundance we estimate the posterior $\pi(N | \lambda)$ by a monte-carlo method.
-Taking $m$ monte-carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
-This is well-defined since $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
-\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
-This allows us to estimate the probability for any specific value of realised abundance $n$ by calculating $\pi(N = n | \lambda)$.
-For reference, \autoref{fig:realized-abundance-comparison} shows a single Poisson PMF with rate parameter equal to the expected abundance.
-
-\autoref{fig:half-normal} shows a summary of the posterior for the detection function, plotting the posterior mean along with 95\% credible intervals.
-We note again that all model summaries shown here are based on monte-carlo sampling of the joint posterior of all model parameters.
-As such all the plots incorporate the uncertainty in the detection process.
-
-\begin{figure}
-	\includegraphics[scale=0.6]{figures/realized_abundance_posterior.png}
-	\caption{Posterior of realized abundance.  The blue line marks the expected abundance}
-	\label{fig:realized-abundance-posterior}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[scale=0.6]{figures/realized_abundance_vs_exp.png}
-	\caption{Solid:  Posterior of realized abundance.  Dashed:  Poisson PMF with rate parameter $\mathbb{E}[\Lambda]$}
-	\label{fig:realized-abundance-comparison}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[scale=0.6]{figures/halfnormal.png}
-	\caption{Posterior detection function}
-	\label{fig:half-normal}
-\end{figure}
-
-\bigskip
-
-\newpage
-
-\section*{Discussion}
-
-\subsection*{General remarks}
-
-The most common method to analyse distance sampling data is a two-stage approach with a generalised additive model that includes a spatial smoothing spline.
-The approach here of specifying a point process model and including an SPDE effect can seem fundamentally different.
-However, these two approaches are related and an analysis of the \akepa{} data using the GAM approach provides similar point estimates.
-There is a correspondence between SPDEs and smoothing splines: the smoothing spline is an estimator of the posterior mean of the SPDE \citep{kimeldorf_spline_2002}.
-The SPDE and the smoothing spline play a similar role within the model and the resulting mean field map looks similar in both approaches.
-In addition, via the point process model, we have avoided the need to bin data into counts, which may at first appear as though we have avoided some loss of information.
-This is not the case since we also applied the standard distance sampling assumption of constant density within each point transect.
-This can be thought of as the point process version of binning data into counts since the location of points within transects does not inform the spatial distribution of points.
-The main advantage of the log-Gaussian Cox process is that the discretisation of space occurs at the level of approximating the GRF.
-This is in contrast to motivating the discretisation by binning the observed data which requires consideration of zero-inflation and dispersion.
-
-We also note that \texttt{inlabru} provides functionality for fitting models with multiple likelihoods \citep{bachl_inlabru_2019}.
-As such our approach is extendible to more complicated contexts such as marked point processes.
-
-We believe the excursions approach is a valuable addition to the literature on communicating uncertainty in estimated surfaces.
-For species distribution modelling this has the potential to be a useful tool for model evaluation and communication with stakeholders.
-In particular the approach requires users to decide (i) a threshold or set of threshold values of interest and (ii) an acceptable probability of error.
-We believe that, in most situations, these decisions should be made by domain experts who have an understanding of the phenomena of interest, the set of possible decisions that may be taken on the basis of the analysis and the consequences of error.
-Once these decisions have been made the statistician can then proceed accordingly, providing outputs  that inform on quantities of interest with a level of uncertainty that stakeholders have agreed is acceptable.
-We note that the excursions approach applied here has the potential to be useful in other methods of species distribution mapping, environmental statistics and spatial statistics in general.
-There is the potential to use excursions to motivate both the \textit{definition} and \textit{estimation} of concepts of interest to ecologists, such as a species' home range or the notion of occupancy.
-
-\subsection*{Possible future directions}
-
-Here we applied the iterated INLA approach in order to estimate the half-normal detection function which is non-linear in its $\log \sigma$ parameter.
-We note that this approach provides a general framework for estimating parametric functions from data and is not restricted to detection functions within a distance sampling model.
-The \texttt{inlabru} software allows users to specify the model in an intuive and modular way.
-Users write individual functions to define (possibly non-linear) model components and specify how these are combined to form the additive predictor.
-By additive we mean a predictor that is a sum of model components, and not necessarily additive in terms of parameters to be estimated.
-
-We presented a simple choice for the detection model.
-The half-normal function has only one parameter to be estimated but other, more flexible, detection functions are common in applications.
-The potential to include further flexibility in the detection model through covariates or random-effects is an open research question and would dramatically increase the usability of our approach presented here.
-Increasing the number of parameters to estimate in the non-linear model components may run into computational challenges.
-This has yet to be explored.
-
-Model selection is a general challenge for models with spatially structured random effects.
-Traditional tools of model assessment and selection do not apply in many contexts.
-For model assessment we advise a pragmatic approach based on comparing summaries of posterior samples to the observed data.
-A moodel selection procedure is useful to users to justify decisions regarding which covariates to include, whether or not to include a spatially structured random effect, and if so, what type of effect should be used.
-
-\bibliographystyle{chicago}
-\bibliography{paper}
+Discuss wisely
 

### 2019-11-22 paper.tex
slight tweaks to old draft

@@
 \begin{abstract}
-Distance sampling methods aim to estimate the size and distribution of animal populations.
-Traditionally this has been achieved using a two-stage approach by using first-stage estimates of detectability as an offset in a second-stage generalised additive model.
+Distance sampling methods aim to estimate the size of animal populations.
+Incorporating spatial models of animal density has traditionally been achieved using a two-stage approach where first-stage estimates of detectability are used as an offset in a second-stage generalised additive model.
 One-stage approaches have typically been implemented using MCMC and data augmentation methods.
@@
 We model the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields.
-We demonstrate the approach by applying the methods to point count survey data of Hawaiian \akepa{}.
+We demonstrate the approach by applying the methods to point count survey data of a forest bird.
 Our Bayesian one-stage approach allows sampling from the joint posterior of all model parameters and so the resulting summaries and plots naturally incorporate uncertainty in the observation model estimates.
@@
 The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
-Distance sampling methods aim to achieve this by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2008, buckland_distance_2015}.
+Distance sampling methods aim to achieve this by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2004, buckland_distance_2015}.
 Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.
-The probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2008}.
+The probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2004}.
 
@@
 Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
-However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004, stoyan_remark_1982, hogmander_random_1991}.
+However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
 We also further develop the point process viewpoint to allow point process models to be applied to data with incomplete location information.
@@
 Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance.
-The relevant distance depends on the survey type.
+The distance measured depends on the type of survey.
 For line transects the perpendicular distance to the transect line is used.
-For point transects the distance to the observer is used.
+For point transects the horizontal distance to the observer is the relevant distance.
 For the remainder of the paper we assume a point transect survey design.
@@
 \label{lgcp-likelihood}
-\pi(\bm{Y} | \theta, \phi) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
+\pi(\bm{Y}) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
 \end{equation}
+where we have suppressed parameter vector notation.
 The integral component of the likelihood does not usually have analytical solutions.
@@
 To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.
-Similarly let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$
-
+Similarly let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$.
 Then the approximate log-likelihood is
-
 \begin{equation}
@@
 \end{equation}
-
 This approximation can be written as a modified Poisson likelihood.  To see this let $\exp \eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
@@
 \end{equation}
-
-This is similar to a Poisson likelihood with an offset term equal to $\log\alpha$ and can be specified using the Poisson likelihood implementation available in \texttt{R-INLA}.
+This is similar to a Poisson likelihood with a $\log\alpha$ offset and can be specified using the Poisson likelihood implementation available in \texttt{R-INLA}.
 
@@
 \end{equation*}
-where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatially indexed covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean Gaussian random field with Mat\'ern covariance
+where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatially indexed covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean GRF with Mat\'ern covariance
 \begin{equation}
@@
 
-We use a Gaussian Markov random field approximation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a stochastic partial differential equation.
+We use a GMRF approximation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.
 Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
-The parameters $\xi_1, \ldots, \xi_L$ form a Gaussian Markov random field with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a stochastic partial differential equation.
-The precision parameter $\tau$ controls the marginal variance of the field and $\kappa$ is an inverse range parameter that controls the rate of decay of the covariance as distance between locations increases.
-Further details of this Gaussian Markov random field approximation are given in the supplementary materials.
+The parameters $\xi_1, \ldots, \xi_L$ form a GMRF with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a SPDE.
+The parameter $\tau$ is a precision parameter and $\kappa$ an inverse range parameter that controls the rate of decay of the covariance as distance between locations increases.
+Further details of the GMRF approximation are given in the supplementary materials.
 
@@
 
-The above specifies a model for all animal locations in the area of interest.
-The log-intensity for observed points is $\log\tl(s) = \log\lambda(s) + \log g(s)$.
+The log-intensity for observed points (the thinned intensity) is $\log\tl(s) = \log\lambda(s) + \log g(s)$.
 This presents a problem since $\log g(s)$ is not typically linear in its parameters.
-However, for any single set of parameter values $g(s)$ can be viewed as acting like an offset.
+However, for any single set of parameter values $g(s)$ can be viewed as an offset term.
 If we could specify $g(s)$ as known then we could proceed with an inference approach suitable for additive linear predictors.
@@
 Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}.
-Information from the HFBS was used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge that was the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
+Information from the HFBS was used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau) that was the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
 Due to the broad-scale coverage and robust design HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.
@@
 During the 20th century \akepa{} declined dramatically due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994},  mosquito-transmitted avian diseases \citep{pratt_avifaunal_1994, atkinson_wildlife_1995}, introduced predators \citep{lepson_akepa_1997}, and food resources competitors \citep{lepson_akepa_1997}.
-\Akepa{} has a global abundance of 16,248 (95\%CI 10,074\textendash25,198) birds that has been restricted to five spatially distinct populations \citep{judge_akepa_2018}.
-Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau), supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}.
+\Akepa{} has a global abundance of approximately 16,200 (95\%CI 10,000\textendash25,200) birds that has been restricted to five spatially distinct populations \citep{judge_akepa_2018}.
+Hakalau supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}.
 Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern.
@@
 For the purposes of our analysis we selected a single survey from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability.
-In 2002, 195 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig.
+In 2002, 289 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig.
 \ref{fig:2002studyareapointspt}).
-On 67 points 152 \akepa{} were detected.
+276 \akepa{} were detected on 121 point transects.
 The number of detections within each sampling unit ranged from zero to 6.
@@
 Note that the \texttt{grf} component is a name we have chosen and not a function.
-The software allows user-specified names for the \texttt{f()} function known to users of \texttt{R-INLA}.
+The software allows user-specified names for model components.
+This feature replaces the \texttt{f()} function known to users of \texttt{R-INLA} and allows the user to keep track of multiple model components rather that calling \texttt{f()} repeatedly.
 Note that \texttt{coordinates} is a keyword that allows users to supply their data in the form of a spatial object using the \texttt{sp} package \citep{pebesma_spatial_2005}.
@@
 To fit the model we specify starting values for the parameters, the domain of possible observed distances and fit the model using the \texttt{lgcp} function
+% UPDATE THE W WHEN THIS CHANGES
 \begin{lstlisting}
@@
 
-The most common method to analyse distance sampling data is a two-stage approach with a generalised additive model that includes a spatial smoothing spline.
+The most common method to analyse combine spatial models and distance sampling data is a two-stage approach with a generalised additive model that includes a spatial smoothing spline.
 The approach here of specifying a point process model and including an SPDE effect can seem fundamentally different.
@@
 For species distribution modelling this has the potential to be a useful tool for model evaluation and communication with stakeholders.
-In particular the approach requires users to decide (i) a threshold or set of threshold values of interest and (ii) an acceptable probability of error.
+In particular the approach requires users to \textit{a priori} decide on (i) a threshold or set of threshold values of interest and (ii) an acceptable probability of error.
 We believe that, in most situations, these decisions should be made by domain experts who have an understanding of the phenomena of interest, the set of possible decisions that may be taken on the basis of the analysis and the consequences of error.

### 2019-11-22 paper-v2.tex
add introduction to new draft

@@
 
+-----
+
+The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
+Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.
+The Hawaiian \akepa{} is an endemic species whose population declined dramatically during the 20th century due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994}.
+The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.
+This information is critical when making decisions about conservation strategies.
+However, answering these questions presents many statistical challenges.
+
+Firstly, as in many ecological surveys, it is impossible to undertake a census.
+The existing population numbers in the thousands and lives in dense forest where logistical and ecological challenges mean a census is not feasible.
+For this reason, a monitoring survey must sub-sample in space and time in some way.
+For the \akepa{} this takes the form of the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986} which is a large-scale, quantitative survey of Hawaiian forest birds.
+Annual surveys of the \akepa{} study-region consist of survey points located along randomly located transects.
+
+Secondly, even at sampled locations, detectability of animals is uncertain.
+The \akepa{} survey attempts to estimate detectability using a distance sampling approach \citep{buckland_distance_2015} where for each observation the distance to the observer is recorded.
+A parametric form of detection function is assumed to decay with increasing distance and is estimated from the observed distances.
+Because of this, answering questions related to the abundance and spatial distribution must involve the estimation of a complex observation process.
+
+Thirdly, due to the spatial sub-sampling, a key aim of the analysis is the generation of spatial predictions.
+However, as is the case in many species distribution models, we usually have reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates.
+For the \akepa{} this takes the form of a north-south gradient that has been investigated but for which no clear-cut explanatory cause has been found [[cite here?]].
+From a statistical perspective this suggests the use of spatially-structured random effects to account for the missing covariates.
+
+Fourthly, the resulting statistical model is complex and it is therefore challenging to effectively communicate the results of the analysis to non-statistically trained stakeholders such as conservation managers and policy officers.
+In particular we note the challenge of communicating uncertainty in maps of predicted animal density.
+
+In this paper we present an analysis that seeks to address each of these issues by presenting
+
+\begin{enumerate}[(i)]
+	\item A model-based spatial point process perspective on point transect distance sampling, representing the observation model as a thinning probability function.
+	\item A one-stage approximate Bayesian approach to inference to simultaneously estimate the observation model and spatial distribution.
+This is based on iterated model fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009}.
+	\item A computationally efficient Gaussian Markov random field (GMRF) spatial effect based on a stochastic partial differential equation approach \citep{lindgren_explicit_2011} to account for missing covariates.
+	\item An approach to model evaluation and communication based on sampling from the joint posterior of all model parameters and excursions methods \citep{bolin_excursion_2015} for investigating uncertainty in spatial predictions.
+\end{enumerate}
+
+We note that these challenges are generic to many types of wildlife survey data, not only the \akepa{} survey, and can be addressed by a range of possible statistical approaches.
+Our analysis here is one possible choice of approach and throughout the remainder of the paper we will highlight differences with existing approaches.
+It is important to note that we do not claim any overall superiority of this approach over others, although we do highlight some areas for which we believe our approach provides a natural setting for addressing the issues above.
+We demonstrate several novel contributions to the problem of species distribution modelling under imperfect detection.
+In particular, the point process perspective with incomplete location information, the one-stage approximate Bayesian inference strategy and focus on communicating uncerainty in spatial predictions using excursions are innovations in this area.
+We also note that although we use a Bayesian approach, maximum-likelihood frameworks could also have been used.
+Finally, we note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analsyis.
+
+The rest of the paper proceeds as follows:  (i) we describe in detail the \akepa{} study-region and survey design; (ii) we present the perspective of distance sampling as a thinned point process; (iii) we describe the GMRF random effect and iterated INLA fitting procedure; (iv) we present the results of the analysis and discuss model evaluation and communication.
+
 \section{Study design}

### 2019-11-22 paper-v2.tex
remove intro sketch

@@
 
-\begin{itemize}
-	\item Estimating population size and spatial distribution is an important question in ecology and conservation
-	\item Here we present the Akepa data:
-		\begin{itemize}
-			\item endangered
-			\item subject of conservation efforts
-			\item require monitoring to understand changes in overall abundance and the spatial distribution to target conservation efforts
-		\end{itemize}
-	\item Estimating the population size and spatial distribution of Akepa is challenging because
-		\begin{enumerate}[(i)]
-			\item cannot do a census (large numbers, dense forest, logistical constraints) and so must subsample in space and time (describe the HFBS design)
-			\item imperfect detection when we do sample - a complex observation process (describe point transect distance sampling approach)
-			\item interest in generating spatial predictions but we usually have missing covariates and use spatial random effects (No covariate for north-south in Akepa data, other studies ambiguous)
-			\item complex models that are difficult to communicate with stakeholders such as conservation managers and policy officers (communicating uncertainty in abundance estimation and prediction maps)
-		\end{enumerate}
-\end{itemize}
-
-In this paper we present an analysis of the Akepa data that seeks to address these issues
-
-\begin{enumerate}[(i)]
-	\item a spatial point process formulation of distance sampling
-	\item a one-stage estimation process to simultaneously estimate the spatial model and the observation process
-	\item a computationally efficient Gaussian Markov random field effect based on a stochastic partial differential equation approach
-	\item model evaluation and communication based on join posterior sampling and excursions package
-\end{enumerate}
-
-We work in a Bayesian context here but note that maximum-likelihood frameworks could also be used.
-
------
-
 The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.

### 2019-11-22 paper-v2.tex
add survey design section

@@
 
-\section*{Introduction}
+\section{Introduction}
 
@@
 Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.
-The Hawaiian \akepa{} is an endemic species whose population declined dramatically during the 20th century due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994}.
+The Hawaiian \akepa{} is an endemic species whose population declined dramatically during the 20th century [[cite??]].
 The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.
@@
 
-\begin{itemize}
-	\item describe survey protocol in detail
-	\item mini lit review - DS, 2-stage, Joyce
-\end{itemize}
+The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.
+Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}.
+Information from the HFBS was used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau) that was the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
+Due to the broad-scale coverage and robust design HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.
 
-\section{Distance Sampling as a Thinned Point Process}
+During the 20th century \akepa{} declined dramatically due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994},  mosquito-transmitted avian diseases \citep{pratt_avifaunal_1994, atkinson_wildlife_1995}, introduced predators \citep{lepson_akepa_1997}, and food resources competitors \citep{lepson_akepa_1997}.
+\Akepa{} has a global abundance of approximately 16,200 (95\%CI 10,000\textendash25,200) birds that has been restricted to five spatially distinct populations \citep{judge_akepa_2018}.
+Hakalau supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}.
+Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern.
+Moreover, having unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.
+
+\subsection{Study area and sampling method}
+
+[[Need to update this and the figure with the extended data area]]
+
+Hakalau was established in 1985 to conserve 15,390-ha of montane forest habitat for native forest birds and rainforest plants.
+Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance.
+Survey points were established along 15 transects following a systematic, random design with points approximately 150 m apart on transects located either 500 or 1,000 m apart.
+Following \cite{camp_population_2010, camp_statespace_2016}, we restrict our study area to the 3,061 ha open-forest stratum of Hakalau.
+The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.
+To the north the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig.
+\ref{fig:2002studyareapointspt}).
+The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of their study area.
+The west side of the study area is bounded by pasture that is dominated by grass and is unsuitable habitat for \akepa.
+
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.5]{figures/2002studyareapoints_pt}
+	\caption{Study area (light blue) showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.}
+	\label{fig:2002studyareapointspt}
+\end{figure}
+
+Surveys used point-transect methods, recording horizontal distances from survey points to individually detected birds.
+Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength \textgreater Baufort scale 3).
+During 8-min counts trained observers recorded the species, exact distance to the nearest meter and detection type for each bird detected, along with the sampling conditions cloud cover, rain, wind strength, gust strength, and time of day each point was surveyed.
+\cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the open-forest study area and the bird surveys.
+
+\subsection{Data description}
+[[ Need to update this with extended data ]]
+
+For the purposes of our analysis we selected a single survey from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability.
+In 2002, 289 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig.
+\ref{fig:2002studyareapointspt}).
+276 \akepa{} were detected on 121 point transects.
+The number of detections within each sampling unit ranged from zero to 6.
+
+\section{Distance sampling as a thinned point process}
 

### 2019-11-22 paper-v2.tex
start DS as PP section

@@
 
+\subsection{Overview of distance sampling methods}
+
+Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2004, buckland_distance_2015}.
+In this section we present a brief historical survey of distance sampling approaches and the challenge of combining distance sampling models with spatially explicit statistical models.
+We end by introducing the most direct predecessor to our method and note differences with previous approaches.
+
+Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size where the probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2004}.
+
+More recently, interest has focused on fully model-based approaches that include a spatially explicit model for animal density and allow the use of non-randomized survey designs.
+These methods also allow density to be associated with spatially-indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
+
+Model-based distance sampling has been implemented in a two-stage modelling framework.
+In the first stage, detectability is estimated within each sampling unit.
+In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset.
+Due to often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.
+Negative-binomial and Tweedie distributions are common choices to model the distribution of counts.
+This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review and software package \texttt{dsm} to fit the models.
+
+A key concern with the two-stage approach is the propagation of the uncertainty from the detectability estimates to the second-stage spatial model.
+Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units as well as the challenge of combining smoothers and spatial bootstraps \citep{bravington_reliable_2018-1}.
+Instead, \cite{bravington_reliable_2018-1}  proposed avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the first-stage maximum-likelihood estimate.
+
+One-stage modelling approaches to distance sampling have also been developed within a Bayesian framework.
+These approaches use data augmentation to model unobserved individuals or groups and were fit using Markov-chain Monte Carlo (MCMC) methods \citep{royle_hierarchical_2008, schmidt_using_2012}.
+More recently  \cite{oedekoven_bayesian_2014} provided a one-stage model that avoids data augmention by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty by applying a reversible jump MCMC algorithm.
+
+Our approach is also one-stage Bayesian but avoids MCMC and data augmentation by using INLA, an approximate Bayesian approach.
+The only paper to use INLA (to the best of our knowledge) to fit a spatially explicit model to distance sampling data is \citet{yuan_point_2017}.
+
+% UP TO HERE - introduce Yuan now.  Shifted some stuff out of here to GMRF / SPDE section.
+
+Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.
+Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
+
+Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
+However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
+We also further develop the point process viewpoint to allow point process models to be applied to data with incomplete location information.
+It is common for distance sampling surveys to only record the distance to the observer and the observer location.
+We show how point process methods can handle this type of data even though the exact location of the point is not known.
+
+We replace the SPDE specification of the detection function in \cite{yuan_point_2017} with a parametric family of detection functions that are more familiar to users of standard distance sampling methods.
+This results in components of the additive predictor that are non-linear in their parameters.
+To deal with this we use a method of iterated model fits based on a Taylor expansion of the non-linear model components.
+We name this approach \textit{iterated INLA}.
+
+\subsection{Model specification}
+
+In this section we introduce the statistical model for the animal locations and observation process, including a detailed description of how this model can be represented as a modified Poisson likelihood.
+We use this formulation to fit the model using the \texttt{R-INLA} implementation of INLA but this likelihood could in principle be used in any statistical software capable of specifying such a modified Poisson form.
+
+We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(s)$.
+The log-Gaussian Cox process is a flexible approach that can include spatially structured random effects on the intensity process to account for unexplained heterogeneity not captured by fixed-effect covariates.
+
+\sloppy For the case with imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected})$.
+A key property of the log-Gaussian Cox process is that a realisation of a point process with intensity process $\lambda(s)$ that is thinned by thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.
+
+Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance.
+The distance measured depends on the type of survey.
+For line transects the perpendicular distance to the transect line is used.
+For point transects the horizontal distance to the observer is the relevant distance.
+For the remainder of the paper we assume a point transect survey design.
+
+The thinning probability function $g(s)$ is specified as a member of a parametric family of functions.
+In this paper we demonstrate our approach using the half-normal formulation, although others such as hazard rate, are also common.
+If $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated.
+The parameter $\sigma$ can only be estimated if an assumption is made about the intensity of the animal locations.
+Without such an assumption detectability and intensity are confounded.
+The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(s)$.
+Thus any observed differences from uniformity can be attributed to detectability and not variation in the intensity.
+
+A point transect distance sampling survey consists of a set of $K$ sampling units.
+The $k$-th sample unit we denote $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region $\Omega = \cup_{k=1}^K \Omega_k$.
+For simplicity we assume that all sampling units are discs with radius $W$ and $\Omega_j \cap \Omega_k = \emptyset$ for all $j \neq k$.
+The thinning probability function $g(s)$ is defined relative to the positions of the sample units.
+The probability of observing a point at location $s$ in the $k$-th sampling unit $\Omega_k$ we denote $g_k(s)$.
+The probability of observing a point outside the surveyed region is zero.
+Since the surveyed regions are non-overlapping each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.
+For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
+The assumption of non-overlapping survey regions can be relaxed by including extra information such as the time of each observation but, for simplicity, we do not consider this here.
+The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for $s \in \Omega_k$.
+
+The thinned log-Gaussian Cox process likelihood with observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
+\begin{equation}
+\label{lgcp-likelihood}
+\pi(\bm{Y}) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
+\end{equation}
+where we have suppressed parameter vector notation.
+The integral component of the likelihood does not usually have analytical solutions.
+Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach of \cite{simpson_going_2016} for a fully observed point pattern.
+
+To evaluate the integral we introduce polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.
+In general the thinning function depends only on $r$ and not on $\theta$.
+We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$.
+It follows that the integral can be rewritten as a sum of weighted one-dimensional integrals $\sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$ using the assumption that $\lambda(s)$ is constant within each sampling unit.
+For each sampling unit we approximate the one-dimensional integral using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
+This gives
+\begin{equation*}
+	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj})
+\end{equation*}
+where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(s_{kj}) = \lambda(s_k) g_k(r_{kj})$.
+
+To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.
+Similarly let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$.
+Then the approximate log-likelihood is
+\begin{equation}
+\label{approx-log-likelihood}
+	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}
+\end{equation}
+This approximation can be written as a modified Poisson likelihood.  To see this let $\exp \eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
+$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $y = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.
+Then the approximate likelihood becomes
+\begin{equation}
+\pi(\bm{Y}) \approx \prod_{i=1}^{KM + n} \eta_i^{y_i}\exp(-\alpha_i\eta_i)
+\end{equation}
+This is similar to a Poisson likelihood with a $\log\alpha$ offset and can be specified using the Poisson likelihood implementation available in \texttt{R-INLA}.
+
+\subsection{Intensities for incomplete data}
+
+In the above we assume the data are complete records of animal location.
+However, in many distance sampling surveys only the location of the observer and the distance to the observer are recorded.
+Data of this type can be analysed within a point process framework by deriving the appropriate intensity function for the incomplete data.
+
+Using the polar coordinates notation given above, for a detected point at location $s_k(r, \theta) \in \Omega_k$ we consider the case where we observe $r$ but not $\theta$.
+It follows that the intensity for points at a distance $r$ observed within sampling unit $\Omega_k$ is
+\begin{align}
+\label{intensity-incomplete}
+\tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(s)g_k(s)\mathrm{d}s \nonumber \\
+&= 2\pi\lambda(s_k)rg_k(r)
+\end{align}
+ where $c_k(r)$ is a circle of radius $r$ centred at $s_k$ and the second line follows from changing to polar coordinates and the assumption of constant intensity within $\Omega_k$.
+This intensity differs from the full data case by including the $2\pi$  term to account for the fact that we do not observe $\theta$ and the additional $r$ term to account for the increasing area surveyed at larger distances.
+
 \section{Iterated INLA}
@@
 \section{GMRF + SPDE}
+
+Yanked from intro:
+
+The reliance on MCMC for one-stage modelling presents a challenge for using spatially-structured random effects, which can be computationally expensive.
+\citet{yuan_point_2017} presented a one-stage approach fitted using integrated nested Laplace approximations (INLA).
+Applying their method to line-transect distance sampling, they simulateously estimated the detectability and spatial distribution of animals.
+
+Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF).
+They consider the data as a realisation of a spatio-temporal log-Gaussian Cox point process, an approach that avoids the need to bin data into counts and sidesteps the issues of zero-inflation and dispersion.
+The observation process is formulated as a thinning of the log-Gaussian Cox process.
+Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.
 

### 2019-11-26 paper-v2.tex
more stuff

@@
 
-Fourthly, the resulting statistical model is complex and it is therefore challenging to effectively communicate the results of the analysis to non-statistically trained stakeholders such as conservation managers and policy officers.
+Fourthly, the resulting statistical model is complex and therefore challenging to effectively communicate the results of the analysis to non-statistically trained stakeholders such as conservation managers and policy officers.
 In particular we note the challenge of communicating uncertainty in maps of predicted animal density.
@@
 Our analysis here is one possible choice of approach and throughout the remainder of the paper we will highlight differences with existing approaches.
-It is important to note that we do not claim any overall superiority of this approach over others, although we do highlight some areas for which we believe our approach provides a natural setting for addressing the issues above.
 We demonstrate several novel contributions to the problem of species distribution modelling under imperfect detection.
@@
 Finally, we note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analsyis.
+
+Our aim is to present a full analysis of the data along with discussion on model specification, inference and model evaluation that reveals key issues in the analysis and our strategies to address them.
+Our end hope is that the analysis we present here may provide other researchers with similar data structures an awareness of the issues involved and some options for strategies to use in their own analysis.
 
@@
 
-Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2004, buckland_distance_2015}.
-In this section we present a brief historical survey of distance sampling approaches and the challenge of combining distance sampling models with spatially explicit statistical models.
-We end by introducing the most direct predecessor to our method and note differences with previous approaches.
+Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer \citep{buckland_advanced_2004, buckland_distance_2015}.
+In this section we present a brief historical survey of distance sampling approaches and the approaches of combining distance sampling methods with spatially explicit statistical models.
+We end by introducing a predecessor to our method and note differences with previous approaches.
 
@@
 
-One-stage modelling approaches to distance sampling have also been developed within a Bayesian framework.
-These approaches use data augmentation to model unobserved individuals or groups and were fit using Markov-chain Monte Carlo (MCMC) methods \citep{royle_hierarchical_2008, schmidt_using_2012}.
+Concerns about uncertainty propagation can be avoided by one-stage modelling approaches.
+In our experience these tend to be developed within a Bayesian framework.
+These approaches use data augmentation and to model unobserved individuals or groups and Markov-chain Monte Carlo (MCMC) methods for inference \citep{royle_hierarchical_2008, schmidt_using_2012}.
 More recently  \cite{oedekoven_bayesian_2014} provided a one-stage model that avoids data augmention by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty by applying a reversible jump MCMC algorithm.
 
-Our approach is also one-stage Bayesian but avoids MCMC and data augmentation by using INLA, an approximate Bayesian approach.
-The only paper to use INLA (to the best of our knowledge) to fit a spatially explicit model to distance sampling data is \citet{yuan_point_2017}.
+The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017} who use the approximate Bayesian method INLA applied to line transect data.
+INLA can be an attractive choice when working with spatially-structured random effects in large or complex spatial domains.
+The SPDE approach \citep{lindgren_explicit_2011} allows the construction of a sparse prior precision matrix which INLA can exploit for gains in computational efficiency and the finite element methods based on triangulation have been shown to be more efficient than regular lattice based approaches \citep{simpson_going_2016}, especially in applications with complex boundaries such as coastlines.
 
-% UP TO HERE - introduce Yuan now.  Shifted some stuff out of here to GMRF / SPDE section.
+\citet{yuan_point_2017} also take a point process perspective and formulate the detection model as a thinning of the full but imperfectly observed point pattern.
+Key to their approach was formulating the detection model using a stochastic partial differential equation.
+Similar to spline approaches, choosing a B-spline basis and finite element methods results in a sparse prior precision matrix for the detection model coefficients which can then be incorporated into the latent-Gaussian model framework required by INLA.
+Downsides to this approach are that it is challenging to mandate a monotonically decreasing detection function.
+The authors suggestion to reject any non-monotonically decreasing functions when sampling from the posterior seems unsatisfactory.
 
-Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.
-Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
+Here we present a similar approach to \citet{yuan_point_2017} in that our models are fitted using INLA.
+However, we avoid the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal.
+This will be more familiar to users of traditional distance samplign approaches.
+This results in components of the additive predictor that are non-linear in their parameters.
+To deal with this we use a method of iterated model fits based on a Taylor expansion of the non-linear model components.
+We name this approach \textit{iterated INLA}.
 
+We also apply the methods to point transect data which require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
 Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
@@
 It is common for distance sampling surveys to only record the distance to the observer and the observer location.
-We show how point process methods can handle this type of data even though the exact location of the point is not known.
-
-We replace the SPDE specification of the detection function in \cite{yuan_point_2017} with a parametric family of detection functions that are more familiar to users of standard distance sampling methods.
-This results in components of the additive predictor that are non-linear in their parameters.
-To deal with this we use a method of iterated model fits based on a Taylor expansion of the non-linear model components.
-We name this approach \textit{iterated INLA}.
+We show how the point process perspective can still handle this type of data even though the exact location of the point cannot be drawn on a map.
 
@@
 
+The log-intensity for observed points (the thinned intensity) is $\log\tl(s) = \log\lambda(s) + \log g(s)$.
+This presents a problem since $\log g(s)$ is not typically linear in its parameters.
+However, for any single set of parameter values $g(s)$ can be viewed as an offset term.
+If we could specify $g(s)$ as known then we could proceed with an inference approach suitable for additive linear predictors.
+
 \begin{itemize}
-	\item mini lit review - communicating uncertainty in spatial models w random effects
-	(should this really be here??  Think about moving)
+ \item Description of iterated INLA here
 \end{itemize}
 
-\section{GMRF + SPDE}
+\section{Gaussian Markov random field spatial effect}
 
-Yanked from intro:
+Using spatially-structured random effects can be computationally expensive, resulting in dense prior precision matrices.
+A key development in this area is the use of Gaussian Markov random field (GMRF) approximations to Gaussian random fields (GRFs).
+The link between them is via a stochastic partial differential equation solved using finite element methods \citep{lindgren_explicit_2011}.
 
-The reliance on MCMC for one-stage modelling presents a challenge for using spatially-structured random effects, which can be computationally expensive.
-\citet{yuan_point_2017} presented a one-stage approach fitted using integrated nested Laplace approximations (INLA).
-Applying their method to line-transect distance sampling, they simulateously estimated the detectability and spatial distribution of animals.
+To model the intensity of animal locations we use a combination of fixed effects of spatial covariates and a spatially structured random effect.
+This allows us to associate our survey data with environmental and ecological conditions that may be relevant to understanding the sdistribution of the species of interest.
+For cases where we have insufficient covariates to fully explain the location of animals the spatially structured random effect accounts for additional heterogeneity in the intensity and is required to assume conditional independence between points, given the intensity.
 
-Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF).
-They consider the data as a realisation of a spatio-temporal log-Gaussian Cox point process, an approach that avoids the need to bin data into counts and sidesteps the issues of zero-inflation and dispersion.
-The observation process is formulated as a thinning of the log-Gaussian Cox process.
-Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.
+The log-intensity of animal locations is given by
+\begin{equation*}
+\log \lambda(s) = \beta_0 + \sum_v \beta_v z_v(s) + \xi(s)
+\end{equation*}
+where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatially indexed covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean GRF with Mat\'ern covariance
+\begin{equation}
+C(s_1,s_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|s_1-s_2\|)^{\nu}K_\nu(\kappa \|s_1-s_2\|)
+\end{equation}
+where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.
+All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which controls the mean-square differentiability of the process.
+We set $\nu = 1$ which is the default value in \texttt{R-INLA}.
+[[Would like to justify $\nu = 1$ somehow?]]
+
+We use a GMRF approximation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.
+Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
+The parameters $\xi_1, \ldots, \xi_L$ form a GMRF with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a SPDE (see Supplementary material).
+The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function.
 
@@
 
+We fitted the models using penalised complexity priors for the hyperparameters of the GMRF \citep{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$ where $\rho$ is the range of the Mat\'ern covariance.
+The value of 130 was selected based on the minimum distance between sampling locations.
+Full code for the analysis can be found in the Supplementary material.
+
+The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.
+This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary materials).
+The CV plot shows that the CV in the posterior intensity field is lower areas with greater sampling effort.
+There is clear preferential sampling and in the north estimated intensities are lower which also contributes to larger CV values.
+\begin{figure}
+	\includegraphics[scale=0.5]{figures/intensity_mean_cv.png}
+	\caption{(left) Posterior mean of the intensity (right) Posterior coefficient of variation of the intensity}
+	\label{fig:intensity-mean-cv}
+\end{figure}
+
 \section{Model Evaluation}
+
+\begin{itemize}
+	\item mini lit review - communicating uncertainty in spatial models w random effects
+\end{itemize}
 

### 2019-11-26 paper-v2.tex
nearly done results

@@
 The value of 130 was selected based on the minimum distance between sampling locations.
-Full code for the analysis can be found in the Supplementary material.
+Full code for the analysis can be found in the Supplementary materials.
 
@@
 There is clear preferential sampling and in the north estimated intensities are lower which also contributes to larger CV values.
+
 \begin{figure}
@@
 
+Since CV values are influenced both by sampling effort and the magnitude of the estimated intensity means other methods we also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
+\begin{figure}[!t]
+	\includegraphics[scale=0.5]{figures/intensity_quantiles.png}
+	\caption{The predicted posterior intensity summarised in three plots by the mean (upper),  0.025 quantile (lower left) and 0.975 quantile (lower right) for each prediction grid cell.
+All three plots use a common colour scale}
+	\label{fig:intensity-quantiles}
+\end{figure}
+
+The posterior detection function \autoref{fig:half-normal} shows that detectability drops to just under 0.25 at the maximum observable distance of 60 metres.
+\begin{figure}[h]
+	\includegraphics[scale=0.6]{figures/halfnormal.png}
+	\caption{Posterior detection function.  Black line - mean posterior estimate, grey shaded area - 95\% credible interval}
+	\label{fig:half-normal}
+\end{figure}
+
+% POSTERIOR MATERN CORRELATION HERE
+Add posterior matern correlation function
+
+\begin{figure}[h]
+	\includegraphics[scale=0.6]{figures/matern_posterior.png}
+	\caption{Posterior Mat\'ern correlation function}
+	\label{fig:half-normal}
+\end{figure}
+
+We plot the posterior abundance estimate using a monte-carlo method.
+Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.
+Integrating the mean of the posterior intensity field provides a point estimate for the abundance - the expected abundance.
+We approximate the $\pi(N | \lambda)$ by a monte-carlo method.
+Taking $m$ monte-carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
+Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
+\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
+This allows us to estimate the probability for any specific value of realised abundance $n$ by calculating $\pi(N = n | \lambda)$.
+\begin{figure}
+	\includegraphics[scale=0.6]{figures/realized_abundance_posterior.png}
+	\caption{Posterior of realized abundance.  The blue line marks the expected abundance}
+	\label{fig:realized-abundance-posterior}
+\end{figure}
+
+We note that this presentation of abundance estimates differs from typical presentations.
+It is common to report a point estimate of the expected abundance and uncertainty in this point estimate.
+However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.
+We find the distinction between expected and realised abundance to be a useful one.
+Here the posterior intensity process is used to generate possible realisations of the intensity, which are then used to generate the abundance estimates.
+
+\clearpage %figure out figure placements later
 \section{Model Evaluation}
@@
 
+\clearpage
 \bibliographystyle{chicago}

### 2019-11-28 analyses/gam_comparison.R
change to inSide() instead of over()

@@ -1,0 +1,119 @@
+# Evaluate fitted model
+library(INLA)
+library(inlabru)
+library(sp)
+
+# set ggplot theme
+theme_set(theme_minimal())
+
+set.seed(9701071)
+
+#### Fitted model  ####
+file_name = "akepa_2002_fitted_model.RDS"
+fit = readRDS(file_name)
+
+study_area = readRDS("data/study_area_extended_no_crs.RDS")
+samplers = readRDS("data/samplers_extended_no_crs.RDS")
+mesh = readRDS("data/mesh_extended_no_crs.RDS")
+
+# intensity
+
+pred.grid <- expand.grid(EASTING=seq(from=255400, to=261200, by=200),
+                         NORTHING=seq(from=2189000, to=2200800, by=200))
+coordinates(pred.grid) = ~ EASTING + NORTHING
+# keep = over(pred.grid, study_area, byid = TRUE)
+# keep[is.na(keep)] = 0
+# keep = as.logical(keep)
+
+# using mgcv inSide() to match Rick's code
+bnd_points = study_area@polygons[[1]]@Polygons[[1]]@coords
+bnd_points = data.frame(EASTING = bnd_points[,1], NORTHING = bnd_points[,2])
+EASTING = pred.grid@coords[,1]
+NORTHING = pred.grid@coords[,2]
+keep = mgcv::inSide(bnd_points, EASTING, NORTHING)
+
+pred.grid.studyarea = pred.grid[keep,]
+
+# to get back as matrix for dsm do 
+pred.grid.matrix = pred.grid.studyarea@coords
+
+ggplot() + gg(pred.grid.studyarea) + gg(study_area)
+
+pred.pxl = SpatialPixels(pred.grid.studyarea)
+pr.int <- predict(fit, pred.pxl, ~ exp(grf + Intercept))
+
+p1 = ggplot() + 
+  gg(study_area) +
+  gg(pr.int) +
+  scale_fill_viridis_c() +
+  ggtitle("Posterior mean intensity") +  
+  xlab("Easting") +
+  ylab("Northing")
+
+p2 = ggplot() +
+  gg(study_area)  +
+  gg(pr.int["sd"]) +
+  gg(samplers, colour = "red") +
+  scale_fill_viridis_c() +
+  ggtitle("Posterior SD of intensity") +  
+  xlab("Easting") +
+  ylab("Northing")
+
[…]

### 2019-11-28 paper-v2.tex
results finished

@@
 Full code for the analysis can be found in the Supplementary materials.
+Since only distance to observer was recorded we used the adjusted intensity given in \eqref{intensity-incomplete}.
 
@@
 Since CV values are influenced both by sampling effort and the magnitude of the estimated intensity means other methods we also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
-\begin{figure}[!t]
+\begin{figure}[h]
 	\includegraphics[scale=0.5]{figures/intensity_quantiles.png}
@@
 
-% POSTERIOR MATERN CORRELATION HERE
-Add posterior matern correlation function
-
+The posterior Mat\'ern correlation function is shown in \autoref{fig:matern-posterior}, the grey regions show the 95\% credible interval. 
 \begin{figure}[h]
@@
 	\caption{Posterior Mat\'ern correlation function}
-	\label{fig:half-normal}
+	\label{fig:matern-posterior}
 \end{figure}
+This shows some correlation even at distances of 10,000 metres.
+However, the northern-most and souther-most boundaries of the study region are approximately 8,000 metres apart.
+The model is estimating positive correlation at distances further apart than were observed.
+[[I'm not sure if this a problem??]]
 
@@
 
+Finally, we note that the one-stage approach is a useful approach when it comes to representing uncertainty.
+All plots here are based on sampling from the joint posterior of all model parameters.
+As such they naturally average over the uncertainty in the detection model.
+
 \clearpage %figure out figure placements later
@@
 	\item mini lit review - communicating uncertainty in spatial models w random effects
+	\item IDEA:  lots of papers don't even both themselves with uncertainty.  Could cite a bunch of recent ones, noting the search term in web of knowledge.
 \end{itemize}

### 2019-12-04 paper-v2.tex
add communication of results section

--- 2019-11-26 paper-v2.tex
@@
 
-Since CV values are influenced both by sampling effort and the magnitude of the estimated intensity means other methods we also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
-\begin{figure}[!t]
+We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
+\begin{figure}[h]
 	\includegraphics[scale=0.5]{figures/intensity_quantiles.png}
@@
 	\caption{Posterior Mat\'ern correlation function}
-	\label{fig:half-normal}
+	\label{fig:matern-posterior}
 \end{figure}
+Note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).
+This is the correlation on the log-intensity scale.
+[[Can I make a better "range of correlation" plot for the posterior intensity process?]]
 
@@
 
-We note that this presentation of abundance estimates differs from typical presentations.
-It is common to report a point estimate of the expected abundance and uncertainty in this point estimate.
-However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.
-We find the distinction between expected and realised abundance to be a useful one.
-Here the posterior intensity process is used to generate possible realisations of the intensity, which are then used to generate the abundance estimates.
+% Add table of estimated model parameters
+% Add 95% CI figures for N
+%% EVAL SPDE AND LGCP POSTERIOR SAMPLING OF barplot k function
 
-\clearpage %figure out figure placements later
-\section{Model Evaluation}
+\clearpage %sort out figure placements later
+
+\section{Communication of Results}
+
+% This is not quite right - model communication?  
+% in Results add a section evaluating SPDE using boxplot k function
 
@@
 
+The results presented above are consistent with a more traditional two-stage model on binned count data.
+Our presentation of the results are broadly consistent with approaches taken in the species distribution modelling literature, although coefficient of variation maps (\autoref{fig:intensity-mean-cv}) and quantile maps (\autoref{fig:intensity-quantiles}) are not always provided.
+We note a worrying tendency in the literature to only provide maps of point estimates without any accompanying communication of uncertainty.
+Where uncertainty maps are provided they tend to be maps of posterior CV, standard deviation (see for example [[CITE THEM]]) and/or posterior or bootstrapped quantiles (see for example [[CITE]]).
+Our intention in this section is to highlight limitations with our above communication of the results and discuss possible remedies.
+
+Whilst summary maps of the posterior intensity process are useful, they can mask certain properties of the fitted model.
+The mapped posterior mean, for example, will be qualitatively smoother than realisations from the posterior.\autoref{fig:intensity-realizations} shows three such realisations.
+Note that each of them has finer-grained spatial structure than is shown in the posterior mean in \autoref{fig:intensity-mean-cv}.
+One interpretation of this is that animals will tend to be more clustered than might be expected if we only looked at the posterior mean.
+
+\begin{figure}
+	\includegraphics[scale=0.35]{figures/intensity_realized.png}
+	\caption{Three realizations of the posterior intensity field}
+	\label{fig:intensity-realizations}
+\end{figure}
+
+During model evaluation we strongly recommend producing plots of realisations as these will more qualitatively resemble spatial structure of the observed data.
+
+More concerning are the problems with the CV, standard deviation and quantile maps.
+The CV map is useful but problematic to interpret in regions of low estimated intensity.
+If the posterior standard deviation is relatively consistent across the survey region, CV values will be higher in regions of low predicted intensity.
+The observed patterns in such maps will just highlight regions of low intensity, not necessarily higher uncertainty.
+This is further confounded in the case with spatially varying posterior standard deviation making the CV map hard to interpet.
+
+The quantile maps also suffer from a potential difficulty in interpretation.
+The temptation is to interpet the maps as showing possible intensity surfaces that could have produced the observed data.
+Presenting these quantiles side by side in a single map obscures the fact that it is vanishingly unlikely for all prediction values to simultaneously achieve the 0.025 or 0.975 quantile.
+We believe most non-statistically trained audiences will struggle with this caveat to interpeting the map.
+To demonstrate the possible consequences of this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile map and 12,400 for the 0.975 quantile map.
+A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance.
+However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
+This demonstrates that the natural tendency to interpet these maps as possible intensities leads to interpretations that are inconsistent with the model that generated the maps.
+
+The standard deviation map is a useful one but care has to be taken to think about what the standard deviation values might imply in the context of a given analysis.
+A default colour scale will more or less always show some spatial variation in uncertainty with some regions of relatively high or low uncertainty.
+What such a map does not show is whether these values actually matter.
+Perhaps the posterior predictions across the entire region are all so uncertain as to the make the predictions operationally useless everywhere.
+Or perhaps differences between high and low uncertainty regions (`high' and `'low' as defined by the colour scale) are small enough to be negligible when it comes to decision making, in which case such a spatially varying map is not a relevant output of the analysis.
+
+Our solution to the problems in the quantile and standard deviation maps is to suggest analysts and stakeholders agree \emph{a priori} on values of the process and uncertainty thresholds that are important in the context of the analysis.
+These values can then be used to construct summary maps.
+We demonstrate this perspective through the use of excursions sets and functions \citep{bolin_excursion_2015}, although there are many possible approaches.
+These techniques require the user to specify thresholds of interest for the process itself and acceptable levels of uncertainty.
+
+This approach to representing variability in stochastic processes specifically considers the joint probability of events across multiple locations.
+The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e.
+the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$.
+For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
+\begin{equation*}
+E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \}
+\end{equation*}
+Note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.
+As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.
+Negative excursion sets can be similarly defined.
+Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.
+An implementation is available in the \texttt{excursions} package \citep{bolin_calculating_2018} available through the Comprehensive R Archive Network \citep{r_2017}.
+
+\autoref{fig:excursions} (left panel) shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95.
+This figure can be interpreted in a natural way.
+It is the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
+\begin{figure}
+	\includegraphics[scale=0.5]{figures/excursions.png}
+	\caption{Left:  The positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.
+Right: The positive excursion function with a level corresponding to 1 bird per hectare}
+	\label{fig:excursions}
+\end{figure}
+To visualise multiple such maps we can use the excursion function $F_u^{+}(s) = \sup \{1 - \alpha ; s \in E_{u,\alpha}^+ \}$, which defines for each location something similar to a p-value.
+For each $s$ it is the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.
+The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursions} (right panel).
+This figure can also be interpreted naturally.
+It shows the largest possible probability for each location to be a member of a set in which the intensity exceeds a threshold simultaneously across all locations in the set.
+It is clear from the figure that regions on the edge of the excursion set would be included if the $\alpha$ value were allowed to increase slightly.
+However, for regions in the north there is essentially no probability level for which those locations would be included in $A_u^{+}(\lambda)$.
+We chose the threshold of 1 bird per hectare and an error probability of 0.05 to demonstrate the approach.  Multiple values for each of these can be considered as a simple extension.
+
 \section{Discussion}
 
-Discuss wisely
+Our intention in this paper was to present a new approach to analysing point transect distance sampling data that incorporates several innovative new methods and considers the entire workflow of model specification, inference, evaluation and communication of results.
+Although each of these areas are substantive enough to be the main focus of a paper, we hope our brief presentation of issues that arise in each area and our chosen solutions will be useful to statisticians who face similar questions and data structures.
+
+spatio-temporal extension. AR1 through time etc.  Citations for this.
+
+This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.
+Three such realizations are in \autoref{fig:intensity-realizations}.
+There are some notable differences when compared to the mean and quantile summary maps.
+Each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.
+In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.
+The summary plots hide the spatial structure the model has been able to estimate from the data.
+In general, the mean of a random field will contain less spatial structure than realisations of the field.
+
+We note that this presentation of abundance estimates differs from typical presentations.
+It is common to report a point estimate of the expected abundance and uncertainty in this point estimate.
+However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.
+We find the distinction between expected and realised abundance to be a useful one.
+Here the posterior intensity process is used to generate possible realisations of the intensity, which are then used to generate the abundance estimates.
 

### 2019-12-05 analyses/eval_model.R
add sd map + tweak some others

@@ -56,30 +56,80 @@ ggsave(filename = "../figures/halfnormal.png",
 # ggplot() + 
 #   gg(dfun)
 
+## Matern plots ##
+
+plot(spde.posterior(fit, "grf", what = "matern.correlation")) +
+  ylab("correlation") + xlab("distance (m)") +
+  ggtitle("Posterior Matern correlation function")
+
+ggsave(filename = "../figures/matern_posterior.png",
+       width = 5, height = 5, units = "in")
+
+# plot 
+ggplot() + 
+  gg(study_area) +
+  gg(mesh) +
+  gg(samplers, colour = "red") +
+  scale_fill_viridis_c() +
+  xlab("Easting") +
+  ylab("Northing") +
+  coord_equal()
+
+plot(mesh, lwd = 0.2, asp =1, main = "")
+
 # intensity
 pxl = pixels(mesh, nx = 100, ny = 100, mask = study_area)
 pr.int <- predict(fit, pxl, ~ exp(grf + Intercept))
 
 p1 = ggplot() + 
   gg(study_area) +
-  gg(pr.int) +
+  gg(pr.int["mean"]) +
   scale_fill_viridis_c() +
   ggtitle("Posterior mean intensity") +  
   xlab("Easting") +
-  ylab("Northing")
+  ylab("Northing") +
+  coord_equal()
 
 p2 = ggplot() +
   gg(study_area)  +
   gg(pr.int["cv"]) +
-  gg(samplers, colour = "red") +
+  # gg(samplers, colour = "red") +
   scale_fill_viridis_c() +
   ggtitle("Posterior CV of intensity") +  
   xlab("Easting") +
-  ylab("Northing")
+  ylab("Northing") +
+  coord_equal()
 
-png(filename = "../figures/intensity_mean_cv.png",
+p3 = ggplot() +
+  gg(study_area)  +
+  gg(pr.int["sd"]) +
+  # gg(samplers, colour = "red") +
+  scale_fill_viridis_c() +
+  ggtitle("Posterior SD of intensity") +  
+  xlab("Easting") +
+  ylab("Northing") +
+  coord_equal()
+
+layout_matrix <- matrix(c(4, 1, 1, 4, 2, 2, 3, 3), nrow = 2, byrow = TRUE)  # triangle arrangement
+
+png(filename = "../figures/intensity_mean_cv_sd.png",
+    width = 10, height = 10, units = "in", res = 100)
[…]

### 2019-12-05 analyses/fit_model.R
tweak starting vals - these ones converge

@@ -44,7 +44,8 @@ g1 <- ggplot() +
   gg(mesh) + 
   gg(study_area) +
   gg(samplers) +
-  gg(realobs, colour = "green")
+  gg(realobs, colour = "green") +
+  coord_equal()
 
 g1
 
@@ -68,7 +69,7 @@ fml <- coordinates + distance ~ grf +
   Intercept
 
 
-starting_values <- data.frame(grf = 0, lsig = 3.4, Intercept = 0)
+starting_values <- data.frame(grf = 0, lsig = 3.36, Intercept = 0)
 
 W <- 60   # transect radius
 distance_domain <- seq(.Machine$double.eps, W, length.out = 30)

### 2019-12-06 analyses/helper_functions.R
add posterior k function stuff

@@ -1,0 +1,62 @@
+#' thin.pp()
+#'
+#' Thin a point pattern using a known detection function and survey design
+#' Point transects only for now - add line transects in future
+#'
+#' Assumes each observation falls within only 1 transect
+#'
+#' TODO:
+#' Change hardcoded "PlotID"
+#' Avoid for loop
+#'
+#' @param pp point pattern SpatialPointsDataFrame
+#' @param samplers SpatialPointsDataFrame of point transect locations   - MUST HAVE column called PlotID, unique identifier of each transect
+#' @param W radius or half strip width
+#' @param detfn  detection function.  First argument must be the vector of distances, second argument list of params
+#' @param snap (default = TRUE) snap sampled points to centre of transects (or to the centre line).  To snap need to have plotID column
+#' @param ... arguments passed to detfn
+#' @return spatial points dataframe
+#' @import sp
+#' @export
+
+
+library(dplyr)
+library(sf)
+
+thin.pp <- function(pp, samplers, W, detfn, snap = TRUE, ...){
+  
+  if (!("plotID" %in% colnames(samplers@data))) stop("samplers needs a column called plotID")
+  samplers_area <- rgeos::gBuffer(samplers, byid = TRUE, width = W)
+  pp_det <- rgeos::gIntersection(pp, samplers_area, byid = TRUE)  # detectable points
+  
+  # Associate each nest with a sampler
+  IDs <- sp::over(pp_det, samplers_area)
+  pp_det$plotID <- IDs$plotID
+  
+  # Calculate distances:
+  distances <- rep(NA, nrow(pp_det))
+  
+  for (i in 1:nrow(pp_det)) {
+    
+    pt <- pp_det[i,]
+    transect <- subset(samplers, plotID == pt$plotID)
+    distances[i] <- rgeos::gDistance(pt, transect)
+  }
+  
+  # add distances column and create prob detected column
+  pdet <- detfn(distances, ...)
+  pp_det$distance <- distances
+  pp_det$pdet <- pdet
+  
+  # subsample detectable to get observed
+  detected <- rbinom(nrow(pp_det), 1, pdet)
+  pp_obs <- subset(pp_det, as.logical(detected))
+  
+  # snap to centre?
+  if (snap) {samplers_sf <- sf::st_as_sf(samplers)  # convert sp to sf
+  test <- dplyr::inner_join(samplers_sf, pp_obs@data, by = "plotID")
[…]

### 2019-12-09 analyses/eval_spde.R
horrible hacky solution to memory issue

@@ -21,19 +21,16 @@ mesh = readRDS("data/mesh_extended_no_crs.RDS")
 obs = readRDS("data/obs_extended_no_crs.RDS")
 
 ### Generate posterior point patterns
-n.pp = 100
+n.pp = 25  
 
 # simulate intensities
 llam = do.call(cbind,
                generate(fit, mesh, ~ grf + Intercept, n.samples = n.pp))
 
-post.pp = sample.lgcp(mesh = mesh, loglambda = llam) # move within for loop if gets too big
-post.pp = post.pp[study_area,]
-
 # create bins and matrix to store counts
 max.ds = max(as.numeric(gDistance(samplers, byid = TRUE)))
 breaks = seq(0, ceiling(max.ds), length.out = 20)
-counts = matrix(NA, nrow = n.pp, ncol = length(breaks)-1)
+
 
 # plot one point pattern
 # a.pp = subset(post.pp, sample == 1)
@@ -54,9 +51,24 @@ hn <- function(distance, lsig) exp(log_hn(distance, lsig))
 # thin.pp needs a plotID columns
 samplers$plotID = samplers$SampleLabel
 
-for (i in 1:n.pp){
-  a.pp = subset(post.pp, sample == i)
-  a.lsig = unlist(generate(fit, data.frame(x=1), ~ lsig, n.samples = 1))
+lsig.post = generate(fit, data.frame(x=1), ~ lsig, n.samples = n.pp)
+
+# memory leakin somewhere in this loop - think rgeos function
+# maybe in thin.pp() ?
+# which time?
+
+# please don't judge me for the code below.
+# I just cant find out what's happening with the memory
+
+counts = matrix(NA, nrow = n.pp, ncol = length(breaks)-1)
+
+# okay run this a few times to get desired sample size.
+# memory problems in the loops.
+for (i in 1:(n.pp)){
+  
+  a.pp = sample.lgcp(mesh = mesh, loglambda = llam[,i])
+  a.pp = a.pp[study_area,]
+  a.lsig = lsig.post[[i]]
   
   a.obs = thin.pp(pp = a.pp, 
                   W = 60, 
@@ -71,22 +83,35 @@ for (i in 1:n.pp){
   a.ds = gDistance(a.obs, byid = TRUE)
   a.ds[upper.tri(a.ds, diag = TRUE)] = NA
   a.ds.vec = as.numeric(a.ds)
[…]

### 2019-12-10 paper-v2.tex
Tweak model communication section

@@
 The CV plot shows that the CV in the posterior intensity field is lower areas with greater sampling effort.
-There is clear preferential sampling and in the north estimated intensities are lower which also contributes to larger CV values.
-
+There is clear preferential sampling with more sampling in the south than the north  (see \autoref{fig:2002studyareapointspt}).
+In the north estimated intensities are lower which also contributes to larger CV values.
 \begin{figure}
-	\includegraphics[scale=0.5]{figures/intensity_mean_cv.png}
-	\caption{(left) Posterior mean of the intensity (right) Posterior coefficient of variation of the intensity}
+	\includegraphics[scale=0.5]{figures/intensity_mean_cv_sd.png}
+	\caption{(left) Posterior mean, CV and SD of the intensity}
 	\label{fig:intensity-mean-cv}
 \end{figure}
-
-We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
+The standard deviation maps shows more overall posterior variability in south where intensity is larger.
+We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location (\autoref{fig:intensity-quantiles}).
 \begin{figure}[h]
@@
 \begin{figure}[h]
-	\includegraphics[scale=0.6]{figures/halfnormal.png}
+	\includegraphics[scale=0.5]{figures/halfnormal.png}
 	\caption{Posterior detection function.  Black line - mean posterior estimate, grey shaded area - 95\% credible interval}
@@
 \begin{figure}[h]
-	\includegraphics[scale=0.6]{figures/matern_posterior.png}
+	\includegraphics[scale=0.5]{figures/matern_posterior.png}
 	\caption{Posterior Mat\'ern correlation function}
@@
 [[Can I make a better "range of correlation" plot for the posterior intensity process?]]
+
+To further investigate the ability of the GMRF component to capture the heterogeneity we generated 100 thinned point patterns from the fitted model and calculated the pairwise distances between all observations \autoref{fig:post-pp-distances}.
+\begin{figure}[h]
+	\includegraphics[scale=0.6]{figures/post_pp_distances.png}
+	\caption{Boxplots of frequency of pairwise distances (tick mark values for end of each bin) based on 100 posterior samples.
+Red circles: observed frequency of pairwise distances between observations}
+	\label{fig:post-pp-distances}
+\end{figure}
+The plot shows that the GMRF component captures the relevant scale of clustering in the observed data.
 
@@
 % Add 95% CI figures for N
-%% EVAL SPDE AND LGCP POSTERIOR SAMPLING OF barplot k function
 
@@
 
-% This is not quite right - model communication?  
-% in Results add a section evaluating SPDE using boxplot k function
+The results presented above are consistent with a more traditional two-stage model on binned count data.
+Our presentation of the results is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (\autoref{fig:intensity-mean-cv} and \autoref{fig:intensity-quantiles}) are not always provided.
+We note a worrying tendency in the species distribution mapping literature to include maps of point estimates without any accompanying communication of uncertainty.
+Where uncertainty maps are provided these tend to be maps of predictive uncertainty such as standard deviation or CV (recent examples include \cite{ fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}) based either on a posterior predictive distribution or bootstrapping.
+Another approach is to map quantiles or the probability of exceeding certain thresholds (recent examples \cite{russell_avoidance_2016, wilson_hierarchical_2010}).
+In the above, we chose to communicate uncertainty through maps of CV, standard deviation, and the 0.025 and 0.975 quantiles.
+Whilst all of these maps are informative and useful in their own way, our intention in this section is to highlight limitations with these maps and suggest ways to address these limitations.
 
-\begin{itemize}
-	\item mini lit review - communicating uncertainty in spatial models w random effects
-	\item IDEA:  lots of papers don't even both themselves with uncertainty.  Could cite a bunch of recent ones, noting the search term in web of knowledge.
-\end{itemize}
-
-\begin{itemize}
-	\item Posterior sampling
-	\item Barplot for random effect
-	\item Excursions
-\end{itemize}
-
-The results presented above are consistent with a more traditional two-stage model on binned count data.
-Our presentation of the results are broadly consistent with approaches taken in the species distribution modelling literature, although coefficient of variation maps (\autoref{fig:intensity-mean-cv}) and quantile maps (\autoref{fig:intensity-quantiles}) are not always provided.
-We note a worrying tendency in the literature to only provide maps of point estimates without any accompanying communication of uncertainty.
-Where uncertainty maps are provided they tend to be maps of posterior CV, standard deviation (see for example [[CITE THEM]]) and/or posterior or bootstrapped quantiles (see for example [[CITE]]).
-Our intention in this section is to highlight limitations with our above communication of the results and discuss possible remedies.
-
-Whilst summary maps of the posterior intensity process are useful, they can mask certain properties of the fitted model.
-The mapped posterior mean, for example, will be qualitatively smoother than realisations from the posterior.\autoref{fig:intensity-realizations} shows three such realisations.
+Whilst summary maps of random fields are useful, they all can mask certain properties of the random field.
+The posterior mean, for example, will be smoother than realisations from the posterior.
+\autoref{fig:intensity-realizations} shows three such realisations.
 Note that each of them has finer-grained spatial structure than is shown in the posterior mean in \autoref{fig:intensity-mean-cv}.
 One interpretation of this is that animals will tend to be more clustered than might be expected if we only looked at the posterior mean.
-
 \begin{figure}
@@
 \end{figure}
+During model evaluation we recommend producing plots of realisations as these will more closely resemble the spatial structure of the observed data.
 
-During model evaluation we strongly recommend producing plots of realisations as these will more qualitatively resemble spatial structure of the observed data.
-
-More concerning are the problems with the CV, standard deviation and quantile maps.
-The CV map is useful but problematic to interpret in regions of low estimated intensity.
+The CV map (\autoref{fig:intensity-mean-cv}) can be hard to interpret in regions of low estimated intensity.
 If the posterior standard deviation is relatively consistent across the survey region, CV values will be higher in regions of low predicted intensity.
-The observed patterns in such maps will just highlight regions of low intensity, not necessarily higher uncertainty.
-This is further confounded in the case with spatially varying posterior standard deviation making the CV map hard to interpet.
+The observed patterns in such maps will highlight regions of relatively lower intensity, not necessarily higher uncertainty.
+This is problem is further confounded in the case with spatially varying posterior standard deviation.
 
@@
 Presenting these quantiles side by side in a single map obscures the fact that it is vanishingly unlikely for all prediction values to simultaneously achieve the 0.025 or 0.975 quantile.
+A similar problem will occur with mapping the marginal probability of exceeding a threshold at each prediction locations.
 We believe most non-statistically trained audiences will struggle with this caveat to interpeting the map.
@@
 However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
-This demonstrates that the natural tendency to interpet these maps as possible intensities leads to interpretations that are inconsistent with the model that generated the maps.
+This demonstrates that the natural tendency to interpet these maps as possible intensities leads to interpretations that are inconsistent with the very model that generated the maps.
 
-The standard deviation map is a useful one but care has to be taken to think about what the standard deviation values might imply in the context of a given analysis.
-A default colour scale will more or less always show some spatial variation in uncertainty with some regions of relatively high or low uncertainty.
-What such a map does not show is whether these values actually matter.
-Perhaps the posterior predictions across the entire region are all so uncertain as to the make the predictions operationally useless everywhere.
-Or perhaps differences between high and low uncertainty regions (`high' and `'low' as defined by the colour scale) are small enough to be negligible when it comes to decision making, in which case such a spatially varying map is not a relevant output of the analysis.
+For the standard deviation map care should be given to what the mapped values imply in the context of a given analysis.
+A default colour scale will more or less always show some spatial variation in uncertainty, with some regions of relatively high or low uncertainty.
+What such a map does not show is whether these values actually matter in the context of the analysis.
+For example, it could be the case that the standard deviation values are so large as to make the predictions operationally useless everywhere.
+Or, on the other end of the scale, perhaps differences between high and low uncertainty regions (`high' and `low' as defined by some default colour scale) are small enough to be negligible when it comes to decision making, in which case such a spatially varying map is not a relevant output of the analysis.
+These are two extreme examples, cases in between suffer from the same problem.
 
-Our solution to the problems in the quantile and standard deviation maps is to suggest analysts and stakeholders agree \emph{a priori} on values of the process and uncertainty thresholds that are important in the context of the analysis.
+Our solution to the problems in the quantile and standard deviation maps is to suggest that consideration should be given \emph{a priori} to values of the process and levels of uncertainty that are important and suitable given the context of the analysis.
 These values can then be used to construct summary maps.
@@
 
+covariates on detection function.  High dimensional taylor falls down at some point?
+
 This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.

### 2019-12-10 paper-v2.tex
add discussion

@@
 In this section we introduce the statistical model for the animal locations and observation process, including a detailed description of how this model can be represented as a modified Poisson likelihood.
-We use this formulation to fit the model using the \texttt{R-INLA} implementation of INLA but this likelihood could in principle be used in any statistical software capable of specifying such a modified Poisson form.
+We use this formulation to fit the model using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension to the \texttt{R-INLA} implementation of INLA.
 
@@
 
-spatio-temporal extension. AR1 through time etc.  Citations for this.
+There are several natural extensions to the model.
+For this analysis we only considered a single survey year.
+Spatio-temporal extensions to this are possible, for example by considering a space-time interaction models.
+\texttt{R-INLA} has the ability to fit space-time effects that can be represented as a kronecker product of the relevant precision matrices.
+An example of this type of space-time interaction effect would be the spatial SPDE above interacting with an AR1 temporal process.
+\cite{blangiardo_spatial_2013} show how this can be done with numerous examples and code snippets.
 
-covariates on detection function.  High dimensional taylor falls down at some point?
+Another extension is to consider more complicated observation processes.
+For example, we did not use any explanatory covariates for the detection function.
+This remains to be investigated further.
+Any additional parameters to the detection function would result in a higher-dimensional outer optimisation problem for iterated INLA.
+This could lead to problem if, for example, identifiability is an issue.
 
-This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.
-Three such realizations are in \autoref{fig:intensity-realizations}.
-There are some notable differences when compared to the mean and quantile summary maps.
-Each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.
-In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.
-The summary plots hide the spatial structure the model has been able to estimate from the data.
-In general, the mean of a random field will contain less spatial structure than realisations of the field.
-
-We note that this presentation of abundance estimates differs from typical presentations.
+We note also that our presentation of abundance estimates differs from typical presentations.
 It is common to report a point estimate of the expected abundance and uncertainty in this point estimate.
@@
 We find the distinction between expected and realised abundance to be a useful one.
-Here the posterior intensity process is used to generate possible realisations of the intensity, which are then used to generate the abundance estimates.
+The credible intervals for the posterior expected abundance will be narrower than the credible intervals for the posterior realised abundance.
+The Bayesian perspective is useful here as well as for the uncertainty in the detection model estimates.
+Since outputs based on sampling from the joint posterior of all model parameters, including the detection model, the Bayesian framework means the resulting model outputs we present here naturally average over the uncertainty in the observation process.
+[[Not really sure if this stuff is worth mentioning or if can be said in a better way]]
+
+Our use of \texttt{inlabru} and \texttt{R-INLA} for inference means we can take advantage of the sparsity of the GMRF precision matrix, a key advantage of the SPDE approach.
+A dense covariance matrix of dimension $n \times n$ typically has an $\mathcal{O}(n^3)$ cost to invert.
+In contrast to this \texttt{R-INLA} makes use of sparse matrix algorithms to invert a sparse 2-dimensional precision matrix at a typical cost of $\mathcal{O}(n^{3/2})$ \citep{lindgren_explicit_2011}.
+This computational efficienty, in addition to the flexibility of the mesh triangulation, can make the SPDE an attractive choice for implementing high-resolution spatial effects in complex domains.
+An additional benefit, as described in detail in \cite{yuan_point_2017}, is that the SPDE can also be implemented on the sphere, making analysis of global datasets computationally feasible.
+
+In our analysis we only briefly touch upon the problem of communicating uncertainty in maps and highlight the excursions methods as a new and positive addition to this area.
+Our example threshold was chosen arbitrarily and it is not obvious to us that there will always be an easy or simple choice of thresholds and levels of uncertainty for a given analysis.
+This requires the input of relevant domain expertise and statistical knowledge so stakeholders understand the consequences of the thresholds and uncertainty levels they wish to consider.
+The most important consequences of an analysis are the possible conclusions and decisions that will be taken on the basis of the results.
+Here we advocate the view that model estimates should be converted into outputs, with associated uncertainty, that are most directly relevant to these conclusions.
+In the above we chose to use excursion sets, excursion functions and the posterior of realised abundance as a demonstration of this.
+It is not always the case that a map of predicted mean abundance will be the best output for a species distribution model.
+
+The point process perspective with spatial SPDE effect can seem fundamentally different to a two-stage GAM approach.
+However, we think that it can be useful to highlight the similarities in approach.
+It can be tempting to think that because with the point process model does not require us to bin the data into counts that we can avoid some loss of information.
+However, we also had to make the same assumption of constant intensity within each transect that is common in more traditional distance sampling approaches.
+This means the exactly location of the points is not used to inform the posterior intensity.
+The assumption of constant intensity within each point transect is essentially the point process version of binning point data into counts.
+
+The SPDE effect can also seem fundamentally different to, say, a spline-based smoother.
+However, there is a fundamental connection between SPDEs and optimal smoothing splines.
+The optimal smoothing spline is a lminimum variance unbiased linear estimator of the posterior mean of an SPDE \citep{kimeldorf_spline_2002}.
+This means a smoothing spline can be interpreted as an SPDE and vice-versa.
+Depending on the methods a user is most familiar with this can be a useful perspective.
+\cite{yue_bayesian_2014} show how to implement common spline penalties in an SPDE framework and \cite{miller_understanding_2019} offer a complementary explanation to interpret and fit SPDE models within standard penalised regression splines software.
 

### 2019-12-10 paper-v2.tex
update study area plot

@@
 	\centering
-	\includegraphics[scale=0.5]{figures/2002studyareapoints_pt}
-	\caption{Study area (light blue) showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.}
+	\includegraphics[scale=0.5]{figures/study_area_design.png}
+	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.}
 	\label{fig:2002studyareapointspt}

### 2019-12-10 figures/make_study_area_fig.R
add study area figure script

@@ -1,0 +1,21 @@
+# Evaluate fitted model
+library(INLA)
+library(inlabru)
+
+# set ggplot theme
+theme_set(theme_minimal())
+
+study_area = readRDS("../analyses/data/study_area_extended_no_crs.RDS")
+samplers = readRDS("../analyses/data/samplers_extended_no_crs.RDS")
+
+png(filename = "study_area_design.png",
+    width = 10, height = 6, units = "in", res = 100)
+
+ggplot() +
+  gg(study_area) +
+  gg(samplers) +
+  coord_equal() +
+  xlab("Easting") +
+  ylab("Northing")
+
+dev.off()

### 2019-12-10 paper-v2.tex
draft sent to janine

@@
 \begin{abstract}
-A New Abstract Goes Here   
+Distance sampling methods aim to estimate the size of animal populations.
+Here we present an analysis of point transect distance sampling data collected on a tropical forest bird, presenting an entire workflow from model conception, inference procedure, model evaluation to communication of results, at each stage discussing the challenges involved and our strategies to address them.
+
+New developments inlcude a new inference method for one-stage distance sampling models within a Bayesian framework based on iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations \citep{rue_approximate_2009} to simultaneously estimate the detectability and spatial distribution of populations of wild animals.
+We take a point process perspective, modelling the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields implemented via a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.
+The Bayesian setting allows sampling from the joint posterior of all model parameters and so the resulting model outputs naturally incorporate uncertainty in the observation model estimates.
+We discuss potential issues with common ways to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels.
+We apply the methods of excursion sets and excursion functions \citep{bolin_excursion_2015} as an example of this approach.
 \end{abstract}
@@
 Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.
-The Hawaiian \akepa{} is an endemic species whose population declined dramatically during the 20th century [[cite??]].
+The Hawaiian \akepa{} is an endemic species whose population declined dramatically during the 20th century [[Rick - citation possible here?]].
 The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.
@@
 However, as is the case in many species distribution models, we usually have reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates.
-For the \akepa{} this takes the form of a north-south gradient that has been investigated but for which no clear-cut explanatory cause has been found [[cite here?]].
+For the \akepa{} this takes the form of a north-south gradient that has been investigated but for which no clear-cut explanatory cause has been found [[Rick - can we cite for this??]].
 From a statistical perspective this suggests the use of spatially-structured random effects to account for the missing covariates.
 
-Fourthly, the resulting statistical model is complex and therefore challenging to effectively communicate the results of the analysis to non-statistically trained stakeholders such as conservation managers and policy officers.
+Fourthly, the resulting statistical model is complex and it is therefore challenging to effectively communicate the results of the analysis to stakeholders with varying levels of statistical training such as conservation managers and policy officers.
 In particular we note the challenge of communicating uncertainty in maps of predicted animal density.
@@
 This is based on iterated model fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009}.
-	\item A computationally efficient Gaussian Markov random field (GMRF) spatial effect based on a stochastic partial differential equation approach \citep{lindgren_explicit_2011} to account for missing covariates.
+	\item A computationally efficient Gaussian Markov random field (GMRF) spatial effect based on a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011} to account for missing covariates.
 	\item An approach to model evaluation and communication based on sampling from the joint posterior of all model parameters and excursions methods \citep{bolin_excursion_2015} for investigating uncertainty in spatial predictions.
@@
 We demonstrate several novel contributions to the problem of species distribution modelling under imperfect detection.
-In particular, the point process perspective with incomplete location information, the one-stage approximate Bayesian inference strategy and focus on communicating uncerainty in spatial predictions using excursions are innovations in this area.
-We also note that although we use a Bayesian approach, maximum-likelihood frameworks could also have been used.
+These include a point process perspective that accounts for incomplete location information; a one-stage approximate Bayesian inference strategy and the use of excursion sets and functions when generating predictive maps.
+Although we use a Bayesian approach, maximum-likelihood frameworks could also have been used for a one-stage analysis.
 Finally, we note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analsyis.
@@
 Our aim is to present a full analysis of the data along with discussion on model specification, inference and model evaluation that reveals key issues in the analysis and our strategies to address them.
-Our end hope is that the analysis we present here may provide other researchers with similar data structures an awareness of the issues involved and some options for strategies to use in their own analysis.
+Our end hope is that the analysis presented may provide an example of the issues involved and some new options for strategies to address them.
 
@@
 \subsection{Data description}
-[[ Need to update this with extended data ]]
 

### 2019-12-10 analyses/summarise_obs.R
summarise_obs.R

@@ -1,0 +1,24 @@
+# Evaluate fitted model
+library(INLA)
+library(inlabru)
+
+study_area = readRDS("data/study_area_extended_no_crs.RDS")
+samplers = readRDS("data/samplers_extended_no_crs.RDS")
+obs = readRDS("data/obs_extended_no_crs.RDS")
+
+# number of transects
+nrow(samplers@coords)
+
+# number of detections
+nrow(obs@coords)
+
+# number of transects with at least one detection
+length(unique(obs$SampleLabel))
+
+# range of counts at each sampling unit
+library(dplyr)
+hmm = data.frame(SampleLabel = obs$SampleLabel)
+hmm %<>% 
+  group_by(SampleLabel) %>% 
+  summarise(n = n())
+max(hmm$n)

### 2019-12-10 paper-v2.tex
actually this one to Janine

@@
 \begin{abstract}
-Distance sampling methods aim to estimate the size of animal populations.
-Here we present an analysis of point transect distance sampling data collected on a tropical forest bird, presenting an entire workflow from model conception, inference procedure, model evaluation to communication of results, at each stage discussing the challenges involved and our strategies to address them.
+Distance sampling methods aim to estimate the size and spatial distribution of populations of wild animals.
+Here we present an analysis of point transect distance sampling data collected on a tropical forest bird.
+We consider an entire workflow, from model conception, inference procedure, model evaluation to communication of results, at each stage discussing the challenges involved and our strategies to address them.
 
-New developments inlcude a new inference method for one-stage distance sampling models within a Bayesian framework based on iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations \citep{rue_approximate_2009} to simultaneously estimate the detectability and spatial distribution of populations of wild animals.
+New developments inlcude an inference method for one-stage distance sampling models within a Bayesian framework based on iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations \citep{rue_approximate_2009}.
+This approach simultaneously estimates the detectability and spatial distribution of populations of wild animals within a joint model.
 We take a point process perspective, modelling the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields implemented via a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.
@@
 
-[[Need to update this and the figure with the extended data area]]
-
 Hakalau was established in 1985 to conserve 15,390-ha of montane forest habitat for native forest birds and rainforest plants.
@@
 Survey points were established along 15 transects following a systematic, random design with points approximately 150 m apart on transects located either 500 or 1,000 m apart.
-Following \cite{camp_population_2010, camp_statespace_2016}, we restrict our study area to the 3,061 ha open-forest stratum of Hakalau.
+We restrict our study area to the open-forest and closed-forest strata of Hakalau, an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016} which only considered did not include the closed-forest strata which was not sampled in the early years of the \akepa{} surveys.
 The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.

### 2020-02-18 paper-v2.tex
some of Janine comments

@@
 The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.
-This information is critical when making decisions about conservation strategies.
+This information is critical when decisions about conservation strategies are made.
 However, answering these questions presents many statistical challenges.
 
-Firstly, as in many ecological surveys, it is impossible to undertake a census.
+Firstly, as in many ecological surveys, it is impossible to undertake a full census.
 The existing population numbers in the thousands and lives in dense forest where logistical and ecological challenges mean a census is not feasible.
-For this reason, a monitoring survey must sub-sample in space and time in some way.
-For the \akepa{} this takes the form of the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986} which is a large-scale, quantitative survey of Hawaiian forest birds.
-Annual surveys of the \akepa{} study-region consist of survey points located along randomly located transects.
+For this reason, a monitoring survey must sub-sample appropriately in space and time.
+For the \akepa{} this takes the form of the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}, which is a large-scale, quantitative survey of Hawaiian forest birds.
+Annual surveys of the \akepa{} study-region consist of a number of point transects located along randomly located line transects.
 
-Secondly, even at sampled locations, detectability of animals is uncertain.
-The \akepa{} survey attempts to estimate detectability using a distance sampling approach \citep{buckland_distance_2015} where for each observation the distance to the observer is recorded.
-A parametric form of detection function is assumed to decay with increasing distance and is estimated from the observed distances.
-Because of this, answering questions related to the abundance and spatial distribution must involve the estimation of a complex observation process.
+Secondly, even at sampled locations, detectability of animals is unknown.
+The \akepa{} survey estimates detectability using a point transect distance sampling approach \citep{buckland_distance_2015} where, for each observation, the distance to the observer is recorded.
+The approach assumes a parametric form of detection function that decays with increasing distance.
+The detection function parameters are estimated from the observed distances, assuming the true density is uniform.
+This combination of spatio-temporal subsampling and unknown detectability means statistical methods to analyse such data must model a complex observation process along with the spatio-temporal distribution of animals.
 
-Thirdly, due to the spatial sub-sampling, a key aim of the analysis is the generation of spatial predictions.
-However, as is the case in many species distribution models, we usually have reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates.
-For the \akepa{} this takes the form of a north-south gradient that has been investigated but for which no clear-cut explanatory cause has been found [[Rick - can we cite for this??]].
-From a statistical perspective this suggests the use of spatially-structured random effects to account for the missing covariates.
+Thirdly, a key aim of the analysis is to predict animal density at un-sampled locations.
+If density is modelled using spatial covariates the usual approach is to assume the relationship between the animals and spatial covariates is the same in the un-sampled locations as it is at the sampled locations.
+Predictions can then be based on estimating this relationship using the observed data.
+However, we usually have reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates available or for which, in principle, no covariate could be constructed (e.g.
+the 'sociality' of the species or other behavioural characteristics).
+For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient that has been investigated but for which no straightforward explanatory cause has been found [[CITE: Ask Rick if poss]].
+From a statistical perspective, this suggests the inclusion of spatially-structured random effects in the model, which necessitates the use of appropriate statistical software capable of fitting such models.
 
-Fourthly, the resulting statistical model is complex and it is therefore challenging to effectively communicate the results of the analysis to stakeholders with varying levels of statistical training such as conservation managers and policy officers.
-In particular we note the challenge of communicating uncertainty in maps of predicted animal density.
+Fourthly, the above challenges mean the resulting statistical model is necessarily complex and it is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.
+The data has been collected with the clear objective of monitoring the population and informing conservation management decisions.
+These decisions will be taken by stakeholders with a range of statistical expertise who should have a good understanding of the full range of possible inferences implied by the model even if they do not understand all the statistical technicalities involved.
+This final step of communicating results is therefore key and should be considered with as much care as the rest of the analysis.
+This requires the input of statisticians to ensure uncertainty is appropriately captured and communicated in the key outputs of the analysis upon which decisions will be based.
+In particular, in the context of species distribution modelling, we note the challenge of communicating uncertainty in maps of predicted animal density and making clear the difference between uncertainty in the estimated \textit{expected} abundance versus uncertainty in the \textit{realised} abundance, the latter of which is likely more relevant to ecological management.
 
-In this paper we present an analysis that seeks to address each of these issues by presenting
+In this paper, we present an analysis that seeks to address each of these issues by presenting
 
 \begin{enumerate}[(i)]
-	\item A model-based spatial point process perspective on point transect distance sampling, representing the observation model as a thinning probability function.
-	\item A one-stage approximate Bayesian approach to inference to simultaneously estimate the observation model and spatial distribution.
-This is based on iterated model fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009}.
-	\item A computationally efficient Gaussian Markov random field (GMRF) spatial effect based on a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011} to account for missing covariates.
-	\item An approach to model evaluation and communication based on sampling from the joint posterior of all model parameters and excursions methods \citep{bolin_excursion_2015} for investigating uncertainty in spatial predictions.
+	\item A model-based spatial point process perspective on point transect distance sampling, representing the observation model as a thinning of a point processes.
+	\item A one-stage approximate Bayesian approach to inference to simultaneously estimate the observation model and spatial distribution based on iterated model fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009}.
+	\item A computationally efficient Gaussian Markov random field (GMRF) spatially structured random effect specified through a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011} to account for unobserved drivers of the spatial distribution.
+	\item Examples of model evaluation and communication of results that take advantage of the one-stage approach by sampling from the joint posterior of all model parameters and introducing excursions methods \citep{bolin_excursion_2015} as a method for investigating uncertainty in spatial predictions of density.
 \end{enumerate}
 
-We note that these challenges are generic to many types of wildlife survey data, not only the \akepa{} survey, and can be addressed by a range of possible statistical approaches.
+We demonstrate several novel contributions to the problem of species distribution modelling under unknown detectability.
+These include a point process perspective that accounts for incomplete location information; a one-stage approximate Bayesian inference strategy and the use of excursion sets and functions to communicate uncertainty in predictive maps.
+We note that the challenges we have addressed are generic to many types of wildlife survey data, not only the \akepa{} survey, and can be tackled using a range of possible statistical approaches.
 Our analysis here is one possible choice of approach and throughout the remainder of the paper we will highlight differences with existing approaches.
-We demonstrate several novel contributions to the problem of species distribution modelling under imperfect detection.
-These include a point process perspective that accounts for incomplete location information; a one-stage approximate Bayesian inference strategy and the use of excursion sets and functions when generating predictive maps.
-Although we use a Bayesian approach, maximum-likelihood frameworks could also have been used for a one-stage analysis.
-Finally, we note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analsyis.
 
-Our aim is to present a full analysis of the data along with discussion on model specification, inference and model evaluation that reveals key issues in the analysis and our strategies to address them.
-Our end hope is that the analysis presented may provide an example of the issues involved and some new options for strategies to address them.
+Our aim is to present a full analysis of the data along with discussion on model specification, inference and evaluation that reveals key issues in the analysis and our strategies to address them.
+Our hope is that the analysis presented here provides a comprehensive example of issues that arise when analysing wildlife survey data along with some novel statistical approaches that practitioners may wish to consider in their own analysis.
 
@@
 Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}.
-Information from the HFBS was used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau) that was the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
-Due to the broad-scale coverage and robust design HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.
+Information from the HFBS is used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau), that is the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
+Due to the broad-scale coverage and robust design, HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.
 
@@
 Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern.
-Moreover, having unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.
+Moreover, unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.
 
-\subsection{Study area and sampling method}
+\subsection{Study area and survey design}
 
@@
 Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance.
-Survey points were established along 15 transects following a systematic, random design with points approximately 150 m apart on transects located either 500 or 1,000 m apart.
-We restrict our study area to the open-forest and closed-forest strata of Hakalau, an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016} which only considered did not include the closed-forest strata which was not sampled in the early years of the \akepa{} surveys.
+Survey points were established along 15 line transects following a systematic, random design with point transects approximately 150 m apart on line transects located either 500 or 1,000 m apart.
+We limit our study area to the open-forest and closed-forest strata of Hakalau (\autoref{fig:2002studyareapointspt}), an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016} who omitted the closed-forest area because it was not sampled in the early years of the \akepa{} surveys.
+For our analysis we consider data from a later year in the survey in which the closed-forest strata was sampled and so we include it in our analysis.
 The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.
-To the north the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig.
+To the north, the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig.
 \ref{fig:2002studyareapointspt}).
-The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of their study area.
+The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of the study area.
 The west side of the study area is bounded by pasture that is dominated by grass and is unsuitable habitat for \akepa.
@@
 
-\subsection{Data description}
-
-For the purposes of our analysis we selected a single survey from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability.
+For the purposes of our analysis we select a single survey year from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability.
 In 2002, 289 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig.
-\ref{fig:2002studyareapointspt}).
-276 \akepa{} were detected on 121 point transects.
-The number of detections within each sampling unit ranged from zero to 6.
+\ref{fig:2002studyareapointspt}) during which 276 \akepa{} were detected on 121 point transects.
+The number of detections within each point transect ranged from zero to 6.
+We select a single year of data to demonstrate our approach in a simplified setting where temporal aspects of analysis are not considered.
+Our approach is easily extendible to spatio-temporal modelling frameworks which we discuss in more detail in the discussion.
 
@@
 
-Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer \citep{buckland_advanced_2004, buckland_distance_2015}.
-In this section we present a brief historical survey of distance sampling approaches and the approaches of combining distance sampling methods with spatially explicit statistical models.
-We end by introducing a predecessor to our method and note differences with previous approaches.
+Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and an assumed detection model to estimate the detectability of animals as a function of distance from transect \citep{buckland_advanced_2004, buckland_distance_2015}.
+Classic distance sampling approaches use a hybrid of design- and model-based inference to estimate population size where the probability of detection is modelled and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2004}.
 
-Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size where the probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2004}.
+More recently, interest has focused on fully model-based approaches that include a spatially explicit model for animal density and allow non-randomized survey designs to be used.
+These methods allow density to be associated with spatially-indexed covariates and animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
+This additional flexibility to make predictions for smaller, user-specified geographic units, has made model-based distance sampling a common approach in the literature (see e.g.
+[[CITE HERE]]).
 
-More recently, interest has focused on fully model-based approaches that include a spatially explicit model for animal density and allow the use of non-randomized survey designs.
-These methods also allow density to be associated with spatially-indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
-
-Model-based distance sampling has been implemented in a two-stage modelling framework.
-In the first stage, detectability is estimated within each sampling unit.
-In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset.
-Due to often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.
-Negative-binomial and Tweedie distributions are common choices to model the distribution of counts.
-This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review and software package \texttt{dsm} to fit the models.
+Model-based distance sampling has most commonly been implemented in a two-stage modelling framework, see the \texttt{dsm} package \citep{miller_spatial_2013}, for example.
+In the first stage, detectability is estimated.
+In the second stage, detectability estimates from the first stage are used as an offset in a generalized additive model framework.
+Detections within sampling units are binned into counts and an appropriate response distribution is chosen.
+Due to the often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.
+Negative-binomial and Tweedie distributions are common choices.
+This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review.
 
@@
 
+Although we use a Bayesian approach, maximum-likelihood frameworks could also have been used for a one-stage analysis.  
+Finally, we note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analsyis.
+
+Discuss selectivity, saturated responses?? Disease modelling?
+
 \clearpage

### 2020-02-18 paper-v2.tex
up to "intensities for incomplete data"

@@
 
-A key concern with the two-stage approach is the propagation of the uncertainty from the detectability estimates to the second-stage spatial model.
-Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units as well as the challenge of combining smoothers and spatial bootstraps \citep{bravington_reliable_2018-1}.
-Instead, \cite{bravington_reliable_2018-1}  proposed avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the first-stage maximum-likelihood estimate.
+A key concern with the two-stage approach is the propagation of uncertainty from the detectability estimates to the second-stage spatial model.
+Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_reliable_2018-1}.
+Instead, \cite{bravington_reliable_2018-1} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the first-stage maximum-likelihood estimate.
 
-Concerns about uncertainty propagation can be avoided by one-stage modelling approaches.
-In our experience these tend to be developed within a Bayesian framework.
-These approaches use data augmentation and to model unobserved individuals or groups and Markov-chain Monte Carlo (MCMC) methods for inference \citep{royle_hierarchical_2008, schmidt_using_2012}.
-More recently  \cite{oedekoven_bayesian_2014} provided a one-stage model that avoids data augmention by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty by applying a reversible jump MCMC algorithm.
+Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.
+These approaches tend to be Bayesian and use data augmentation and to model unobserved individuals or groups and Markov-chain Monte Carlo (MCMC) methods for inference (see )\citep{royle_hierarchical_2008, schmidt_using_2012}, for example).
+\cite{oedekoven_bayesian_2014} demonstrate a one-stage model that avoids data augmentation by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty using reversible jump MCMC.
 
-The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017} who use the approximate Bayesian method INLA applied to line transect data.
-INLA can be an attractive choice when working with spatially-structured random effects in large or complex spatial domains.
-The SPDE approach \citep{lindgren_explicit_2011} allows the construction of a sparse prior precision matrix which INLA can exploit for gains in computational efficiency and the finite element methods based on triangulation have been shown to be more efficient than regular lattice based approaches \citep{simpson_going_2016}, especially in applications with complex boundaries such as coastlines.
+The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use the approximate Bayesian method of integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009} applied to line transect distance sampling data.
+INLA is a popular choice for fitting models with spatially-structured random effects in large or complex spatial domains due to the implementation of a computationally efficient approach based on the construction of a sparse prior precision matrix for the random effect parameters.
+This sparse precision matrix specifies a GMRF representation of a Gaussian random field defined as the stationary solutions to an SPDE \citep{lindgren_explicit_2011} (see Section \ref{GMRFexplanation} for more details on this approach).
+The finite element methods involved use a flexible triangulation of the spatial domain and have been shown to be more efficient than regular lattice based approaches \citep{simpson_going_2016}, a useful feature for applications with complex boundaries which appear in wildlife survey data such as boundaries formed by coastlines, rivers and other natural features.
 
-\citet{yuan_point_2017} also take a point process perspective and formulate the detection model as a thinning of the full but imperfectly observed point pattern.
-Key to their approach was formulating the detection model using a stochastic partial differential equation.
-Similar to spline approaches, choosing a B-spline basis and finite element methods results in a sparse prior precision matrix for the detection model coefficients which can then be incorporated into the latent-Gaussian model framework required by INLA.
-Downsides to this approach are that it is challenging to mandate a monotonically decreasing detection function.
-The authors suggestion to reject any non-monotonically decreasing functions when sampling from the posterior seems unsatisfactory.
+\citet{yuan_point_2017} also take a point process perspective and formulate the detection model as a thinning of the full, but imperfectly observed, point pattern.
+Key to their approach was formulating the detection model as the solution to an SPDE.
+Choosing a B-spline basis and applying finite element methods results in a sparse prior precision matrix for the detection model coefficients which can then be incorporated into INLA using the same machinery as already exists for latent Gaussian precision matrices.
+A major downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing - usually a key feature of detection functions.
+The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior seems unsatisfactory.
+It is also not clear how covariate effects on the detection function could be incorporated in this approach.
 
-Here we present a similar approach to \citet{yuan_point_2017} in that our models are fitted using INLA.
-However, we avoid the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal.
-This will be more familiar to users of traditional distance samplign approaches.
-This results in components of the additive predictor that are non-linear in their parameters.
-To deal with this we use a method of iterated model fits based on a Taylor expansion of the non-linear model components.
+Here we present a related approach to the one in \citet{yuan_point_2017} in that our models are fitted using INLA.
+However, we avoid these issues with the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal, as a user-written R function.
+This approach of choosing a parametric family of detection functions will be more familiar to users of classical distance sampling methods than the SPDE specification.
+However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in \text{R-INLA}.
+We address this by using a method of iterated model fits based on a Taylor expansion of the non-linear model components.
 We name this approach \textit{iterated INLA}.
 
-We also apply the methods to point transect data which require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
-Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
-However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
-We also further develop the point process viewpoint to allow point process models to be applied to data with incomplete location information.
-It is common for distance sampling surveys to only record the distance to the observer and the observer location.
-We show how the point process perspective can still handle this type of data even though the exact location of the point cannot be drawn on a map.
+We apply these methods to point transect \akepa data which differ from line transect methods to account for the fact that the area surveyed increases with increasing distance from the transect.
+Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
+However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982, waagepetersen_likelihood-based_2006}.
 
@@
 
-In this section we introduce the statistical model for the animal locations and observation process, including a detailed description of how this model can be represented as a modified Poisson likelihood.
-We use this formulation to fit the model using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension to the \texttt{R-INLA} implementation of INLA.
+In this section we introduce the statistical model for the spatial pattern of animal locations and the observation process, including a detailed description of how this model can be written as a modified Poisson likelihood.
+We use this formulation to fit the model using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension to the \texttt{R-INLA} implementation of INLA, which implements the iterated INLA approach.
 
-We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(s)$.
-The log-Gaussian Cox process is a flexible approach that can include spatially structured random effects on the intensity process to account for unexplained heterogeneity not captured by fixed-effect covariates.
+We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(.)$, i.e.
+the intensity at location $s$ is a random variable $\lambda(s)$.
+The log-Gaussian Cox process is a flexible approach that can include spatial covariates to model the mean intensity and a mean-zero spatially structured random effect to account for unexplained heterogeneity not captured by the covariates \citep{moller_log_1998}.
 
-\sloppy For the case with imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected})$.
+To account for the imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected})$.
 A key property of the log-Gaussian Cox process is that a realisation of a point process with intensity process $\lambda(s)$ that is thinned by thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.
@@
 Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance.
-The distance measured depends on the type of survey.
+The type of distance measured depends on the type of survey.
 For line transects the perpendicular distance to the transect line is used.
 For point transects the horizontal distance to the observer is the relevant distance.
-For the remainder of the paper we assume a point transect survey design.
+For the remainder of the paper we assume a point transect survey design and hence horizontal distance to the observer located at the centre of the point transect.
 
-The thinning probability function $g(s)$ is specified as a member of a parametric family of functions.
-In this paper we demonstrate our approach using the half-normal formulation, although others such as hazard rate, are also common.
-If $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated.
-The parameter $\sigma$ can only be estimated if an assumption is made about the intensity of the animal locations.
+The thinning probability function is specified as a parametric family of functions with parameter vector.
+For example, if $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$, where $\sigma$ is a variance parameter to be estimated.
+Other types of detection function exist, such as hazard-rate, negative exponential, and they may or may not be adjusted using series expansions to add some flexibility to obtain a better fit to a given dataset \citep{buckland_distance_2015}.
+In this analysis we assume a half-normal detection function with no covariates on detection parameters.
+
+The detection function parameters can only be estimated if an assumption is made about the intensity of the animal locations.
 Without such an assumption detectability and intensity are confounded.
 The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(s)$.
-Thus any observed differences from uniformity can be attributed to detectability and not variation in the intensity.
+Thus any observed deviations from uniformity can be attributed to detectability and not to variation in the intensity.
 
-A point transect distance sampling survey consists of a set of $K$ sampling units.
-The $k$-th sample unit we denote $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region $\Omega = \cup_{k=1}^K \Omega_k$.
-For simplicity we assume that all sampling units are discs with radius $W$ and $\Omega_j \cap \Omega_k = \emptyset$ for all $j \neq k$.
-The thinning probability function $g(s)$ is defined relative to the positions of the sample units.
-The probability of observing a point at location $s$ in the $k$-th sampling unit $\Omega_k$ we denote $g_k(s)$.
+A point transect distance sampling survey consists of a set of $K$ point transects.
+We denote $k$-th subset of space covered by the $k-th$ point transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.
+For simplicity, we assume that all point transects are non-overlapping discs with radius $W$, as is the case in the HFBS survey.
+The probability of observing a point at location $s \in \Omega_k$ we denote $g_k(s)$.
 The probability of observing a point outside the surveyed region is zero.
-Since the surveyed regions are non-overlapping each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.
+Since the point transects are non-overlapping, each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.
 For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
-The assumption of non-overlapping survey regions can be relaxed by including extra information such as the time of each observation but, for simplicity, we do not consider this here.
+In practice the observer is always located at the centre of the point transect.
+The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation but this is not a requirement here.
 The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for $s \in \Omega_k$.
@@
 \end{equation}
-where we have suppressed parameter vector notation.
-The integral component of the likelihood does not usually have analytical solutions.
-Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach of \cite{simpson_going_2016} for a fully observed point pattern.
+where $\lambda(.)$ and $g(.)$ both depend on parameter vectors omitted for readability.
 
-To evaluate the integral we introduce polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.
-In general the thinning function depends only on $r$ and not on $\theta$.
+\subsection{Writing the model as a modified Poisson likelihood}
+
+Here we show how to write the thinned log Gaussian Cox process model as a modified Poisson likelihood.
+The integral component $\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s$ in \autoref{lgcp-likelihood} does not have analytical solutions.
+Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach developed by \cite{simpson_going_2016} for a fully observed point pattern.
+
+To evaluate the integral we use polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.
+In general, the thinning function depends only on $r$ and not on $\theta$.
 We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$.
-It follows that the integral can be rewritten as a sum of weighted one-dimensional integrals $\sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$ using the assumption that $\lambda(s)$ is constant within each sampling unit.
+The integral can then be rewritten as a sum of weighted one-dimensional integrals $\sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$ using the assumption that $\lambda(s)$ is constant within each sampling unit.
 For each sampling unit we approximate the one-dimensional integral using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
@@
 \begin{equation*}
-	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj})
+	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj}) \;\;\; ,
 \end{equation*}
@@
 To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.
-Similarly let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$.
+Similarly, let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$.
 Then the approximate log-likelihood is
@@
 \end{equation}
-This approximation can be written as a modified Poisson likelihood.  To see this let $\exp \eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
+This approximation can be expressed as a modified Poisson likelihood.  To see this let $\exp \eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
 $\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $y = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.
@@
 \end{equation}
-This is similar to a Poisson likelihood with a $\log\alpha$ offset and can be specified using the Poisson likelihood implementation available in \texttt{R-INLA}.
+This is similar to a Poisson likelihood with offset $\log\alpha$ and can be specified using the Poisson likelihood implementated in \texttt{R-INLA} and \texttt{inlabru}.
+The approach presented here is similar to the Berman-Turner device \citep{berman_approximating_1992, baddeley_practical_2000} although we relax the assumption that the observed points form part of the quadrature scheme.
 
@@
 
-In the above we assume the data are complete records of animal location.
+In the above, we assume the data are complete records of animal locations.
 However, in many distance sampling surveys only the location of the observer and the distance to the observer are recorded.
-Data of this type can be analysed within a point process framework by deriving the appropriate intensity function for the incomplete data.
+Data of this type can be analysed within a point process framework by deriving the appropriate intensity function for the incomplete data as outlined below.
 
@@
 \tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(s)g_k(s)\mathrm{d}s \nonumber \\
-&= 2\pi\lambda(s_k)rg_k(r)
+&= 2\pi\lambda(s_k)rg_k(r) \;\;\; ,
 \end{align}
@@
 \section{Gaussian Markov random field spatial effect}
+\label{GMRFexplanation}
 
@@
 Although we use a Bayesian approach, maximum-likelihood frameworks could also have been used for a one-stage analysis.  
-Finally, we note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analsyis.
+Finally, we note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analysis.
 
@@
 \clearpage
-\bibliographystyle{chicago}
+\bibliographystyle{plainnat}
 \bibliography{paper}

### 2020-02-19 paper-v2.tex
up to end of results

@@
 The thinning probability function is specified as a parametric family of functions with parameter vector.
-For example, if $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$, where $\sigma$ is a variance parameter to be estimated.
+For example, if $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$, where $\sigma^2 > 0$ is a variance parameter to be estimated.
 Other types of detection function exist, such as hazard-rate, negative exponential, and they may or may not be adjusted using series expansions to add some flexibility to obtain a better fit to a given dataset \citep{buckland_distance_2015}.
@@
  where $c_k(r)$ is a circle of radius $r$ centred at $s_k$ and the second line follows from changing to polar coordinates and the assumption of constant intensity within $\Omega_k$.
-This intensity differs from the full data case by including the $2\pi$  term to account for the fact that we do not observe $\theta$ and the additional $r$ term to account for the increasing area surveyed at larger distances.
+This intensity differs from the full data case through the $2\pi$  term that accounts for the fact that we do not observe $\theta$ and the additional $r$ term that accounts for the increasing area surveyed at larger distances.
+The log-intensity for observed points (the thinned intensity) is thus $\log\tl(s) = \log 2 \pi r + \log\lambda(s) + \log g(s)$.
+As we note above, the detection function $g(s)$ is non-linear in its parameters and this component of the additive predictor requires a novel approach to parameter estimation.
 
@@
 
-The log-intensity for observed points (the thinned intensity) is $\log\tl(s) = \log\lambda(s) + \log g(s)$.
-This presents a problem since $\log g(s)$ is not typically linear in its parameters.
-However, for any single set of parameter values $g(s)$ can be viewed as an offset term.
-If we could specify $g(s)$ as known then we could proceed with an inference approach suitable for additive linear predictors.
+The half-normal detection function depends on the strictly positive parameter $\sigma^2$.
+We choose a log link and consider $g(s | \phi)$, where $\log\sigma^2 = \phi$.
+Our approach is based on the observation that for any single choice of $\phi$, the $\log g(s | \phi)$ term acts essentially as an offset.
+i.e If we could specify $g(s)$ as known then we could proceed with an inference approach suitable for additive linear predictors.
 
@@
 
-\section{Gaussian Markov random field spatial effect}
+\section{Spatially Structured Random Effect}
 \label{GMRFexplanation}
 
-Using spatially-structured random effects can be computationally expensive, resulting in dense prior precision matrices.
-A key development in this area is the use of Gaussian Markov random field (GMRF) approximations to Gaussian random fields (GRFs).
-The link between them is via a stochastic partial differential equation solved using finite element methods \citep{lindgren_explicit_2011}.
-
-To model the intensity of animal locations we use a combination of fixed effects of spatial covariates and a spatially structured random effect.
-This allows us to associate our survey data with environmental and ecological conditions that may be relevant to understanding the sdistribution of the species of interest.
-For cases where we have insufficient covariates to fully explain the location of animals the spatially structured random effect accounts for additional heterogeneity in the intensity and is required to assume conditional independence between points, given the intensity.
+Using spatially-structured random effects can be computationally expensive, often resulting in dense prior precision matrices which typically have $\mathcal{O}(n^3)$ cost to factorise for an effect with $n$ parameters - sometimes referred to as the `big $n$ problem'.
+A key development in this area is the use of Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs).
+This representation is constructed via a stochastic partial differential equation solved using finite element methods \citep{lindgren_explicit_2011}.
 
@@
 \end{equation*}
-where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatially indexed covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean GRF with Mat\'ern covariance
+where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatial covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean GRF with Mat\'ern covariance
 \begin{equation}
@@
 where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.
-All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which controls the mean-square differentiability of the process.
+All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which specifies the mean-square differentiability of the process.
 We set $\nu = 1$ which is the default value in \texttt{R-INLA}.
-[[Would like to justify $\nu = 1$ somehow?]]
 
-We use a GMRF approximation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.
+We use a GMRF representation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.
 Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
@@
 
+The SPDE approach can at first appear to be fundamentally different to other spatial random effect approaches such as penalised regression splines.
+However, the two methods are closely related and the GRF model component in our model plays a similar role to a tensor product spline on $s$ with an appropriate smoothing penalty.
+See \cite{yue_bayesian_2014, miller_understanding_2019} for more details on the similarites between penalised regression splines and the SPDE approach.
+
 \section{Results}
 
-We fitted the models using penalised complexity priors for the hyperparameters of the GMRF \citep{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$ where $\rho$ is the range of the Mat\'ern covariance.
+We fitted the models using penalised complexity priors for the hyperparameters of the GMRF \citep{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$, where $\rho$ is the range parameter of the Mat\'ern covariance.
 The value of 130 was selected based on the minimum distance between sampling locations.
-Full code for the analysis can be found in the Supplementary materials.
+The complete code for the analysis can be found in the Supplementary materials.
 Since only distance to observer was recorded we used the adjusted intensity given in \eqref{intensity-incomplete}.
@@
 The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.
-This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary materials).
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[scale=0.4]{figures/intensity_mean_cv_sd.png}
+		\caption{Posterior mean, CV and SD of the intensity}
+		\label{fig:intensity-mean-cv}
+	\end{center}
+\end{figure}
+This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary material).
 The CV plot shows that the CV in the posterior intensity field is lower areas with greater sampling effort.
-There is clear preferential sampling with more sampling in the south than the north  (see \autoref{fig:2002studyareapointspt}).
-In the north estimated intensities are lower which also contributes to larger CV values.
-\begin{figure}
-	\includegraphics[scale=0.5]{figures/intensity_mean_cv_sd.png}
-	\caption{(left) Posterior mean, CV and SD of the intensity}
-	\label{fig:intensity-mean-cv}
-\end{figure}
-The standard deviation maps shows more overall posterior variability in south where intensity is larger.
-We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location (\autoref{fig:intensity-quantiles}).
+There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north  (see \autoref{fig:2002studyareapointspt}).
+In the north, estimated intensities are lower, which also contributes to larger CV values.
+The standard deviation map shows stronger overall posterior variability in the south where intensity is larger.
+We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location.
+[[Say something about these - hard to know what to say about them?]]   (\autoref{fig:intensity-quantiles}).
 \begin{figure}[h]
-	\includegraphics[scale=0.5]{figures/intensity_quantiles.png}
+	\begin{center}
+		\includegraphics[scale=0.4]{figures/intensity_quantiles.png}
 		\caption{The predicted posterior intensity summarised in three plots by the mean (upper),  0.025 quantile (lower left) and 0.975 quantile (lower right) for each prediction grid cell.
@@
 		\label{fig:intensity-quantiles}
+	\end{center}
 \end{figure}
@@
 \begin{figure}[h]
-	\includegraphics[scale=0.5]{figures/halfnormal.png}
+	\begin{center}
+		\includegraphics[scale=0.4]{figures/halfnormal.png}
 		\caption{Posterior detection function.  Black line - mean posterior estimate, grey shaded area - 95\% credible interval}
 		\label{fig:half-normal}
+	\end{center}
 \end{figure}
-
 The posterior Mat\'ern correlation function is shown in \autoref{fig:matern-posterior}, the grey regions show the 95\% credible interval. 
 \begin{figure}[h]
-	\includegraphics[scale=0.5]{figures/matern_posterior.png}
+	\begin{center}
+		\includegraphics[scale=0.4]{figures/matern_posterior.png}
 		\caption{Posterior Mat\'ern correlation function}
 		\label{fig:matern-posterior}
+	\end{center}
 \end{figure}
 Note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).
-This is the correlation on the log-intensity scale.
-[[Can I make a better "range of correlation" plot for the posterior intensity process?]]
+This is the correlation range on the log-intensity scale.
+[[Can I make a better "range of correlation" plot for the posterior intensity process?
+How can I convert this to something on the intensity scale?
+For process $Z$ with covariance $C$ is there an implied covariance for $exp(Z)$?]]
 
-To further investigate the ability of the GMRF component to capture the heterogeneity we generated 100 thinned point patterns from the fitted model and calculated the pairwise distances between all observations \autoref{fig:post-pp-distances}.
+To further investigate the ability of the GMRF component to capture the heterogeneity we generated 100 thinned point patterns from the fitted model and calculated the pairwise distances between all observations for each of these patterns \autoref{fig:post-pp-distances}.
 \begin{figure}[h]
+	\begin{center}
 		\includegraphics[scale=0.6]{figures/post_pp_distances.png}
@@
 		\label{fig:post-pp-distances}
+	\end{center}
 \end{figure}
-The plot shows that the GMRF component captures the relevant scale of clustering in the observed data.
+The plot shows that the GMRF component broadly captures the relevant scale of clustering in the observed data.
+There is some evidence of over-predicting the strength of the clustering at distances between 1,000 and 2,500 m.
 
-We plot the posterior abundance estimate using a monte-carlo method.
+We plot the posterior abundance estimate using a Monte Carlo method.
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[scale=0.4]{figures/realized_abundance_posterior.png}
+		\caption{Posterior of realized abundance.  The blue line marks the expected abundance}
+		\label{fig:realized-abundance-posterior}
+	\end{center}
+\end{figure}
 Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.
-Integrating the mean of the posterior intensity field provides a point estimate for the abundance - the expected abundance.
-We approximate the $\pi(N | \lambda)$ by a monte-carlo method.
-Taking $m$ monte-carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
+Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e.
+the expected abundance.
+We approximate the posterior $\pi(N | \lambda)$ by a Monte Carlo method.
+Taking $m$ Monte Carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
 Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
@@
 This allows us to estimate the probability for any specific value of realised abundance $n$ by calculating $\pi(N = n | \lambda)$.
-\begin{figure}
-	\includegraphics[scale=0.6]{figures/realized_abundance_posterior.png}
-	\caption{Posterior of realized abundance.  The blue line marks the expected abundance}
-	\label{fig:realized-abundance-posterior}
-\end{figure}
 
 % Add table of estimated model parameters
-% Add 95% CI figures for N
+% Add 95% CI figures for N?
 

### 2020-02-20 analyses/eval_model.R
update figures

@@ -1,6 +1,8 @@
 # Evaluate fitted model
 library(INLA)
 library(inlabru)
+library(scales)
+library(cowplot)
 
 # set ggplot theme
 theme_set(theme_minimal())
@@ -16,6 +18,9 @@ fit = readRDS(file_name)
 study_area = readRDS("data/study_area_extended_no_crs.RDS")
 samplers = readRDS("data/samplers_extended_no_crs.RDS")
 mesh = readRDS("data/mesh_extended_no_crs.RDS")
+
+#### ggplot theme ####
+theme_set(theme_classic())
 
 #### Specify detection functions  ####
 
@@ -43,78 +48,141 @@ W = 60   # transect radius
 distdf <- data.frame(distance = seq(1,W,length=100))
 
 hnpred <- predict(fit, distdf, ~hn(distance,lsig))
-ggplot() + 
-  gg(hnpred) +
-  ggtitle("Posterior half-normal detection function")
-
-ggsave(filename = "../figures/halfnormal.png",
-       width = 5, height = 5, units = "in")
-
-# half-normal polar coords?  Is that what this is?  
-# distdf <- data.frame(distance = seq(1,W,length=100))
-# dfun <- predict(fit, distdf, ~ dsamp(distance,lsig))
-# ggplot() + 
-#   gg(dfun)
+ghn = ggplot() + 
+        gg(hnpred) +
+        ylab("probability of detection\n") +
+        xlab("\ndistance (m)") +
+        theme(axis.title = element_text(size = 16),
+        axis.text = element_text(size = 16))
 
 ## Matern plots ##
 
-plot(spde.posterior(fit, "grf", what = "matern.correlation")) +
-  ylab("correlation") + xlab("distance (m)") +
-  ggtitle("Posterior Matern correlation function")
+blah = spde.posterior(fit, "grf", what = "matern.correlation")
+blah$x = blah$x / 1000
+gm = ggplot() +
+      gg(blah) +
+      ylab("correlation\n") + 
+      xlab("\ndistance (km)") +
+      theme(axis.title = element_text(size = 16, vjust = 0.3),
+            axis.text = element_text(size = 16))
+      
+gm
 
-ggsave(filename = "../figures/matern_posterior.png",
-       width = 5, height = 5, units = "in")
+png(filename = "../figures/detfn_and_matern.png",
[…]

### 2020-02-20 analyses/excursions.R

@@ -30,6 +30,7 @@ n.mc = 500     # lower this when fiddling with things
 pxl = pixels(mesh, nx = 100, ny = 100, mask = study_area)
 X = matrix(NA, nrow = nrow(pxl@coords), ncol = n.mc)
 for (i in 1:n.mc){
+  print(paste("i = ", i))
   int_draw <- predict(fit, pxl, ~ exp(grf + Intercept), n.samples = 1)
   X[,i] = int_draw@data$mean
 }
@@ -56,24 +57,43 @@ Fpxl$F = ex$F
 # png(filename = "../figures/excursion_function.png",
 #     width = 4.5, height = 5.5, units = "cm", res = 1000)
 p2 = ggplot() + 
+  gg(study_area) +
   gg(Fpxl) +
-  scale_fill_viridis_c() +
-  ggtitle("Excursion function for > 1 bird per hectare") +
-  xlab("Easting") +
-  ylab("Northing")
+  scale_fill_viridis_c(limits = c(0,1), 
+                       breaks = c(0, 0.5, 1),
+                       labels = c(0, 0.5, 1))+
+  coord_equal() +
+  theme_void() +
+  theme(legend.position = "bottom", 
+        legend.direction = "horizontal") +
+        #plot.margin = margin(0.1, 0, 0.1, 0, "in")) +
+  guides(fill = guide_colourbar(title.vjust = 0.95,
+                                title.theme = element_blank(),
+                                label.theme = element_text(size = 16, hjust = 0.5)))
 
 p2
+
 # ggsave(filename = "../figures/excursion_function.png",
 #        width = 5, height = 6, units = "in")
 
 Epxl = pxl
 Epxl$E = as.factor(ex$E)
+Epxl = Epxl[Epxl$E == 1,]
 p1 = ggplot() +
+  gg(study_area) +
   gg(Epxl, aes(group = E)) +
-  ggtitle("Excursion set for > 1 bird per hectare,\nalpha = 0.05") +
-  scale_fill_manual(values = c("gray", "blue")) +
-  xlab("Easting") +
-  ylab("Northing")
+  coord_equal() +
+  # ggtitle("Excursion set for > 1 bird per hectare,\nalpha = 0.05") +
+  scale_fill_manual(values = c("blue"),
+                    labels = c("Excursion set")) +
+  theme_void() +
+  theme(legend.position = "bottom", 
+        legend.direction = "vertical",
+        plot.margin = margin(0.1, 0, 0.1, 0, "in"),
[…]

### 2020-02-20 paper-v2.tex
up to para 2 in discussion

@@
 
-\usepackage{hyperref} % auto detect ref type
+\usepackage[hidelinks]{hyperref} % auto detect ref type
 \usepackage{natbib}
@@
 This requires the input of statisticians to ensure uncertainty is appropriately captured and communicated in the key outputs of the analysis upon which decisions will be based.
-In particular, in the context of species distribution modelling, we note the challenge of communicating uncertainty in maps of predicted animal density and making clear the difference between uncertainty in the estimated \textit{expected} abundance versus uncertainty in the \textit{realised} abundance, the latter of which is likely more relevant to ecological management.
+In particular, in the context of species distribution modelling, we note the challenge of communicating uncertainty in maps of predicted animal density.
 
-In this paper, we present an analysis that seeks to address each of these issues by presenting
+The most common method to communicate uncertainty in a spatial estimate tend is to produce maps of predictive uncertainty such as standard deviation or CV (e.g.
+\cite{ fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}) based either on a posterior predictive distribution or bootstrapping.
+Another approach is to map quantiles or the probability of exceeding certain thresholds (e.g.
+\cite{russell_avoidance_2016, wilson_hierarchical_2010}).
+In Section \ref{results} we produce maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of a posterior spatially structured random field.
+Whilst maps that summarise spatial predictions are useful, they all mask certain properties we may care about when communicating the results of an analysis.
+For example, maps of the posterior mean will always be smoother than realisations of the posterior random field itself.
+This can obscure the spatial structure we would expect if we observed a realisation of process itself.
+Maps of the posterior coefficient of variation are hard to interpret when estimated densities are close to zero and posterior quantiles are tempting to interpret jointly when they should be viewed independently.
+This challenges mean careful consideration must be given to the communication.
+
+In this paper, we present an analysis that seeks to address these statistical challenges by presenting
 
@@
 	\item Examples of model evaluation and communication of results that take advantage of the one-stage approach by sampling from the joint posterior of all model parameters and introducing excursions methods \citep{bolin_excursion_2015} as a method for investigating uncertainty in spatial predictions of density.
+These approaches are based on considering the joint probability of events at multiple spatial locations and avoid the interpretability issues of quantile or exceedance probability maps.
 \end{enumerate}
@@
 \section{Results}
+\label{results}
 
@@
 	\begin{center}
-		\includegraphics[scale=0.4]{figures/intensity_mean_cv_sd.png}
-		\caption{Posterior mean, CV and SD of the intensity}
+		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
+		\caption{The A) mean, B) CV and C) SD of the posterior intensity field}
 		\label{fig:intensity-mean-cv}
@@
 	\begin{center}
-		\includegraphics[scale=0.4]{figures/intensity_quantiles.png}
-		\caption{The predicted posterior intensity summarised in three plots by the mean (upper),  0.025 quantile (lower left) and 0.975 quantile (lower right) for each prediction grid cell.
-All three plots use a common colour scale}
+		\includegraphics[scale=0.525]{figures/intensity_quantiles.png}
+		\caption{The predicted posterior intensity quantiles:  A) 0.025 quantile B) 0.975 quantile.  Both plots plots use the same colour scale}
 		\label{fig:intensity-quantiles}
@@
 	\begin{center}
-		\includegraphics[scale=0.4]{figures/halfnormal.png}
-		\caption{Posterior detection function.  Black line - mean posterior estimate, grey shaded area - 95\% credible interval}
+		\includegraphics[scale=0.525]{figures/detfn_and_matern.png}
+		\caption{A) posterior half-normal detection function.
+B) posterior Mat\'ern correlation function.
+The black line is the mean posterior estimate and the grey shaded area is 95\% credible interval in both plots}
 		\label{fig:half-normal}
@@
 The posterior Mat\'ern correlation function is shown in \autoref{fig:matern-posterior}, the grey regions show the 95\% credible interval.
-\begin{figure}[h]
-	\begin{center}
-		\includegraphics[scale=0.4]{figures/matern_posterior.png}
-		\caption{Posterior Mat\'ern correlation function}
-		\label{fig:matern-posterior}
-	\end{center}
-\end{figure}
+
 Note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).
@@
 	\begin{center}
-		\includegraphics[scale=0.4]{figures/realized_abundance_posterior.png}
-		\caption{Posterior of realized abundance.  The blue line marks the expected abundance}
+		\includegraphics[scale=0.525]{figures/realized_abundance_posterior.png}
+		\caption{Posterior abundance}
 		\label{fig:realized-abundance-posterior}
@@
 
-The results presented above are consistent with a more traditional two-stage model on binned count data.
-Our presentation of the results is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (\autoref{fig:intensity-mean-cv} and \autoref{fig:intensity-quantiles}) are not always provided.
-We note a worrying tendency in the species distribution mapping literature to include maps of point estimates without any accompanying communication of uncertainty.
-Where uncertainty maps are provided these tend to be maps of predictive uncertainty such as standard deviation or CV (recent examples include \cite{ fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}) based either on a posterior predictive distribution or bootstrapping.
-Another approach is to map quantiles or the probability of exceeding certain thresholds (recent examples \cite{russell_avoidance_2016, wilson_hierarchical_2010}).
-In the above, we chose to communicate uncertainty through maps of CV, standard deviation, and the 0.025 and 0.975 quantiles.
-Whilst all of these maps are informative and useful in their own way, our intention in this section is to highlight limitations with these maps and suggest ways to address these limitations.
+The results presented above are consistent with those resulting from a two-stage model fitted to binned count data (see Supplementary material).
+Our presentation of the results is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (\autoref{fig:intensity-mean-cv} and \autoref{fig:intensity-quantiles}) are not always provided there and it is common to see maps of point estimates without any accompanying communication of uncertainty.
+Our intention in this section is to highlight limitations with these maps, including those presented in Section \ref{results}, and suggest ways to address these limitations.
 
-Whilst summary maps of random fields are useful, they all can mask certain properties of the random field.
-The posterior mean, for example, will be smoother than realisations from the posterior.
-\autoref{fig:intensity-realizations} shows three such realisations.
-Note that each of them has finer-grained spatial structure than is shown in the posterior mean in \autoref{fig:intensity-mean-cv}.
-One interpretation of this is that animals will tend to be more clustered than might be expected if we only looked at the posterior mean.
+\subsection{Limitations of mapped summaries}
+
+The mapped posterior mean is the most common map one encounters in the literature.  However, the mean will always be smoother than realisations from the posterior random field.
 \begin{figure}
-	\includegraphics[scale=0.35]{figures/intensity_realized.png}
+	\includegraphics[scale=0.525]{figures/intensity_realized.png}
 	\caption{Three realizations of the posterior intensity field}
@@
 \end{figure}  
-During model evaluation we recommend producing plots of realisations as these will more closely resemble the spatial structure of the observed data.
+\autoref{fig:intensity-realizations} shows three such realisations from the posterior intensity field.
+Note that each realisation has a finer-grained spatial structure than is shown in the posterior mean (\autoref{fig:intensity-mean-cv}A).
+Hence, considering this finer-grained structure, our interpretation is that the clustering of animals is stronger than might be expected if we only looked at the posterior mean.
+Similarly for model evaluation we recommend plotting multiple realisations of the posterior field as these will more closely resemble the spatial structure of the observed data.
+Noting differences between realisations is also an effective way to communicate uncertainty.
 
-The CV map (\autoref{fig:intensity-mean-cv}) can be hard to interpret in regions of low estimated intensity.
-If the posterior standard deviation is relatively consistent across the survey region, CV values will be higher in regions of low predicted intensity.
-The observed patterns in such maps will highlight regions of relatively lower intensity, not necessarily higher uncertainty.
-This is problem is further confounded in the case with spatially varying posterior standard deviation.
+The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.
+However, CV values will be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.
+The observed patterns in such maps will highlight regions of relatively lower intensity, not necessarily higher overall uncertainty.
+This is clearly the case in \autoref{fig:intensity-mean-cv}B.
+
+The standard deviation map (\autoref{fig:intensity-mean-cv}C) does show clearly the spatially varying uncertainty.
+However,  care should be given to what the mapped values imply in the context of a given analysis.
+A default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high or low uncertainty.
+What such a map does not show is whether these differences actually matter in the context of the analysis.
+For example, it could be the case that the standard deviation values are so large as to make the predictions have an unacceptably high uncertainty everywhere.
+Or, on the other hand, perhaps differences between high and low uncertainty regions (`high' and `low' as defined by the default colour scale) are small enough to be negligible when it comes to making conclusions or decisions, in which case such a spatially varying map is not a relevant output of the analysis.
+These are two extreme examples.
+However, cases in between suffer from the same problem.
 
 The quantile maps also suffer from a potential difficulty in interpretation.
-The temptation is to interpet the maps as showing possible intensity surfaces that could have produced the observed data.
-Presenting these quantiles side by side in a single map obscures the fact that it is vanishingly unlikely for all prediction values to simultaneously achieve the 0.025 or 0.975 quantile.
-A similar problem will occur with mapping the marginal probability of exceeding a threshold at each prediction locations.
-We believe most non-statistically trained audiences will struggle with this caveat to interpeting the map.
-To demonstrate the possible consequences of this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile map and 12,400 for the 0.975 quantile map.
+The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data.
+However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for all prediction values to simultaneously achieve the 0.025 or 0.975 quantile.
+A similar problem will occur if mapping the marginal probability of exceeding a threshold at each prediction location.
+We believe these caveats make the map difficult to use for most non-statistically trained audiences.
+To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile map and 12,400 for the 0.975 quantile map.
 A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance.
 However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
-This demonstrates that the natural tendency to interpet these maps as possible intensities leads to interpretations that are inconsistent with the very model that generated the maps.
+This demonstrates that the tendency to interpret these maps as possible intensities leads to interpretations that are inconsistent with the very model that generated the maps.
 
-For the standard deviation map care should be given to what the mapped values imply in the context of a given analysis.
-A default colour scale will more or less always show some spatial variation in uncertainty, with some regions of relatively high or low uncertainty.
-What such a map does not show is whether these values actually matter in the context of the analysis.
-For example, it could be the case that the standard deviation values are so large as to make the predictions operationally useless everywhere.
-Or, on the other end of the scale, perhaps differences between high and low uncertainty regions (`high' and `low' as defined by some default colour scale) are small enough to be negligible when it comes to decision making, in which case such a spatially varying map is not a relevant output of the analysis.
-These are two extreme examples, cases in between suffer from the same problem.
+\subsection{Excursion sets and excursion functions}
 
-Our solution to the problems in the quantile and standard deviation maps is to suggest that consideration should be given \emph{a priori} to values of the process and levels of uncertainty that are important and suitable given the context of the analysis.
+Our solution to these problems with quantile and standard deviation maps is to suggest that consideration should be given \emph{a priori} to relevant values of the process and acceptable levels of uncertainty that are suitable given the context and aims of the analysis.
 These values can then be used to construct summary maps.
-We demonstrate this perspective through the use of excursions sets and functions \citep{bolin_excursion_2015}, although there are many possible approaches.
-These techniques require the user to specify thresholds of interest for the process itself and acceptable levels of uncertainty.
+We demonstrate this perspective using excursions sets and excursion functions \citep{bolin_excursion_2015}.
+Although there are other possible approaches [[cite here]], excursions methods are a natural choice for Gaussian random fields.
+These methods require the user to specify thresholds of interest for the random field and acceptable levels of uncertainty.
+For this reason they avoid the issue of a default colour scale playing too much of a role in our interpretation of results.
+Excursion sets and excursion functions are also based on the joint probability of events across a set of locations.
+For this reason they avoid the interpretability issues of the quantile maps.
 
-This approach to representing variability in stochastic processes specifically considers the joint probability of events across multiple locations.
 The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e.
-the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$.
+the set of all locations in $\Omega$ where $f$ exceeds a threshold value $u$.
 For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
 \begin{equation*}
-E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \}
+E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \} \;\;\; .
 \end{equation*}
@@
 As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.
-Negative excursion sets can be similarly defined.
+Negative excursion sets are similarly defined.
 Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.
@@
 \autoref{fig:excursions} (left panel) shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95.
-This figure can be interpreted in a natural way.
-It is the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
+This figure can be interpreted in a natural way as the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
 \begin{figure}
@@
 \end{figure}
-To visualise multiple such maps we can use the excursion function $F_u^{+}(s) = \sup \{1 - \alpha ; s \in E_{u,\alpha}^+ \}$, which defines for each location something similar to a p-value.
-For each $s$ it is the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.
+To visualise multiple such maps we can use the excursion function $F_u^{+}(s) = \sup \{1 - \alpha ; s \in E_{u,\alpha}^+ \}$, which defines for each location largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.
+i.e.
+if we allow greater uncertainty, which locations would be included in the excursion set.
 The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursions} (right panel).
@@
 
-Our intention in this paper was to present a new approach to analysing point transect distance sampling data that incorporates several innovative new methods and considers the entire workflow of model specification, inference, evaluation and communication of results.
-Although each of these areas are substantive enough to be the main focus of a paper, we hope our brief presentation of issues that arise in each area and our chosen solutions will be useful to statisticians who face similar questions and data structures.
+This paper presents a new approach to analysing point transect distance sampling data that incorporates several innovative methods and considers the bulk of an applied statistician's workflow of model specification, inference procedure, model evaluation and communication of results.
+Although any of these areas on its own would be substantive enough to be the main focus of a paper, our brief presentation of issues in sequence along with our chosen solutions will be useful to statisticians who face similar questions and data structures.
 
@@
 Spatio-temporal extensions to this are possible, for example by considering a space-time interaction models.
-\texttt{R-INLA} has the ability to fit space-time effects that can be represented as a kronecker product of the relevant precision matrices.
-An example of this type of space-time interaction effect would be the spatial SPDE above interacting with an AR1 temporal process.
+\texttt{R-INLA} has the ability to fit space-time effects that can be represented as a Kronecker product of the relevant precision matrices (e.g.
+\citep{yuan_point_2017}).
+An example of this type of space-time interaction effect is an interaction between the spatial SPDE with an auto-regressive temporal process.
 \cite{blangiardo_spatial_2013} show how this can be done with numerous examples and code snippets.
 
-Another extension is to consider more complicated observation processes.
+Another natural extension is to consider more complicated observation processes.
 For example, we did not use any explanatory covariates for the detection function.
-This remains to be investigated further.
-Any additional parameters to the detection function would result in a higher-dimensional outer optimisation problem for iterated INLA.
-This could lead to problem if, for example, identifiability is an issue.
+Any additional parameters to non-linear model components will become part of the outer-optimisation step of iterated INLA.
 

### 2020-02-21 paper-v2.tex
end of Janine's comments

@@
 Here we present an analysis of point transect distance sampling data collected on a tropical forest bird.
-We consider an entire workflow, from model conception, inference procedure, model evaluation to communication of results, at each stage discussing the challenges involved and our strategies to address them.
-
-New developments inlcude an inference method for one-stage distance sampling models within a Bayesian framework based on iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations \citep{rue_approximate_2009}.
-This approach simultaneously estimates the detectability and spatial distribution of populations of wild animals within a joint model.
-We take a point process perspective, modelling the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields implemented via a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.
-The Bayesian setting allows sampling from the joint posterior of all model parameters and so the resulting model outputs naturally incorporate uncertainty in the observation model estimates.
-We discuss potential issues with common ways to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels.
-We apply the methods of excursion sets and excursion functions \citep{bolin_excursion_2015} as an example of this approach.
+We consider the full workflow from model specification, inference procedure, model evaluation and communication of results, at each stage discussing the statistical challenges involved and our strategies to address them.
+We take a novel point process perspective on point transect data, viewing observations as a thinned log-Gaussian Cox process with a spatially structured random effect specified via a computationally efficient Gaussian Markov random field approximation to Gaussian random field \citep{lindgren_explicit_2011}.
+One-stage inference is achieved within a Bayesian framework by using an approach based on iterated fitting of latent Gaussian models using integrated nested Laplace approximations \citep{rue_approximate_2009} to account for non-linear model components, simultaneously estimating detectability and spatial distribution.
+Model outputs are based on samples from the joint posterior of all model parameters and so naturally incorporate uncertainty due to detectability, a key advantage of the one-stage Bayesian approach.
+We discuss limitations of methods to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels by using excursions methods \citep{bolin_excursion_2015} to avoid these issues.
+The methods presented are general, flexible and applicable in a wide range of areas in spatial statistics beyond just species distribution mapping.
+\bigskip
 \end{abstract}
 
-%\begin{keyword}
-%Distance sampling \sep Stochastic partial differential equations \sep Integrated nested Laplace approximation \sep Generalized additive model
-%\end{keyword}
+\begin{keyword}
+Distance sampling \sep Species distribution modelling \sep Estimating abundance \sep Stochastic partial differential equations \sep Integrated nested Laplace approximation
+\bigskip
+\end{keyword}
 \end{frontmatter}
@@
 Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.
-The Hawaiian \akepa{} is an endemic species whose population declined dramatically during the 20th century [[Rick - citation possible here?]].
+The Hawaiian \akepa{} is an endemic species whose population declined dramatically during the 20th century [[CITE]].
 The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.
@@
 the 'sociality' of the species or other behavioural characteristics).
-For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient that has been investigated but for which no straightforward explanatory cause has been found [[CITE: Ask Rick if poss]].
+For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient that has been investigated but for which no straightforward explanatory cause has been found [[CITE]].
 From a statistical perspective, this suggests the inclusion of spatially-structured random effects in the model, which necessitates the use of appropriate statistical software capable of fitting such models.
@@
 This challenges mean careful consideration must be given to the communication.
+
+The most important consequences of an analysis are the possible conclusions and decisions that will be taken on the basis of the results.
+Here we advocate the view that model estimates should be converted into outputs, with associated uncertainty, that are most directly relevant to these conclusions and decisions.
+In this paper we provide an example based on a scenario where the most important aim of the analysis is identification of areas where animal density is above a certain threshold.
+It is not always the case that the most appropriate output of a species distribution model is a map of the predicted mean density.
 
@@
 
-Using spatially-structured random effects can be computationally expensive, often resulting in dense prior precision matrices which typically have $\mathcal{O}(n^3)$ cost to factorise for an effect with $n$ parameters - sometimes referred to as the `big $n$ problem'.
-A key development in this area is the use of Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs).
+Using spatially-structured random effects can be computationally expensive, often resulting in dense prior precision matrices which typically have $\mathcal{O}(n^3)$ cost to factorise for an effect with $n$ parameters.
+A key development in this area is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs) and the \texttt{R-INLA} software that makes use of the sparsity to factorise the precision matrix at a cost of $\mathcal{O}(n^{3/2})$.
 This representation is constructed via a stochastic partial differential equation solved using finite element methods \citep{lindgren_explicit_2011}.
@@
 
-The results presented above are consistent with those resulting from a two-stage model fitted to binned count data (see Supplementary material).
+The results above are consistent with those resulting from a two-stage model fitted to binned count data (see Supplementary material).
 Our presentation of the results is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (\autoref{fig:intensity-mean-cv} and \autoref{fig:intensity-quantiles}) are not always provided there and it is common to see maps of point estimates without any accompanying communication of uncertainty.
-Our intention in this section is to highlight limitations with these maps, including those presented in Section \ref{results}, and suggest ways to address these limitations.
+It is also common to report a point estimate of the expected abundance along with uncertainty in this point estimate.
+However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.
+For this reason we chose to show the full abundance posterior in \autoref{fig:realized-abundance-posterior}.
+
+Our intention in this section is to highlight limitations with mapped summaries of model predictions, including those presented in Section \ref{results}, and suggest ways to address these limitations.
+We note that all model outputs are based on sampling from the joint posterior of all model parameters which naturally average over the uncertainty in the observation process, a key advantage of the one-stage approach.
 
@@
 Another natural extension is to consider more complicated observation processes.
-For example, we did not use any explanatory covariates for the detection function.
-Any additional parameters to non-linear model components will become part of the outer-optimisation step of iterated INLA.
+There is the potential to include explanatory covariates on detection function parameters.
+Any additional parameters to the non-linear model components will be estimated using the outer-optimisation step of iterated INLA.
+[[What can I say about this?
+Speculate on how many parameters would work?]]
 
-We note also that our presentation of abundance estimates differs from typical presentations.
-It is common to report a point estimate of the expected abundance and uncertainty in this point estimate.
-However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.
-We find the distinction between expected and realised abundance to be a useful one.
-The credible intervals for the posterior expected abundance will be narrower than the credible intervals for the posterior realised abundance.
-The Bayesian perspective is useful here as well as for the uncertainty in the detection model estimates.
-Since outputs based on sampling from the joint posterior of all model parameters, including the detection model, the Bayesian framework means the resulting model outputs we present here naturally average over the uncertainty in the observation process.
-[[Not really sure if this stuff is worth mentioning or if can be said in a better way]]
+The iterated INLA inference procedure is a general approach that could be applied to other non-linear model components that arise in the area of spatial ecology.
+We note two promising areas in which our flexible inference procedure may be useful:
 
-Our use of \texttt{inlabru} and \texttt{R-INLA} for inference means we can take advantage of the sparsity of the GMRF precision matrix, a key advantage of the SPDE approach.
-A dense covariance matrix of dimension $n \times n$ typically has an $\mathcal{O}(n^3)$ cost to invert.
-In contrast to this \texttt{R-INLA} makes use of sparse matrix algorithms to invert a sparse 2-dimensional precision matrix at a typical cost of $\mathcal{O}(n^{3/2})$ \citep{lindgren_explicit_2011}.
-This computational efficienty, in addition to the flexibility of the mesh triangulation, can make the SPDE an attractive choice for implementing high-resolution spatial effects in complex domains.
-An additional benefit, as described in detail in \cite{yuan_point_2017}, is that the SPDE can also be implemented on the sphere, making analysis of global datasets computationally feasible.
+\begin{itemize}
+	\item \textbf{Selectivity analysis} In surveys of the marine environment, the size, length and weight of fish affects the likelihood of being caught in a trawl net.
+Thus any analysis of these data must account for this.
+The probability of capture is typically modelled as a logistic function with parameters that depend on morphological traits \citep{herrmann_understanding_2016, madsen_selectivity_2007, galbraith_demersal_1994}, which could be estimated jointly with the spatial distribution within our inference framework.
+	\item \textbf{Functional responses} In the area of understanding species-habitat or species-species interactions, the concept of a `functional response' captures the idea that the abundance of a resource or risk will affect the response to it \citep{holling_some_1959}.
+These varying responses can be described mathematically using parametric equations that have the desired properties, e.g.
+a horizontal asymptote so that beyond a certain threshold no extra amount of a resource will affect the response to it.
+These types of responses are typically not considered in species distribution models but our framework allows them to be estimated alongside the more common linear or quadratic fixed effects.
+This flexibility in specifying model components could also be more widely applicable in areas such as public health and spatial epidemiology, where responses to conditions may be non-linear.
+
+\end{itemize}
+We note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analysis
 
 In our analysis we only briefly touch upon the problem of communicating uncertainty in maps and highlight the excursions methods as a new and positive addition to this area.
-Our example threshold was chosen arbitrarily and it is not obvious to us that there will always be an easy or simple choice of thresholds and levels of uncertainty for a given analysis.
-This requires the input of relevant domain expertise and statistical knowledge so stakeholders understand the consequences of the thresholds and uncertainty levels they wish to consider.
-The most important consequences of an analysis are the possible conclusions and decisions that will be taken on the basis of the results.
-Here we advocate the view that model estimates should be converted into outputs, with associated uncertainty, that are most directly relevant to these conclusions.
-In the above we chose to use excursion sets, excursion functions and the posterior of realised abundance as a demonstration of this.
-It is not always the case that a map of predicted mean abundance will be the best output for a species distribution model.
+In our example, the threshold was chosen to illustrate the method but we believe that discussions with relevant stakeholders will lead to agreement on a (possibly large) set of thresholds that are relevant given the aims and context of the analysis.
+This requires the input of both the relevant domain experts and statisticians so stakeholders understand the consequences of the thresholds and uncertainty levels they wish to consider.
+We also note the potential for excursions to be used to define the notion of occupancy (and thus map occupied areas, with a given probability of error).
 
-The point process perspective with spatial SPDE effect can seem fundamentally different to a two-stage GAM approach.
-However, we think that it can be useful to highlight the similarities in approach.
-It can be tempting to think that because with the point process model does not require us to bin the data into counts that we can avoid some loss of information.
-However, we also had to make the same assumption of constant intensity within each transect that is common in more traditional distance sampling approaches.
-This means the exactly location of the points is not used to inform the posterior intensity.
-The assumption of constant intensity within each point transect is essentially the point process version of binning point data into counts.
+We end with a note on the similarity between what we have done and existing two stage approach.
+The one-stage thinned point process perspective can seem fundamentally different to a two-stage GAM approach fitted to count data.
+It can be tempting to think that, because the point process model does not require us to bin the data into counts, we can avoid some loss of information.
+However, we also made the same assumption of constant intensity within each transect as in more traditional distance sampling approaches.
+This leads to the same loss of information since we do not use the exact location of points within each transect, except to account for detectability.
 
-The SPDE effect can also seem fundamentally different to, say, a spline-based smoother.
-However, there is a fundamental connection between SPDEs and optimal smoothing splines.
-The optimal smoothing spline is a lminimum variance unbiased linear estimator of the posterior mean of an SPDE \citep{kimeldorf_spline_2002}.
-This means a smoothing spline can be interpreted as an SPDE and vice-versa.
-Depending on the methods a user is most familiar with this can be a useful perspective.
-\cite{yue_bayesian_2014} show how to implement common spline penalties in an SPDE framework and \cite{miller_understanding_2019} offer a complementary explanation to interpret and fit SPDE models within standard penalised regression splines software.
+In summary, we have presented a novel framework for analysing point transect distance sampling data that introduces new approaches to model specification, inference procedure and communication of results.
+These developments will be of interest to those with similar data structures in spatial ecology as well as researchers in other areas who we hope will explore the full flexibility and generality of these developments in their own work.
 
-Although we use a Bayesian approach, maximum-likelihood frameworks could also have been used for a one-stage analysis.  
-Finally, we note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analysis.
-
-Discuss selectivity, saturated responses?? Disease modelling?
+% spp has potential to provide general framework? [Janine comment]
+% Mention modelling on sphere is possible without projecting on to R^2?
 
 \clearpage
-\bibliographystyle{plainnat}
+\bibliographystyle{humannat}
 \bibliography{paper}

### 2020-03-02 paper-v2.tex
Rick comments

@@
 Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.
-The Hawaiian \akepa{} is an endemic species whose population declined dramatically during the 20th century [[CITE]].
+The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.
 The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.
@@
 
-Firstly, as in many ecological surveys, it is impossible to undertake a full census.
+Firstly, as in many ecological surveys, it is impossible to undertake a full census (i.e.
+the complete enumeration of all individuals within a defined study region).
 The existing population numbers in the thousands and lives in dense forest where logistical and ecological challenges mean a census is not feasible.
 For this reason, a monitoring survey must sub-sample appropriately in space and time.
-For the \akepa{} this takes the form of the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}, which is a large-scale, quantitative survey of Hawaiian forest birds.
+For the \akepa{} this takes the form of the \hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}, which is a large-scale, quantitative survey of Hawaiian forest birds.
 Annual surveys of the \akepa{} study-region consist of a number of point transects located along randomly located line transects.
@@
 the 'sociality' of the species or other behavioural characteristics).
-For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient that has been investigated but for which no straightforward explanatory cause has been found [[CITE]].
+For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient that has been investigated but for which no straightforward explanatory cause has been found [[CITE: Rick accepted Ecography paper - no doi available yet]].
 From a statistical perspective, this suggests the inclusion of spatially-structured random effects in the model, which necessitates the use of appropriate statistical software capable of fitting such models.
 
-Fourthly, the above challenges mean the resulting statistical model is necessarily complex and it is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.
+Lastly, the above challenges mean the resulting statistical model is necessarily complex and it is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.
 The data has been collected with the clear objective of monitoring the population and informing conservation management decisions.
@@
 
-The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.
-Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}.
+The \hawaii{} \akepa{} is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.
+Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the HFBS \citep{scott_HFBS_1986}.
 Information from the HFBS is used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau), that is the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
@@
 Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance.
-Survey points were established along 15 line transects following a systematic, random design with point transects approximately 150 m apart on line transects located either 500 or 1,000 m apart.
-We limit our study area to the open-forest and closed-forest strata of Hakalau (\autoref{fig:2002studyareapointspt}), an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016} who omitted the closed-forest area because it was not sampled in the early years of the \akepa{} surveys.
-For our analysis we consider data from a later year in the survey in which the closed-forest strata was sampled and so we include it in our analysis.
+Survey points were established along 14 line transects following a systematic, random design with point transects approximately 150 m apart on line transects located either 500 or 1,000 m apart.
+We limit our study area to the open-forest and closed-forest strata of Hakalau (\autoref{fig:2002studyareapointspt}), an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016} who omitted the closed-forest stratum because it was not sampled in the early years of the \akepa{} surveys.
+For our analysis we consider data from a later year in the survey in which the closed-forest stratum was sampled and so we include it in our analysis.
 The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.
+The closed-forest stratum was least modified by grazing and is relatively intact forest habitat.
 To the north, the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig.
@@
 During 8-min counts trained observers recorded the species, exact distance to the nearest meter and detection type for each bird detected, along with the sampling conditions cloud cover, rain, wind strength, gust strength, and time of day each point was surveyed.
-\cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the open-forest study area and the bird surveys.
+\cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the study area and the bird surveys.
 
 For the purposes of our analysis we select a single survey year from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability.
-In 2002, 289 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig.
+In 2002, 289 points were sampled using point-transect distance sampling methods within the 4,603 ha study area of Hakalau (Fig.
 \ref{fig:2002studyareapointspt}) during which 276 \akepa{} were detected on 121 point transects.
-The number of detections within each point transect ranged from zero to 6.
+The number of detections within each point transect ranged from zero to six.
 We select a single year of data to demonstrate our approach in a simplified setting where temporal aspects of analysis are not considered.
@@
 This additional flexibility to make predictions for smaller, user-specified geographic units, has made model-based distance sampling a common approach in the literature (see e.g.
-[[CITE HERE]]).
+\cite{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}).
 
@@
 
-A key concern with the two-stage approach is the propagation of uncertainty from the detectability estimates to the second-stage spatial model.
-Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_reliable_2018-1}.
+A key concern with the two-stage approach is the propagation of uncertainty from the detection model to the second-stage spatial model.
+Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_reliable_2018-1, williams_chilean_2011}.
 Instead, \cite{bravington_reliable_2018-1} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the first-stage maximum-likelihood estimate.
@@
 Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.
-These approaches tend to be Bayesian and use data augmentation and to model unobserved individuals or groups and Markov-chain Monte Carlo (MCMC) methods for inference (see )\citep{royle_hierarchical_2008, schmidt_using_2012}, for example).
+These approaches tend to be Bayesian and use data augmentation to model unobserved individuals or groups and Markov-chain Monte Carlo (MCMC) methods for inference (see )\citep{royle_hierarchical_2008, schmidt_using_2012}, for example).
 \cite{oedekoven_bayesian_2014} demonstrate a one-stage model that avoids data augmentation by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty using reversible jump MCMC.
@@
 This approach of choosing a parametric family of detection functions will be more familiar to users of classical distance sampling methods than the SPDE specification.
-However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in \text{R-INLA}.
-We address this by using a method of iterated model fits based on a Taylor expansion of the non-linear model components.
+However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in the \text{R-INLA} \citep{rue_approximate_2009} package implementation of INLA.
+We address this by using a method of iterated model fits based on a Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019} available through the .
 We name this approach \textit{iterated INLA}.
@@
 In this section we introduce the statistical model for the spatial pattern of animal locations and the observation process, including a detailed description of how this model can be written as a modified Poisson likelihood.
-We use this formulation to fit the model using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension to the \texttt{R-INLA} implementation of INLA, which implements the iterated INLA approach.
+We use this formulation to fit the model using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension to the \texttt{R-INLA} implementation of INLA, which implements the iterated INLA approach and is available through the Comprehensive R Archive Network \citep{r_2017}.
 
@@
 This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary material).
-The CV plot shows that the CV in the posterior intensity field is lower areas with greater sampling effort.
+The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort.
 There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north  (see \autoref{fig:2002studyareapointspt}).
@@
 
-The posterior detection function \autoref{fig:half-normal} shows that detectability drops to just under 0.25 at the maximum observable distance of 60 metres.
+The posterior detection function \autoref{fig:det-matern} shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres.
 \begin{figure}[h]
@@
 The black line is the mean posterior estimate and the grey shaded area is 95\% credible interval in both plots}
-		\label{fig:half-normal}
+		\label{fig:det-matern}
 	\end{center}
 \end{figure}
-The posterior Mat\'ern correlation function is shown in \autoref{fig:matern-posterior}, the grey regions show the 95\% credible interval.
-
+The posterior Mat\'ern correlation function is shown in \autoref{fig:det-matern}, the grey regions show the 95\% credible interval.
 Note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).
@@
 How can I convert this to something on the intensity scale?
-For process $Z$ with covariance $C$ is there an implied covariance for $exp(Z)$?]]
+For a field $Z$ with covariance function $C$, is there an implied covariance for $\exp(Z)$?]]
 
@@
 We demonstrate this perspective using excursions sets and excursion functions \citep{bolin_excursion_2015}.
-Although there are other possible approaches [[cite here]], excursions methods are a natural choice for Gaussian random fields.
+Although there are other possible approaches [[cite here?]], excursions methods are a natural choice for Gaussian random fields.
 These methods require the user to specify thresholds of interest for the random field and acceptable levels of uncertainty.
-For this reason they avoid the issue of a default colour scale playing too much of a role in our interpretation of results.
+For this reason they avoid the issue of a default colour scale potentially affecting our interpretation of results.
 Excursion sets and excursion functions are also based on the joint probability of events across a set of locations.
@@
 
-The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e.
+The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$, i.e.
 the set of all locations in $\Omega$ where $f$ exceeds a threshold value $u$.
@@
 Note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.
-As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.
+As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set, with a chosen probability.
 Negative excursion sets are similarly defined.
@@
 There is the potential to include explanatory covariates on detection function parameters.
+For example, we did not make use of the detection type (audio or visual) in our analysis, although this was recorded and detectability may vary depending on whether a bird is seen or heard.
+Other factors could be considered, such as weather conditions, observer expertise, animal behaviour and morphological traits.
 Any additional parameters to the non-linear model components will be estimated using the outer-optimisation step of iterated INLA.
-[[What can I say about this?
-Speculate on how many parameters would work?]]
+[[Can I speculate here about how iterated INLA might perform as parameters are added to non-linear components?
+Is, higher-dimensional optimisation slower/infeasible/unstable at a certain point?]]
 

### 2020-03-02 analyses/eval_model.R

@@ -44,7 +44,7 @@ summary(fit)
 
 # half-normal
 
-W = 60   # transect radius
+W = 58   # transect radius
 distdf <- data.frame(distance = seq(1,W,length=100))
 
 hnpred <- predict(fit, distdf, ~hn(distance,lsig))

### 2020-03-02 analyses/excursions.R

@@ -89,8 +89,8 @@ p1 = ggplot() +
   theme_void() +
   theme(legend.position = "bottom", 
         legend.direction = "vertical",
-        plot.margin = margin(0.1, 0, 0.1, 0, "in"),
-        legend.box.margin = margin(-10,0,0,0)) +
+        plot.margin = margin(0.1, 0, 0.3, 0, "in"),
+        legend.box.margin = margin(-4,0,0,0)) +
   guides(fill = guide_legend(title.vjust = 0.95,
                              title.theme = element_blank(),
                              label.theme = element_text(size = 16, hjust = 0.5)))

### 2020-11-06 paper-v2.tex
start working through Finn comments

@@
 
-The most common method to communicate uncertainty in a spatial estimate tend is to produce maps of predictive uncertainty such as standard deviation or CV (e.g.
-\cite{ fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}) based either on a posterior predictive distribution or bootstrapping.
-Another approach is to map quantiles or the probability of exceeding certain thresholds (e.g.
-\cite{russell_avoidance_2016, wilson_hierarchical_2010}).
+The most common method to communicate uncertainty in a spatial estimate tend is to produce maps of predictive uncertainty such as standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014} based either on a posterior predictive distribution or bootstrapping.
+Another approach is to map quantiles or the probability of exceeding certain thresholds \citep{russell_avoidance_2016, wilson_hierarchical_2010}.
 In Section \ref{results} we produce maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of a posterior spatially structured random field.
@@
 These methods allow density to be associated with spatially-indexed covariates and animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
-This additional flexibility to make predictions for smaller, user-specified geographic units, has made model-based distance sampling a common approach in the literature (see e.g.
-\cite{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}).
+This additional flexibility to make predictions for smaller, user-specified geographic units, has made model-based distance sampling a common approach in the literature \citep{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}.
 
@@
 Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.
-These approaches tend to be Bayesian and use data augmentation to model unobserved individuals or groups and Markov-chain Monte Carlo (MCMC) methods for inference (see )\citep{royle_hierarchical_2008, schmidt_using_2012}, for example).
-\cite{oedekoven_bayesian_2014} demonstrate a one-stage model that avoids data augmentation by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty using reversible jump MCMC.
+These approaches tend to be Bayesian and use data augmentation to model unobserved individuals or groups and Markov-chain Monte Carlo (MCMC) methods for inference \citep{royle_hierarchical_2008, schmidt_using_2012}.
+\citet{oedekoven_bayesian_2014} demonstrate a one-stage model that avoids data augmentation by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty using reversible jump MCMC.
 
@@
 A major downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing - usually a key feature of detection functions.
-The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior seems unsatisfactory.
-It is also not clear how covariate effects on the detection function could be incorporated in this approach.
+The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful.
 
@@
 However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in the \text{R-INLA} \citep{rue_approximate_2009} package implementation of INLA.
-We address this by using a method of iterated model fits based on a Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019} available through the .
-We name this approach \textit{iterated INLA}.
+We address this by using a method of iterated model fits based on a Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019}.
 
@@
 Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
-However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982, waagepetersen_likelihood-based_2006}.
+However, the point process viewpoint is not new and has been taken numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
 
@@
 
-To account for the imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected})$.
+To account for the imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected | a point is at $s$})$.
 A key property of the log-Gaussian Cox process is that a realisation of a point process with intensity process $\lambda(s)$ that is thinned by thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.
@@
 For simplicity, we assume that all point transects are non-overlapping discs with radius $W$, as is the case in the HFBS survey.
-The probability of observing a point at location $s \in \Omega_k$ we denote $g_k(s)$.
+The probability of observing a point at location $s \in \Omega_k$ given it is at $s \in \Omega_k$ we denote $g_k(s)$.
 The probability of observing a point outside the surveyed region is zero.
@@
 \label{lgcp-likelihood}
-\pi(\bm{Y}) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
+\pi(\bm{Y}) = \exp\left( |\Omega| -\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
 \end{equation}
-where $\lambda(.)$ and $g(.)$ both depend on parameter vectors omitted for readability.
+where $\lambda(.)$ and $g(.)$ both depend on parameter vectors omitted for readability and $|\Omega|$ is the area of $\Omega$.
 
@@
 
-Here we show how to write the thinned log Gaussian Cox process model as a modified Poisson likelihood.
+Here we show how to approximate the thinned log Gaussian Cox process model with modified Poisson likelihood.
 The integral component $\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s$ in \autoref{lgcp-likelihood} does not have analytical solutions.
-Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach developed by \cite{simpson_going_2016} for a fully observed point pattern.
+Replacing the integral with a weighted sum allows the log-likelihood to be approximated as a weighted Poisson log-likelihood, which we describe below, adapting the approach developed by \cite{simpson_going_2016} for a fully observed point pattern.
 

### 2020-11-09 paper-v2.tex
section 3 comments done

@@
 Since the point transects are non-overlapping, each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.
-For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
+For example, for an observer at location $s_k \in \Omega_k$, the centre of the disc, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
 In practice the observer is always located at the centre of the point transect.
@@
 \end{equation}
-where $\lambda(.)$ and $g(.)$ both depend on parameter vectors omitted for readability and $|\Omega|$ is the area of $\Omega$.
+where $\lambda(.)$ and $g(.)$ both depend on parameter vectors omitted for readability and $|\Omega|$ is the area of the total surveyed region.
 
@@
 
-Here we show how to approximate the thinned log Gaussian Cox process model with modified Poisson likelihood.
+\sloppy Here we show how to approximate the thinned log Gaussian Cox process model with a modified Poisson likelihood.
 The integral component $\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s$ in \autoref{lgcp-likelihood} does not have analytical solutions.
@@
 We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$.
-The integral can then be rewritten as a sum of weighted one-dimensional integrals $\sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$ using the assumption that $\lambda(s)$ is constant within each sampling unit.
-For each sampling unit we approximate the one-dimensional integral using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
+The integral in the likelihood can be simplified using an assumption that $\lambda(s)$ is linear within each sampling unit.
+This assumption implies that $\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) = 2\lambda(s_k)$ and so
+\begin{align*}
+	\int_0^{2\pi} \lambda(s(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \left[\lambda(s(r, \theta)) + \lambda(s(r, \theta + \pi)) \right] g_k(r)\mathrm{d}\theta \\
+	&= 2\pi \lambda(s_k)g(r)
+\end{align*}
+We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.
+A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piece-wise linear basis functions we use to model $\lambda$ (see section 5).
+Integrating with respect to $r$ and accounting for the change of variables gives $\int \lambda(s)g(s) \mathrm{d}s = \sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$.
+For each sampling unit we approximate the one-dimensional integral with respect to $r$ using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
 This gives
@@
 \end{equation}
-This approximation can be expressed as a modified Poisson likelihood.  To see this let $\exp \eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
-$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $y = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.
+This approximation can be expressed as a modified Poisson likelihood.  To see this let $\eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
+$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $z = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.
 Then the approximate likelihood becomes
 \begin{equation}
-\pi(\bm{Y}) \approx \prod_{i=1}^{KM + n} \eta_i^{y_i}\exp(-\alpha_i\eta_i)
+\pi(\bm{Y}) \approx C \prod\limits_{i=1}^{KM + n} \eta_i^{z_i}\exp(-\alpha_i\eta_i)
 \end{equation}
-This is similar to a Poisson likelihood with offset $\log\alpha$ and can be specified using the Poisson likelihood implementated in \texttt{R-INLA} and \texttt{inlabru}.
-The approach presented here is similar to the Berman-Turner device \citep{berman_approximating_1992, baddeley_practical_2000} although we relax the assumption that the observed points form part of the quadrature scheme.
+where $C$ is a constant.
+This approximate likelihood is implemented in \texttt{inlabru} as a \texttt{"cp"} likelihood.
+This is similar to the Berman-Turner device \citep{berman_approximating_1992, baddeley_practical_2000} although without the assumption that the locations of observed points form a part of the quadrature scheme.
 
@@
 \label{intensity-incomplete}
-\tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(s)g_k(s)\mathrm{d}s \nonumber \\
+\tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(s)g_k(r)\mathrm{d}s \nonumber \\
 &= 2\pi\lambda(s_k)rg_k(r) \;\;\; ,
 \end{align}
- where $c_k(r)$ is a circle of radius $r$ centred at $s_k$ and the second line follows from changing to polar coordinates and the assumption of constant intensity within $\Omega_k$.
+ where $c_k(r)$ is a circle of radius $r$ centred at $s_k$ and the second line follows from changing to polar coordinates and the assumption of linear intensity within $\Omega_k$.
 This intensity differs from the full data case through the $2\pi$  term that accounts for the fact that we do not observe $\theta$ and the additional $r$ term that accounts for the increasing area surveyed at larger distances.
-The log-intensity for observed points (the thinned intensity) is thus $\log\tl(s) = \log 2 \pi r + \log\lambda(s) + \log g(s)$.
-As we note above, the detection function $g(s)$ is non-linear in its parameters and this component of the additive predictor requires a novel approach to parameter estimation.
+The log-intensity for observed points (the thinned intensity) is thus $\log\tl(s) = \log 2 \pi r + \log\lambda(s) + \log g(s)$ and so can be implemented during fitting by using an offset.
+As we noted above, the detection function $g(s)$ is non-linear in its parameters and this component of the additive predictor requires a novel approach to parameter estimation.
 

### 2020-11-09 paper-v2.tex
Rest of Finn comments except posterior N and Cov(exp(grf))

@@
 \begin{align*}
-	\int_0^{2\pi} \lambda(s(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \left[\lambda(s(r, \theta)) + \lambda(s(r, \theta + \pi)) \right] g_k(r)\mathrm{d}\theta \\
-	&= 2\pi \lambda(s_k)g(r)
+	\int_0^{2\pi} \lambda(s_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \left[\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) \right] g_k(r)\mathrm{d}\theta \\
+	&= 2\pi \lambda(s_k)g_k(r)
 \end{align*}
@@
 The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function.
+In order to specify priors on the Mat\'ern covariance below we use a reparameterisation to a range and variance parameter.
+When $\nu = 1$ and the domain is of dimension two the reparameterisation is $\rho = \sqrt{8} / \kappa$ and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ \citep{blangiardo_spatial_2013}.
 
 The SPDE approach can at first appear to be fundamentally different to other spatial random effect approaches such as penalised regression splines.
-However, the two methods are closely related and the GRF model component in our model plays a similar role to a tensor product spline on $s$ with an appropriate smoothing penalty.
-See \cite{yue_bayesian_2014, miller_understanding_2019} for more details on the similarites between penalised regression splines and the SPDE approach.
+However, the two methods are closely related and the GRF model component in our model plays a similar role to a tensor product spline with an appropriate smoothing penalty.
+See \cite{yue_bayesian_2014, miller_understanding_2019} for more details on the similarities and differences between penalised regression splines and the SPDE approach.
 
@@
 		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
-		\caption{The A) mean, B) CV and C) SD of the posterior intensity field}
+		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$}
 		\label{fig:intensity-mean-cv}
@@
 the expected abundance.
-We approximate the posterior $\pi(N | \lambda)$ by a Monte Carlo method.
-Taking $m$ Monte Carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
+We approximate the posterior $\pi(N | \bm{Y})$ by a Monte Carlo method.
+Taking $m$ Monte Carlo samples  $\lambda^{(1)}, \ldots, \lambda^{(m)}$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)})$.
 Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
 \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
-This allows us to estimate the probability for any specific value of realised abundance $n$ by calculating $\pi(N = n | \lambda)$.
+This allows us to estimate the probability for any specific value of realised abundance $n$ by estimated $\pi(N = n | \lambda)$.
+Strictly speaking
 
@@
 Spatio-temporal extensions to this are possible, for example by considering a space-time interaction models.
-\texttt{R-INLA} has the ability to fit space-time effects that can be represented as a Kronecker product of the relevant precision matrices (e.g.
-\citep{yuan_point_2017}).
+\texttt{R-INLA} has the ability to fit space-time effects that can be represented as a Kronecker product of the relevant precision matrices \citep{blangiardo_spatial_2013, yuan_point_2017}.
 An example of this type of space-time interaction effect is an interaction between the spatial SPDE with an auto-regressive temporal process.

### 2020-11-25 analyses/fit_model.R
add additional fit with int.strategy = "auto"

@@ -6,13 +6,12 @@ library(INLA)  # need testing version of inla
 # library(rgeos)
 # library(raster)  # raster::crs() easy way to remove CRS information
 # library(maptools)
-
-init.tutorial()
+# init.tutorial()
 
 #### Save Fitted Model Object? ####
 
 #### WARNING!!!! THIS WILL OVERWRITE EXISTING OBJECT!!!!
-to_save = FALSE
+to_save = TRUE
 file_name = "akepa_2002_fitted_model_new_inlabru.RDS"
 
 
@@ -71,15 +70,8 @@ fml <- coordinates + distance ~ grf +
   log(2*pi) +   # 2*pi offset for not knowing angle theta
   Intercept
 
-
-
 W <- 58/1000   # transect radius
 distance_domain <- inla.mesh.1d(seq(.Machine$double.eps, W, length.out = 30))
-
-
-# inlabru:::iinla.setOption(iinla.verbose = TRUE)
-# inlabru:::iinla.getOption("iinla.verbose")
-
 starting_values <- list(lsig = 3.36 - log(1000))
 
 fit <-  lgcp(components = cmp, 
@@ -90,13 +82,25 @@ fit <-  lgcp(components = cmp,
              formula = fml,
              options = list(bru_max_iter = 40,
                             bru_result = starting_values,
-                            # bru_method = list(taylor = "legacy"),
-                            bru_verbose = 3))
+                            bru_verbose = 3,
+                            control.inla = list(int.strategy = "eb")))
+
+# now fit once with int.strategy = "auto"
+fit2 = lgcp(components = cmp, 
+           data = realobs,
+           samplers = samplers,
+           domain = list(coordinates = mesh,
+                         distance = distance_domain),
+           formula = fml,
+           options = list(bru_max_iter = 1,
+                          bru_result = fit,
+                          control.inla = list(int.strategy = "auto")))
 
 summary(fit)
+summary(fit2)
 
 #### Save model object ####
 
-if (to_save) saveRDS(object = fit, file = file_name)
+if (to_save) saveRDS(object = fit2, file = file_name)
 
   

### 2020-11-25 analyses/info.R
info(s)

@@ -1,0 +1,70 @@
+# Evaluate fitted model
+library(INLA)
+library(inlabru)
+library(sp)
+library(raster)
+library(patchwork)
+
+#### Fitted model  ####
+file_name = "akepa_2002_fitted_model_new_inlabru.RDS"
+fit = readRDS(file_name)
+
+study_area = readRDS("data/study_area_extended_no_crs.RDS")
+samplers = readRDS("data/samplers_extended_no_crs.RDS")
+mesh = readRDS("data/mesh_extended_no_crs.RDS")
+
+# generate samples from joint posterior of Intercept and grf st dev
+n.mc =  1000
+ps = generate(fit, 
+              formula = ~ data.frame(Intercept_latent, Stdev_for_grf),
+              n.samples = n.mc)
+
+ps = do.call(rbind, ps)
+e1 = exp(2*(ps[,1] + ps[,2]))
+e2 = exp(ps[,1] + ps[,2]/2)
+
+A0 = 1 / n.mc * sum(e1) - (1 / n.mc * sum(e2))^2
+A0
+
+# regular posterior variance
+pxl = pixels(mesh, mask = study_area)
+pred = predict(fit, pxl, ~ exp(Intercept + grf))
+A1 = pred$sd^2
+max(A1)
+
+# Info and plot it
+Info = 1 - A1 / A0
+pxl$Info = Info
+
+ggplot() +
+  gg(pxl) +
+  gg(samplers, colour = "red") +
+  coord_equal() +
+  scale_fill_viridis_c() +
+  theme_minimal()
+
+# This looks like the inverse of the intensity map.  In the north
+# sampling effort is lower and almost no birds detected.
+# In the south sampling effort is higher and more birds detected
+
+# The above is comparing variances, what about comparing sd's?
+
+Info_sd = 1 - sqrt(A1) / sqrt(A0)
+pxl$Info_sd = Info_sd
+g1 = ggplot() +
+  gg(pxl, mapping = aes(fill = Info_sd)) +
+  gg(samplers, colour = "red") +
+  coord_equal() +
+  scale_fill_viridis_c() +
+  theme_minimal()
+
+pxl$sd = pred$sd
+g2 = ggplot() +
+  gg(pxl, mapping = aes(fill = sd)) +
+  gg(samplers, colour = "red") +
+  coord_equal() +
+  scale_fill_viridis_c() +
+  theme_minimal()
+
+g1 + g2
+

### 2020-11-26 paper-v2.tex
add description of iterated INLA

@@
 We choose a log link and consider $g(s | \phi)$, where $\log\sigma^2 = \phi$.
-Our approach is based on the observation that for any single choice of $\phi$, the $\log g(s | \phi)$ term acts essentially as an offset.
-i.e If we could specify $g(s)$ as known then we could proceed with an inference approach suitable for additive linear predictors.
+However, this cannot be estimated in standard INLA formulation as the predictor is not linear in its parameters.
+To address this we use an approximate method based on a first-order Taylor approximation of the non-linear predictor.
+Let $\eta(\bm{u}) = \log \tilde{\lambda}(\bm{u})$ where $\bm{u}$ is the parameter vector of the predictor (which includes $\phi$ as well as any parameters for fixed and random effects).
+Denote by $\bar{\eta}_{\bm{u}_0}$ the first-order Taylor approximation of $\eta(\bm{u})$ at the point $\bm{u}_0$.
+That is
+\begin{equation*}
+\bar{\eta}_{\bm{u}_0} = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0)
+\end{equation*}
+where $\bm{B}_{\bm{u}_0}$ is the derivative matrix of $\eta(\bm{u})$ evaluated at $\bm{u}_0$.
+Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the linear Taylor approximation $\bar{\eta}_{\bm{u_0}}$.
+The question then is how to choose the linearisation point $\bm{u}_0$.
+The approach taken is via a fixed point iteration scheme.
+There are many possible choices here and we describe the approach in the current implementation of \texttt{inlabru} \citep{bachl_inlabru_2019}.
+However, more options are likely to be added in the future.
+We denote by $\bar{p}_{\bm{v}}$ the posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$ and define the functional
+\begin{equation}
+	f(\bar{p}_{\bm{v}}) = \left\{ \argmax_{\bm{u}_i} \: \bar{p}_{\bm{v}}(u_i | \bm{Y}), \quad i = 1, \ldots, n_{\bm{u}} \right\}
+\end{equation}  
+where $n_{\bm{u}}$ is the length of $\bm{u}$.  Each $\bar{p}_{\bm{v}} (u_i | \bm{Y})$ is approximated and the mode estimated during the model fitting using the standard INLA method.
+We then seek a fixed point of this functional so that $\bm{u}_0 = f(\bar{p}_{\bm{u}_0})$.
+This fixed point is identified via an iterative scheme.
+Given a current choice of linearisation point $\bm{v}$, estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.
+Iterate this procedure until a fixed point is identified within a chosen tolerance.
 
-\begin{itemize}
- \item Description of iterated INLA here
-\end{itemize}
+In practice the point $\bm{u}_*$ may be problematic and result in a linearised model configuration that is challenging to fit.
+Instead this point is considered only as an initial candidate point.
+The next linearisation point is chosen by setting $\bm{u}_{\alpha} = (1 - \alpha)\bm{v} + \alpha \bm{u}_*$ and finding the value of $\alpha$ that minimises $ \lVert \eta(\bm{u}_{\alpha}) - \bar{\eta}_{\bm{v}} \rVert$.
+This means that the new linearisation point remains close to a feasible region for fitting, assuming the current point $\bm{v}$ is reasonable.
+We estimate $\alpha$ using an approximate line search method that avoids many potentially expensive evaluations of the non-linear predictor.
+This iterative method is implemented in the R package \texttt{inlabru} \citep{bachl_inlabru_2019}.
 

### 2020-11-27 analyses/data/create_extended_mesh.R
update data scripts to use here()

@@ -3,8 +3,9 @@ library(inlabru)
 library(INLA)  # need testing version of inla
 library(rgeos)
 
-point_transects <- readRDS("samplers_extended_no_crs.RDS")
-study_area <- readRDS("study_area_extended_no_crs.RDS")
+data_path = here::here("analyses", "data")
+point_transects <- readRDS(here::here(data_path, "samplers_no_crs.RDS"))
+study_area <- readRDS(here::here(data_path,"study_area_extended_no_crs.RDS"))
 outer_boundary = gBuffer(study_area, width = 20)
 
 # meshbuilder()
@@ -36,4 +37,4 @@ ggplot() + gg(mesh) +
   coord_equal() + xlim(c(250, 267)) + ylim(c(2183, 2210))
 
 # save the mesh
-saveRDS(object = mesh, file = "mesh_extended_no_crs.RDS")
+saveRDS(object = mesh, file = here::here(data_path,"mesh_extended_no_crs.RDS"))

### 2020-11-27 analyses/data/create_mesh.R

@@ -2,8 +2,9 @@
 library(inlabru)
 library(INLA)  # need testing version of inla
 
-point_transects <- readRDS("samplers_no_crs.RDS")
-study_area <- readRDS("study_area_no_crs.RDS")
+data_path = here::here("analyses", "data")
+point_transects <- readRDS(here::here(data_path, "samplers_no_crs.RDS"))
+study_area <- readRDS(here::here(data_path,"study_area_no_crs.RDS"))
 
 # meshbuilder()
 
@@ -27,4 +28,4 @@ mesh <- inla.mesh.2d(boundary=boundary,
 ggplot() + gg(mesh) + gg(point_transects, colour = "red")
 
 # save the mesh
-saveRDS(object = mesh, file = "mesh_no_crs.RDS")
+saveRDS(object = mesh, file = here::here(data_path,"mesh_no_crs.RDS"))

### 2020-11-27 analyses/data/create_study_area.R

@@ -5,23 +5,24 @@ library(sp)
 library(rgeos)
 library(inlabru)
 
-samplers <- readRDS("samplers_extended_no_crs.RDS")
+data_path = here::here("analyses", "data")
+samplers <- readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
 samplersnew = samplers[samplers$Stratum == "CF",]
 # old_study_area = readRDS("study_area_no_crs.RDS")
 # chull = gConvexHull(samplersnew)
 # chull = gBuffer(chull, width = 500, capStyle = "SQUARE")
-# 
+#
 # ggplot() +
 #   gg(old_study_area) +
 #   gg(chull) +
 #   gg(samplers) +
 #   gg(samplersnew, colour = "red") +
 #   coord_equal()
-# 
+#
 # study_area_new = gUnion(old_study_area, chull)
 
 # bnd.csv = read.csv("bnd_forest2.csv")
-bnd.csv = read.csv("OpenClosedForestBoundary.csv")
+bnd.csv = read.csv(here::here(data_path, "OpenClosedForestBoundary.csv"))
 bnd.xy = data.frame(x = bnd.csv$EASTING / 1000,
                     y = bnd.csv$NORTHING / 1000)
 study_area_new <- SpatialPolygons(list(Polygons(list(Polygon(bnd.xy)), ID=1)))
@@ -32,4 +33,4 @@ ggplot() +
   gg(samplersnew, colour = "red") +
   coord_equal()
 
-saveRDS(study_area_new, file = "study_area_extended_no_crs.RDS")
+saveRDS(study_area_new, file = here::here(data_path, "study_area_extended_no_crs.RDS"))

### 2020-11-27 analyses/data/process_AnalysisDataset2002.R

@@ -4,7 +4,9 @@ library(dplyr)
 library(sp)
 library(inlabru)   # for quick plotting
 
-newdata = read.csv("AnalysisDataset2002.csv", stringsAsFactors = FALSE)
+data_path = here::here("analyses", "data")
+newdata = read.csv(here::here(data_path, "AnalysisDataset2002.csv"),
+                   stringsAsFactors = FALSE)
 
 # old_obs = readRDS("obs_2002_no_crs.RDS")
 # old_samplers = readRDS("samplers_no_crs.RDS")
@@ -15,7 +17,7 @@ newdata = read.csv("AnalysisDataset2002.csv", stringsAsFactors = FALSE)
 # RF = reforested    - no akepa here
 
 newdata %>%
-  select(SampleLabel, Easting, Northing, Stratum, HabCode) %>% 
+  select(SampleLabel, Easting, Northing, Stratum, HabCode) %>%
   filter(Stratum %in% c("OF", "CF")) %>%
   distinct() -> samplers
 
@@ -24,10 +26,10 @@ samplers$x = samplers$Easting/1000
 samplers$y = samplers$Northing/1000
 coordinates(samplers) = ~ x + y
 
-obs = newdata %>% 
-  select(SampleLabel, Distance, Easting, Northing, Stratum, HabCode) %>% 
-  filter(Stratum %in% c("OF", "CF"), !is.na(Distance)) %>% 
-  rename(distance = Distance) -> obs
+obs = newdata %>%
+  select(SampleLabel, Distance, Easting, Northing, Stratum, HabCode) %>%
+  filter(Stratum %in% c("OF", "CF"), !is.na(Distance)) %>%
+  rename(distance = Distance)
 
 obs$x = obs$Easting/1000
 obs$y = obs$Northing/1000
@@ -41,5 +43,5 @@ ggplot() +
   gg(samplers) +
   gg(obs, col = "green")
 
-saveRDS(samplers, file = "samplers_extended_no_crs.RDS")
-saveRDS(obs, file = "obs_extended_no_crs.RDS")
+saveRDS(samplers, file = here::here(data_path, "samplers_extended_no_crs.RDS"))
+saveRDS(obs, file = here::here(data_path, "obs_extended_no_crs.RDS"))

### 2020-11-27 analyses/eval_model.R
update analyses scripts with here()  [to do:  eval_spde.R, its a mess]

@@ -3,6 +3,7 @@ library(INLA)
 library(inlabru)
 library(scales)
 library(cowplot)
+library(rgeos)
 
 # set ggplot theme
 theme_set(theme_minimal())
@@ -12,12 +13,17 @@ set.seed(9701071)
 #### WARNING - as it stands this script will overwrite figures for the paper
 
 #### Fitted model  ####
-file_name = "akepa_2002_fitted_model.RDS"
-fit = readRDS(file_name)
+model_path = here::here("analyses", "akepa_2002_fitted_model_new_inlabru.RDS")
+fit = readRDS(model_path)
 
-study_area = readRDS("data/study_area_extended_no_crs.RDS")
-samplers = readRDS("data/samplers_extended_no_crs.RDS")
-mesh = readRDS("data/mesh_extended_no_crs.RDS")
+#### Other things ####
+data_path = here::here("analyses", "data")
+study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
+samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
+mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
+
+#### Where to save figures ####
+fig_path = here::here("figures")
 
 #### ggplot theme ####
 theme_set(theme_classic())
@@ -25,7 +31,7 @@ theme_set(theme_classic())
 #### Specify detection functions  ####
 
 # Log half-normal
-log_hn = function(distance, lsig){ 
+log_hn = function(distance, lsig){
   -0.5*(distance/exp(lsig))^2
 }
 
@@ -33,7 +39,7 @@ log_hn = function(distance, lsig){
 hn <- function(distance, lsig) exp(log_hn(distance, lsig))
 
 # Include r for switching to polar coords
-dsamp = function(distance, lsig){ 
+dsamp = function(distance, lsig){
   log(distance) + log_hn(distance, lsig)
 }
 
@@ -48,27 +54,31 @@ W = 58   # transect radius
 distdf <- data.frame(distance = seq(1,W,length=100))
 
 hnpred <- predict(fit, distdf, ~hn(distance,lsig))
-ghn = ggplot() + 
+ghn = ggplot() +
         gg(hnpred) +
         ylab("probability of detection\n") +
         xlab("\ndistance (m)") +
         theme(axis.title = element_text(size = 16),
[…]

### 2020-11-27 analyses/excursions.R

@@ -12,12 +12,17 @@ theme_set(theme_minimal())
 set.seed(1525)
 
 #### Fitted model  ####
-file_name = "akepa_2002_fitted_model.RDS"
-fit = readRDS(file_name)
+model_path = here::here("analyses", "akepa_2002_fitted_model_new_inlabru.RDS")
+fit = readRDS(model_path)
 
-study_area = readRDS("data/study_area_extended_no_crs.RDS")
-samplers = readRDS("data/samplers_extended_no_crs.RDS")
-mesh = readRDS("data/mesh_extended_no_crs.RDS")
+#### Other things ####
+data_path = here::here("analyses", "data")
+study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
+samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
+mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
+
+#### Where to save figures ####
+fig_path = here::here("figures")
 
 #### monte-carlo samples of intensity ####
 
@@ -25,6 +30,8 @@ mesh = readRDS("data/mesh_extended_no_crs.RDS")
 
 # There must be a way to get the actual values from
 # predict with n.samples > 1 ?
+
+# yes update this to use generate
 
 n.mc = 500     # lower this when fiddling with things
 pxl = pixels(mesh, nx = 100, ny = 100, mask = study_area)
@@ -56,15 +63,15 @@ Fpxl$F = ex$F
 
 # png(filename = "../figures/excursion_function.png",
 #     width = 4.5, height = 5.5, units = "cm", res = 1000)
-p2 = ggplot() + 
+p2 = ggplot() +
   gg(study_area) +
   gg(Fpxl) +
-  scale_fill_viridis_c(limits = c(0,1), 
+  scale_fill_viridis_c(limits = c(0,1),
                        breaks = c(0, 0.5, 1),
                        labels = c(0, 0.5, 1))+
   coord_equal() +
   theme_void() +
-  theme(legend.position = "bottom", 
+  theme(legend.position = "bottom",
         legend.direction = "horizontal") +
         #plot.margin = margin(0.1, 0, 0.1, 0, "in")) +
   guides(fill = guide_colourbar(title.vjust = 0.95,
@@ -87,7 +94,7 @@ p1 = ggplot() +
   scale_fill_manual(values = c("blue"),
                     labels = c("Excursion set")) +
   theme_void() +
[…]

### 2020-11-27 analyses/fit_hn_covariate.R

@@ -6,10 +6,11 @@ init.tutorial()
 
 #### Load Data ####
 
-realobs <- readRDS("data/obs_extended_no_crs.RDS")
-samplers <- readRDS("data/samplers_extended_no_crs.RDS")
-study_area <- readRDS("data/study_area_extended_no_crs.RDS")
-mesh <- readRDS("data/mesh_extended_no_crs.RDS")
+data_path = here::here("analyses", "data")
+realobs <- readRDS(here::here(data_path, "obs_extended_no_crs.RDS"))
+study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
+samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
+mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
 
 ### Plot detections for different habitat types
 
@@ -18,11 +19,11 @@ df$HabCode = as.factor(df$HabCode)
 df$Stratum = as.factor(df$Stratum)
 
 ### Plot detections by Stratum
-df %>% 
-  group_by(Stratum) %>% 
+df %>%
+  group_by(Stratum) %>%
   summarise(n = n())
 
-df %>% 
+df %>%
   ggplot() +
   geom_histogram(aes(x = distance, y = after_stat(density)), bins = 15) +
   facet_wrap(~ Stratum)
@@ -44,7 +45,7 @@ samplers$habcode = ifelse(samplers$HabCode == "Mesic", 1, 0)
 # lsig_int is for x == 0
 
 # Log half-normal
-log_hn = function(distance, x, lsig_int, lsig_x){ 
+log_hn = function(distance, x, lsig_int, lsig_x){
   -0.5*(distance/exp(lsig_int + lsig_x*x))^2
 }
 
@@ -55,33 +56,33 @@ hn <- function(distance, x, lsig_int, lsig_x) {
 
 # Include r for switching to polar coords
 # 2*pi for not knowing angle theta
-dsamp = function(distance, x, lsig_int, lsig_x){ 
+dsamp = function(distance, x, lsig_int, lsig_x){
   log(2*pi) + log(distance) + log_hn(distance, x, lsig_int, lsig_x)
 }
 
 
-# define distance domain and starting values 
+# define distance domain and starting values
 # for detection scale param intercept
 W <- 58/1000   # transect radius
 distance_domain <- inla.mesh.1d(seq(.Machine$double.eps, W, length.out = 30))
[…]

### 2020-11-27 analyses/fit_model.R

@@ -12,20 +12,20 @@ library(INLA)  # need testing version of inla
 
 #### WARNING!!!! THIS WILL OVERWRITE EXISTING OBJECT!!!!
 to_save = TRUE
-file_name = "akepa_2002_fitted_model_new_inlabru.RDS"
-
+model_path = here::here("analyses", "akepa_2002_fitted_model_new_inlabru.RDS")
 
 #### Load Data ####
 
-realobs <- readRDS("data/obs_extended_no_crs.RDS")
-samplers <- readRDS("data/samplers_extended_no_crs.RDS")
-study_area <- readRDS("data/study_area_extended_no_crs.RDS")
-mesh <- readRDS("data/mesh_extended_no_crs.RDS")
+data_path = here::here("analyses", "data")
+realobs <- readRDS(here::here(data_path, "obs_extended_no_crs.RDS"))
+study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
+samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
+mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
 
 #### Specify detection function  ####
 
 # Log half-normal
-log_hn = function(distance, lsig){ 
+log_hn = function(distance, lsig){
   -0.5*(distance/exp(lsig))^2
 }
 
@@ -33,14 +33,14 @@ log_hn = function(distance, lsig){
 hn <- function(distance, lsig) exp(log_hn(distance, lsig))
 
 # Include r for switching to polar coords
-dsamp = function(distance, lsig){ 
+dsamp = function(distance, lsig){
   log(distance) + log_hn(distance, lsig)
 }
 
 
 # Plot the data
 g1 <- ggplot() +
-  gg(mesh) + 
+  gg(mesh) +
   gg(study_area) +
   gg(samplers) +
   gg(realobs, colour = "green") +
@@ -57,16 +57,16 @@ g1
 # (yrange <- study_area@bbox["y",2] - study_area@bbox["y",1])
 # sort(gDistance(pts, byid = TRUE), decreasing = TRUE)
 
-matern <- inla.spde2.pcmatern(mesh, 
-                              prior.sigma = c(2, 0.01),    
+matern <- inla.spde2.pcmatern(mesh,
+                              prior.sigma = c(2, 0.01),
                               prior.range = c(300/1000, 0.01))   # was 300 before, now km units so divide
 
[…]

### 2020-11-27 analyses/info.R

@@ -6,16 +6,17 @@ library(raster)
 library(patchwork)
 
 #### Fitted model  ####
-file_name = "akepa_2002_fitted_model_new_inlabru.RDS"
-fit = readRDS(file_name)
+model_path = here::here("analyses", "akepa_2002_fitted_model_new_inlabru.RDS")
+data_path = here::here("analyses", "data")
+fit = readRDS(model_path)
 
-study_area = readRDS("data/study_area_extended_no_crs.RDS")
-samplers = readRDS("data/samplers_extended_no_crs.RDS")
-mesh = readRDS("data/mesh_extended_no_crs.RDS")
+study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
+samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
+mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
 
 # generate samples from joint posterior of Intercept and grf st dev
 n.mc =  1000
-ps = generate(fit, 
+ps = generate(fit,
               formula = ~ data.frame(Intercept_latent, Stdev_for_grf),
               n.samples = n.mc)
 

### 2020-11-27 analyses/summarise_obs.R

@@ -2,9 +2,10 @@
 library(INLA)
 library(inlabru)
 
-study_area = readRDS("data/study_area_extended_no_crs.RDS")
-samplers = readRDS("data/samplers_extended_no_crs.RDS")
-obs = readRDS("data/obs_extended_no_crs.RDS")
+data_path = here::here("analyses", "data")
+study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
+samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
+mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
 
 # number of transects
 nrow(samplers@coords)
@@ -18,7 +19,7 @@ length(unique(obs$SampleLabel))
 # range of counts at each sampling unit
 library(dplyr)
 hmm = data.frame(SampleLabel = obs$SampleLabel)
-hmm %<>% 
-  group_by(SampleLabel) %>% 
+hmm %<>%
+  group_by(SampleLabel) %>%
   summarise(n = n())
 max(hmm$n)

### 2020-11-27 analyses/posterior_N.R
move posterior abundance to own script + start new way

@@ -1,0 +1,104 @@
+library(INLA)
+library(inlabru)
+library(rgeos)
+
+# set ggplot theme
+theme_set(theme_minimal())
+
+#### Fitted model  ####
+model_path = here::here("analyses", "akepa_2002_fitted_model_new_inlabru.RDS")
+fit = readRDS(model_path)
+
+#### Other things ####
+data_path = here::here("analyses", "data")
+study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
+samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
+mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
+
+#### Where to save figures ####
+fig_path = here::here("figures")
+
+#### A new way: don't predict within transects ####
+# Unfinished - need to create my own buffer around the point locations
+# gBuffer doesn't create things that are close enough to a circle for my liking
+
+# Total surveyed region as % of study area?
+nrow(samplers@data)*pi*58^2 / gArea(study_area) * 100  # 58m radius for point transects
+
+# Buffer samplers to create transects
+samplers_buffered = gBuffer(samplers, width = 58, byid = TRUE)
+gArea(samplers_buffered) - nrow(samplers@coords)*pi*58^2   # this doesn't look great
+
+# Remove transects from study area
+study_area_no_samplers = gDifference(study_area, samplers_buffered)
+gArea(study_area) - gArea(study_area_no_samplers) - gArea(samplers_buffered)  # should be zero but isn't
+
+
+
+#### The original way of doing it ####
+# Posterior abundance
+
+# try the new way:
+ips = ipoints(study_area_no_samplers, domain = mesh)
+ggplot() +
+  gg(study_area) +
+  gg(ips) +
+  coord_equal()
+
+# how does this compare?
+ips2 = ipoints(study_area, domain = mesh)
+nrow(ips)
+nrow(ips2)
+
+# source("AS.Nposterior.R")
+predpts = ipoints(study_area, mesh)
+# abnc = AS.Nposterior(fit, predpts,
+#                      n.sd = 5,
+#                      n.grid = 100,
+#                      n.samples = 100)   # n.samples = 1000 default
+
[…]

### 2020-12-07 analyses/eval_spde.R
avoid rgeos memory leaks in posterior distance checks

@@ -1,4 +1,4 @@
-# Evaluate spde by posterior sampling
+ # Evaluate spde by posterior sampling
 
 library(INLA)
 library(inlabru)
@@ -12,25 +12,36 @@ set.seed(9701071)
 #### WARNING - as it stands this script will overwrite figures for the paper
 
 #### Fitted model  ####
-file_name = "akepa_2002_fitted_model.RDS"
-fit = readRDS(file_name)
+model_path = here::here("analyses", "akepa_2002_fitted_model_new_inlabru.RDS")
+fit = readRDS(model_path)
 
-study_area = readRDS("data/study_area_extended_no_crs.RDS")
-samplers = readRDS("data/samplers_extended_no_crs.RDS")
-mesh = readRDS("data/mesh_extended_no_crs.RDS")
-obs = readRDS("data/obs_extended_no_crs.RDS")
+data_path = here::here("analyses", "data")
+obs <- readRDS(here::here(data_path, "obs_extended_no_crs.RDS"))
+study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
+samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
+W = 58/1000
+samplers_buffered = gBuffer(samplers, width = W, byid = TRUE)
+mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
 
 ### Generate posterior point patterns
-n.pp = 25  
+n.pp = 100
 
-# simulate intensities
-llam = do.call(cbind,
-               generate(fit, mesh, ~ grf + Intercept, n.samples = n.pp))
+# simulate intensities and detection functions
+# each column of llam is a realisation of the log-intensity at each mesh node
+post.sample = generate(fit,
+                formula = ~ {c(
+                  llam = grf_latent + Intercept_latent,
+                  lsig = lsig_latent)
+                },
+                n.samples = n.pp)
+
+llam = post.sample[1:mesh$n, ]
+lsig.post = post.sample["lsig",]
 
 # create bins and matrix to store counts
-max.ds = max(as.numeric(gDistance(samplers, byid = TRUE)))
-breaks = seq(0, ceiling(max.ds), length.out = 20)
-
+# max.ds = max(as.numeric(gDistance(samplers, byid = TRUE)))
+# breaks = seq(0, ceiling(max.ds), by = round(ceiling(max.ds)) / 15)
[…]

### 2020-12-07 analyses/helper_functions.R
don't think I need this anymore

@@ -1,62 +1,0 @@
-#' thin.pp()
-#'
-#' Thin a point pattern using a known detection function and survey design
-#' Point transects only for now - add line transects in future
-#'
-#' Assumes each observation falls within only 1 transect
-#'
-#' TODO:
-#' Change hardcoded "PlotID"
-#' Avoid for loop
-#'
-#' @param pp point pattern SpatialPointsDataFrame
-#' @param samplers SpatialPointsDataFrame of point transect locations   - MUST HAVE column called PlotID, unique identifier of each transect
-#' @param W radius or half strip width
-#' @param detfn  detection function.  First argument must be the vector of distances, second argument list of params
-#' @param snap (default = TRUE) snap sampled points to centre of transects (or to the centre line).  To snap need to have plotID column
-#' @param ... arguments passed to detfn
-#' @return spatial points dataframe
-#' @import sp
-#' @export
-
-
-library(dplyr)
-library(sf)
-
-thin.pp <- function(pp, samplers, W, detfn, snap = TRUE, ...){
-  
-  if (!("plotID" %in% colnames(samplers@data))) stop("samplers needs a column called plotID")
-  samplers_area <- rgeos::gBuffer(samplers, byid = TRUE, width = W)
-  pp_det <- rgeos::gIntersection(pp, samplers_area, byid = TRUE)  # detectable points
-  
-  # Associate each nest with a sampler
-  IDs <- sp::over(pp_det, samplers_area)
-  pp_det$plotID <- IDs$plotID
-  
-  # Calculate distances:
-  distances <- rep(NA, nrow(pp_det))
-  
-  for (i in 1:nrow(pp_det)) {
-    
-    pt <- pp_det[i,]
-    transect <- subset(samplers, plotID == pt$plotID)
-    distances[i] <- rgeos::gDistance(pt, transect)
-  }
-  
-  # add distances column and create prob detected column
-  pdet <- detfn(distances, ...)
-  pp_det$distance <- distances
-  pp_det$pdet <- pdet
-  
-  # subsample detectable to get observed
-  detected <- rbinom(nrow(pp_det), 1, pdet)
-  pp_obs <- subset(pp_det, as.logical(detected))
-  
-  # snap to centre?
-  if (snap) {samplers_sf <- sf::st_as_sf(samplers)  # convert sp to sf
-  test <- dplyr::inner_join(samplers_sf, pp_obs@data, by = "plotID")
[…]

### 2020-12-07 analyses/eval_model.R
update figures produced by eval_model.R

@@ -50,10 +50,14 @@ summary(fit)
 
 # half-normal
 
-W = 58   # transect radius
-distdf <- data.frame(distance = seq(1,W,length=100))
+W = 58/1000
+Wm = 58    # transect radius in metres
+distdf <- data.frame(distance = seq(.Machine$double.eps, Wm, length=100))
 
-hnpred <- predict(fit, distdf, ~hn(distance,lsig))
+# adjust for change of units (km in model fit, metres here)
+hnpred <- predict(fit, distdf, ~ hn(distance, lsig + log(1000)),
+                  n.samples = 100)
+
 ghn = ggplot() +
         gg(hnpred) +
         ylab("probability of detection\n") +
@@ -61,14 +65,10 @@ ghn = ggplot() +
         theme(axis.title = element_text(size = 16),
         axis.text = element_text(size = 16))
 
-# detectability per transect
-df = ipoints(c(0, W), 50, name = "distance")
-dpred = predict(fit, df, ~ 2/W^2*sum(weight*exp(dsamp(distance, lsig))))
-
 # Matern plots ##
 
 blah = spde.posterior(fit, "grf", what = "matern.correlation")
-blah$x = blah$x / 1000
+# blah$x = blah$x / 1000
 gm = ggplot() +
       gg(blah) +
       ylab("correlation\n") +
@@ -86,41 +86,23 @@ plot_grid(NULL, ghn, NULL, gm, NULL,
           ncol = 5)
 dev.off()
 
-# ggsave(filename = "../figures/matern_posterior.png",
-#        width = 5, height = 5, units = "in")
-
-
-# plot
-# ggplot() +
-#   gg(study_area) +
-#   gg(mesh) +
-#   gg(samplers, colour = "red") +
-#   scale_fill_viridis_c() +
-#   xlab("Easting") +
-#   ylab("Northing") +
-#   coord_equal()
-#
-# plot(mesh, lwd = 0.2, asp =1, main = "")
-
 # intensity plots
-pxl = pixels(mesh, nx = 100, ny = 100, mask = study_area)
+pxl = pixels(mesh, nx = 300, ny = 300, mask = study_area)
+nrow(pxl@coords)
 pr.int <- predict(fit, pxl, ~ exp(grf + Intercept))
 
 # mean
 lower = min(pr.int["mean"]$mean)
-lower = signif(lower, digits = 3)
-lower
-
 upper = max(pr.int["mean"]$mean)
-upper = signif(upper, digits = 3)
[…]

### 2020-12-07 analyses/excursions.R
update excursions with new model

@@ -2,6 +2,7 @@
 library(INLA)
 library(inlabru)
 library(excursions)
+library(cowplot)
 source("gg.R")   # small edit of gg.SpatialPixelsDataFrame
 
 ## todo - replace predict loop with generate() ?
@@ -34,22 +35,25 @@ fig_path = here::here("figures")
 # yes update this to use generate
 
 n.mc = 500     # lower this when fiddling with things
-pxl = pixels(mesh, nx = 100, ny = 100, mask = study_area)
-X = matrix(NA, nrow = nrow(pxl@coords), ncol = n.mc)
-for (i in 1:n.mc){
-  print(paste("i = ", i))
-  int_draw <- predict(fit, pxl, ~ exp(grf + Intercept), n.samples = 1)
-  X[,i] = int_draw@data$mean
-}
+pxl = pixels(mesh, nx = 300, ny = 300, mask = study_area)
+
+X = generate(fit,
+             pxl,
+             ~ exp(grf + Intercept),
+             n.samples = n.mc)
+
+# X = matrix(NA, nrow = nrow(pxl@coords), ncol = n.mc)
+# for (i in 1:n.mc){
+#   print(paste("i = ", i))
+#   int_draw <- predict(fit, pxl, ~ exp(grf + Intercept), n.samples = 1)
+#   X[,i] = int_draw@data$mean
+# }
 
 #### Try excursions ####
 
-# re-scale X.  Currently per m^2
-# per km^2:
-Xkm = X*1000000
-
-# per hectare:
-Xhec = X*10000
+# re-scale X (per km)
+# o per hectare:
+Xhec = X/100
 
 # per hectare threshold:
 u = 1
@@ -116,16 +120,16 @@ plot_grid(NULL, p1, NULL, p2, NULL,
 dev.off()
 
 # contour map - how to visualise this?  ?
-cm = contourmap.mc(Xhec, levels = c(0.1, 0.5, 0.8, 1.5, 2, 4), alpha = alpha)
-str(cm)
-
-Gpxl = pxl
-Gpxl$G = cm$G
-ggplot() +
-  gg(Gpxl)
-
-cm = contourmap.mc(Xhec, levels = c(1), alpha = alpha)
-Gpxl = pxl
-Gpxl$G = cm$G
-ggplot() +
-  gg(Gpxl)
[…]

### 2020-12-07 paper-v2.tex
change one sentence in iterated inla description
change back to chicago style biblography????
Add project file to help avoid spurious whitespace differences.
fix typo in linearisation equation
update N post figure ref

@@
 We choose a log link and consider $g(s | \phi)$, where $\log\sigma^2 = \phi$.
-However, this cannot be estimated in standard INLA formulation as the predictor is not linear in its parameters.
+However, this model cannot be directly fitted using INLA since $\log g(s | \phi)$ is not linear in its parameters.
 To address this we use an approximate method based on a first-order Taylor approximation of the non-linear predictor.
@@
 \begin{equation*}
-\bar{\eta}_{\bm{u}_0} = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0)
+\bar{\eta}_{\bm{u}_0} (\bm{u}) = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0)
 \end{equation*}
@@
 	\begin{center}
-		\includegraphics[scale=0.525]{figures/realized_abundance_posterior.png}
-		\caption{Posterior abundance}
+		\includegraphics[scale=0.525]{figures/N_posterior.png}
+		\caption{Posterior abundance.  Grey area marks two Monte-carlo standard errors above and below the mean}
 		\label{fig:realized-abundance-posterior}
@@
 We believe these caveats make the map difficult to use for most non-statistically trained audiences.
-To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile map and 12,400 for the 0.975 quantile map.
+To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 2,800 for the 0.025 quantile map and 10,200 for the 0.975 quantile map.
 A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance.
 However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
-This demonstrates that the tendency to interpret these maps as possible intensities leads to interpretations that are inconsistent with the very model that generated the maps.
+This demonstrates that the tendency to interpret these maps as possible intensities leads to interpretations that are inconsistent with the model that generated the maps.
 
@@
 \clearpage
-\bibliographystyle{humannat}
+\bibliographystyle{chicago}
 \bibliography{paper}

### 2020-12-07 extras.tex
create extras.tex for stuff that will go in thesis but not paper

\documentclass[12pt]{article}
\usepackage{setspace}
\doublespacing

\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{enumerate}
\usepackage{caption}
\usepackage{textcomp}    % for Hawaii characters

% shortcuts
\newcommand{\bbeta}{\boldsymbol{\beta}}
\newcommand{\blambda}{\boldsymbol{\lambda}}
\newcommand{\T}{\intercal}
\newcommand{\bS}{\mathbf{S}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bSigma}{\boldsymbol{\Sigma}}
\newcommand{\intd}{\mathrm{d}}
\newcommand{\bm}{\boldsymbol}  % bold maths symbols
\newcommand{\bs}{\boldsymbol{s}}
\newcommand{\tl}{\tilde{\lambda}}   % thinned little lambda
\newcommand{\tL}{\tilde{\Lambda}}  % thinned big lambda

% RJC 09/08/2019 Added shortcuts for Hawaiian words
\newcommand{\akepa}{\textquotesingle\={a}kepa}  % adds Hawaiian diacritical marks
\newcommand{\Akepa}{\textquotesingle\={A}kepa}  % adds Hawaiian diacritical marks
\newcommand{\hawaii}{Hawai\textquotesingle i}   % adds Hawaiian diacritical marks
\DeclareMathOperator*{\argmax}{arg\,max}  % * means _ puts thing beneath operator

\title{Extras}
\date{}

\begin{document}

\maketitle

These things might end up in supplementary materials and my thesis but probably not in any paper submission.

\section*{Estimating abundance}

In the main paper the posterior for abundance is estimated by taking posterior realisations of the intensity at every location within the study area.  This predicts the intensity within point transects in the same manner as outside the sampled region.  However, we could use the fact that we have observed counts within point transects.  If we denote the entire study region as $A$ and denote the random variable of abundance within a bounded subset $B \subset A$ as $N(B) \sim \text{Poisson}(\Lambda(B))$ where, $\Lambda(B) = \int_{\bs \in B} \lambda(\bs) \intd \boldsymbol{s}$.  Recall also that the entire surveyed region is denoted $\Omega$, the union of all point transects.

The approach in the paper is to simulate intensities $\lambda^{(1)}, \ldots, \lambda^{(M)}$ over the entire study region $A$, where each $\lambda^{(m)}$ is sampled from the posterior distribution of $\lambda | \bm{Y}$, and then to approximate the posterior abundance via a Monte-carlo method as $N(A | \bm{Y}) \approx 1/M \sum_{m=1}^M N(A | \lambda = \lambda^{(m)}, \bm{Y})$.  However, an alternative approach is to apply the above method to only the un-sampled region $A \setminus \Omega$ and to predict only the \textit{undetected} counts within each transect.  The distribution of undetected counts across all transects is Poisson with rate parameter $ \tilde{\Lambda}(\Omega) := \int_{\bs \in \Omega}(1 - g(\bs))\lambda(\bs) \intd \bs$.  Using this approach the distribution of all undetected birds, whether through not being within the transects (and so in principle undetectable) or within a transect but not detected, is Poisson with rate parameter $\Lambda(A \setminus \Omega) + \tilde{\Lambda}(\Omega)$.  Then the distribution of total abundance can be derived by simply adding observed count to this probability distribution.  In theory, since this now uses the observed counts instead of predicting them, this should have lower variance compared to the \textit{predict everywhere} approach.  However, the point transects cover only $6.6\%$ of the total study area so the effect may be small.

To investigate whether this alternative approach makes a substantial difference to the results I compared the approximate posteriors using both methods defined using the same numerical integration scheme to ensure that any differences were not due to different schemes. Recall each point transect is denoted $\Omega_k$ with centroid $\bs_k$ and associated detection function $g_k(\bs)$. Then we can approximate the rate parameter for birds undetected within transects as 
\begin{align*}
 \tilde{\Lambda}(\Omega) &= \int_{\bs \in \Omega}(1 - g(\bs))\lambda(\bs) \intd \bs \\
 &= \sum_k \int_{\bs \in \Omega_k} (1 - g(\bs))\lambda(\bs) \intd \bs \\
 &\approx \sum_k 2 \pi \lambda(\bs_k) \int_0^W r(1-g(r))\intd r \\
 &\approx \sum_k \sum_j 2 \pi r_{kj} \alpha_{kj} \lambda(\bs_k)(1 - g(r_{kj}))
\end{align*}  
where locations $r_{kj}$ and weights $\alpha_{kj}$ are constructed following a mid-point integration scheme.

The integration scheme for $\Lambda(A \setminus \Omega)$ is defined via projection of integration locations to mesh nodes, taking into account the fact that some mesh nodes may be within or near to point transects and so should have a lower weight than in comparison to the full mesh integration scheme.  This can be achieved within \texttt{inlabru} by using the \texttt{samplers} argument in the \texttt{ipoints()} function.  Figure \ref{fig:int_scheme_raw} shows the raw integration points for the integration scheme along side those for one that also integrates within transects for comparison.  These raw weights are then projected to mesh nodes which is shown in Figure \ref{fig:int_scheme_raw}
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figures/compare_int_scheme_raw.png}
	\caption{Raw integration weights for integration over the whole study region (left) and integration over the whole region minus the transects (right).  The black circles are point transects.}
	\label{fig:int_scheme_raw}
\end{figure}
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figures/compare_int_scheme_projected.png}
	\caption{Projected integration weights for integration over the whole study region (left) and integration over the whole region minus the transects (right).  The black circles are point transects.}
	\label{fig:int_scheme_projected}
\end{figure}
where one can see the lower weight at mesh nodes within or near point transects.  I use this integration scheme to integrate over $A \setminus \Omega$. For comparison I use this same integration scheme except do not thin within point transects to predict the abundance everywhere.  The posteriors estimated by both approaches are shown in Figure \ref{fig:compare_post_N}.
\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{figures/compare_N_posteriors.png}
	\caption{comparing approximate posteriors.  Shaded regions are 2 monte-carlo standard errors above and below the mean}
	\label{fig:compare_post_N}
\end{figure}
This plot shows there is not a great deal of difference between the two approaches.  This may well be because the observed counts are so low compared to predicted abundance (153 compared to a range of 4000-8000 or so) and that detectability was fairly high.  Coupled with the fact that the sampled region is only $6.6\%$ of the total study region, it is perhaps not surprising that the results are almost identical.  I conclude that there is no much benefit to go through with the extra complication of using the observed counts over predicting everywhere.  However, for other populations this difference may well be large enough to be worth accounting for.  \texttt{inlabru} provides useful tools for this type of high-resolution quadrature that was necessary here given that transects were so small compared to the resolution of the mesh and the size of the study area as a whole.

\section*{Understanding where we have gained information}

Given we have a posterior random field describing the density of birds, it is natural to ask questions along the lines of ``which areas of space have we learned more/less about the distribution of birds?".  Given that there is a correlation that decays with distance, we have clearly learned more than just the density within each point transect, but how can we say exactly where we have learned the most, given the locations of the point transects and the observations made there.  In this section I consider an approach to answering a question these types of questions by defining various metrics that seek to quantify this idea.  However, we will see that the resulting metrics are largely driven by the assumptions of the model and, although inspired by similar approaches in the geostatistical literature, they may not easily carry over into the Poisson point process setting.

\section*{Spatio-temporal models}

Need to get other years of data from Rick to start looking into this.

\end{document}

### 2020-12-07 analyses/posterior_N.R
add int scheme comparison figures

@@ -41,14 +41,42 @@ abs(gArea(study_area) - (gArea(study_area_no_samplers) + gArea(samplers_buffered
 
 # For integration points before they are projected to mesh nodes
 # use method = "direct"
-#
-# ips_unsurveyed = ipoints(study_area_no_samplers,
-#                          domain = mesh,
-#                          int.args = list(method = "direct"))
-#
-# ips_full = ipoints(study_area,
-#                    domain = mesh,
-#                    int.args = list(method = "direct"))
+
+ips_unsurveyed = ipoints(study_area_no_samplers,
+                         domain = mesh,
+                         int.args = list(method = "direct"))
+
+ips_full = ipoints(study_area,
+                   domain = mesh,
+                   int.args = list(method = "direct"))
+
+g1 = ggplot() +
+  gg(mesh) +
+  gg(ips_full, aes(colour = weight), cex = 0.3) +
+  gg(study_area_no_samplers, alpha = 0.01) +
+  coord_equal(xlim = c(257.75, 258.25),
+              ylim = c(2190.75, 2191.25)) +
+  scale_colour_viridis_c() +
+  ggtitle("whole study area") +
+  xlab("") +
+  ylab("") +
+  guides(colour = FALSE)
+
+g2 = ggplot() +
+  gg(mesh) +
+  gg(ips_unsurveyed, aes(colour = weight), cex = 0.3) +
+  gg(study_area_no_samplers, alpha = 0.01) +
+  coord_equal(xlim = c(257.75, 258.25),
+              ylim = c(2190.75, 2191.25)) +
+  scale_colour_viridis_c() +
+  ggtitle("unsurveyed region only") +
+  xlab("") +
+  ylab("")
+
+g1 + g2 + plot_layout(guides = "collect")
+
+ggsave(filename = here::here(fig_path, "compare_int_scheme_raw.png"),
+       width = 10, height = 5, units = "in", dpi = 150)
 
 # Create integration points for full and for unsurveyed region
 ips_unsurveyed = ipoints(study_area_no_samplers,
@@ -62,22 +90,30 @@ g1 = ggplot() +
   gg(mesh) +
   gg(ips_full, aes(size = weight, colour = weight)) +
   gg(study_area_no_samplers, alpha = 0.01) +
-  coord_equal(xlim = c(258, 259),
-              ylim = c(2191, 2192)) +
[…]

### 2020-12-08 extras.tex
create figures for info(s) and write up in extras

@@
 In this section I consider an approach to answering a question these types of questions by defining various metrics that seek to quantify this idea.
-However, we will see that the resulting metrics are largely driven by the assumptions of the model and, although inspired by similar approaches in the geostatistical literature, they may not easily carry over into the Poisson point process setting.
+However, we will see that the resulting metrics are largely driven by the assumptions of the model and, although inspired by similar approaches in the geostatistical literature, they do not easily carry over into the Poisson process setting.
+The basic intuition is to compare the posterior random field informed by the data $\lambda | \bm{Y}, \theta$ and the random field conditioned only on the posterior covariance structure $\lambda | \theta$, where $\theta$ is the hyper-parameters of the SPDE as well as the intercept variable.
+The intercept should be included since the GRF and the intercept are not independent.
+We consider two quantities
+\begin{align}
+   V_0(\bs) &= \mathbb{E}_{\theta | \bm{Y}} \left[ \text{Var}(\lambda(\bs) | \theta) \right] + \text{Var}_{\theta | \bm{Y}}\left[\mathbb{E}(\lambda(\bs) | \theta)\right] \\  
+   V_{\bm{Y}}(\bs) &= \mathbb{E}_{\theta | \bm{Y}} \left[ \text{Var}(\lambda(\bs) | \theta, \bm{Y}) \right] + \text{Var}_{\theta | \bm{Y}}\left[\mathbb{E}(\lambda(\bs) | \theta, \bm{Y})\right].
+\end{align}
+Note that by the law of total variance $V_{\bm{Y}}(\bs) = \text{Var}(\lambda(\bs) | \bm{Y})$, the usual the usual posterior variance.
+$V_0(\bs)$ is a adjusted version of these where the random field is considered only conditional on $\theta$ but not on the data directly.
+This is a form of null model describing a random field with the same posterior covariance structure but no observations directly informing $\lambda$.
+Note that $V_0(\bs) = V_{\bm{Y}}(\bs)$ and so by constructing metrics of the form
+\begin{equation*}
+		I_{\text{Var}}(\bs) = 1 - \frac{V_{\bm{Y}}(\bs)}{V_0(\bs)}
+\end{equation*}  
+we get a measure between 0 and 1 that describes how much the observations $\bm{Y}$ have informed the posterior random field at location $\bs$.
+The subscript here indicates this is using the variance measure directly.
+I also consider two alternatives:
+\begin{align*}
+	I_{\text{sd}}(\bs) &= 1 - \frac{\sqrt{V_{\bm{Y}}(\bs)}}{\sqrt{V_0(\bs)}} \\ \\
+	I_{\text{cv}}(\bs) &= 1 - \frac{\sqrt{V_{\bm{Y}}(\bs)}/ \mu_{\bm{Y}}(\bs)}{\sqrt{V_0(\bs)}/\mu_0} \\
+\end{align*}
+where $\mu_{\bm{Y}}$ is the mean of $\lambda(\bs) | \bm{Y}, \theta$ and $\mu_0$ is the mean of $\lambda | \theta$.
+These are the standard deviation and the coefficient of variation versions of the metric based on the variance.
+The idea for considering all three is that the variance and standard deviation have a strong relationship with the mean due to the Poisson assumption.
+
+The usual posterior variance, standard deviation and coefficient of variation are all estimable by posterior sampling in \texttt{inlabru}.
+I describe now a way to estimate $V_0(\bs)$ using Monte-carlo sampling based on the moments of the log-normal distribution.
+This estimate can then be transformed as required for the $I_{\text{sd}}$ and $I_{\text{cv}}$ metrics.
+We have $\log \lambda | \theta \sim N(m, \sigma^2)$ where the parameters $m$ and $\sigma^2$ are either included in or derivable from the parameter vector $\theta$.
+Therefore,
+\begin{align*}
+   V_0(\bs) &= \mathbb{E}_{\theta | \bm{Y}} \left[ \text{Var}(\lambda(\bs) | \theta) \right] + \text{Var}_{\theta | \bm{Y}}\left[\mathbb{E}(\lambda(\bs) | \theta)\right] \\
+   &= \mathbb{E}_{\theta | \bm{Y}} \left[ \exp(2m + \sigma^2)(\exp(\sigma^2) - 1) \right] + \text{Var}_{\theta | \bm{Y}}\left[ \exp(m + \sigma^2/2) \right]
+\end{align*}
+Given Monte-carlo samples $\theta^{(k)} \sim \theta | \bm{Y}$ for $k = 1, \ldots, K$ this can be estimated as
+\begin{align*}
+V_0(\bs) &\approx \frac{1}{K} \sum_k \exp(2m_k + \sigma^2_k)(\exp(\sigma_k^2) - 1) + \phantom{y} \\ 
+&\phantom{\approx} \;\;\frac{1}{K} \sum_k \exp(2m_k + \sigma_k^2) - \frac{1}{K^2}\left[\sum_k \exp(m_k + \sigma_k^2 / 2) \right]^2 \\
+&= \frac{1}{K} \sum_k \exp(2m_k + 2\sigma^2_k) - \frac{1}{K^2}\left[\sum_k \exp(m_k + \sigma_k^2 / 2) \right]^2
+\end{align*}
+The three metrics are shown in Figure \ref{fig:info.png}.
+\begin{figure}
+	\centering
+	\includegraphics[width=\textwidth]{figures/info.png}
+	\caption{Comparing different information metrics.  From left to right:  $I_{\text{Var}}$,  $I_{\text{sd}}$ and $I_{\text{cv}}$.}.
+	\label{fig:info.png}
+\end{figure}
+The variance and standard deviation maps are dominated by the Poisson assumption.
+Compare these to the posterior intensity and the look extremely similar.
+The standard deviation metric looks to me more natural, in this example $V_0$ proved to be quite large (on order of 600,000) and so the metric is extremely close to 1 almost everywhere.
+Taking the square root of this (about 775) provided a number more comparable with the scale of the variation in the posterior intensity.
+Perhaps the most useful plot is $I_{\text{cv}}$.
+This shows an intuitive map - in the north, where sampling effort is lower, we have less information than in the south, where sampling effort is higher.
+One thing to note is that the posterior range of the random field is very large relative to the size of the study region.
+For random fields where this range of correlation is lower, one might expect to see ``bumps" of higher information around each sampled location.
+In this example the posterior field is too correlated for such an effect to be apparent.
+It would be interested to try out these metrics in other settings as a useful way of communicating which regions of space we have the most information about.
+The exact scale of each metric is hard to interpret, for now I am just looking at these differences in a relative sense (``more" or ``less" information compared to other regions).
 

### 2020-12-08 analyses/info.R

@@ -11,6 +11,9 @@ model_path = here::here("analyses", "akepa_2002_fitted_model_new_inlabru.RDS")
 data_path = here::here("analyses", "data")
 fit = readRDS(model_path)
 
+#### Where to save figures ####
+fig_path = here::here("figures")
+
 study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
 samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
 mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
@@ -20,8 +23,12 @@ n.mc =  1000
 ps = generate(fit,
               formula = ~ data.frame(Intercept_latent, Stdev_for_grf),
               n.samples = n.mc)
-
 ps = do.call(rbind, ps)
+head(ps)
+range(ps$Intercept_latent)
+range(ps$Stdev_for_grf)
+hist(ps$Stdev_for_grf)
+
 e1 = exp(2*(ps[,1] + ps[,2]))
 e2 = exp(ps[,1] + ps[,2]/2)
 
@@ -42,8 +49,10 @@ g1 = ggplot() +
   gg(pxl, mapping = aes(fill = Info_var)) +
   gg(samplers, colour = "red") +
   coord_equal() +
-  scale_fill_viridis_c() +
+  scale_fill_viridis_c(name = "I_var") +
   theme_minimal()
+
+g1
 
 # This looks like the inverse of the intensity map.  In the north
 # sampling effort is lower and almost no birds detected.
@@ -55,11 +64,12 @@ Info_sd = 1 - sqrt(A1) / sqrt(A0)
 pxl$Info_sd = Info_sd
 g2 = ggplot() +
   gg(pxl, mapping = aes(fill = Info_sd)) +
-  gg(samplers, colour = "red") +
+  gg(samplers, colour = "red", pch = 1) +
   coord_equal() +
-  scale_fill_viridis_c() +
+  scale_fill_viridis_c(name = "I_sd") +
   theme_minimal()
 
+g2
 
 # Coefficient of variation version:
 A1_cv = pred$cv
@@ -70,19 +80,24 @@ pxl$Info_cv = Info_cv
 
 g3 = ggplot() +
   gg(pxl, mapping = aes(fill = Info_cv)) +
-  gg(samplers, colour = "red") +
+  gg(samplers, colour = "red", pch = 1) +
   coord_equal() +
[…]

### 2020-12-14 paper.tex
add citations for data release and Rick's paper

@@
 \begin{abstract}
-Distance sampling methods aim to estimate the size of animal populations.
-Incorporating spatial models of animal density has traditionally been achieved using a two-stage approach where first-stage estimates of detectability are used as an offset in a second-stage generalised additive model.
-One-stage approaches have typically been implemented using MCMC and data augmentation methods.
-Here we present a new approach to one-stage distance sampling models, using iterated fits of latent Gaussian models fitted using integrated nested Laplace approximations (INLA) to simultaneously estimate the detectability and spatial distribution.
-We model the data as a thinned log-Gaussian Cox process using computationally efficient Gaussian Markov random field approximations to Gaussian random fields.
-We demonstrate the approach by applying the methods to point count survey data of a forest bird.
-Our Bayesian one-stage approach allows sampling from the joint posterior of all model parameters and so the resulting summaries and plots naturally incorporate uncertainty in the observation model estimates.
+Distance sampling methods aim to estimate the size and spatial distribution of populations of wild animals.
+Here we present an analysis of point transect distance sampling data collected on a tropical forest bird.
+We consider the full workflow from model specification, inference procedure, model evaluation and communication of results, at each stage discussing the statistical challenges involved and our strategies to address them.
+We take a novel point process perspective on point transect data, viewing observations as a thinned log-Gaussian Cox process with a spatially structured random effect specified via a computationally efficient Gaussian Markov random field approximation to Gaussian random field \citep{lindgren_explicit_2011}.
+One-stage inference is achieved within a Bayesian framework by using an approach based on iterated fitting of latent Gaussian models using integrated nested Laplace approximations \citep{rue_approximate_2009} to account for non-linear model components, simultaneously estimating detectability and spatial distribution.
+Model outputs are based on samples from the joint posterior of all model parameters and so naturally incorporate uncertainty due to detectability, a key advantage of the one-stage Bayesian approach.
+We discuss limitations of methods to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels by using excursions methods \citep{bolin_excursion_2015} to avoid these issues.
+The methods presented are general, flexible and applicable in a wide range of areas in spatial statistics beyond just species distribution mapping.
+\bigskip
 \end{abstract}
 
-%\begin{keyword}
-%Distance sampling \sep Stochastic partial differential equations \sep Integrated nested Laplace approximation \sep Generalized additive model
-%\end{keyword}
+\begin{keyword}
+Distance sampling \sep Species distribution modelling \sep Estimating abundance \sep Stochastic partial differential equations \sep Integrated nested Laplace approximation
+\bigskip
+\end{keyword}
 \end{frontmatter}
 
-\section*{Introduction}
+\section{Introduction}
 
-The estimation of the size of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
-Distance sampling methods aim to achieve this by using a spatially explicit sampling design and assumed detection model to estimate the detectability of animals as a function of distance from observer (in the case of point transects) or perpendicular distance to a transect line (for line transects) \citep{buckland_advanced_2004, buckland_distance_2015}.
-Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.
-The probability of detection is modelled as a function of distance and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2004}.
+The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
+Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.
+The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.
+The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.
+This information is critical when decisions about conservation strategies are made.
+However, answering these questions presents many statistical challenges.
 
-More recently, interest has focused on fully model-based approaches that include a spatially explicit model for animal density and allow the use of non-randomized survey designs.
-These methods also allow density to be associated with spatially-indexed covariates and abundance can, in principle, be estimated for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
+Firstly, as in many ecological surveys, it is impossible to undertake a full census (i.e.
+the complete enumeration of all individuals within a defined study region).
+The existing population numbers in the thousands and lives in dense forest where logistical and ecological challenges mean a census is not feasible.
+For this reason, a monitoring survey must sub-sample appropriately in space and time.
+For the \akepa{} this takes the form of the \hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}, which is a large-scale, quantitative survey of Hawaiian forest birds.
+Annual surveys of the \akepa{} study-region consist of a number of point transects located along randomly located line transects.
 
-Model-based distance sampling has been implemented in a two-stage modelling framework.
-In the first stage, detectability is estimated within each sampling unit.
-In the second stage, detections within sampling units are binned into counts and used as the response variable in a generalized additive model framework where the detectability estimates from the first stage are used as an offset.
-Due to often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.
-Negative-binomial and Tweedie distributions are common choices to model the distribution of counts.
-This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review and software package \texttt{dsm} to fit the models.
+Secondly, even at sampled locations, detectability of animals is unknown.
+The \akepa{} survey estimates detectability using a point transect distance sampling approach \citep{buckland_distance_2015} where, for each observation, the distance to the observer is recorded.
+The approach assumes a parametric form of detection function that decays with increasing distance.
+The detection function parameters are estimated from the observed distances, assuming the true density is uniform.
+This combination of spatio-temporal subsampling and unknown detectability means statistical methods to analyse such data must model a complex observation process along with the spatio-temporal distribution of animals.
 
-A key concern with the two-stage approach is the propagation of the uncertainty from the detectability estimates to the second-stage spatial model.
-Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units as well as the challenge of combining smoothers (a Bayesian/random effects concept) and boostraps (a frequentist concept) \citep{bravington_reliable_2018-1}.
-Instead, \cite{bravington_reliable_2018-1}  proposed avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the first-stage maximum-likelihood estimate.
+Thirdly, a key aim of the analysis is to predict animal density at un-sampled locations.
+If density is modelled using spatial covariates the usual approach is to assume the relationship between the animals and spatial covariates is the same in the un-sampled locations as it is at the sampled locations.
+Predictions can then be based on estimating this relationship using the observed data.
+However, we usually have reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates available or for which, in principle, no covariate could be constructed (e.g.
+the 'sociality' of the species or other behavioural characteristics).
+For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient that has been investigated but for which no straightforward explanatory cause has been found [[CITE: Rick accepted Ecography paper - no doi available yet]].
+From a statistical perspective, this suggests the inclusion of spatially-structured random effects in the model, which necessitates the use of appropriate statistical software capable of fitting such models.
 
-One-stage modelling approaches to distance sampling have also been developed within a Bayesian framework.
-The first efforts in this area used data augmentation approaches to account for unobserved individuals or groups and were fit using Markov-chain Monte Carlo (MCMC) methods \citep{royle_hierarchical_2008, schmidt_using_2012}.
-More recently  \cite{oedekoven_bayesian_2014} provided a one-stage model that avoids data augmention by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty by applying a reversible jump MCMC algorithm.
+Lastly, the above challenges mean the resulting statistical model is necessarily complex and it is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.
+The data has been collected with the clear objective of monitoring the population and informing conservation management decisions.
+These decisions will be taken by stakeholders with a range of statistical expertise who should have a good understanding of the full range of possible inferences implied by the model even if they do not understand all the statistical technicalities involved.
+This final step of communicating results is therefore key and should be considered with as much care as the rest of the analysis.
+This requires the input of statisticians to ensure uncertainty is appropriately captured and communicated in the key outputs of the analysis upon which decisions will be based.
+In particular, in the context of species distribution modelling, we note the challenge of communicating uncertainty in maps of predicted animal density.
 
-The reliance on MCMC for one-stage modelling presents a challenge for using spatially-structured random effects, which can be computationally expensive.
-\citet{yuan_point_2017} presented a one-stage approach fitted using integrated nested Laplace approximations (INLA).
-Applying their method to line-transect distance sampling, they simulateously estimated the detectability and spatial distribution of animals.
-Key to their approach is the use of a sparse Gaussian Markov random field (GMRF) approximation to a continously-indexed Gaussian random field (GRF).
-They consider the data as a realisation of a spatio-temporal log-Gaussian Cox point process, an approach that avoids the need to bin data into counts and sidesteps the issues of zero-inflation and dispersion.
-The observation process is formulated as a thinning of the log-Gaussian Cox process.
-Both the log-intensity of points and the detection function were modelled using a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011}.
+The most common method to communicate uncertainty in a spatial estimate tend is to produce maps of predictive uncertainty such as standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014} based either on a posterior predictive distribution or bootstrapping.
+Another approach is to map quantiles or the probability of exceeding certain thresholds \citep{russell_avoidance_2016, wilson_hierarchical_2010}.
+In Section \ref{results} we produce maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of a posterior spatially structured random field.
+Whilst maps that summarise spatial predictions are useful, they all mask certain properties we may care about when communicating the results of an analysis.
+For example, maps of the posterior mean will always be smoother than realisations of the posterior random field itself.
+This can obscure the spatial structure we would expect if we observed a realisation of process itself.
+Maps of the posterior coefficient of variation are hard to interpret when estimated densities are close to zero and posterior quantiles are tempting to interpret jointly when they should be viewed independently.
+This challenges mean careful consideration must be given to the communication.
 
-Here we present a similar approach to \citet{yuan_point_2017} but instead focus on the case of point transect data.
-Point transects require special treatment to account for the fact that the area surveyed increases with increasing distance from the observer.
+The most important consequences of an analysis are the possible conclusions and decisions that will be taken on the basis of the results.
+Here we advocate the view that model estimates should be converted into outputs, with associated uncertainty, that are most directly relevant to these conclusions and decisions.
+In this paper we provide an example based on a scenario where the most important aim of the analysis is identification of areas where animal density is above a certain threshold.
+It is not always the case that the most appropriate output of a species distribution model is a map of the predicted mean density.
 
-Our case study is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
-However, the point process viewpoint is not new and has been applied numerous times to line transect data \citep{buckland_model-based_2016, johnson_model-based_2010, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
-We also further develop the point process viewpoint to allow point process models to be applied to data with incomplete location information.
-It is common for distance sampling surveys to only record the distance to the observer and the observer location.
-We show how point process methods can handle this type of data even though the exact location of the point is not known.
+In this paper, we present an analysis that seeks to address these statistical challenges by presenting
 
-We replace the SPDE specification of the detection function in \cite{yuan_point_2017} with a parametric family of detection functions that are more familiar to users of standard distance sampling methods.
-This results in components of the additive predictor that are non-linear in their parameters.
-To deal with this we use a method of iterated model fits based on a Taylor expansion of the non-linear model components.
-We name this approach \textit{iterated INLA}.
+\begin{enumerate}[(i)]
+	\item A model-based spatial point process perspective on point transect distance sampling, representing the observation model as a thinning of a point processes.
+	\item A one-stage approximate Bayesian approach to inference to simultaneously estimate the observation model and spatial distribution based on iterated model fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009}.
+	\item A computationally efficient Gaussian Markov random field (GMRF) spatially structured random effect specified through a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011} to account for unobserved drivers of the spatial distribution.
+	\item Examples of model evaluation and communication of results that take advantage of the one-stage approach by sampling from the joint posterior of all model parameters and introducing excursions methods \citep{bolin_excursion_2015} as a method for investigating uncertainty in spatial predictions of density.
+These approaches are based on considering the joint probability of events at multiple spatial locations and avoid the interpretability issues of quantile or exceedance probability maps.
+\end{enumerate}
 
-To the best of our knowledge we also demonstrate the first use of a sparse GMRF spatial random effect with point transect distance sampling data.
-The sparsity of the GMRF is useful for computational efficiency and the finite element mesh motivated by the SPDE allows for efficient handling of complex spatial domains such as coastlines \citep{simpson_going_2016}.
-The flexibility of the mesh allows a higher-resolution discretisation of space where complexity requires it but also lower resolutions in areas where this is not required.
-As collecting high-resolution data and incorporating complex spatial domains (broadly defined here as domains where regular lattice discretisations of space lead to a computational bottleneck) becomes more common we believe this approximate-Bayesian GMRF approach will become a valuable addition to the spatial ecology toolbox.
+We demonstrate several novel contributions to the problem of species distribution modelling under unknown detectability.
+These include a point process perspective that accounts for incomplete location information; a one-stage approximate Bayesian inference strategy and the use of excursion sets and functions to communicate uncertainty in predictive maps.
+We note that the challenges we have addressed are generic to many types of wildlife survey data, not only the \akepa{} survey, and can be tackled using a range of possible statistical approaches.
+Our analysis here is one possible choice of approach and throughout the remainder of the paper we will highlight differences with existing approaches.
 
-% DRASTICALLY REWORK NEXT TWO PARAGRAPHS
+Our aim is to present a full analysis of the data along with discussion on model specification, inference and evaluation that reveals key issues in the analysis and our strategies to address them.
+Our hope is that the analysis presented here provides a comprehensive example of issues that arise when analysing wildlife survey data along with some novel statistical approaches that practitioners may wish to consider in their own analysis.
 
-We apply the methods to point transect survey data of Hawaiian \akepa{}, an endangered bird that is the focus of targeted conservation efforts.
-We simultaneously estimate the detectability and spatial distribution of \akepa{} in a region of Hawaii island.
-Model evaluation focuses on uncertainty in model outputs that are based on sampling from the joint posterior of all model parameters.
-This provides a natural way to produce useful outputs from the analysis that average over detection model uncertainty.
-The posterior intensity field is investigated in detail and its uncertainty visualised using excursion sets and excursion functions \citep{bolin_excursion_2015, bolin_calculating_2018}, approaches which deal with the problem of multiple hypothesis testing when mapping prediction quantiles.
+The rest of the paper proceeds as follows:  (i) we describe in detail the \akepa{} study-region and survey design; (ii) we present the perspective of distance sampling as a thinned point process; (iii) we describe the GMRF random effect and iterated INLA fitting procedure; (iv) we present the results of the analysis and discuss model evaluation and communication.
 
-The article proceeds as follows:
-\begin{enumerate}[(i)]
-\item We present point transect distance sampling as a thinned log-Gaussian Cox process, describing how this can be formulated as a latent Gaussian model with a spatial GMRF random effect that can be fitted using INLA.
-We also show how the point process model can be applied in the case of incomplete location information.
-\item We describe how the model can be specified in a modified Poisson log-likelihood form and fitted using the iterated INLA approach.
-\item A case-study using point transect distance sampling survey of the Hawaiian \akepa{}.
-We give example code showing how the above methods can be specified and fit using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension of the \texttt{R-INLA} package with additional functionality for iterated INLA.
-\end{enumerate}
-Full scripts for the analysis can be found in the supplementary materials.
+\section{Study design}
 
-\section*{Distance Sampling as a Thinned Point Process}
-
-We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(s)$.
-The log-Gaussian Cox process is a flexible approach that can include spatially structured random effects on the intensity process to account for unexplained heterogeneity not captured by fixed-effect covariates.
-
-\sloppy For the case with imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected})$.
-A key property of the log-Gaussian Cox process is that a realisation of a point process with intensity process $\lambda(s)$ that is thinned by thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.
-
-Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance.
-The distance measured depends on the type of survey.
-For line transects the perpendicular distance to the transect line is used.
-For point transects the horizontal distance to the observer is the relevant distance.
-For the remainder of the paper we assume a point transect survey design.
-
-The thinning probability function $g(s)$ is specified as a member of a parametric family of functions.
-In this paper we demonstrate our approach using the half-normal formulation, although others such as hazard rate, are also common.
-If $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$ where $\sigma$ is a variance parameter to be estimated.
-The parameter $\sigma$ can only be estimated if an assumption is made about the intensity of the animal locations.
-Without such an assumption detectability and intensity are confounded.
-The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(s)$.
-Thus any observed differences from uniformity can be attributed to detectability and not variation in the intensity.
-
-A point transect distance sampling survey consists of a set of $K$ sampling units.
-The $k$-th sample unit we denote $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region $\Omega = \cup_{k=1}^K \Omega_k$.
-For simplicity we assume that all sampling units are discs with radius $W$ and $\Omega_j \cap \Omega_k = \emptyset$ for all $j \neq k$.
-The thinning probability function $g(s)$ is defined relative to the positions of the sample units.
-The probability of observing a point at location $s$ in the $k$-th sampling unit $\Omega_k$ we denote $g_k(s)$.
-The probability of observing a point outside the surveyed region is zero.
-Since the surveyed regions are non-overlapping each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.
-For example, for an observer at location $s_k \in \Omega_k$, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
-The assumption of non-overlapping survey regions can be relaxed by including extra information such as the time of each observation but, for simplicity, we do not consider this here.
-The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for $s \in \Omega_k$.
-
-The thinned log-Gaussian Cox process likelihood with observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
-\begin{equation}
-\label{lgcp-likelihood}
-\pi(\bm{Y}) = \exp\left(-\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
-\end{equation}
-where we have suppressed parameter vector notation.
-The integral component of the likelihood does not usually have analytical solutions.
-Replacing the integral with a weighted sum allows the log-likelihood to be rewritten as a weighted Poisson log-likelihood, which we describe below, adapting the approach of \cite{simpson_going_2016} for a fully observed point pattern.
-
-To evaluate the integral we introduce polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.
-In general the thinning function depends only on $r$ and not on $\theta$.
-We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$.
-It follows that the integral can be rewritten as a sum of weighted one-dimensional integrals $\sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$ using the assumption that $\lambda(s)$ is constant within each sampling unit.
-For each sampling unit we approximate the one-dimensional integral using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
-This gives
-\begin{equation*}
-	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj})
-\end{equation*}
-where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(s_{kj}) = \lambda(s_k) g_k(r_{kj})$.
-
-To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.
-Similarly let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$.
-Then the approximate log-likelihood is
-\begin{equation}
-\label{approx-log-likelihood}
-	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}
-\end{equation}
-This approximation can be written as a modified Poisson likelihood.  To see this let $\exp \eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
-$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $y = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.
-Then the approximate likelihood becomes
-\begin{equation}
-\pi(\bm{Y}) \approx \prod_{i=1}^{KM + n} \eta_i^{y_i}\exp(-\alpha_i\eta_i)
-\end{equation}
-This is similar to a Poisson likelihood with a $\log\alpha$ offset and can be specified using the Poisson likelihood implementation available in \texttt{R-INLA}.
-
-\section*{Intensities for incomplete data}
-
-In the above we assume the data are complete records of animal location.
-However, in many distance sampling surveys only the location of the observer and the distance to the observer are recorded.
-Data of this type can be analysed within a point process framework by deriving the appropriate intensity function for the incomplete data.
-
-Using the polar coordinates notation given above, for a detected point at location $s_k(r, \theta) \in \Omega_k$ we consider the case where we observe $r$ but not $\theta$.
-It follows that the intensity for points at a distance $r$ observed within sampling unit $\Omega_k$ is
-\begin{align}
-\label{intensity-incomplete}
-\tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(s)g_k(s)\mathrm{d}s \nonumber \\
-&= 2\pi\lambda(s_k)rg_k(r)
-\end{align}
- where $c_k(r)$ is a circle of radius $r$ centred at $s_k$ and the second line follows from changing to polar coordinates and the assumption of constant intensity within $\Omega_k$.
-This intensity differs from the full data case by including the $2\pi$  term to account for the fact that we do not observe $\theta$ and the additional $r$ term to account for the increasing area surveyed at larger distances.
-
-\section*{Modelling $\lambda(s)$ using the stochastic partial differential equation approach}
-
-To model the intensity of animal locations we use a combination of fixed effects of spatial covariates and a spatially structured random effect.
-This allows us to associate distance sampling data with environmental and ecological conditions that may be relevant to understanding the ecology of the species of interest.
-For cases where we have insufficient covariates to fully explain the location of animals the spatially structured random effect accounts for additional heterogeneity in the intensity and is required to assume conditional independence between points, given the intensity.
-
-The log-intensity of animal locations is given by
-\begin{equation*}
-\log \lambda(s) = \beta_0 + \sum_v \beta_v z_v(s) + \xi(s)
-\end{equation*}
-where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatially indexed covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean GRF with Mat\'ern covariance
-\begin{equation}
-C(s_1,s_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|s_1-s_2\|)^{\nu}K_\nu(\kappa \|s_1-s_2\|)
-\end{equation}
-where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.
-All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which controls the mean-square differentiability of the process.
-We set $\nu = 1$ which is the default value in \texttt{R-INLA}.
-[[Would like to justify $\nu = 1$ somehow?]]
-
-We use a GMRF approximation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.
-Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
-The parameters $\xi_1, \ldots, \xi_L$ form a GMRF with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a SPDE.
-The parameter $\tau$ is a precision parameter and $\kappa$ an inverse range parameter that controls the rate of decay of the covariance as distance between locations increases.
-Further details of the GMRF approximation are given in the supplementary materials.
-
-\section*{Thinned Intensities}
-
-The log-intensity for observed points (the thinned intensity) is $\log\tl(s) = \log\lambda(s) + \log g(s)$.
-This presents a problem since $\log g(s)$ is not typically linear in its parameters.
-However, for any single set of parameter values $g(s)$ can be viewed as an offset term.
-If we could specify $g(s)$ as known then we could proceed with an inference approach suitable for additive linear predictors.
-
-\begin{itemize}
- \item Description of iterated INLA here
-\end{itemize}
-
-\section*{Case-Study:  Hawaiian Akepa Survey}
-
-% RJC 09/08/2019 Added text describing the akepa data, survey method, key questions and what we care about from ecology/conservation point of view.
-This text provides a background on Hawaiian forest birds, why it is important to monitor their populations and the need for unbiased and precise estimates.
-I am wondering if we need to strengthen the 'key questions' text.
-
-\bigskip
-
-% RJC 09/08/2019 Added case study text between % addtextstart and % addtextend markers.
-% addtextstart
-In this section, we apply the above methods to point transect data on the \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}), an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.
-Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the Hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}.
-Information from the HFBS was used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau) that was the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
-Due to the broad-scale coverage and robust design HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.
+The \hawaii{} \akepa{} is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.
+Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the HFBS \citep{scott_HFBS_1986}.
+Information from the HFBS is used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau), that is the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
+Due to the broad-scale coverage and robust design, HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.
 
@@
 Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern.
-Moreover, having unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.
+Moreover, unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.
 
-\subsection*{Study area and sampling method}
-
-[[Need to update this and the figure with the extended data area]]
+\subsection{Study area and survey design}
 
@@
 Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance.
-Survey points were established along 15 transects following a systematic, random design with points approximately 150 m apart on transects located either 500 or 1,000 m apart.
-Following \cite{camp_population_2010, camp_statespace_2016}, we restrict our study area to the 3,061 ha open-forest stratum of Hakalau.
+Survey points were established along 14 line transects following a systematic, random design with point transects approximately 150 m apart on line transects located either 500 or 1,000 m apart.
+We limit our study area to the open-forest and closed-forest strata of Hakalau (\autoref{fig:2002studyareapointspt}), an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016} who omitted the closed-forest stratum because it was not sampled in the early years of the \akepa{} surveys.
+For our analysis we consider data from a later year in the survey in which the closed-forest stratum was sampled and so we include it in our analysis.
 The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.
-To the north the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig.
+The closed-forest stratum was least modified by grazing and is relatively intact forest habitat.
+To the north, the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig.
 \ref{fig:2002studyareapointspt}).
-The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of their study area.
+The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of the study area.
 The west side of the study area is bounded by pasture that is dominated by grass and is unsuitable habitat for \akepa.
@@
 During 8-min counts trained observers recorded the species, exact distance to the nearest meter and detection type for each bird detected, along with the sampling conditions cloud cover, rain, wind strength, gust strength, and time of day each point was surveyed.
-\cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the open-forest study area and the bird surveys.
+\cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the study area and the bird surveys.
 
-\subsection*{Data description}
-[[ Need to update this with extended data ]]
+The survey data is available in the open forest stratum is available for download \citep{camp_datarelease_2002} and the full time series data was analysed in a spatio-temporal context in \citep{camp_dsm_2020}.
+For the purposes of our analysis we select a single survey year from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability.
+In 2002, 289 points were sampled using point-transect distance sampling methods within the 4,603 ha study area of Hakalau (Fig.
+\ref{fig:2002studyareapointspt}) during which 276 \akepa{} were detected on 121 point transects.
+The number of detections within each point transect ranged from zero to six.
+We select a single year of data to demonstrate our approach in a simplified setting where temporal aspects of analysis are not considered.
+Our approach is readily extendible to spatio-temporal modelling frameworks which we discuss in more detail in the discussion.
 
-For the purposes of our analysis we selected a single survey from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability.
-In 2002, 289 points were sampled using point-transect distance sampling methods within the 3,061 ha open-forest study area of Hakalau (Fig.
-\ref{fig:2002studyareapointspt}).
-276 \akepa{} were detected on 121 point transects.
-The number of detections within each sampling unit ranged from zero to 6.
+\section{Distance sampling as a thinned point process}
 
-%\begin{figure}
-%	\centering
-%	\includegraphics[scale=0.5]{figures/2002counts_pt}
-%	\caption{Counts of \akepa{} by point during the 2002 survey.}
-%	\label{fig:2002countspt}
-%\end{figure}
+\subsection{Overview of distance sampling methods}
 
-% addtextend
+Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and an assumed detection model to estimate the detectability of animals as a function of distance from transect \citep{buckland_advanced_2004, buckland_distance_2015}.
+Classic distance sampling approaches use a hybrid of design- and model-based inference to estimate population size where the probability of detection is modelled and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2004}.
 
-\subsection*{Model specification in \texttt{inlabru}}
+More recently, interest has focused on fully model-based approaches that include a spatially explicit model for animal density and allow non-randomized survey designs to be used.
+These methods allow density to be associated with spatially-indexed covariates and animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
+This additional flexibility to make predictions for smaller, user-specified geographic units, has made model-based distance sampling a common approach in the literature \citep{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}.
 
-In this section we introduce \texttt{inlabru} syntax and provide example code to demonstrate how to specify the model.  First, we specify the SPDE model component as
-\begin{lstlisting}
-matern <- inla.spde2.pcmatern(mesh, 
-                              prior.sigma = c(2, 0.01),    
-                              prior.range = c(130, 0.01))
-\end{lstlisting}
-which uses penalised complexity priors for the hyperparameters of the GMRF \citep{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$ where $\rho$ is the range of the Mat\'ern covariance.
+Model-based distance sampling has most commonly been implemented in a two-stage modelling framework, see the \texttt{dsm} package \citep{miller_spatial_2013}, for example.
+In the first stage, detectability is estimated.
+In the second stage, detectability estimates from the first stage are used as an offset in a generalized additive model framework.
+Detections within sampling units are binned into counts and an appropriate response distribution is chosen.
+Due to the often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.
+Negative-binomial and Tweedie distributions are common choices.
+This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review.
+
+A key concern with the two-stage approach is the propagation of uncertainty from the detection model to the second-stage spatial model.
+Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_reliable_2018-1, williams_chilean_2011}.
+Instead, \cite{bravington_reliable_2018-1} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the first-stage maximum-likelihood estimate.
+
+Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.
+These approaches tend to be Bayesian and use data augmentation to model unobserved individuals or groups and Markov-chain Monte Carlo (MCMC) methods for inference \citep{royle_hierarchical_2008, schmidt_using_2012}.
+\citet{oedekoven_bayesian_2014} demonstrate a one-stage model that avoids data augmentation by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty using reversible jump MCMC.
+
+The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use the approximate Bayesian method of integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009} applied to line transect distance sampling data.
+INLA is a popular choice for fitting models with spatially-structured random effects in large or complex spatial domains due to the implementation of a computationally efficient approach based on the construction of a sparse prior precision matrix for the random effect parameters.
+This sparse precision matrix specifies a GMRF representation of a Gaussian random field defined as the stationary solutions to an SPDE \citep{lindgren_explicit_2011} (see Section \ref{GMRFexplanation} for more details on this approach).
+The finite element methods involved use a flexible triangulation of the spatial domain and have been shown to be more efficient than regular lattice based approaches \citep{simpson_going_2016}, a useful feature for applications with complex boundaries which appear in wildlife survey data such as boundaries formed by coastlines, rivers and other natural features.
+
+\citet{yuan_point_2017} also take a point process perspective and formulate the detection model as a thinning of the full, but imperfectly observed, point pattern.
+Key to their approach was formulating the detection model as the solution to an SPDE.
+Choosing a B-spline basis and applying finite element methods results in a sparse prior precision matrix for the detection model coefficients which can then be incorporated into INLA using the same machinery as already exists for latent Gaussian precision matrices.
+A major downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing - usually a key feature of detection functions.
+The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful.
+
+Here we present a related approach to the one in \citet{yuan_point_2017} in that our models are fitted using INLA.
+However, we avoid these issues with the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal, as a user-written R function.
+This approach of choosing a parametric family of detection functions will be more familiar to users of classical distance sampling methods than the SPDE specification.
+However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in the \text{R-INLA} \citep{rue_approximate_2009} package implementation of INLA.
+We address this by using a method of iterated model fits based on a Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019}.
+
+We apply these methods to point transect \akepa data which differ from line transect methods to account for the fact that the area surveyed increases with increasing distance from the transect.
+Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
+However, the point process viewpoint is not new and has been taken numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
+
+\subsection{Model specification}
+
+In this section we introduce the statistical model for the spatial pattern of animal locations and the observation process, including a detailed description of how this model can be written as a modified Poisson likelihood.
+We use this formulation to fit the model using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension to the \texttt{R-INLA} implementation of INLA, which implements the iterated INLA approach and is available through the Comprehensive R Archive Network \citep{r_2017}.
+
+We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(.)$, i.e.
+the intensity at location $s$ is a random variable $\lambda(s)$.
+The log-Gaussian Cox process is a flexible approach that can include spatial covariates to model the mean intensity and a mean-zero spatially structured random effect to account for unexplained heterogeneity not captured by the covariates \citep{moller_log_1998}.
+
+To account for the imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected | a point is at $s$})$.
+A key property of the log-Gaussian Cox process is that a realisation of a point process with intensity process $\lambda(s)$ that is thinned by thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.
+
+Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance.
+The type of distance measured depends on the type of survey.
+For line transects the perpendicular distance to the transect line is used.
+For point transects the horizontal distance to the observer is the relevant distance.
+For the remainder of the paper we assume a point transect survey design and hence horizontal distance to the observer located at the centre of the point transect.
+
+The thinning probability function is specified as a parametric family of functions with parameter vector.
+For example, if $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$, where $\sigma^2 > 0$ is a variance parameter to be estimated.
+Other types of detection function exist, such as hazard-rate, negative exponential, and they may or may not be adjusted using series expansions to add some flexibility to obtain a better fit to a given dataset \citep{buckland_distance_2015}.
+In this analysis we assume a half-normal detection function with no covariates on detection parameters.
+
+The detection function parameters can only be estimated if an assumption is made about the intensity of the animal locations.
+Without such an assumption detectability and intensity are confounded.
+The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(s)$.
+Thus any observed deviations from uniformity can be attributed to detectability and not to variation in the intensity.
+
+A point transect distance sampling survey consists of a set of $K$ point transects.
+We denote $k$-th subset of space covered by the $k-th$ point transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.
+For simplicity, we assume that all point transects are non-overlapping discs with radius $W$, as is the case in the HFBS survey.
+The probability of observing a point at location $s \in \Omega_k$ given it is at $s \in \Omega_k$ we denote $g_k(s)$.
+The probability of observing a point outside the surveyed region is zero.
+Since the point transects are non-overlapping, each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.
+For example, for an observer at location $s_k \in \Omega_k$, the centre of the disc, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
+In practice the observer is always located at the centre of the point transect.
+The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation but this is not a requirement here.
+The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for $s \in \Omega_k$.
+
+The thinned log-Gaussian Cox process likelihood with observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
+\begin{equation}
+\label{lgcp-likelihood}
+\pi(\bm{Y}) = \exp\left( |\Omega| -\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
+\end{equation}
+where $\lambda(.)$ and $g(.)$ both depend on parameter vectors omitted for readability and $|\Omega|$ is the area of the total surveyed region.
+
+\subsection{Writing the model as a modified Poisson likelihood}
+
+\sloppy Here we show how to approximate the thinned log Gaussian Cox process model with a modified Poisson likelihood.
+The integral component $\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s$ in \autoref{lgcp-likelihood} does not have analytical solutions.
+Replacing the integral with a weighted sum allows the log-likelihood to be approximated as a weighted Poisson log-likelihood, which we describe below, adapting the approach developed by \cite{simpson_going_2016} for a fully observed point pattern.
+
+To evaluate the integral we use polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.
+In general, the thinning function depends only on $r$ and not on $\theta$.
+We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$.
+The integral in the likelihood can be simplified using an assumption that $\lambda(s)$ is linear within each sampling unit.
+This assumption implies that $\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) = 2\lambda(s_k)$ and so
+\begin{align*}
+	\int_0^{2\pi} \lambda(s_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \left[\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) \right] g_k(r)\mathrm{d}\theta \\
+	&= 2\pi \lambda(s_k)g_k(r)
+\end{align*}
+We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.
+A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piece-wise linear basis functions we use to model $\lambda$ (see section 5).
+Integrating with respect to $r$ and accounting for the change of variables gives $\int \lambda(s)g(s) \mathrm{d}s = \sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$.
+For each sampling unit we approximate the one-dimensional integral with respect to $r$ using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
+This gives
+\begin{equation*}
+	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj}) \;\;\; ,
+\end{equation*}
+where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(s_{kj}) = \lambda(s_k) g_k(r_{kj})$.
+
+To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.
+Similarly, let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$.
+Then the approximate log-likelihood is
+\begin{equation}
+\label{approx-log-likelihood}
+	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}
+\end{equation}
+This approximation can be expressed as a modified Poisson likelihood.  To see this let $\eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
+$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $z = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.
+Then the approximate likelihood becomes
+\begin{equation}
+\pi(\bm{Y}) \approx C \prod\limits_{i=1}^{KM + n} \eta_i^{z_i}\exp(-\alpha_i\eta_i)
+\end{equation}
+where $C$ is a constant.
+This approximate likelihood is implemented in \texttt{inlabru} as a \texttt{"cp"} likelihood.
+This is similar to the Berman-Turner device \citep{berman_approximating_1992, baddeley_practical_2000} although without the assumption that the locations of observed points form a part of the quadrature scheme.
+
+\subsection{Intensities for incomplete data}
+
+In the above, we assume the data are complete records of animal locations.
+However, in many distance sampling surveys only the location of the observer and the distance to the observer are recorded.
+Data of this type can be analysed within a point process framework by deriving the appropriate intensity function for the incomplete data as outlined below.
+
+Using the polar coordinates notation given above, for a detected point at location $s_k(r, \theta) \in \Omega_k$ we consider the case where we observe $r$ but not $\theta$.
+It follows that the intensity for points at a distance $r$ observed within sampling unit $\Omega_k$ is
+\begin{align}
+\label{intensity-incomplete}
+\tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(s)g_k(r)\mathrm{d}s \nonumber \\
+&= 2\pi\lambda(s_k)rg_k(r) \;\;\; ,
+\end{align}
+ where $c_k(r)$ is a circle of radius $r$ centred at $s_k$ and the second line follows from changing to polar coordinates and the assumption of linear intensity within $\Omega_k$.
+This intensity differs from the full data case through the $2\pi$  term that accounts for the fact that we do not observe $\theta$ and the additional $r$ term that accounts for the increasing area surveyed at larger distances.
+The log-intensity for observed points (the thinned intensity) is thus $\log\tl(s) = \log 2 \pi r + \log\lambda(s) + \log g(s)$ and so can be implemented during fitting by using an offset.
+As we noted above, the detection function $g(s)$ is non-linear in its parameters and this component of the additive predictor requires a novel approach to parameter estimation.
+
+\section{Iterated INLA}
+
+The half-normal detection function depends on the strictly positive parameter $\sigma^2$.
+We choose a log link and consider $g(s | \phi)$, where $\log\sigma^2 = \phi$.
+However, this model cannot be directly fitted using INLA since $\log g(s | \phi)$ is not linear in its parameters.
+To address this we use an approximate method based on a first-order Taylor approximation of the non-linear predictor.
+Let $\eta(\bm{u}) = \log \tilde{\lambda}(\bm{u})$ where $\bm{u}$ is the parameter vector of the predictor (which includes $\phi$ as well as any parameters for fixed and random effects).
+Denote by $\bar{\eta}_{\bm{u}_0}$ the first-order Taylor approximation of $\eta(\bm{u})$ at the point $\bm{u}_0$.
+That is
+\begin{equation*}
+\bar{\eta}_{\bm{u}_0} (\bm{u}) = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0)
+\end{equation*}
+where $\bm{B}_{\bm{u}_0}$ is the derivative matrix of $\eta(\bm{u})$ evaluated at $\bm{u}_0$.
+Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the linear Taylor approximation $\bar{\eta}_{\bm{u_0}}$.
+The question then is how to choose the linearisation point $\bm{u}_0$.
+The approach taken is via a fixed point iteration scheme.
+There are many possible choices here and we describe the approach in the current implementation of \texttt{inlabru} \citep{bachl_inlabru_2019}.
+However, more options are likely to be added in the future.
+We denote by $\bar{p}_{\bm{v}}$ the posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$ and define the functional
+\begin{equation}
+	f(\bar{p}_{\bm{v}}) = \left\{ \argmax_{\bm{u}_i} \: \bar{p}_{\bm{v}}(u_i | \bm{Y}), \quad i = 1, \ldots, n_{\bm{u}} \right\}
+\end{equation}
+where $n_{\bm{u}}$ is the length of $\bm{u}$.  Each $\bar{p}_{\bm{v}} (u_i | \bm{Y})$ is approximated and the mode estimated during the model fitting using the standard INLA method.
+We then seek a fixed point of this functional so that $\bm{u}_0 = f(\bar{p}_{\bm{u}_0})$.
+This fixed point is identified via an iterative scheme.
+Given a current choice of linearisation point $\bm{v}$, estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.
+Iterate this procedure until a fixed point is identified within a chosen tolerance.
+
+In practice the point $\bm{u}_*$ may be problematic and result in a linearised model configuration that is challenging to fit.
+Instead this point is considered only as an initial candidate point.
+The next linearisation point is chosen by setting $\bm{u}_{\alpha} = (1 - \alpha)\bm{v} + \alpha \bm{u}_*$ and finding the value of $\alpha$ that minimises $ \lVert \eta(\bm{u}_{\alpha}) - \bar{\eta}_{\bm{v}} \rVert$.
+This means that the new linearisation point remains close to a feasible region for fitting, assuming the current point $\bm{v}$ is reasonable.
+We estimate $\alpha$ using an approximate line search method that avoids many potentially expensive evaluations of the non-linear predictor.
+This iterative method is implemented in the R package \texttt{inlabru} \citep{bachl_inlabru_2019}.
+
+\section{Spatially Structured Random Effect}
+\label{GMRFexplanation}
+
+Using spatially-structured random effects can be computationally expensive, often resulting in dense prior precision matrices which typically have $\mathcal{O}(n^3)$ cost to factorise for an effect with $n$ parameters.
+A key development in this area is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs) and the \texttt{R-INLA} software that makes use of the sparsity to factorise the precision matrix at a cost of $\mathcal{O}(n^{3/2})$.
+This representation is constructed via a stochastic partial differential equation solved using finite element methods \citep{lindgren_explicit_2011}.
+
+The log-intensity of animal locations is given by
+\begin{equation*}
+\log \lambda(s) = \beta_0 + \sum_v \beta_v z_v(s) + \xi(s)
+\end{equation*}
+where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatial covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean GRF with Mat\'ern covariance
+\begin{equation}
+C(s_1,s_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|s_1-s_2\|)^{\nu}K_\nu(\kappa \|s_1-s_2\|)
+\end{equation}
+where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.
+All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which specifies the mean-square differentiability of the process.
+We set $\nu = 1$ which is the default value in \texttt{R-INLA}.
+
+We use a GMRF representation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.
+Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
+The parameters $\xi_1, \ldots, \xi_L$ form a GMRF with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a SPDE (see Supplementary material).
+The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function.
+In order to specify priors on the Mat\'ern covariance below we use a reparameterisation to a range and variance parameter.
+When $\nu = 1$ and the domain is of dimension two the reparameterisation is $\rho = \sqrt{8} / \kappa$ and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ \citep{blangiardo_spatial_2013}.
+
+The SPDE approach can at first appear to be fundamentally different to other spatial random effect approaches such as penalised regression splines.
+However, the two methods are closely related and the GRF model component in our model plays a similar role to a tensor product spline with an appropriate smoothing penalty.
+See \cite{yue_bayesian_2014, miller_understanding_2019} for more details on the similarities and differences between penalised regression splines and the SPDE approach.
+
+\section{Results}
+\label{results}
+
+We fitted the models using penalised complexity priors for the hyperparameters of the GMRF \citep{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$, where $\rho$ is the range parameter of the Mat\'ern covariance.
 The value of 130 was selected based on the minimum distance between sampling locations.
-
-Second, we specify the thinning probability function model component.
-The data contains only distances from the observer so we use equation \eqref{intensity-incomplete} to specify the log thinning probability function as
-\begin{lstlisting}
-thinning_function <- function(distance, lsig){
-    log(2*pi) + log(distance) + log_halfnormal(distance, lsig)
-    }
-\end{lstlisting}
-where \texttt{log\_halfnormal} is the log half-normal detection function $\log g(r | \sigma) = -1/2(r/\sigma)^2$ and the argument \texttt{lsig} is $\log(\sigma)$.
-The dataset contains a column named \texttt{distance} that contains the observed distances.
-The software will search for the appropriate columns using the names of the arguments specified in this function.
-To allow the software to distinguish between named covariates and parameters - which both appear as arguments of the function to be estimated - we must also specify a model components object
-\begin{lstlisting}
-cmp <- ~ Intercept + grf(map = coordinates, model = matern) + lsig 
-\end{lstlisting} 
-This object names all model components to be estimated.
-Note that the \texttt{grf} component is a name we have chosen and not a function.
-The software allows user-specified names for model components.
-This feature replaces the \texttt{f()} function known to users of \texttt{R-INLA} and allows the user to keep track of multiple model components rather that calling \texttt{f()} repeatedly.
-Note that \texttt{coordinates} is a keyword that allows users to supply their data in the form of a spatial object using the \texttt{sp} package \citep{pebesma_spatial_2005}.
-
-We can now specify the predictor formula using the components defined above.  
-\begin{lstlisting}
-fml <- coordinates + distance ~ Intercept + grf +
-	thinning_function(distance, lsig)
-\end{lstlisting}
-To fit the model we specify starting values for the parameters, the domain of possible observed distances and fit the model using the \texttt{lgcp} function
-% UPDATE THE W WHEN THIS CHANGES
-\begin{lstlisting}
-starting_values <- data.frame(grf = 0, lsig = 3.65, Intercept = 0)
-W <- 60   # transect radius
-distance_domain <- seq(.Machine$double.eps, W, length.out = 30)
-lgcp(components = cmp, 
-     data = obs,
-     samplers = samplers,
-     domain = list(distance = distance_domain),
-     formula = fml,
-     options = list(result = starting_values, max.iter = 30))
-\end{lstlisting}
-where our observed data is contained in a \texttt{SpatialPointsDataFrame} object named \texttt{obs}.
-We also specify the \texttt{samplers} object, also a \texttt{SpatialPointsDataFrame}, which contains the locations of all sampling units.
-Full scripts for the analysis are available in the supplementary materials.
-
-\subsection*{Results}
+The complete code for the analysis can be found in the Supplementary materials.
+Since only distance to observer was recorded we used the adjusted intensity given in \eqref{intensity-incomplete}.
 
 The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.
-This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary materials).
-The CV plot shows that the CV in the posterior intensity field is lower areas with greater sampling effort.
-There is clear preferential sampling and in the north estimated intensities are lower which also contributes to larger CV values.
-\begin{figure}
-	\includegraphics[scale=0.5]{figures/intensity_mean_cv.png}
-	\caption{(left) Posterior mean of the intensity (right) Posterior coefficient of variation of the intensity}
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
+		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$}
 		\label{fig:intensity-mean-cv}
+	\end{center}
+\end{figure}
+This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary material).
+The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort.
+There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north  (see \autoref{fig:2002studyareapointspt}).
+In the north, estimated intensities are lower, which also contributes to larger CV values.
+The standard deviation map shows stronger overall posterior variability in the south where intensity is larger.
+We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location.
+[[Say something about these - hard to know what to say about them?]]   (\autoref{fig:intensity-quantiles}).
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[scale=0.525]{figures/intensity_quantiles.png}
+		\caption{The predicted posterior intensity quantiles:  A) 0.025 quantile B) 0.975 quantile.  Both plots plots use the same colour scale}
+		\label{fig:intensity-quantiles}
+	\end{center}
 \end{figure}
 
-The fact that CV values are influenced both by sampling effort and the estimated intensity means other methods to visualise uncertainty in the posterior intensity field are also often considered.
-A common approach is to map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction cell (\autoref{fig:intensity-quantiles}).
+The posterior detection function \autoref{fig:det-matern} shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres.
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[scale=0.525]{figures/detfn_and_matern.png}
+		\caption{A) posterior half-normal detection function.
+B) posterior Mat\'ern correlation function.
+The black line is the mean posterior estimate and the grey shaded area is 95\% credible interval in both plots}
+		\label{fig:det-matern}
+	\end{center}
+\end{figure}
+The posterior Mat\'ern correlation function is shown in \autoref{fig:det-matern}, the grey regions show the 95\% credible interval.
+Note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).
+This is the correlation range on the log-intensity scale.
+[[Can I make a better "range of correlation" plot for the posterior intensity process?
+How can I convert this to something on the intensity scale?
+For a field $Z$ with covariance function $C$, is there an implied covariance for $\exp(Z)$?]]
+
+To further investigate the ability of the GMRF component to capture the heterogeneity we generated 100 thinned point patterns from the fitted model and calculated the pairwise distances between all observations for each of these patterns \autoref{fig:post-pp-distances}.
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[scale=0.6]{figures/post_pp_distances.png}
+		\caption{Boxplots of frequency of pairwise distances (tick mark values for end of each bin) based on 100 posterior samples.
+Red circles: observed frequency of pairwise distances between observations}
+		\label{fig:post-pp-distances}
+	\end{center}
+\end{figure}
+The plot shows that the GMRF component broadly captures the relevant scale of clustering in the observed data.
+There is some evidence of over-predicting the strength of the clustering at distances between 1,000 and 2,500 m.
+
+We plot the posterior abundance estimate using a Monte Carlo method.
+\begin{figure}[h]
+	\begin{center}
+		\includegraphics[scale=0.525]{figures/N_posterior.png}
+		\caption{Posterior abundance.  Grey area marks two Monte-carlo standard errors above and below the mean}
+		\label{fig:realized-abundance-posterior}
+	\end{center}
+\end{figure}
+Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.
+Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e.
+the expected abundance.
+We approximate the posterior $\pi(N | \bm{Y})$ by a Monte Carlo method.
+Taking $m$ Monte Carlo samples  $\lambda^{(1)}, \ldots, \lambda^{(m)}$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)})$.
+Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
+\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
+This allows us to estimate the probability for any specific value of realised abundance $n$ by estimated $\pi(N = n | \lambda)$.
+Strictly speaking
+
+% Add table of estimated model parameters
+% Add 95% CI figures for N?
+
+\clearpage %sort out figure placements later
+
+\section{Communication of Results}
+
+The results above are consistent with those resulting from a two-stage model fitted to binned count data (see Supplementary material).
+Our presentation of the results is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (\autoref{fig:intensity-mean-cv} and \autoref{fig:intensity-quantiles}) are not always provided there and it is common to see maps of point estimates without any accompanying communication of uncertainty.
+It is also common to report a point estimate of the expected abundance along with uncertainty in this point estimate.
+However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.
+For this reason we chose to show the full abundance posterior in \autoref{fig:realized-abundance-posterior}.
+
+Our intention in this section is to highlight limitations with mapped summaries of model predictions, including those presented in Section \ref{results}, and suggest ways to address these limitations.
+We note that all model outputs are based on sampling from the joint posterior of all model parameters which naturally average over the uncertainty in the observation process, a key advantage of the one-stage approach.
+
+\subsection{Limitations of mapped summaries}
+
+The mapped posterior mean is the most common map one encounters in the literature.  However, the mean will always be smoother than realisations from the posterior random field.
 \begin{figure}
-	\includegraphics[scale=0.5]{figures/intensity_quantiles.png}
-	\caption{The predicted posterior intensity summarised in three plots by the mean (upper),  0.025 quantile (lower left) and 0.975 quantile (lower right) for each prediction grid cell.
-All three plots use a common colour scale}
-	\label{fig:intensity-quantiles}
+	\includegraphics[scale=0.525]{figures/intensity_realized.png}
+	\caption{Three realizations of the posterior intensity field}
+	\label{fig:intensity-realizations}
 \end{figure}
-Although commonly used, we believe quantile plots like those in \autoref{fig:intensity-quantiles} to be challenging to interpret and potentially misleading.
-The temptation is to interpet these plots as showing possible intensity surfaces that produced the observed data.
-Presenting these quantiles side by side in a single map obscures the problem of multiple hypothesis testing.
-To demonstrate the fallacy of this we treated the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 3,300 for the 0.025 quantile and 12,400 for the 0.975 quantile.
-A naive (and tempting) interpretation of these numbers is as upper and lower limits of the 95\% credible interval for abundance.
+\autoref{fig:intensity-realizations} shows three such realisations from the posterior intensity field.
+Note that each realisation has a finer-grained spatial structure than is shown in the posterior mean (\autoref{fig:intensity-mean-cv}A).
+Hence, considering this finer-grained structure, our interpretation is that the clustering of animals is stronger than might be expected if we only looked at the posterior mean.
+Similarly for model evaluation we recommend plotting multiple realisations of the posterior field as these will more closely resemble the spatial structure of the observed data.
+Noting differences between realisations is also an effective way to communicate uncertainty.
+
+The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.
+However, CV values will be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.
+The observed patterns in such maps will highlight regions of relatively lower intensity, not necessarily higher overall uncertainty.
+This is clearly the case in \autoref{fig:intensity-mean-cv}B.
+
+The standard deviation map (\autoref{fig:intensity-mean-cv}C) does show clearly the spatially varying uncertainty.
+However,  care should be given to what the mapped values imply in the context of a given analysis.
+A default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high or low uncertainty.
+What such a map does not show is whether these differences actually matter in the context of the analysis.
+For example, it could be the case that the standard deviation values are so large as to make the predictions have an unacceptably high uncertainty everywhere.
+Or, on the other hand, perhaps differences between high and low uncertainty regions (`high' and `low' as defined by the default colour scale) are small enough to be negligible when it comes to making conclusions or decisions, in which case such a spatially varying map is not a relevant output of the analysis.
+These are two extreme examples.
+However, cases in between suffer from the same problem.
+
+The quantile maps also suffer from a potential difficulty in interpretation.
+The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data.
+However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for all prediction values to simultaneously achieve the 0.025 or 0.975 quantile.
+A similar problem will occur if mapping the marginal probability of exceeding a threshold at each prediction location.
+We believe these caveats make the map difficult to use for most non-statistically trained audiences.
+To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 2,800 for the 0.025 quantile map and 10,200 for the 0.975 quantile map.
+A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance.
 However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
+This demonstrates that the tendency to interpret these maps as possible intensities leads to interpretations that are inconsistent with the model that generated the maps.
 
-As an alternative to the quantile maps we use excursion sets and excursion functions \citep{bolin_excursion_2015}.
-This is a recent addition to the literature so we take some time here to define and elaborate on the concepts used.
-This approach to representing variability in stochastic processes specifically considers the joint probability of events across multiple locations.
-The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$ i.e.
-the set of all locations in $\Omega$ where $f$ exceeds the a threshold value $u$.
+\subsection{Excursion sets and excursion functions}
+
+Our solution to these problems with quantile and standard deviation maps is to suggest that consideration should be given \emph{a priori} to relevant values of the process and acceptable levels of uncertainty that are suitable given the context and aims of the analysis.
+These values can then be used to construct summary maps.
+We demonstrate this perspective using excursions sets and excursion functions \citep{bolin_excursion_2015}.
+Although there are other possible approaches [[cite here?]], excursions methods are a natural choice for Gaussian random fields.
+These methods require the user to specify thresholds of interest for the random field and acceptable levels of uncertainty.
+For this reason they avoid the issue of a default colour scale potentially affecting our interpretation of results.
+Excursion sets and excursion functions are also based on the joint probability of events across a set of locations.
+For this reason they avoid the interpretability issues of the quantile maps.
+
+The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$, i.e.
+the set of all locations in $\Omega$ where $f$ exceeds a threshold value $u$.
 For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
-% Somehow keep this information somewhere in the paper:
-%This difference between summaries of the posterior and possible intensity surfaces that are consistent with the data becomes clear if one generates realizations of the posterior intensity field.
-Three such realizations are in \autoref{fig:intensity-realizations}.
-There are some notable differences when compared to the mean and quantile summary maps.
-Each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.
-In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only look at the posterior mean of the field.
-The summary plots hide the spatial structure the model has been able to estimate from the data.
-In general, the mean of a random field will contain less spatial structure than realisations of the field.
-%
-
 \begin{equation*}
-E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \}
+E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \} \;\;\; .
 \end{equation*}
 Note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.
-As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set.
-Negative excursion sets can be similarly defined.
+As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set, with a chosen probability.
+Negative excursion sets are similarly defined.
 Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.
@@
 However, for regions in the north there is essentially no probability level for which those locations would be included in $A_u^{+}(\lambda)$.
-In the above we used 1 bird per hectare as an illustrative threshold.
-In applications multiple thresholds can be considered - based on the research question at hand.
-An attractive feature of the excursions approach is that it forces researchers to identify values that are of interest and choose an acceptable probability of error given the wider context of the research.
+We chose the threshold of 1 bird per hectare and an error probability of 0.05 to demonstrate the approach.  Multiple values for each of these can be considered as a simple extension.
 
-Given the issues of interpreting maps that simultaneously show summaries of the posterior random field it can often be informative to look directly at realisations themselves.
-Three such realizations are in \autoref{fig:intensity-realizations}.
-\begin{figure}
-	\includegraphics[scale=0.35]{figures/intensity_realized.png}
-	\caption{Three realizations of the posterior intensity field}
-	\label{fig:intensity-realizations}
-\end{figure}
-There are some notable differences when compared to the mean and quantile summary maps.
-Note each realization has finer-grained spatial structure than the mean or quantile plots of the posterior intensity.
-In practice, this indicates that we should expect more spatial structure in our collected data than what we might expect if we only considered the posterior mean of the field.
-The summary maps hide the spatial structure the model has been able to estimate from the data.
+\section{Discussion}
 
-Another key aim of performing distance sampling surveys is to estimate the abundance of the population of interest.
-We describe here how to estimate the posterior for the abundance.
-Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.
-Integrating the mean of the posterior intensity field provides a point estimate for the abundance - the expected abundance.
-We approximate the posterior $\pi(N | \lambda)$ by a monte-carlo method.
-Taking $m$ monte-carlo samples  $\lambda_1, \ldots, \lambda_m$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \lambda) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda_i)$.
-Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
-\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
-This allows us to estimate the probability for any specific value of realised abundance $n$ by calculating $\pi(N = n | \lambda)$.
-For reference, \autoref{fig:realized-abundance-comparison} shows a single Poisson PMF with rate parameter equal to the expected abundance.
-\begin{figure}
-	\includegraphics[scale=0.6]{figures/realized_abundance_posterior.png}
-	\caption{Posterior of realized abundance.  The blue line marks the expected abundance}
-	\label{fig:realized-abundance-posterior}
-\end{figure}
-\begin{figure}
-	\includegraphics[scale=0.6]{figures/realized_abundance_vs_exp.png}
-	\caption{Solid:  Posterior of realized abundance.  Dashed:  Poisson PMF with rate parameter $\mathbb{E}[\Lambda]$}
-	\label{fig:realized-abundance-comparison}
-\end{figure}
-We note here that in a GAM approach it is common to only report the estimated expected abundance and the uncertainty in this estimate.
-This does not include the additional variability that comes from the Poisson distribution with a rate parameter equal to the expected abundance.
+This paper presents a new approach to analysing point transect distance sampling data that incorporates several innovative methods and considers the bulk of an applied statistician's workflow of model specification, inference procedure, model evaluation and communication of results.
+Although any of these areas on its own would be substantive enough to be the main focus of a paper, our brief presentation of issues in sequence along with our chosen solutions will be useful to statisticians who face similar questions and data structures.
 
-Finally, we show that the one-stage model has estimated a plausible detection function.
-\autoref{fig:half-normal} shows a summary of the posterior for the detection function, plotting the posterior mean along with 95\% credible intervals.
-We note again that all model summaries shown in this section are based on monte-carlo sampling of the joint posterior of all model parameters.
-As such all the plots incorporate the uncertainty in the detection process.
-\begin{figure}
-	\includegraphics[scale=0.6]{figures/halfnormal.png}
-	\caption{Posterior detection function}
-	\label{fig:half-normal}
-\end{figure}
+There are several natural extensions to the model.
+For this analysis we only considered a single survey year.
+Spatio-temporal extensions to this are possible, for example by considering a space-time interaction models.
+\texttt{R-INLA} has the ability to fit space-time effects that can be represented as a Kronecker product of the relevant precision matrices \citep{blangiardo_spatial_2013, yuan_point_2017}.
+An example of this type of space-time interaction effect is an interaction between the spatial SPDE with an auto-regressive temporal process.
+\cite{blangiardo_spatial_2013} show how this can be done with numerous examples and code snippets.
 
-\bigskip
+Another natural extension is to consider more complicated observation processes.
+There is the potential to include explanatory covariates on detection function parameters.
+For example, we did not make use of the detection type (audio or visual) in our analysis, although this was recorded and detectability may vary depending on whether a bird is seen or heard.
+Other factors could be considered, such as weather conditions, observer expertise, animal behaviour and morphological traits.
+Any additional parameters to the non-linear model components will be estimated using the outer-optimisation step of iterated INLA.
+[[Can I speculate here about how iterated INLA might perform as parameters are added to non-linear components?
+Is, higher-dimensional optimisation slower/infeasible/unstable at a certain point?]]
 
-\newpage
+The iterated INLA inference procedure is a general approach that could be applied to other non-linear model components that arise in the area of spatial ecology.
+We note two promising areas in which our flexible inference procedure may be useful:
 
-\section*{Discussion}
+\begin{itemize}
+	\item \textbf{Selectivity analysis} In surveys of the marine environment, the size, length and weight of fish affects the likelihood of being caught in a trawl net.
+Thus any analysis of these data must account for this.
+The probability of capture is typically modelled as a logistic function with parameters that depend on morphological traits \citep{herrmann_understanding_2016, madsen_selectivity_2007, galbraith_demersal_1994}, which could be estimated jointly with the spatial distribution within our inference framework.
+	\item \textbf{Functional responses} In the area of understanding species-habitat or species-species interactions, the concept of a `functional response' captures the idea that the abundance of a resource or risk will affect the response to it \citep{holling_some_1959}.
+These varying responses can be described mathematically using parametric equations that have the desired properties, e.g.
+a horizontal asymptote so that beyond a certain threshold no extra amount of a resource will affect the response to it.
+These types of responses are typically not considered in species distribution models but our framework allows them to be estimated alongside the more common linear or quadratic fixed effects.
+This flexibility in specifying model components could also be more widely applicable in areas such as public health and spatial epidemiology, where responses to conditions may be non-linear.
 
-\subsection*{General remarks}
+\end{itemize}
+We note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analysis
 
-The most common method to analyse combine spatial models and distance sampling data is a two-stage approach with a generalised additive model that includes a spatial smoothing spline.
-The approach here of specifying a point process model and including an SPDE effect can seem fundamentally different.
-However, these two approaches are related and an analysis of the \akepa{} data using the GAM approach provides similar point estimates.
-There is a correspondence between SPDEs and smoothing splines: the smoothing spline is an estimator of the posterior mean of the SPDE \citep{kimeldorf_spline_2002}.
-The SPDE and the smoothing spline play a similar role within the model and the resulting mean field map looks similar in both approaches.
-In addition, it can appear as though the point process model avoids the need to bin data into counts and so avoids some loss of information.
-This is not the case since we also applied the standard distance sampling assumption of constant intensity within each point transect.
-This can be thought of as the point process version of binning data into counts since the location of points within transects does not inform the spatial distribution of points across the whole study region.
-The main advantage of the log-Gaussian Cox process is that the discretisation of space occurs during the step of approximating the GRF.
-This is in contrast to motivating the discretisation by the act of binning the data into counts so as to create a response variable that can be analysed using a GAM model.
-We also note that \texttt{inlabru} provides functionality for fitting models with multiple likelihoods \citep{bachl_inlabru_2019}.
-As such our approach is extendible to more complicated contexts such as marked point processes.
+In our analysis we only briefly touch upon the problem of communicating uncertainty in maps and highlight the excursions methods as a new and positive addition to this area.
+In our example, the threshold was chosen to illustrate the method but we believe that discussions with relevant stakeholders will lead to agreement on a (possibly large) set of thresholds that are relevant given the aims and context of the analysis.
+This requires the input of both the relevant domain experts and statisticians so stakeholders understand the consequences of the thresholds and uncertainty levels they wish to consider.
+We also note the potential for excursions to be used to define the notion of occupancy (and thus map occupied areas, with a given probability of error).
 
-We believe the excursions approach is a valuable addition to the literature on communicating uncertainty in estimated surfaces.
-For species distribution modelling this has the potential to be a useful tool for model evaluation and communication with stakeholders.
-In particular the approach requires users to \textit{a priori} decide on (i) a threshold or set of threshold values of interest and (ii) an acceptable probability of error.
-We believe that, in most situations, these decisions should be made by domain experts who have an understanding of the phenomena of interest, the set of possible decisions that may be taken on the basis of the analysis and the consequences of error.
-Once these decisions have been made the statistician can then proceed accordingly, providing outputs  that inform on quantities of interest with a level of uncertainty that stakeholders have agreed is acceptable.
-We note that the excursions approach applied here has the potential to be useful in other methods of species distribution mapping, environmental statistics and spatial statistics in general.
-There is the potential to use excursions to motivate both the \textit{definition} and \textit{estimation} of concepts of interest to ecologists, such as a species' home range or the notion of occupancy.
+We end with a note on the similarity between what we have done and existing two stage approach.
+The one-stage thinned point process perspective can seem fundamentally different to a two-stage GAM approach fitted to count data.
+It can be tempting to think that, because the point process model does not require us to bin the data into counts, we can avoid some loss of information.
+However, we also made the same assumption of constant intensity within each transect as in more traditional distance sampling approaches.
+This leads to the same loss of information since we do not use the exact location of points within each transect, except to account for detectability.
 
-\subsection*{Possible future research}
+In summary, we have presented a novel framework for analysing point transect distance sampling data that introduces new approaches to model specification, inference procedure and communication of results.
+These developments will be of interest to those with similar data structures in spatial ecology as well as researchers in other areas who we hope will explore the full flexibility and generality of these developments in their own work.
 
-Here we applied the iterated INLA approach in order to estimate the half-normal detection function which is non-linear in its $\log \sigma$ parameter.
-We note that this approach provides a general framework for estimating parametric functions from data and is not restricted to detection functions in a distance sampling model.
-The \texttt{inlabru} software allows users to specify the model in an intuive and modular way.
-Users write individual functions to define (possibly non-linear) model components and specify how these are combined to form the additive predictor.
-By additive we mean a predictor that is a sum of model components, and not necessarily additive in terms of parameters to be estimated.
+% spp has potential to provide general framework? [Janine comment]
+% Mention modelling on sphere is possible without projecting on to R^2?
 
-We presented a simple choice for the detection model.
-The half-normal function has only one parameter to be estimated but other, more flexible, detection functions are common in applications.
-The potential to include further flexibility in the detection model through covariates or random-effects is an open research question and would dramatically increase the usability of our approach presented here.
-Increasing the number of parameters to estimate in the non-linear model components may run into computational challenges.
-This has yet to be explored.
-
-Model selection is a general challenge for models with spatially structured random effects.
-Traditional tools of model assessment and selection do not apply in many contexts.
-For model assessment we advise a pragmatic approach based on comparing summaries of posterior samples to the observed data.
-A moodel selection procedure is useful to users to justify decisions regarding which covariates to include, whether or not to include a spatially structured random effect, and if so, what type of effect should be used.
-
+\clearpage
 \bibliographystyle{chicago}

### 2020-12-14 analyses/info.R
info(s) figures for extras.pdf (draft)

@@ -47,7 +47,7 @@ pxl$Info_var = Info
 
 g1 = ggplot() +
   gg(pxl, mapping = aes(fill = Info_var)) +
-  gg(samplers, colour = "red") +
+  gg(samplers, colour = "red", pch = 1) +
   coord_equal() +
   scale_fill_viridis_c(name = "I_var") +
   theme_minimal()
@@ -87,11 +87,9 @@ g3 = ggplot() +
 
 g3
 
-
-g1 + g2 + g3
+g1 + g2 + g3 + plot_annotation(tag_levels = "A")
 ggsave(filename = here::here(fig_path, "info.png"),
        width = 10, height = 5, dpi = 150)
-
 
 # Regular sd plot
 pxl$sd = pred$sd

### 2021-09-01 analyses/posterior_N.R
update figures for thesis

@@ -101,10 +101,14 @@ g1 = ggplot() +
   coord_equal(xlim = c(257.75, 258.25),
               ylim = c(2190.75, 2191.25)) +
   scale_colour_viridis_c() +
-  ggtitle("whole study area") +
   xlab("") +
   ylab("") +
-  guides(size = FALSE)
+  guides(size = FALSE) +
+  theme(axis.text = element_text(size = 12),
+        axis.title = element_text(size = 12),
+        legend.title = element_text(size = 12, vjust = 0.75),
+        legend.text = element_text(size = 12),
+        legend.key.size = unit(2, "line"))
 
 g2 = ggplot() +
   gg(mesh) +
@@ -113,12 +117,16 @@ g2 = ggplot() +
   coord_equal(xlim = c(257.75, 258.25),
               ylim = c(2190.75, 2191.25)) +
   scale_colour_viridis_c() +
-  ggtitle("unsurveyed region only") +
   xlab("") +
   ylab("") +
-  guides(size = FALSE, colour = FALSE)
+  guides(size = FALSE, colour = FALSE) +
+  theme(axis.text = element_text(size = 12),
+        axis.title = element_text(size = 12),
+        legend.title = element_text(size = 12, vjust = 0.75),
+        legend.text = element_text(size = 12),
+        legend.key.size = unit(2, "line"))
 
-g1 + g2 + plot_layout(guides = "collect")
+g1 + g2 + plot_layout(guides = "collect") + plot_annotation(tag_levels = "A")
 
 ggsave(filename = here::here(fig_path, "compare_int_scheme_projected.png"),
        width = 10, height = 5, units = "in", dpi = 150)

### 2021-09-01 paper.tex
use feedback branch for feedback
RSS style files and start editing for final draft before submission

@@
-%% Template article for Elsevier's document class `elsarticle'
-%% with numbered style bibliographic references
-\documentclass[preprint,12pt]{elsarticle}
+\documentclass{stylefile16/statsoc}
+\usepackage[a4paper]{geometry}   % otherwise statsoc class doesn't compile to A4 correctly
 
-% remove preprint footnote
-\makeatletter
-\def\ps@pprintTitle{%
- \let\@oddhead\@empty
- \let\@evenhead\@empty
- \def\@oddfoot{\centerline{\thepage}}%
- \let\@evenfoot\@oddfoot}
-\makeatother
-
-\usepackage{setspace}
-\doublespacing
+%\usepackage{setspace}
+%\doublespacing
 
@@
 \usepackage{natbib}
-\setcitestyle{authoryear}
-
-%% Use the option review to obtain double line spacing
-%% \documentclass[preprint,review,12pt]{elsarticle}
-
-%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
-%% for a journal layout:
-%% \documentclass[final,1p,times]{elsarticle}
-%% \documentclass[final,1p,times,twocolumn]{elsarticle}
-%% \documentclass[final,3p,times]{elsarticle}
-%% \documentclass[final,3p,times,twocolumn]{elsarticle}
-%% \documentclass[final,5p,times]{elsarticle}
-%% \documentclass[final,5p,times,twocolumn]{elsarticle}
+\setcitestyle{stylefile16/rss}
 
@@
 \usepackage{textcomp}    % for Hawaii characters
-%\usepackage{listings}
-%\lstset{language=R}
-\usepackage[T1]{fontenc}  % tilde in middle in lstlisting
-\usepackage[formats]{listings}
-\lstset{language=R,
-		basicstyle=\ttfamily,
-		columns=fixed,
-		basewidth=0.5em,
-		literate={~}{{$\sim$}}1
-		}
-
-\biboptions{comma,round}
 
@@
 
+\title[One-stage point transect distance sampling using iterated INLA]{One-stage point transect distance sampling using iterated integrated nested Laplace approximations}
+
+\author[Andrew E. Seaton {\it et al.}]{Andrew E. Seaton}
+\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, 
+	University of St Andrews, 
+	St Andrews,
+	UK}
+\email{aes22@st-andrews.ac.uk}
+
+\author{Janine B. Illian}
+\address{School of Mathematics and Statistics, University of Glasgow, Glasgow, UK}
+
+\author{Finn Lindgren}
+\address{School of Mathematics, University of Edinburgh, Edinburgh, UK}
+
+\author{Richard J. Camp}
+\address{U. S. Geological Survey, Pacific Island Ecosystems Research Center, P.O. Box 44, \hawaii{} National Park, HI 96718, U.S.A.}
+
+% seems like this document class uses short author from the last author
+% macro in the preamble?  
+\author[Andrew E. Seaton \textit{et al.}]{Steve J. Kendall}
+\address{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}
+
 \begin{document}
-\begin{frontmatter}
-\title{One-stage point transect distance sampling using iterated integrated nested Laplace approximations}
-
-%% use the tnoteref command within \title for footnotes;
-%% use the tnotetext command for the associated footnote;
-%% use the fnref command within \author or \address for footnotes;
-%% use the fntext command for the associated footnote;
-%% use the corref command within \author for corresponding author footnotes;
-%% use the cortext command for the associated footnote;
-%% use the ead command for the email address,
-%% and the form \ead[url] for the home page:
-%%
-%% \title{Title\tnoteref{label1}}
-%% \tnotetext[label1]{}
-%% \author{Name\corref{cor1}\fnref{label2}}
-%% \ead{email address}
-%% \ead[url]{home page}
-%% \fntext[label2]{}
-%% \cortext[cor1]{}
-%% \address{Address\fnref{label3}}
-%% \fntext[label3]{}
-
-%% use optional labels to link authors explicitly to addresses:
-%% \author[label1,label2]{<author name>}
-%% \address[label1]{<address>}
-%% \address[label,2]{<address>}
-
-% RJC 09/08/2019 Added co-authors and affiliations. Note that I used \affil[]{} instead of \address[]{}
-\author[1,*]{Andrew E Seaton}
-\author[1,2]{Richard J Camp}
-\author[3]{Finn Lindgren}
-\author[1]{Janine B Illian}
-\author[1]{David L Borchers}
-\author[1]{David L Miller}      % Ask about co-authoring the manuscript
-\author[1]{Len Thomas}          % Ask about co-authoring the manuscript
-\author[1]{Stephen T Buckland}  % Ask about co-authoring the manuscript
-\author[4]{Steve J Kendall}     % It is Steve's data we are using and I have already mentioned this manuscript to him.
-
-%\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}
-\address[1]{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, University of St Andrews, St Andrews, Fife, Scotland}
-\address[2]{U. S. Geological Survey, Pacific Island Ecosystems Research Center, P.O. Box 44, \hawaii{} National Park, HI 96718, U.S.A.}
-\address[3]{School of Mathematics, University of Edinburgh, Edinburgh, Scotland}
-\address[4]{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}
-\address[*]{Correspondence: Andrew E Seaton, Email: aes22@st-andrews.ac.uk}
 
@@
 Distance sampling methods aim to estimate the size and spatial distribution of populations of wild animals.
-Here we present an analysis of point transect distance sampling data collected on a tropical forest bird.
-We consider the full workflow from model specification, inference procedure, model evaluation and communication of results, at each stage discussing the statistical challenges involved and our strategies to address them.
-We take a novel point process perspective on point transect data, viewing observations as a thinned log-Gaussian Cox process with a spatially structured random effect specified via a computationally efficient Gaussian Markov random field approximation to Gaussian random field \citep{lindgren_explicit_2011}.
-One-stage inference is achieved within a Bayesian framework by using an approach based on iterated fitting of latent Gaussian models using integrated nested Laplace approximations \citep{rue_approximate_2009} to account for non-linear model components, simultaneously estimating detectability and spatial distribution.
-Model outputs are based on samples from the joint posterior of all model parameters and so naturally incorporate uncertainty due to detectability, a key advantage of the one-stage Bayesian approach.
-We discuss limitations of methods to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels by using excursions methods \citep{bolin_excursion_2015} to avoid these issues.
-The methods presented are general, flexible and applicable in a wide range of areas in spatial statistics beyond just species distribution mapping.
-\bigskip
+Here we present an analysis of point transect distance sampling data collected on an endangered tropical forest bird.
+We consider a workflow from model specification, inference method, model evaluation and communication of results, at each stage discussing the statistical challenges involved and our strategies to address them.
+We take a point process perspective on point transect data, viewing observations of birds as a thinned log-Gaussian Cox process with a spatially structured random effect describing the spatial distribution of animals specified via a computationally efficient Gaussian Markov random field approximation to Gaussian random field \citep{lindgren_explicit_2011}.
+
+We use a novel approximate Bayesian inference approach based on iterated fitting of latent Gaussian models using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009} to account for non-linear model components and jointly estimate detectability and spatial distribution, a process that is usually separated into two models.
+Outputs from the model are based on samples from the joint posterior of all model parameters and so naturally incorporate uncertainty due to detectability, a key advantage of the one-stage Bayesian approach.
+We discuss limitations of methods to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels by using excursions methods \citep{bolin_excursion_2015}.
+
+We present an example of a computationally efficient one-stage Bayesian approach to analysing distance sampling data.
+However, the inference method presented is general, flexible and applicable in a wide range of areas in statistics beyond modelling species distributions.
+
 \end{abstract}
 
-\begin{keyword}
-Distance sampling \sep Species distribution modelling \sep Estimating abundance \sep Stochastic partial differential equations \sep Integrated nested Laplace approximation
-\bigskip
-\end{keyword}
-\end{frontmatter}
+\keywords{Distance sampling, Species distribution modelling, Estimating abundance, Stochastic partial differential equations, Integrated nested Laplace approximation, SPDE approach}
 
@@
 The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
-Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.
+Here we present an analysis of wildlife survey data collected on a critically endangered \hawaii{}an forest bird that aims to meet this objective.
 The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.
@@
 This information is critical when decisions about conservation strategies are made.
-However, answering these questions presents many statistical challenges.
+However, estimating the abundance and spatial distribution of wild populations of animals presents many statistical challenges.
 
-Firstly, as in many ecological surveys, it is impossible to undertake a full census (i.e.
+First, as in many ecological surveys, it is impossible to undertake a full census (i.e.
 the complete enumeration of all individuals within a defined study region).
-The existing population numbers in the thousands and lives in dense forest where logistical and ecological challenges mean a census is not feasible.
+The existing population numbers in the thousands and lives in dense tropical forest where logistical and ecological challenges mean a census is not feasible.
 For this reason, a monitoring survey must sub-sample appropriately in space and time.
@@
 
-Secondly, even at sampled locations, detectability of animals is unknown.
+Second, even at surveyed locations, the detectability of animals is unknown.
 The \akepa{} survey estimates detectability using a point transect distance sampling approach \citep{buckland_distance_2015} where, for each observation, the distance to the observer is recorded.
-The approach assumes a parametric form of detection function that decays with increasing distance.
-The detection function parameters are estimated from the observed distances, assuming the true density is uniform.
-This combination of spatio-temporal subsampling and unknown detectability means statistical methods to analyse such data must model a complex observation process along with the spatio-temporal distribution of animals.
+The probability of detecting a bird is modelled using a parametric detection function that decays with increasing distance.
+The detection function parameters are estimated from the observed distances by assuming the true density is uniform with respect to distance and attributing deviations from uniformity to imperfect detectability.
+The combination of spatio-temporal subsampling and unknown detectability means statistical methods to analyse distance sampling data must model a complex observation process alongside the spatial distribution of animals.
 
-Thirdly, a key aim of the analysis is to predict animal density at un-sampled locations.
-If density is modelled using spatial covariates the usual approach is to assume the relationship between the animals and spatial covariates is the same in the un-sampled locations as it is at the sampled locations.
-Predictions can then be based on estimating this relationship using the observed data.
-However, we usually have reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates available or for which, in principle, no covariate could be constructed (e.g.
-the 'sociality' of the species or other behavioural characteristics).
-For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient that has been investigated but for which no straightforward explanatory cause has been found [[CITE: Rick accepted Ecography paper - no doi available yet]].
-From a statistical perspective, this suggests the inclusion of spatially-structured random effects in the model, which necessitates the use of appropriate statistical software capable of fitting such models.
+Third, a key aim of the analysis is to predict animal density at unsurveyed locations.
+If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed location.
+However, in many ecological contexts there can be good reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates available or for which, in principle, no covariate could be constructed (e.g.
+the `sociality' of the species).
+For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient for which no straightforward explanatory cause has been found \cite{camp_dsm_2020}.
+From a statistical perspective, this heterogeneity that cannot be explained by available covariates suggests the inclusion of spatially structured random effects in the model.
+Such effects require a careful consideration of the model structure and inference methods to ensure that model fitting remains computationally feasible.
 
-Lastly, the above challenges mean the resulting statistical model is necessarily complex and it is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.
-The data has been collected with the clear objective of monitoring the population and informing conservation management decisions.
-These decisions will be taken by stakeholders with a range of statistical expertise who should have a good understanding of the full range of possible inferences implied by the model even if they do not understand all the statistical technicalities involved.
-This final step of communicating results is therefore key and should be considered with as much care as the rest of the analysis.
+Lastly, the above challenges mean the resulting statistical model is necessarily complex.
+It is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.
+The \akepa{} data is collected with the clear objective of monitoring the population and informing conservation management decisions.
+These decisions will be taken by stakeholders with a range of statistical expertise.
+This final step of communicating results should therefore be considered with as much care as the rest of the analysis.
 This requires the input of statisticians to ensure uncertainty is appropriately captured and communicated in the key outputs of the analysis upon which decisions will be based.
 In particular, in the context of species distribution modelling, we note the challenge of communicating uncertainty in maps of predicted animal density.
-
-The most common method to communicate uncertainty in a spatial estimate tend is to produce maps of predictive uncertainty such as standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014} based either on a posterior predictive distribution or bootstrapping.
-Another approach is to map quantiles or the probability of exceeding certain thresholds \citep{russell_avoidance_2016, wilson_hierarchical_2010}.
-In Section \ref{results} we produce maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of a posterior spatially structured random field.
-Whilst maps that summarise spatial predictions are useful, they all mask certain properties we may care about when communicating the results of an analysis.
-For example, maps of the posterior mean will always be smoother than realisations of the posterior random field itself.
-This can obscure the spatial structure we would expect if we observed a realisation of process itself.
-Maps of the posterior coefficient of variation are hard to interpret when estimated densities are close to zero and posterior quantiles are tempting to interpret jointly when they should be viewed independently.
-This challenges mean careful consideration must be given to the communication.
-
-The most important consequences of an analysis are the possible conclusions and decisions that will be taken on the basis of the results.
-Here we advocate the view that model estimates should be converted into outputs, with associated uncertainty, that are most directly relevant to these conclusions and decisions.
-In this paper we provide an example based on a scenario where the most important aim of the analysis is identification of areas where animal density is above a certain threshold.
-It is not always the case that the most appropriate output of a species distribution model is a map of the predicted mean density.
 
@@
 
+The most common method to communicate uncertainty in a spatial estimate tend is to produce maps of predictive uncertainty such as standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014} based either on a posterior predictive distribution or bootstrapping.
+Another approach is to map quantiles or the probability of exceeding certain thresholds \citep{russell_avoidance_2016, wilson_hierarchical_2010}.
+In Section \ref{results} we produce maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of a posterior spatially structured random field.
+Whilst maps that summarise spatial predictions are useful, they all mask certain properties we may care about when communicating the results of an analysis.
+For example, maps of the posterior mean will always be smoother than realisations of the posterior random field itself.
+This can obscure the spatial structure we would expect if we observed a realisation of process itself.
+Maps of the posterior coefficient of variation are hard to interpret when estimated densities are close to zero and posterior quantiles are tempting to interpret jointly when they should be viewed independently.
+This challenges mean careful consideration must be given to the communication.
+
+The most important consequences of an analysis are the possible conclusions and decisions that will be taken on the basis of the results.
+Here we advocate the view that model estimates should be converted into outputs, with associated uncertainty, that are most directly relevant to these conclusions and decisions.
+In this paper we provide an example based on a scenario where the most important aim of the analysis is identification of areas where animal density is above a certain threshold.
+It is not always the case that the most appropriate output of a species distribution model is a map of the predicted mean density.
+
 The mapped posterior mean is the most common map one encounters in the literature.  However, the mean will always be smoother than realisations from the posterior random field.
@@
 \clearpage
-\bibliographystyle{chicago}
+\bibliographystyle{stylefile16/rss}
 \bibliography{paper}

### 2021-09-01 paper.tex
edits up to end of intro

@@
 
-In this paper, we present an analysis that seeks to address these statistical challenges by presenting
+We present an analysis that addresses these statistical challenges by using the following approaches:
 
 \begin{enumerate}[(i)]
-	\item A model-based spatial point process perspective on point transect distance sampling, representing the observation model as a thinning of a point processes.
-	\item A one-stage approximate Bayesian approach to inference to simultaneously estimate the observation model and spatial distribution based on iterated model fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009}.
+	\item A spatial point process perspective on point transect distance sampling and representing the detection model as a thinning of a point processes.
+	\item A one-stage approximate Bayesian approach to inference based on repeated model fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009}.
+This allows for the simultaneous estimation of the spatial distribution and detectability of animals.
 	\item A computationally efficient Gaussian Markov random field (GMRF) spatially structured random effect specified through a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011} to account for unobserved drivers of the spatial distribution.
-	\item Examples of model evaluation and communication of results that take advantage of the one-stage approach by sampling from the joint posterior of all model parameters and introducing excursions methods \citep{bolin_excursion_2015} as a method for investigating uncertainty in spatial predictions of density.
-These approaches are based on considering the joint probability of events at multiple spatial locations and avoid the interpretability issues of quantile or exceedance probability maps.
+	\item A discussion on model evaluation and communication of results that takes advantage of the one-stage approach by sampling from the joint posterior of all model parameters.
+We advocate for excursions methods \citep{bolin_excursion_2015} as a way to handle uncertainty in spatial predictions of density.
 \end{enumerate}
@@
 We demonstrate several novel contributions to the problem of species distribution modelling under unknown detectability.
-These include a point process perspective that accounts for incomplete location information; a one-stage approximate Bayesian inference strategy and the use of excursion sets and functions to communicate uncertainty in predictive maps.
-We note that the challenges we have addressed are generic to many types of wildlife survey data, not only the \akepa{} survey, and can be tackled using a range of possible statistical approaches.
-Our analysis here is one possible choice of approach and throughout the remainder of the paper we will highlight differences with existing approaches.
+In particular, we present, to the best of our knowledge, the first analysis of point transect distance sampling data using a point process perspective which requires a modified intensity function that accounts for incomplete location information for animal sightings.
+This is also the first application of a novel approximate Bayesian inference method to estimate the detection function jointly with the spatial model.
+This method is readily generalisable to other applications beyond distance sampling.
 
-Our aim is to present a full analysis of the data along with discussion on model specification, inference and evaluation that reveals key issues in the analysis and our strategies to address them.
-Our hope is that the analysis presented here provides a comprehensive example of issues that arise when analysing wildlife survey data along with some novel statistical approaches that practitioners may wish to consider in their own analysis.
-
-The rest of the paper proceeds as follows:  (i) we describe in detail the \akepa{} study-region and survey design; (ii) we present the perspective of distance sampling as a thinned point process; (iii) we describe the GMRF random effect and iterated INLA fitting procedure; (iv) we present the results of the analysis and discuss model evaluation and communication.
+% Update to section refs
+The rest of the paper proceeds as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the GMRF random effect; and Sections \ref{sec-results} and \ref{sec-comms} present the results of the analysis and discuss model evaluation and communication.
 
 \section{Study design}
+\label{sec-study-design}
 
@@
 \section{Distance sampling as a thinned point process}
+\label{sec-ds-pp}
 
@@
 INLA is a popular choice for fitting models with spatially-structured random effects in large or complex spatial domains due to the implementation of a computationally efficient approach based on the construction of a sparse prior precision matrix for the random effect parameters.
-This sparse precision matrix specifies a GMRF representation of a Gaussian random field defined as the stationary solutions to an SPDE \citep{lindgren_explicit_2011} (see Section \ref{GMRFexplanation} for more details on this approach).
+This sparse precision matrix specifies a GMRF representation of a Gaussian random field defined as the stationary solutions to an SPDE \citep{lindgren_explicit_2011} (see Section \ref{sec-gmrf} for more details on this approach).
 The finite element methods involved use a flexible triangulation of the spatial domain and have been shown to be more efficient than regular lattice based approaches \citep{simpson_going_2016}, a useful feature for applications with complex boundaries which appear in wildlife survey data such as boundaries formed by coastlines, rivers and other natural features.
@@
 
-\subsection{Intensities for incomplete data}
+\subsection{Intensity functions for incomplete data}
 
@@
 \section{Iterated INLA}
+\label{sec-iinla}
 
@@
 \section{Spatially Structured Random Effect}
-\label{GMRFexplanation}
+\label{sec-gmrf}
 
@@
 \section{Results}
-\label{results}
+\label{sec-results}
 
@@
 \section{Communication of Results}
+\label{sec-comms}
 
@@
 \section{Discussion}
+\label{sec-discussion}
 

### 2021-09-02 paper.tex
edits up to 2.1

@@
 We demonstrate several novel contributions to the problem of species distribution modelling under unknown detectability.
-In particular, we present, to the best of our knowledge, the first analysis of point transect distance sampling data using a point process perspective which requires a modified intensity function that accounts for incomplete location information for animal sightings.
+This is, to the best of our knowledge, the first analysis of point transect distance sampling data using a point process perspective.
+This requires a modified intensity function that accounts for incomplete location information for animal sightings.
 This is also the first application of a novel approximate Bayesian inference method to estimate the detection function jointly with the spatial model.
-This method is readily generalisable to other applications beyond distance sampling.
+This inference method is readily generalisable to other applications beyond distance sampling models.
+As such, this analysis will be of interest to the broad applied statistics community in general, as well as those specifically interested in modelling species distributions.
 
@@
 Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the HFBS \citep{scott_HFBS_1986}.
-Information from the HFBS is used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau), that is the first wildlife refuge with the primary purpose to protect, conserve and manage native forests for threatened and endangered bird and plant species.
-Due to the broad-scale coverage and robust design, HFBS has become the baseline to determine changes in bird species distributions, population sizes and trends in density patterns over time.
+Information from the HFBS is used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau).
+Hakalau is the first wildlife refuge established with the primary purpose to protect and manage native forests for the conservation of threatened and endangered bird and plant species.
+Due to its broad-scale coverage and robust design, the HFBS has become an invaluable dataset used to determine changes in bird species distributions, population sizes and trends in density patterns over time.
 
-During the 20th century \akepa{} declined dramatically due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994},  mosquito-transmitted avian diseases \citep{pratt_avifaunal_1994, atkinson_wildlife_1995}, introduced predators \citep{lepson_akepa_1997}, and food resources competitors \citep{lepson_akepa_1997}.
-\Akepa{} has a global abundance of approximately 16,200 (95\%CI 10,000\textendash25,200) birds that has been restricted to five spatially distinct populations \citep{judge_akepa_2018}.
-Hakalau supports the largest \akepa{} population that in 2012 was estimated at more than 11,000 birds \citep{camp_statespace_2016}.
+During the 20th century, the \akepa{} population declined dramatically due to habitat modification \citep{scott_HFBS_1986, pratt_avifaunal_1994},  mosquito-transmitted avian diseases \citep{pratt_avifaunal_1994, atkinson_wildlife_1995}, introduced predators \citep{lepson_akepa_1997}, and food resources competitors \citep{lepson_akepa_1997}.
+\Akepa{} has a global abundance of approximately 16,200 (95\%CI 10,000\textendash25,200) birds that is now restricted to five spatially distinct populations \citep{judge_akepa_2018}.
+Hakalau supports the largest remaining distinct \akepa{} population which, in 2012, was estimated at more than 11,000 birds \citep{camp_statespace_2016}.
 Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern.
-Moreover, unbiased and precise abundance estimates are required by land and resource managers for evaluating management actions and establishing management planning, and policy makers for decision-making processes.
+Unbiased and precise abundance estimates are therefore required by land and resource managers for the purposes of evaluating management actions and developing future management plans.
 

### 2021-09-02 paper.tex
edits up to section 3

@@
 
-Hakalau was established in 1985 to conserve 15,390-ha of montane forest habitat for native forest birds and rainforest plants.
-Annual forest bird surveys were initiated in 1987 to determine population status and track trends in abundance.
-Survey points were established along 14 line transects following a systematic, random design with point transects approximately 150 m apart on line transects located either 500 or 1,000 m apart.
-We limit our study area to the open-forest and closed-forest strata of Hakalau (\autoref{fig:2002studyareapointspt}), an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016} who omitted the closed-forest stratum because it was not sampled in the early years of the \akepa{} surveys.
-For our analysis we consider data from a later year in the survey in which the closed-forest stratum was sampled and so we include it in our analysis.
-The open-forest stratum was previously heavily grazed, and since the removal of cattle in 1988 regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.
-The closed-forest stratum was least modified by grazing and is relatively intact forest habitat.
-To the north, the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig.
+Hakalau was established in 1985 to conserve 15,390 ha of montane forest habitat for native forest birds and rainforest plants.
+Annual forest bird surveys were initiated in 1987 to determine the population status and track trends in abundance.
+Survey points were established along 14 line transects following a systematic, random design with point transects located approximately 150m apart on line transects located either 500m or 1,000m apart.
+We limit our study area to the open-forest and closed-forest strata of Hakalau (\autoref{fig:2002studyareapointspt}), an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016}, who omitted the closed-forest stratum because it was not sampled in the early years of the surveys.
+For our analysis, we use data from a later year in which the closed-forest stratum was sampled and so we are able to include it.
+The open-forest stratum was previously heavily grazed, and, since the removal of cattle in 1988, regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.
+The closed-forest stratum was historically least modified by grazing and is relatively intact forest habitat.
+The northern boundary of the study area follows the refuge boundary while to the east it is bounded by a fence line (Fig.
 \ref{fig:2002studyareapointspt}).
-The southern boundary was modified from \cite{camp_population_2010} to exclude the non-sampled forested portion of the study area.
-The west side of the study area is bounded by pasture that is dominated by grass and is unsuitable habitat for \akepa.
+The southern boundary is the same as that in \cite{camp_population_2010}, chosen to exclude unsurveyed regions to the south of Hakalau but does not represent a physical boundary.
+To the west of the study area is pasture that is dominated by grass and is unsuitable habitat for \akepa{} and the boundary here marks the edge of the forest.
 
@@
 	\includegraphics[scale=0.5]{figures/study_area_design.png}
-	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.}
+	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.  Eastings and Northings are in kilometres.}
 	\label{fig:2002studyareapointspt}
@@
 
-Surveys used point-transect methods, recording horizontal distances from survey points to individually detected birds.
-Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength \textgreater Baufort scale 3).
-During 8-min counts trained observers recorded the species, exact distance to the nearest meter and detection type for each bird detected, along with the sampling conditions cloud cover, rain, wind strength, gust strength, and time of day each point was surveyed.
+The survey follows a point-transect distance sampling design, recording the horizontal distances from observers to detected birds.
+Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength greater than Beaufort scale 3).
+During 8-minute counts, trained observers recorded the species, distance to the nearest metre and detection type for each bird detected, along with the sampling conditions cloud cover, rain, wind strength, gust strength, and time of day each point was surveyed.
 \cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the study area and the bird surveys.
 
-The survey data is available in the open forest stratum is available for download \citep{camp_datarelease_2002} and the full time series data was analysed in a spatio-temporal context in \citep{camp_dsm_2020}.
-For the purposes of our analysis we select a single survey year from the \akepa{} time series that contained on broad sampling of the study area with sufficient numbers of detections to estimate detectability.
+The survey data in the open forest stratum is available for download \citep{camp_datarelease_2002} and the full time series data was analysed in a spatio-temporal context in \citep{camp_dsm_2020}.
+For the purposes of our analysis we select a single survey year from the \akepa{} time series (2002) that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability.
 In 2002, 289 points were sampled using point-transect distance sampling methods within the 4,603 ha study area of Hakalau (Fig.
@@
 The number of detections within each point transect ranged from zero to six.
-We select a single year of data to demonstrate our approach in a simplified setting where temporal aspects of analysis are not considered.
-Our approach is readily extendible to spatio-temporal modelling frameworks which we discuss in more detail in the discussion.
+We selected data from a single year to demonstrate our approach in a simplified setting that does not require a temporal model component.
+The choice of year is somewhat arbitrary and the model has also been tested analysing data from other years with similar results to those presented here.
+However, our approach is readily extendible to a multi-year analysis, similar to that in \cite{camp_dsm_2020}, which we discuss in more detail in Section \ref{sec-discussion}.
 

### 2021-09-02 paper.tex
edits up to 3.2

@@
 
-Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and an assumed detection model to estimate the detectability of animals as a function of distance from transect \citep{buckland_advanced_2004, buckland_distance_2015}.
-Classic distance sampling approaches use a hybrid of design- and model-based inference to estimate population size where the probability of detection is modelled and a randomized sampling design allows the construction of Horvitz-Thompson-like estimators of animal density \citep{horvitz_generalization_1952,  buckland_advanced_2004}.
+Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and a parametric detection function to estimate the detectability of animals as a function of distance from transect or observer \citep{buckland_advanced_2004, buckland_distance_2015}.
+Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.
+The probability of detection is modelled using a parametric equation and, given estimates of detectability, a randomized sampling design allows for the construction of Horvitz-Thompson-like estimators of animal density \citep{ buckland_advanced_2004, horvitz_generalization_1952}.
+Under this approach, animal density is uniform within each strata of the survey.
 
-More recently, interest has focused on fully model-based approaches that include a spatially explicit model for animal density and allow non-randomized survey designs to be used.
-These methods allow density to be associated with spatially-indexed covariates and animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
-This additional flexibility to make predictions for smaller, user-specified geographic units, has made model-based distance sampling a common approach in the literature \citep{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}.
+More recently, interest has focused on fully model-based approaches that model a spatially varying animal density and allow for the use of non-random survey designs.
+These methods allow animal density to be associated with spatially-indexed covariates and non-parametric random effects.
+Therefore, animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
+As such, varying density estimates are not restricted by the stratification design required for Horvitz-Thompson-like estimators.
+This additional flexibility to make predictions for smaller, user-specified geographic units, and extrapolate relationships to unsurveyed regions has made model-based distance sampling a common approach in the literature \citep{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}.
 
-Model-based distance sampling has most commonly been implemented in a two-stage modelling framework, see the \texttt{dsm} package \citep{miller_spatial_2013}, for example.
-In the first stage, detectability is estimated.
+Model-based distance sampling has most commonly been implemented in a two-stage modelling framework.
+In the first stage, detectability is estimated using the frequency of observed distances to fit the detection function.
 In the second stage, detectability estimates from the first stage are used as an offset in a generalized additive model framework.
-Detections within sampling units are binned into counts and an appropriate response distribution is chosen.
-Due to the often sparse nature of wildlife survey data this may require consideration of over- or under-dispersion and zero-inflated distributions.
-Negative-binomial and Tweedie distributions are common choices.
+Detections are are binned into counts within user-defined sampling units (usually point transects or segments of line transects) and an appropriate response distribution is chosen to model these counts.
+The \texttt{dsm} package \citep{miller_spatial_2013}, for example, provides tools to do this using the \texttt{R} package \texttt{mgcv} \citep{wood_gam_2017} to implement the spatial count model.
+Due to the often sparse nature of wildlife survey data, this may require consideration of over- or under-dispersion and zero-inflated distributions and the negative-binomial or Tweedie distributions are common choices.
 This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review.
 
-A key concern with the two-stage approach is the propagation of uncertainty from the detection model to the second-stage spatial model.
-Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties of the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_reliable_2018-1, williams_chilean_2011}.
-Instead, \cite{bravington_reliable_2018-1} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of detectability around the first-stage maximum-likelihood estimate.
+A key concern with the two-stage approach is the propagation of uncertainty from the detection model to the spatial model.
+Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties with the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_VariancePropagationDensity_2021, williams_chilean_2011}.
+Instead, \cite{bravington_VariancePropagationDensity_2021} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of the detection function at the first-stage maximum-likelihood estimate.
 
 Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.
-These approaches tend to be Bayesian and use data augmentation to model unobserved individuals or groups and Markov-chain Monte Carlo (MCMC) methods for inference \citep{royle_hierarchical_2008, schmidt_using_2012}.
-\citet{oedekoven_bayesian_2014} demonstrate a one-stage model that avoids data augmentation by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty using reversible jump MCMC.
+One-stage analyses of distance sampling data have tended to be Bayesian and use Markov chain Monte Carlo (MCMC) methods for inference, usually requiring data augmentation to model unobserved individuals or groups
+% problem with this Royle reference here.  Not sure what I was intending to cite.
+\citep{royle_HierarchicalModelSpatial_2008, schmidt_using_2012}.
+\citet{oedekoven_bayesian_2014} present a one-stage model that avoids data augmentation and incorporate model uncertainty using reversible jump MCMC.
 
-The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use the approximate Bayesian method of integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009} applied to line transect distance sampling data.
-INLA is a popular choice for fitting models with spatially-structured random effects in large or complex spatial domains due to the implementation of a computationally efficient approach based on the construction of a sparse prior precision matrix for the random effect parameters.
-This sparse precision matrix specifies a GMRF representation of a Gaussian random field defined as the stationary solutions to an SPDE \citep{lindgren_explicit_2011} (see Section \ref{sec-gmrf} for more details on this approach).
-The finite element methods involved use a flexible triangulation of the spatial domain and have been shown to be more efficient than regular lattice based approaches \citep{simpson_going_2016}, a useful feature for applications with complex boundaries which appear in wildlife survey data such as boundaries formed by coastlines, rivers and other natural features.
+The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use INLA \citep{rue_approximate_2009} and present an application on line transect distance sampling data.
+\citet{yuan_point_2017} take a point process perspective and formulate the detection model as a thinning of a point pattern.
+Key to their approach is formulating the detection model as the solution to an SPDE.
+An approximate solution the SPDE is constructed using a B-spline basis and finite element methods which results in a sparse prior precision matrix for the detection model parameters which can be incorporated into the model in INLA.
 
-\citet{yuan_point_2017} also take a point process perspective and formulate the detection model as a thinning of the full, but imperfectly observed, point pattern.
-Key to their approach was formulating the detection model as the solution to an SPDE.
-Choosing a B-spline basis and applying finite element methods results in a sparse prior precision matrix for the detection model coefficients which can then be incorporated into INLA using the same machinery as already exists for latent Gaussian precision matrices.
-A major downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing - usually a key feature of detection functions.
+A major downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, usually a key feature of detection functions.
 The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful.
+Representing the observation model as an SPDE also differs substantially from standard distance sampling models that use a parametric model for the detection function.
+This may have hindered a wider adoption of using INLA for distance sampling analyses, \cite{yuan_point_2017} remains the only example we are aware of.
+Furthermore, parametric forms of detection functions are readily extendible by including covariates and random effects to explain variation in detectability.
+It is not clear how to do this the SPDE formulation.
+These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA and the benefits of a one-stage modelling process.
 
-Here we present a related approach to the one in \citet{yuan_point_2017} in that our models are fitted using INLA.
-However, we avoid these issues with the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal, as a user-written R function.
-This approach of choosing a parametric family of detection functions will be more familiar to users of classical distance sampling methods than the SPDE specification.
-However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in the \text{R-INLA} \citep{rue_approximate_2009} package implementation of INLA.
-We address this by using a method of iterated model fits based on a Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019}.
+Here we present an approach that is closely related to \citet{yuan_point_2017}.
+We also fit the model using INLA and take advantage of the benefits the increased computational efficiency compared to MCMC.
+However, we avoid the issues with the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal detection function, as will be more familiar to users of classical distance sampling methods.
+However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in the \texttt{R-INLA} implementation of INLA \citep{rue_approximate_2009}.
+We address this by using a method of iterated model fits based on a first-order Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019}.
 
-We apply these methods to point transect \akepa data which differ from line transect methods to account for the fact that the area surveyed increases with increasing distance from the transect.
+We fit the model to point transect \akepa{} data, which differ from line transect methods in that there is a need to account for the increasing area surveyed as distance increases from the transect.
 Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.

### 2021-09-02 paper.tex
edits up to 33

@@
 \newcommand{\T}{\intercal}
+\newcommand{\bs}{\mathbf{s}}
 \newcommand{\bS}{\mathbf{S}}
@@
 
-\begin{figure}
+\begin{figure}[!htb]
 	\centering
@@
 For the purposes of our analysis we select a single survey year from the \akepa{} time series (2002) that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability.
-In 2002, 289 points were sampled using point-transect distance sampling methods within the 4,603 ha study area of Hakalau (Fig.
-\ref{fig:2002studyareapointspt}) during which 276 \akepa{} were detected on 121 point transects.
+In 2002, there were a total of 289 point transects sampled within the 4,603 ha study area of Hakalau.
+(Fig.
+\ref{fig:2002studyareapointspt}) A total of 276 \akepa{} were detected on 121 point transects.
 The number of detections within each point transect ranged from zero to six.
@@
 However, our approach is readily extendible to a multi-year analysis, similar to that in \cite{camp_dsm_2020}, which we discuss in more detail in Section \ref{sec-discussion}.
+
+\vfill
 
@@
 
+\vfill
+
 \subsection{Model specification}
 
-In this section we introduce the statistical model for the spatial pattern of animal locations and the observation process, including a detailed description of how this model can be written as a modified Poisson likelihood.
-We use this formulation to fit the model using the software package \texttt{inlabru} \citep{bachl_inlabru_2019}, an extension to the \texttt{R-INLA} implementation of INLA, which implements the iterated INLA approach and is available through the Comprehensive R Archive Network \citep{r_2017}.
+This section describes the statistical model for the spatial pattern of observed animal locations.
+We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda$, i.e.
+the intensity at location $\bs$ is a random variable denoted $\lambda(\bs)$.
+The log-Gaussian Cox process is a flexible model that can include spatial covariates to model the mean intensity as well a mean-zero spatially structured random effect to account for unexplained heterogeneity not captured by the covariates \citep{moller_log_1998}.
+To account for the imperfect detection of points we specify a thinning probability function $g(\bs) = \mathbb{P}(\text{a point at $\bs$ is detected } |\text{ a point is at $\bs$})$.
+A key property of the log-Gaussian Cox process is that the point pattern of \textit{detected} points itself follows a log-Gaussian Cox process with intensity process $\tl(\bs) := \lambda(\bs)g(\bs)$.
+Using a log link this implies that $\log g(\bs)$ is included as an additive term to the model for $\log \lambda(\bs)$.
 
-We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda(.)$, i.e.
-the intensity at location $s$ is a random variable $\lambda(s)$.
-The log-Gaussian Cox process is a flexible approach that can include spatial covariates to model the mean intensity and a mean-zero spatially structured random effect to account for unexplained heterogeneity not captured by the covariates \citep{moller_log_1998}.
+Distance sampling models specify the thinning function $g(\bs)$ as a function that decays with increasing distance.
+The type of distance measured depends on the type of survey.
+For line transects the perpendicular distance to the transect line is used whereas for point transects it is the horizontal distance to the observer.
+For the remainder of the paper we assume a point transect survey design and hence horizontal distance to an observer located at the centre of each point transect.
 
-To account for the imperfect detection of points we specify a thinning probability function $g(s) = \mathbb{P}(\text{a point at $s$ is detected | a point is at $s$})$.
-A key property of the log-Gaussian Cox process is that a realisation of a point process with intensity process $\lambda(s)$ that is thinned by thinning probability function $g(s)$ also follows a log-Gaussian Cox process with intensity given by $\tl(s) = \lambda(s)g(s)$.
+The thinning probability function is specified as a parametric family of functions.
+For example, if $r(\bs)$ denotes the distance of a point at $\bs$ from an observer, the half-normal thinning probability function is $g(\bs | \sigma) = \exp(-r(\bs)^2 / 2\sigma^2)$, where $\sigma^2 > 0$ is a variance parameter to be estimated.
+Other types of detection function, such as hazard-rate, negative exponential, can also be used.
+Their parameters themselves can be modelled using appropriate link functions to incorporate covariates that may affect detectability.
+In addition to these parametric forms, additional adjustments have been proposed to increase the flexibility of such models by including trigonometric series expansions \citep{buckland_distance_2015}.
 
-Standard distance sampling approaches specify $g(s)$ as a function that decays with increasing distance.
-The type of distance measured depends on the type of survey.
-For line transects the perpendicular distance to the transect line is used.
-For point transects the horizontal distance to the observer is the relevant distance.
-For the remainder of the paper we assume a point transect survey design and hence horizontal distance to the observer located at the centre of the point transect.
-
-The thinning probability function is specified as a parametric family of functions with parameter vector.
-For example, if $r(s)$ denotes the distance of a point at $s$ from the observer, the half-normal thinning probability function is $g(s | \sigma) = \exp(-r(s)^2 / 2\sigma^2)$, where $\sigma^2 > 0$ is a variance parameter to be estimated.
-Other types of detection function exist, such as hazard-rate, negative exponential, and they may or may not be adjusted using series expansions to add some flexibility to obtain a better fit to a given dataset \citep{buckland_distance_2015}.
-In this analysis we assume a half-normal detection function with no covariates on detection parameters.
-
-The detection function parameters can only be estimated if an assumption is made about the intensity of the animal locations.
-Without such an assumption detectability and intensity are confounded.
-The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(s)$.
-Thus any observed deviations from uniformity can be attributed to detectability and not to variation in the intensity.
+In this analysis we assume a half-normal detection function with no covariates on detection parameters to demonstrate the approach in a simplified setting.
+The detection function parameters can only be estimated if an assumption is made about $\lambda(\bs)$ otherwise $g(\bs)$ and the intensity are confounded.
+The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(\bs)$.
+The usual approach for two stage models is to assume $\lambda(\bs)$ is constant within each spatial unit, however we are able to relax this assumption slightly to allow $\lambda(\bs)$ to be a linear function within in each transect.
+Given such an assumption, any observed deviations from uniformity can be attributed to detectability and not to variation in the intensity.
 
 A point transect distance sampling survey consists of a set of $K$ point transects.
-We denote $k$-th subset of space covered by the $k-th$ point transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.
-For simplicity, we assume that all point transects are non-overlapping discs with radius $W$, as is the case in the HFBS survey.
-The probability of observing a point at location $s \in \Omega_k$ given it is at $s \in \Omega_k$ we denote $g_k(s)$.
+We denote $k$th subset of space covered by the $k$th point transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.
+For simplicity, we assume that all point transects are non-overlapping discs with equal radius $W$, as is the case in the HFBS survey.
+It would be straightforward to relax this assumption.
+The probability of observing a point at location $\bs \in \Omega_k$ given it is at $\bs \in \Omega_k$ we denote $g_k(\bs)$.
 The probability of observing a point outside the surveyed region is zero.
-Since the point transects are non-overlapping, each location $s \in \Omega$ is unambigiously associated with a single thinning probability function $g_k$.
-For example, for an observer at location $s_k \in \Omega_k$, the centre of the disc, the half-normal thinning probability function is $g_k(s) = \exp(-\lVert s - s_k \rVert_2^2 / 2\sigma^2)$.
-In practice the observer is always located at the centre of the point transect.
-The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation but this is not a requirement here.
-The thinning probability function for any $s \in \Omega$ is then given by $g(s) = g_{k(s)}(s)$ where $k(s) = k$ for $s \in \Omega_k$.
+Since the point transects are non-overlapping, each location $\bs \in \Omega$ is unambiguously associated with a single thinning probability function $g_k$.
+For example, letting $\bs_k$ denote the centre of transect $\Omega_k$, the half-normal thinning probability function for points in $\Omega_k$ is $g_k(\bs) = \exp(-\lVert \bs - \bs_k \rVert_2^2 / 2\sigma^2)$.
+The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation.
+The thinning probability function for any $\bs \in \Omega$ is then given by $g(\bs) = g_{k(\bs)}(\bs)$ where $k(\bs)$ is an indexing function such that $k(\bs) = k$ for $\bs \in \Omega_k$.
 
-The thinned log-Gaussian Cox process likelihood with observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is then
+The thinned log-Gaussian Cox process likelihood for observed points at locations $\bm{Y} = (s_1, \ldots, s_n)^\intercal$ is
 \begin{equation}
 \label{lgcp-likelihood}
-\pi(\bm{Y}) = \exp\left( |\Omega| -\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s \right)\prod_{i=1}^n \lambda(s_i)g(s_i)
+\pi(\bm{Y}) = \exp\left( |\Omega| -\int\displaylimits_{\bs \in \Omega} \lambda(\bs) g(\bs) \mathrm{d}\bs \right)\prod_{i=1}^n \lambda(\bs_i)g(\bs_i)
 \end{equation}
-where $\lambda(.)$ and $g(.)$ both depend on parameter vectors omitted for readability and $|\Omega|$ is the area of the total surveyed region.
+where $|\Omega|$ is the area of the total surveyed region and $\lambda(\bs)$ and $g(\bs)$ both depend on parameter vectors omitted for readability.
 
@@
 The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\begin{center}
@@
 [[Say something about these - hard to know what to say about them?]]   (\autoref{fig:intensity-quantiles}).
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\begin{center}
@@
 The posterior detection function \autoref{fig:det-matern} shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres.
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\begin{center}
@@
 To further investigate the ability of the GMRF component to capture the heterogeneity we generated 100 thinned point patterns from the fitted model and calculated the pairwise distances between all observations for each of these patterns \autoref{fig:post-pp-distances}.
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\begin{center}
@@
 We plot the posterior abundance estimate using a Monte Carlo method.
-\begin{figure}[h]
+\begin{figure}[!htb]
 	\begin{center}
@@
 The mapped posterior mean is the most common map one encounters in the literature.  However, the mean will always be smoother than realisations from the posterior random field.
-\begin{figure}
+\begin{figure}[!htb]
 	\includegraphics[scale=0.525]{figures/intensity_realized.png}
@@
 This figure can be interpreted in a natural way as the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
-\begin{figure}
+\begin{figure}[!htb]
 	\includegraphics[scale=0.5]{figures/excursions.png}

### 2021-09-03 paper.tex
edits up to 3.4

@@
 A key property of the log-Gaussian Cox process is that the point pattern of \textit{detected} points itself follows a log-Gaussian Cox process with intensity process $\tl(\bs) := \lambda(\bs)g(\bs)$.
-Using a log link this implies that $\log g(\bs)$ is included as an additive term to the model for $\log \lambda(\bs)$.
+Using a log link $\log g(\bs)$ is therefore included in the model as an additive component to the model for $\log \lambda(\bs)$.
 
@@
 We denote $k$th subset of space covered by the $k$th point transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.
-For simplicity, we assume that all point transects are non-overlapping discs with equal radius $W$, as is the case in the HFBS survey.
+For simplicity, we assume that all point transects are non-overlapping disks with equal radius $W$, as is the case in the HFBS survey.
 It would be straightforward to relax this assumption.
@@
 
-\sloppy Here we show how to approximate the thinned log Gaussian Cox process model with a modified Poisson likelihood.
-The integral component $\int\displaylimits_{s \in \Omega} \lambda(s) g(s) \mathrm{d}s$ in \autoref{lgcp-likelihood} does not have analytical solutions.
-Replacing the integral with a weighted sum allows the log-likelihood to be approximated as a weighted Poisson log-likelihood, which we describe below, adapting the approach developed by \cite{simpson_going_2016} for a fully observed point pattern.
+\sloppy The likelihood \eqref{lgcp-likelihood} contains an integral that cannot be solved analytically.
+A common approach for fitting log-Gaussian Cox processes in practice is to approximate the likelihood by replacing the integral with a weighted sum.
+In this section we describe this approximate likelihood and show how it can be written as a modified Poisson likelihood.
+We follow closely with the approach given in \cite{simpson_going_2016} for a fully observed log-Gaussian Cox process but adapted for the point transect distance sampling case.
 
-To evaluate the integral we use polar coordinates notation $s_k(r, \theta) = s_k + r\left[\cos\theta, \sin\theta \right]^T$ to represent locations in each sampling unit $\Omega_k$.
-In general, the thinning function depends only on $r$ and not on $\theta$.
-We assume this is the case in the following and use the shorthand $g_k(r) = g(s_k(r, \theta))$.
-The integral in the likelihood can be simplified using an assumption that $\lambda(s)$ is linear within each sampling unit.
-This assumption implies that $\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) = 2\lambda(s_k)$ and so
-\begin{align*}
-	\int_0^{2\pi} \lambda(s_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \left[\lambda(s_k(r, \theta)) + \lambda(s_k(r, \theta + \pi)) \right] g_k(r)\mathrm{d}\theta \\
-	&= 2\pi \lambda(s_k)g_k(r)
-\end{align*}
+To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.
+In a slight abuse of notation $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes any location within the transect.
+It should be clear from context which is intended by $\bs_k$.
+The thinning function $g_k(\bs_k(r, \theta))$ depends only on $r$ and not on $\theta$ and so we use the shorthand notation $ g(\bs_k(r, \theta)) = g_k(r)$.
+
+The integral in the likelihood can be simplified using an assumption that $\lambda(\bs)$ is linear within each sampling unit.
+This implies that $\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi)) = 2\lambda(\bs_k)$.
+Therefore
+\begin{align}
+\label{eq-linear-intensity}
+	\int_0^{2\pi} \lambda(\bs_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \{\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi) \} g_k(r)\mathrm{d}\theta \nonumber \\
+	&= 2\pi \lambda(\bs_k)g_k(r)
+\end{align}
 We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.
-A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piece-wise linear basis functions we use to model $\lambda$ (see section 5).
-Integrating with respect to $r$ and accounting for the change of variables gives $\int \lambda(s)g(s) \mathrm{d}s = \sum_{k=1}^K 2\pi \lambda(s_k) \int_0^W r g_k(r)\mathrm{d}r$.
-For each sampling unit we approximate the one-dimensional integral with respect to $r$ using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
+A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piecewise linear basis functions used for the GMRF spatial effect (Section \ref{sec-gmrf}).
+
+It follows that $\int \lambda(\bs)g(\bs) \mathrm{d}\bs = \sum_{k=1}^K 2\pi \lambda(\bs_k) \int_0^W r g_k(r)\mathrm{d}r$ by applying the change of variables formula for integration and substituting in \eqref{eq-linear-intensity}.
+Note that this integral now only involves the one dimensional integral with respect to $r$.
+For each sampling unit we approximate this integral using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
 This gives
 \begin{equation*}
-	\int\displaylimits_{s \in \Omega} \lambda(s)g(s)\mathrm{d}s \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(s_{kj}) \;\;\; ,
+	\int\displaylimits_{\bs \in \Omega} \lambda(\bs)g(\bs)\mathrm{d}\bs \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(\bs_{kj}) \;\;\; ,
 \end{equation*}
-where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(s_{kj}) = \lambda(s_k) g_k(r_{kj})$.
+where $\tilde{\alpha}_{kj} = 2\pi \alpha_{kj}r_{kj}$ and $\tl(\bs_{kj}) = \lambda(\bs_k) g_k(r_{kj})$.
 
-To simplify notation below we let $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.
-Similarly, let $\tl_k = (\tl(s_{k1}), \ldots, \tl(s_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$ and denote the intensity evaluated at the observed locations as $\tl_{obs} = (\tl(s_1), \ldots, \tl(s_n))^\intercal$.
-Then the approximate log-likelihood is
+To simplify notation below we gather all integration weights into the vectors $\tilde{\alpha}_{k} = (\alpha_{k1}, \ldots, \alpha_{kM})^\intercal$ and $\tilde{\alpha} = (\alpha_1^\intercal, \ldots, \alpha_K^\intercal)^\intercal$.
+Similarly, let $\tl_k = (\tl(\bs_{k1}), \ldots, \tl(\bs_{kM}))^\intercal$, $\tl_{int} = (\tl_1^\intercal, \ldots, \tl_K^\intercal)^\intercal$, the thinned intensity evaluated at each integration location, and $\tl_{obs} = (\tl(\bs_1), \ldots, \tl(\bs_n))^\intercal$, the thinned intensity evaluated at each observation location.
+Then the approximate log-likelihood can be written as
 \begin{equation}
@@
 \end{equation}
-This approximation can be expressed as a modified Poisson likelihood.  To see this let $\eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
-$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and construct a vector of pseudo-observations $z = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.
-Then the approximate likelihood becomes
+This approximate likelihood can be expressed as a modified Poisson likelihood by letting $\eta = (\tl_{int}^\intercal, \tl_{obs}^\intercal)^\intercal$,
+$\alpha = (\tilde{\alpha}^\intercal, 0_{n \times 1}^\intercal)^\intercal$ and constructing a vector of pseudo-observations $z = (0_{KM\times 1}^\intercal, 1_{n \times 1}^\intercal)^\intercal$.
+Then the approximate likelihood is
 \begin{equation}
@@
 \end{equation}
-where $C$ is a constant.
-This approximate likelihood is implemented in \texttt{inlabru} as a \texttt{"cp"} likelihood.
-This is similar to the Berman-Turner device \citep{berman_approximating_1992, baddeley_practical_2000} although without the assumption that the locations of observed points form a part of the quadrature scheme.
+where $C$ is a constant.  This approximate likelihood is implemented in \texttt{inlabru} as a \texttt{"cp"} likelihood.
+
+The approach taken here is similar to the Berman-Turner device \citep{berman_approximating_1992, baddeley_practical_2000} although without the assumption that the locations of observed points form a part of the quadrature scheme.
 

### 2021-09-03 paper.tex
edits up to section 5

@@
 
-\begin{enumerate}[(i)]
+\begin{enumerate}[(a)]
 	\item A spatial point process perspective on point transect distance sampling and representing the detection model as a thinning of a point processes.
@@
 \label{lgcp-likelihood}
-\pi(\bm{Y}) = \exp\left( |\Omega| -\int\displaylimits_{\bs \in \Omega} \lambda(\bs) g(\bs) \mathrm{d}\bs \right)\prod_{i=1}^n \lambda(\bs_i)g(\bs_i)
+\pi(\bm{Y}) = \exp\left( |\Omega| -\int\displaylimits_{\bs \in \Omega} \lambda(\bs) g(\bs) \mathrm{d}\bs \right)\prod_{i=1}^n \lambda(\bs_i)g(\bs_i),
 \end{equation}
@@
 	\int_0^{2\pi} \lambda(\bs_k(r, \theta))g_k(r)\mathrm{d}\theta &= \int_0^\pi \{\lambda(\bs_k(r, \theta)) + \lambda(\bs_k(r, \theta + \pi) \} g_k(r)\mathrm{d}\theta \nonumber \\
-	&= 2\pi \lambda(\bs_k)g_k(r)
+	&= 2\pi \lambda(\bs_k)g_k(r).
 \end{align}
@@
 \begin{equation*}
-	\int\displaylimits_{\bs \in \Omega} \lambda(\bs)g(\bs)\mathrm{d}\bs \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(\bs_{kj}) \;\;\; ,
+	\int\displaylimits_{\bs \in \Omega} \lambda(\bs)g(\bs)\mathrm{d}\bs \approx \sum_{k=1}^K \sum_{j=1}^M \tilde{\alpha}_{kj} \tl(\bs_{kj}) ,
 \end{equation*}
@@
 \label{approx-log-likelihood}
-	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}
+	\log \pi(\bm{Y}) \approx - \tilde{\alpha}^\intercal \tl_{int} + 1^\intercal\log\tl_{obs}.
 \end{equation}
@@
 \begin{equation}
-\pi(\bm{Y}) \approx C \prod\limits_{i=1}^{KM + n} \eta_i^{z_i}\exp(-\alpha_i\eta_i)
+\pi(\bm{Y}) \approx C \prod\limits_{i=1}^{KM + n} \eta_i^{z_i}\exp(-\alpha_i\eta_i),
 \end{equation}
@@
 In the above, we assume the data are complete records of animal locations.
+That is, each observation has a known position $\bs_{k(\bs_i)} = \bs_{k(\bs_i)}(r_i, \theta_i)$.
 However, in many distance sampling surveys only the location of the observer and the distance to the observer are recorded.
-Data of this type can be analysed within a point process framework by deriving the appropriate intensity function for the incomplete data as outlined below.
-
-Using the polar coordinates notation given above, for a detected point at location $s_k(r, \theta) \in \Omega_k$ we consider the case where we observe $r$ but not $\theta$.
+Data of this type can be analysed within a point process framework by deriving the appropriate intensity function.
+We consider the case where $r$ is recorded but not $\theta$.
 It follows that the intensity for points at a distance $r$ observed within sampling unit $\Omega_k$ is
@@
 \label{intensity-incomplete}
-\tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(s)g_k(r)\mathrm{d}s \nonumber \\
-&= 2\pi\lambda(s_k)rg_k(r) \;\;\; ,
+\tl_k(r) &= \oint\displaylimits_{c_k(r)} \lambda(\bs)g_k(r)\mathrm{d}\bs \nonumber \\
+&= 2\pi\lambda(\bs_k)rg_k(r),
 \end{align}
- where $c_k(r)$ is a circle of radius $r$ centred at $s_k$ and the second line follows from changing to polar coordinates and the assumption of linear intensity within $\Omega_k$.
+where $c_k(r)$ is a circle of radius $r$ centred at $\bs_k$ and the second line follows from change of variables and the assumption of linear intensity within $\Omega_k$.
 This intensity differs from the full data case through the $2\pi$  term that accounts for the fact that we do not observe $\theta$ and the additional $r$ term that accounts for the increasing area surveyed at larger distances.
-The log-intensity for observed points (the thinned intensity) is thus $\log\tl(s) = \log 2 \pi r + \log\lambda(s) + \log g(s)$ and so can be implemented during fitting by using an offset.
-As we noted above, the detection function $g(s)$ is non-linear in its parameters and this component of the additive predictor requires a novel approach to parameter estimation.
+We note that a similar equation is derived from the perspective of a triangular distribution for point transect distance sampling in \cite{buckland_advanced_2004}.
+
+The log-intensity for observed points (the thinned intensity) is $\log\tl(\bs) = \log 2 \pi r + \log\lambda(\bs) + \log g(\bs)$.
+As we noted above, the detection function $g(\bs)$ is non-linear in its parameters and this component of the additive predictor requires a novel approach to parameter estimation.
 
@@
 The half-normal detection function depends on the strictly positive parameter $\sigma^2$.
-We choose a log link and consider $g(s | \phi)$, where $\log\sigma^2 = \phi$.
-However, this model cannot be directly fitted using INLA since $\log g(s | \phi)$ is not linear in its parameters.
-To address this we use an approximate method based on a first-order Taylor approximation of the non-linear predictor.
+Since it is not possible to restrict the range of parameters to be strictly positive within \texttt{R-INLA} this model cannot be fitted using the standard INLA methodology.
+Instead, we choose a log link and consider $g(\bs | \phi)$, where $\log\sigma^2 = \phi$.
+This is a non-linear function of $\phi$ which cannot be fitted using standard GLM methods.
+To address this we use a novel approximate method based on a first-order Taylor approximation of the non-linear predictor.
+
 Let $\eta(\bm{u}) = \log \tilde{\lambda}(\bm{u})$ where $\bm{u}$ is the parameter vector of the predictor (which includes $\phi$ as well as any parameters for fixed and random effects).
@@
 \begin{equation*}
-\bar{\eta}_{\bm{u}_0} (\bm{u}) = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0)
+\bar{\eta}_{\bm{u}_0} (\bm{u}) = \eta(\bm{u_0}) + \bm{B}_{\bm{u}_0}(\bm{u} - \bm{u}_0),
 \end{equation*}
@@
 Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the linear Taylor approximation $\bar{\eta}_{\bm{u_0}}$.
+
 The question then is how to choose the linearisation point $\bm{u}_0$.
-The approach taken is via a fixed point iteration scheme.
-There are many possible choices here and we describe the approach in the current implementation of \texttt{inlabru} \citep{bachl_inlabru_2019}.
-However, more options are likely to be added in the future.
-We denote by $\bar{p}_{\bm{v}}$ the posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$ and define the functional
+To find an `optimal' linearisation point we use a fixed point iteration scheme.
+There are various possible choices to define an `optimal' choice and this affects the properties of the approximate inference method.
+Here we describe the criteria used when fitting the model that produces the results presented in this paper.
+However, more recently, other options have been added to \texttt{inlabru}, the software package that implements these methods.
+
+We denote by $\bar{p}_{\bm{v}}$ the posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.
+New linearisation points are found by considering the functional
 \begin{equation}
@@
 \end{equation}
-where $n_{\bm{u}}$ is the length of $\bm{u}$.  Each $\bar{p}_{\bm{v}} (u_i | \bm{Y})$ is approximated and the mode estimated during the model fitting using the standard INLA method.
+where $n_{\bm{u}}$ is the length of $\bm{u}$.
+Each marginal posterior $\bar{p}_{\bm{v}} (u_i | \bm{Y})$ and their modes are estimated during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost for each fit of the linearised model.
 We then seek a fixed point of this functional so that $\bm{u}_0 = f(\bar{p}_{\bm{u}_0})$.
@@
 Given a current choice of linearisation point $\bm{v}$, estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.
-Iterate this procedure until a fixed point is identified within a chosen tolerance.
+We iterate this procedure until a fixed point is identified within a chosen tolerance.
 
@@
 This means that the new linearisation point remains close to a feasible region for fitting, assuming the current point $\bm{v}$ is reasonable.
+In practice, we have found that this requires some care when choosing starting values for the initial model fit.
 We estimate $\alpha$ using an approximate line search method that avoids many potentially expensive evaluations of the non-linear predictor.
-This iterative method is implemented in the R package \texttt{inlabru} \citep{bachl_inlabru_2019}.
+This iterative method is implemented in the R package \texttt{inlabru} \citep{bachl_inlabru_2019} which is both a wrapper and an extension to \texttt{R-INLA}.
 

### 2021-09-06 paper.tex
edits up to section 6

@@
 
-The log-intensity for observed points (the thinned intensity) is $\log\tl(\bs) = \log 2 \pi r + \log\lambda(\bs) + \log g(\bs)$.
+The log-intensity for detected points is therefore $\log\tl(\bs) = \log 2 \pi r + \log\lambda(\bs) + \log g(\bs)$ and so it is straightforward to account for incomplete location information by including an offset term.
 As we noted above, the detection function $g(\bs)$ is non-linear in its parameters and this component of the additive predictor requires a novel approach to parameter estimation.
@@
 
-Using spatially-structured random effects can be computationally expensive, often resulting in dense prior precision matrices which typically have $\mathcal{O}(n^3)$ cost to factorise for an effect with $n$ parameters.
-A key development in this area is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs) and the \texttt{R-INLA} software that makes use of the sparsity to factorise the precision matrix at a cost of $\mathcal{O}(n^{3/2})$.
-This representation is constructed via a stochastic partial differential equation solved using finite element methods \citep{lindgren_explicit_2011}.
+This section describes the spatial model for the intensity process.  
+As mentioned above, the \akepa{} population exhibits spatially heterogeneity for which there is no known explanatory covariate.
+Therefore we use a spatially structured random effect to explain this variation.
+
+Using spatially-structured random effects can be computationally expensive.
+Often this is due to parameterisations and approximations that lead to dense prior precision matrices which have an  $\mathcal{O}(k^3)$ cost to factorise for an effect with $k$ basis parameters.
+A key development in the development of computationally efficient methods is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs).
+These result in a sparse precision matrix which can be factorised with a cost of $\mathcal{O}(k^{3/2})$.
+We use a GMRF representation based on a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.
+This representation is constructed by applying finite element methods to solve the SPDE numerically.
 
@@
 \begin{equation*}
-\log \lambda(s) = \beta_0 + \sum_v \beta_v z_v(s) + \xi(s)
+\log \lambda(s) = \beta_0 + \xi(\bs)
 \end{equation*}
-where $\beta_0$ is an intercept parameter, the $z_v(s)$ are spatial covariates with effect parameters $\beta_v$ and $\xi(s)$ is a zero-mean GRF with Mat\'ern covariance
+where $\beta_0$ is an intercept parameter and $\xi(\bs)$ is a mean zero GRF with Mat\'ern covariance.  The Mat\'ern covariance function in two dimensions can be written as 
 \begin{equation}
-C(s_1,s_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|s_1-s_2\|)^{\nu}K_\nu(\kappa \|s_1-s_2\|)
+C(\bs_1,\bs_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|\bs_1-\bs_2\|)^{\nu}K_\nu(\kappa \|\bs_1-\bs_2\|)
 \end{equation}
 where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.
-All three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which specifies the mean-square differentiability of the process.
+The three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which specifies the mean-square differentiability of the process.
 We set $\nu = 1$ which is the default value in \texttt{R-INLA}.
@@
 We use a GMRF representation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.
-Given a finite element mesh on the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$ we represent the GRF as $\xi(s) = \sum_l \xi_l \phi_l(s)$.
-The parameters $\xi_1, \ldots, \xi_L$ form a GMRF with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices the elements of which are defined via an approximate solution to a SPDE (see Supplementary material).
-The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function.
-In order to specify priors on the Mat\'ern covariance below we use a reparameterisation to a range and variance parameter.
-When $\nu = 1$ and the domain is of dimension two the reparameterisation is $\rho = \sqrt{8} / \kappa$ and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ \citep{blangiardo_spatial_2013}.
+Given a finite element mesh defined over the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$, we represent the GRF as $\xi(\bs) = \sum_l \beta_l \phi_l(\bs)$.
+The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices (see Appendix [BLAH]).
+The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function as distance between locations increases.
+
+In order to specify priors on the Mat\'ern covariance below we use a reparameterisation of $\kappa$ and $\tau$ to a range and variance parameter.
+When $\nu = 1$ and the domain is two dimensional then the reparameterisation is $\rho = \sqrt{8} / \kappa$ for the range and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ for the marginal variance (\cite{blangiardo_spatial_2013}, p196).
 
 The SPDE approach can at first appear to be fundamentally different to other spatial random effect approaches such as penalised regression splines.
-However, the two methods are closely related and the GRF model component in our model plays a similar role to a tensor product spline with an appropriate smoothing penalty.
-See \cite{yue_bayesian_2014, miller_understanding_2019} for more details on the similarities and differences between penalised regression splines and the SPDE approach.
+However, the two methods are closely related and the GRF model component in our model plays a similar role to a penalised smoothing spline on space.
+\cite{yue_bayesian_2014} show how spline models can be viewed through the perspective of SPDEs and \cite{ miller_understanding_2019} do the reverse and show how to implement the SPDE as a penalised spline model in \texttt{mgcv} \citep{wood_gam_2017}.
 

### 2021-09-06 paper.tex
really up to 6 this time

@@
 
-The SPDE approach can at first appear to be fundamentally different to other spatial random effect approaches such as penalised regression splines.
+This parameterisation is used to set penalised complexity priors \citep{simpson_penalising_2017} on $\rho$ and $\sigma^2$ which allow the GRMF effect to shrink towards zero.
+For the Mat\'ern field, the base model is the limiting case with $\sigma^2 = 0$ and $\rho \rightarrow \infty$, which defines a random field that is almost surely zero everywhere.
+The purpose of penalised complexity priors is to place sufficient prior probability on a simple model to allow the random effect to be `turned off' if the model component is not required.
+
+To specify the penalised complexity priors on the SPDE effect we set $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$.
+The value of 130 for the range parameter was selected based on the minimum distance between sampling locations.
+Since only the distance to the observer was recorded, and not the exact location of \akepa{}, we used the adjusted intensity given in equation \eqref{intensity-incomplete}.
+
+The SPDE approach can at first appear to be fundamentally different  other approaches such as penalised regression splines.
 However, the two methods are closely related and the GRF model component in our model plays a similar role to a penalised smoothing spline on space.
-\cite{yue_bayesian_2014} show how spline models can be viewed through the perspective of SPDEs and \cite{ miller_understanding_2019} do the reverse and show how to implement the SPDE as a penalised spline model in \texttt{mgcv} \citep{wood_gam_2017}.
+\cite{yue_bayesian_2014} show how spline models can be viewed through the perspective of SPDEs and \cite{ miller_understanding_2019} do the reverse to show how to implement the Mat\'ern SPDE as a penalised spline model in \texttt{mgcv} \citep{wood_gam_2017}.
 
@@
 \label{sec-results}
-
-We fitted the models using penalised complexity priors for the hyperparameters of the GMRF \citep{simpson_penalising_2017}, setting $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$, where $\rho$ is the range parameter of the Mat\'ern covariance.
-The value of 130 was selected based on the minimum distance between sampling locations.
-The complete code for the analysis can be found in the Supplementary materials.
-Since only distance to observer was recorded we used the adjusted intensity given in \eqref{intensity-incomplete}.
 
@@
 The standard deviation map shows stronger overall posterior variability in the south where intensity is larger.
-We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location.
-[[Say something about these - hard to know what to say about them?]]   (\autoref{fig:intensity-quantiles}).
+We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location (\autoref{fig:intensity-quantiles}).
 \begin{figure}[!htb]

### 2021-09-07 paper.tex
edits up to 6.1

@@
 
-The log-intensity of animal locations is given by
+Since only the distance to the observer was recorded, and not the exact location of \akepa{}, we used the adjusted intensity given in equation \eqref{intensity-incomplete}.
+The log-intensity of animal locations, detected or undetected, is given by
 \begin{equation*}
-\log \lambda(s) = \beta_0 + \xi(\bs)
+\log \lambda(\bs) = \beta_0 + \xi(\bs)
 \end{equation*}
@@
 We use a GMRF representation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.
-Given a finite element mesh defined over the spatial domain with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$, we represent the GRF as $\xi(\bs) = \sum_l \beta_l \phi_l(\bs)$.
-The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q}_{\xi} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices (see Appendix [BLAH]).
+We define a finite element mesh with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$.
+The GRF is represented as $\xi(\bs) = \sum_l \beta_l \phi_l(\bs)$.
+The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices (see Appendix [BLAH]).
 The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function as distance between locations increases.
+We used a finite element mesh with 3748 nodes defined over the entire study region and extended over a buffer region to avoid the effects of boundary conditions within the region of inference.
 
@@
 When $\nu = 1$ and the domain is two dimensional then the reparameterisation is $\rho = \sqrt{8} / \kappa$ for the range and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ for the marginal variance (\cite{blangiardo_spatial_2013}, p196).
+This parameterisation is used to set penalised complexity priors \citep{simpson_penalising_2017} on $\rho$ and $\sigma^2$ which allow the GRMF effect to shrink towards zero.
+For the Mat\'ern field, the base model is the limiting case with $\sigma^2 \rightarrow 0$ and $\rho \rightarrow \infty$, which defines a random field that is almost surely zero everywhere.
+The purpose of penalised complexity priors is to place sufficient prior probability on a simple model to allow the random effect to be essentially `turned off' if the model component is not supported by the data.
 
-This parameterisation is used to set penalised complexity priors \citep{simpson_penalising_2017} on $\rho$ and $\sigma^2$ which allow the GRMF effect to shrink towards zero.
-For the Mat\'ern field, the base model is the limiting case with $\sigma^2 = 0$ and $\rho \rightarrow \infty$, which defines a random field that is almost surely zero everywhere.
-The purpose of penalised complexity priors is to place sufficient prior probability on a simple model to allow the random effect to be `turned off' if the model component is not required.
-
-To specify the penalised complexity priors on the SPDE effect we set $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$.
+To specify penalised complexity priors on the SPDE effect we set $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$.
 The value of 130 for the range parameter was selected based on the minimum distance between sampling locations.
-Since only the distance to the observer was recorded, and not the exact location of \akepa{}, we used the adjusted intensity given in equation \eqref{intensity-incomplete}.
 
@@
 
-The predicted mean of the posterior intensity is shown in \autoref{fig:intensity-mean-cv} along with a map of the coefficient of variation (CV) of the posterior intensity field on a regular prediction grid with cell area of approximately 1.7 hectares.
+The model fits 40 to 60 seconds on a laptop with 10 Gb RAM and a 2.50 GHz CPU when INLA is run using the `empirical Bayes' option which does not incorporate uncertainty in hyper parameters.
+The model fits in minutes in full INLA mode including hyper parameter uncertainty.
+We suggest using the `empirical Bayes' mode for model development and full INLA on a smaller subset of candidate models.
+
+The predicted mean of the posterior intensity field is shown in \autoref{fig:intensity-mean-cv}A along with a map of the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid with each prediction cell area approximately 1.7 hectares.
 \begin{figure}[!htb]
@@
 \end{figure}
-This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data (see Supplementary material).
-The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort.
-There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north  (see \autoref{fig:2002studyareapointspt}).
-In the north, estimated intensities are lower, which also contributes to larger CV values.
-The standard deviation map shows stronger overall posterior variability in the south where intensity is larger.
-We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location (\autoref{fig:intensity-quantiles}).
+This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data \citep{camp_dsm_2020}.
+The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort and higher intensity.
+There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north.
+In the north, estimated intensities are lower, which also contributes to larger CV values and makes the CV map hard to interpret.
+The standard deviation map shows stronger overall posterior variability in the south where intensity is larger, most likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.
+
+We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location (Figure \ref{fig:intensity-quantiles}).
+As we discuss below, these quantiles are independent summary statistics of the posterior intensity field at each prediction location.
 \begin{figure}[!htb]
@@
 
-The posterior detection function \autoref{fig:det-matern} shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres.
+The posterior detection function (\autoref{fig:det-matern}) shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres and also broadly agrees with a two-stage analysis.
 \begin{figure}[!htb]
@@
 		\includegraphics[scale=0.525]{figures/detfn_and_matern.png}
-		\caption{A) posterior half-normal detection function.
-B) posterior Mat\'ern correlation function.
+		\caption{A) Posterior half-normal detection function.
+B) Posterior Mat\'ern correlation function.
 The black line is the mean posterior estimate and the grey shaded area is 95\% credible interval in both plots}
@@
 \end{figure}
-The posterior Mat\'ern correlation function is shown in \autoref{fig:det-matern}, the grey regions show the 95\% credible interval.
-Note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).
-This is the correlation range on the log-intensity scale.
-[[Can I make a better "range of correlation" plot for the posterior intensity process?
-How can I convert this to something on the intensity scale?
-For a field $Z$ with covariance function $C$, is there an implied covariance for $\exp(Z)$?]]
+The posterior Mat\'ern correlation function is shown in \autoref{fig:det-matern}; the grey regions show the 95\% credible interval.
+We note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).
+This is because this is the correlation range on the log-intensity scale.
+On the intensity scale the empirical range of correlation is shorter.
 
-To further investigate the ability of the GMRF component to capture the heterogeneity we generated 100 thinned point patterns from the fitted model and calculated the pairwise distances between all observations for each of these patterns \autoref{fig:post-pp-distances}.
+To evaluate the ability of the SPDE effect to capture the spatial heterogeneity in the observed data we generated 1,000 point patterns using the posterior intensity field and thinned them with the posterior detection function to create 1,000 simulated datasets to compare with the observed data.
+We calculated the pairwise distances between all observations for each of these datasets and compared the frequency of pairwise distances with those in the observed data.
 \begin{figure}[!htb]
@@
 		\includegraphics[scale=0.6]{figures/post_pp_distances.png}
-		\caption{Boxplots of frequency of pairwise distances (tick mark values for end of each bin) based on 100 posterior samples.
-Red circles: observed frequency of pairwise distances between observations}
+		\caption{Boxplots of the frequency of pairwise distances (tick mark values for end of each bin) based on 1,000 simulated datasets.
+Red circles indicate the frequency of pairwise distances in the observed data.}
 		\label{fig:post-pp-distances}
@@
 \end{figure}
-The plot shows that the GMRF component broadly captures the relevant scale of clustering in the observed data.
-There is some evidence of over-predicting the strength of the clustering at distances between 1,000 and 2,500 m.
+Figure \ref{fig:post-pp-distances} shows that the SPDE effect broadly captures the spatial clustering in the observed data.
+There is some evidence of under-predicting the strength of the clustering at shorter distances between 1,000 and 2,500 m, with the simulated data showing some evidence of negative bias.
+This could be due to over-smoothing in some areas which can happen if using a stationary model.
+Large areas of the study region having low density, which has a larger correlation range than higher density areas, and so the effect will average between these.
 
-We plot the posterior abundance estimate using a Monte Carlo method.
+Figure \ref{fig:post-pp-distances} is similar in spirit to Ripley's $K$-function \citep{ripley_SecondorderAnalysisStationary_1976}, which is a common approach to investigating spatial structure in point patterns.
+The pairwise distances in the observed data is a function of spatial heterogeneity as well as the spatially restricted sampling scheme.
+Other methods of comparison could also be considered such as the average difference between predicted and observed counts at each point transect.
+Working with posterior predictive datasets is a flexible approach to model evaluation that allows for innovative methods of comparison with observed data that can focus on particular features of the model.
+
+To plot the posterior abundance estimate for the \akepa{} in Hakalau reserve we use a Monte Carlo method based on sampling realisations of the posterior intensity field.
+Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(\bs)\mathrm{d}\bs$ its corresponding random variable within our Bayesian framework.
+Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e.
+the expected abundance.
+To generate a full posterior estimate for abundance we approximate the posterior $\pi(N | \bm{Y})$ by taking $m$ Monte Carlo samples of the posterior intensity field, $\lambda^{(1)}, \ldots, \lambda^{(m)}$ and approximate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)}, \bm{Y})$.
+Each $\pi(N | \lambda = \lambda^{(i)}, \bm{Y})$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda^{(i)}(\bs)\mathrm{d}\bs$.
 \begin{figure}[!htb]
@@
 \end{figure}
-Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(s)\mathrm{d}s$ its corresponding random variable within our Bayesian framework.
-Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e.
-the expected abundance.
-We approximate the posterior $\pi(N | \bm{Y})$ by a Monte Carlo method.
-Taking $m$ Monte Carlo samples  $\lambda^{(1)}, \ldots, \lambda^{(m)}$ of the posterior intensity field we estimate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)})$.
-Each $\pi(N | \lambda = \lambda_i)$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda_i(s)\mathrm{d}s$.
+
 \autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
-This allows us to estimate the probability for any specific value of realised abundance $n$ by estimated $\pi(N = n | \lambda)$.
-Strictly speaking
+This allows us to estimate the probability for any specific value of realised abundance $n$ which may be of interest to conservation managers.
+This approach can also be used to generate credible intervals and other posterior summaries of interest.
 
-% Add table of estimated model parameters
-% Add 95% CI figures for N?
+The abundance estimates broadly agree with a standard two-stage analysis of the data \citep{camp_dsm_2020} but the attractive feature of the one-stage model is that, by using Monte Carlo samples from the joint posterior, model summaries naturally incorporate uncertainty in the detection model.
+There is no additional step required to incorporate detection function uncertainty beyond working with samples from the joint posterior.
 
-\clearpage %sort out figure placements later
-
-\section{Communication of Results}
-\label{sec-comms}
-
-The results above are consistent with those resulting from a two-stage model fitted to binned count data (see Supplementary material).
-Our presentation of the results is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (\autoref{fig:intensity-mean-cv} and \autoref{fig:intensity-quantiles}) are not always provided there and it is common to see maps of point estimates without any accompanying communication of uncertainty.
-It is also common to report a point estimate of the expected abundance along with uncertainty in this point estimate.
-However, uncertainty in a point estimate will tend to have lower variance than variance of the random variable of interest.
-For this reason we chose to show the full abundance posterior in \autoref{fig:realized-abundance-posterior}.
-
-Our intention in this section is to highlight limitations with mapped summaries of model predictions, including those presented in Section \ref{results}, and suggest ways to address these limitations.
-We note that all model outputs are based on sampling from the joint posterior of all model parameters which naturally average over the uncertainty in the observation process, a key advantage of the one-stage approach.
+\vfill
 

### 2021-09-07 paper.tex
more edits

@@
 
-The most common method to communicate uncertainty in a spatial estimate tend is to produce maps of predictive uncertainty such as standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014} based either on a posterior predictive distribution or bootstrapping.
-Another approach is to map quantiles or the probability of exceeding certain thresholds \citep{russell_avoidance_2016, wilson_hierarchical_2010}.
-In Section \ref{results} we produce maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of a posterior spatially structured random field.
-Whilst maps that summarise spatial predictions are useful, they all mask certain properties we may care about when communicating the results of an analysis.
-For example, maps of the posterior mean will always be smoother than realisations of the posterior random field itself.
-This can obscure the spatial structure we would expect if we observed a realisation of process itself.
-Maps of the posterior coefficient of variation are hard to interpret when estimated densities are close to zero and posterior quantiles are tempting to interpret jointly when they should be viewed independently.
-This challenges mean careful consideration must be given to the communication.
+Our presentation of the results above is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (\autoref{fig:intensity-mean-cv} and \autoref{fig:intensity-quantiles}) are not always provided.
+It is common to see maps of point estimates (e.g.
+a map of expected density) without any accompanying communication of uncertainty.
+It is also common to report a point estimate of the expected abundance along with uncertainty in the point estimate.
+However, uncertainty in a point estimate will have lower variance than the variance of the posterior random variable.
+For this reason we chose to present the full abundance posterior in \autoref{fig:realized-abundance-posterior} as opposed to uncertainty of a `best guess' for abundance.
+Our intention in this section is to highlight further limitations with the model outputs presented above and to suggest ways to address these limitations.
 
-The most important consequences of an analysis are the possible conclusions and decisions that will be taken on the basis of the results.
-Here we advocate the view that model estimates should be converted into outputs, with associated uncertainty, that are most directly relevant to these conclusions and decisions.
-In this paper we provide an example based on a scenario where the most important aim of the analysis is identification of areas where animal density is above a certain threshold.
-It is not always the case that the most appropriate output of a species distribution model is a map of the predicted mean density.
+The most common method to communicate uncertainty in a maps of animal density is to produce maps of some measure of predictive uncertainty, such as the standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014} that can be derived from the posterior predictive distribution for the model or bootstrapping in a maximum likelihood context.
+Another approach is to map quantiles or the probability of exceeding certain thresholds for each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.
+In Section \ref{sec-results}, we present maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of the posterior intensity field that are all intended to communicate uncertainty in the spatial distribution of \akepa{} (Figures \ref{fig:intensity-mean-cv}B, \ref{fig:intensity-mean-cv}C, \ref{fig:intensity-quantiles}A, and \ref{fig:intensity-quantiles}B, respectively).
+Whilst maps that summarise spatial predictions are useful, they all mask certain properties that may be important when communicating the results of the analysis.
 
-The mapped posterior mean is the most common map one encounters in the literature.  However, the mean will always be smoother than realisations from the posterior random field.
+A map showing the posterior predicted mean across the study region (Figure \ref{fig:intensity-mean-cv}A) is often the key output of a species distribution model.
+However, even this relatively innocuous summary statistic can mask important features of the random field since the posterior mean will always be smoother than realisations of the field.
 \begin{figure}[!htb]
@@
 Note that each realisation has a finer-grained spatial structure than is shown in the posterior mean (\autoref{fig:intensity-mean-cv}A).
-Hence, considering this finer-grained structure, our interpretation is that the clustering of animals is stronger than might be expected if we only looked at the posterior mean.
-Similarly for model evaluation we recommend plotting multiple realisations of the posterior field as these will more closely resemble the spatial structure of the observed data.
-Noting differences between realisations is also an effective way to communicate uncertainty.
+Hence, considering this finer-grained structure, our interpretation is that the clustering of animals is stronger than might be expected if we only looked at the map of the posterior mean.
+This is one example of how a summary statistic can mask important features of the posterior random field.
+
+For model evaluation we recommend plotting multiple realisations of the posterior field as these will more closely resemble the spatial structure of the observed data and thus what we might encounter in the field.
+The mean is only a summary of the posterior intensity field and can give a misleading sense of homogeneity, particularly in the high density area in the south, when compared to realisations of the mean.
+Presenting multiple realisations, perhaps as an animation, is also an effective way to communicate uncertainty.
+\cite{bowman_GraphicsUncertainty_2019a} provide an interpolation method that preserves the mean and covariance structure of realisations that smoothly interpolates between realisations.
+This avoids abrupt changes that can occur when animating the raw realisations.
 
@@
 However, CV values will be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.
-The observed patterns in such maps will highlight regions of relatively lower intensity, not necessarily higher overall uncertainty.
+The observed higher CV areas in such maps will highlight regions of relatively lower intensity but not necessarily higher overall uncertainty.
 This is clearly the case in \autoref{fig:intensity-mean-cv}B.
+However, there is also variable sampling effort within the study region and it is not clear what impact this has on the CV map.
+This means that the resulting map is hard to interpret.
 
-The standard deviation map (\autoref{fig:intensity-mean-cv}C) does show clearly the spatially varying uncertainty.
-However,  care should be given to what the mapped values imply in the context of a given analysis.
-A default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high or low uncertainty.
+The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values.
+However, care should be taken as to what the mapped values imply in the context of a given analysis.
+A default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.
 What such a map does not show is whether these differences actually matter in the context of the analysis.
 For example, it could be the case that the standard deviation values are so large as to make the predictions have an unacceptably high uncertainty everywhere.
-Or, on the other hand, perhaps differences between high and low uncertainty regions (`high' and `low' as defined by the default colour scale) are small enough to be negligible when it comes to making conclusions or decisions, in which case such a spatially varying map is not a relevant output of the analysis.
+Or, on the other hand, perhaps differences between high and low uncertainty regions (where `high' and `low' are defined by the default colour scale for the map) are small enough to be negligible when it comes to making conclusions or decisions, in which case such a spatially varying map may not be a useful output to communicate to decision makers.
 These are two extreme examples.
-However, cases in between suffer from the same problem.
+However, even in less extreme situations, the actual practical relevance of the SD values should be considered in order to be able to interpret the map.
 
-The quantile maps also suffer from a potential difficulty in interpretation.
-The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data.
-However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for all prediction values to simultaneously achieve the 0.025 or 0.975 quantile.
-A similar problem will occur if mapping the marginal probability of exceeding a threshold at each prediction location.
-We believe these caveats make the map difficult to use for most non-statistically trained audiences.
-To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they were intensities and integrated them to obtain an expected abundance estimate of approximately 2,800 for the 0.025 quantile map and 10,200 for the 0.975 quantile map.
+The SD map suffers from another weakness which is that the SD values are linked to the predicted mean since the model assumes a log-Gaussian relationship.
+This means that the SD map usually looks similar to the mean intensity map.
+This relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort.
+One might expect uncertainty to be lower in areas with greater survey effort, however for the \akepa{} data the opposite is the case.
+This is another limitation of the SD map.
+
+The quantile maps (Figure \ref{fig:intensity-quantiles}) are also potentially difficult to interpret.
+The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data with one possible intensity surface (0.025 quantile map) that signifies lower abundance and another surface (0.975 quantile map) that signifies higher abundance.
+However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for all prediction locations to \textit{simultaneously} achieve their 0.025 or 0.975 quantiles.
+A similar problem will occur for maps of the marginal probability of exceeding a threshold at each prediction location.
+
+Presenting independent quantiles jointly in a single image is risky.
+We believe these caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps if they are not careful.
+To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they are intensity functions and integrate them to obtain an expected abundance estimate of approximately 2,800 for the 0.025 quantile map and 10,200 for the 0.975 quantile map.
 A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance.
 However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
-This demonstrates that the tendency to interpret these maps as possible intensities leads to interpretations that are inconsistent with the model that generated the maps.
+This demonstrates that the tendency to interpret these maps as showing possible intensity surfaces that are consistent with the observed data can lead to interpretations that are inconsistent with the very model that generated the maps.
+
+We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult or impossible to interpret.
+We next present some alternatives that avoid some, but not all, of these problems.
 

### 2021-09-10 paper.tex
edits up to 6.2

@@
 		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
-		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$}
+		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$.}
 		\label{fig:intensity-mean-cv}
@@
 
-The most common method to communicate uncertainty in a maps of animal density is to produce maps of some measure of predictive uncertainty, such as the standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014} that can be derived from the posterior predictive distribution for the model or bootstrapping in a maximum likelihood context.
-Another approach is to map quantiles or the probability of exceeding certain thresholds for each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.
-In Section \ref{sec-results}, we present maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of the posterior intensity field that are all intended to communicate uncertainty in the spatial distribution of \akepa{} (Figures \ref{fig:intensity-mean-cv}B, \ref{fig:intensity-mean-cv}C, \ref{fig:intensity-quantiles}A, and \ref{fig:intensity-quantiles}B, respectively).
-Whilst maps that summarise spatial predictions are useful, they all mask certain properties that may be important when communicating the results of the analysis.
+The most common method to communicate uncertainty in a maps of animal density is to produce maps of some measure of predictive uncertainty, such as the standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}.
+These can be derived from the posterior predictive distribution for the model in a Bayesian framework and bootstrapping in a maximum likelihood context.
+Another approach is to map quantiles or the probability of exceeding certain thresholds at each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.
+In Section \ref{sec-results}, we present maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of the posterior intensity field (Figures \ref{fig:intensity-mean-cv}B, \ref{fig:intensity-mean-cv}C, \ref{fig:intensity-quantiles}A, and \ref{fig:intensity-quantiles}B, respectively) that are all intended to communicate uncertainty in the estimate of the \akepa{} spatial distributions.
+Whilst such maps that summarise spatial predictions are useful, they all mask certain properties that may be important when communicating the results of the analysis.
 
 A map showing the posterior predicted mean across the study region (Figure \ref{fig:intensity-mean-cv}A) is often the key output of a species distribution model.
-However, even this relatively innocuous summary statistic can mask important features of the random field since the posterior mean will always be smoother than realisations of the field.
+However, the posterior mean will always be smoother than realisations of the field itself, highlighting that even this relatively innocuous summary statistic can mask important features of the random field.
+\autoref{fig:intensity-realizations} shows three such realisations from the posterior intensity field.
+Note that each realisation has a finer-grained spatial structure than is shown in the posterior mean (\autoref{fig:intensity-mean-cv}A).
+Our interpretation is that the clustering of animals is stronger than might be expected if we only looked at the map of the posterior mean.
 \begin{figure}[!htb]
@@
 \end{figure}
-\autoref{fig:intensity-realizations} shows three such realisations from the posterior intensity field.
-Note that each realisation has a finer-grained spatial structure than is shown in the posterior mean (\autoref{fig:intensity-mean-cv}A).
-Hence, considering this finer-grained structure, our interpretation is that the clustering of animals is stronger than might be expected if we only looked at the map of the posterior mean.
-This is one example of how a summary statistic can mask important features of the posterior random field.
 
-For model evaluation we recommend plotting multiple realisations of the posterior field as these will more closely resemble the spatial structure of the observed data and thus what we might encounter in the field.
-The mean is only a summary of the posterior intensity field and can give a misleading sense of homogeneity, particularly in the high density area in the south, when compared to realisations of the mean.
+For model evaluation we recommend plotting multiple realisations of the posterior field as these will more closely resemble the spatial structure of the observed data and thus what ecologists might encounter in the field.
+The mean can give a sense of homogeneity, particularly in the high density area in the south, when compared to realisations of the mean.
 Presenting multiple realisations, perhaps as an animation, is also an effective way to communicate uncertainty.
@@
 The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.
-However, CV values will be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.
-The observed higher CV areas in such maps will highlight regions of relatively lower intensity but not necessarily higher overall uncertainty.
-This is clearly the case in \autoref{fig:intensity-mean-cv}B.
-However, there is also variable sampling effort within the study region and it is not clear what impact this has on the CV map.
-This means that the resulting map is hard to interpret.
+However, CV values will be generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.
+Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.
+However, there is also variable sampling effort within the study region, with higher CV values in the region with lower effort.
+It is not clear what impact this has on the CV values and this makes the map hard to interpret.
 
-The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values.
+The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values also intended to indicate uncertainty.
 However, care should be taken as to what the mapped values imply in the context of a given analysis.
 A default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.
-What such a map does not show is whether these differences actually matter in the context of the analysis.
-For example, it could be the case that the standard deviation values are so large as to make the predictions have an unacceptably high uncertainty everywhere.
-Or, on the other hand, perhaps differences between high and low uncertainty regions (where `high' and `low' are defined by the default colour scale for the map) are small enough to be negligible when it comes to making conclusions or decisions, in which case such a spatially varying map may not be a useful output to communicate to decision makers.
-These are two extreme examples.
-However, even in less extreme situations, the actual practical relevance of the SD values should be considered in order to be able to interpret the map.
+However, whether these relative differences matter is dependent on the overall context of the analysis.
+Some levels of variability may be acceptable in some contexts and unacceptable in others.
 
-The SD map suffers from another weakness which is that the SD values are linked to the predicted mean since the model assumes a log-Gaussian relationship.
-This means that the SD map usually looks similar to the mean intensity map.
+The SD values also reflect a modelling \textit{assumption} but not necessarily variability due to varying survey effort.
+The SD is linked to the predicted mean by the assumed log-Gaussian relationship which means the mean and variance are directly connected, with higher mean values necessitating higher SD values.
+This means that the SD map will produce a pattern similar to the mean intensity map, as is the case here (Figures 2A and 2B).
 This relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort.
-One might expect uncertainty to be lower in areas with greater survey effort, however for the \akepa{} data the opposite is the case.
-This is another limitation of the SD map.
+One might expect uncertainty to be lower in areas with greater survey effort, however for the \akepa{} data the opposite is the case when interpreting the SD map.
 
-The quantile maps (Figure \ref{fig:intensity-quantiles}) are also potentially difficult to interpret.
-The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data with one possible intensity surface (0.025 quantile map) that signifies lower abundance and another surface (0.975 quantile map) that signifies higher abundance.
-However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for all prediction locations to \textit{simultaneously} achieve their 0.025 or 0.975 quantiles.
-A similar problem will occur for maps of the marginal probability of exceeding a threshold at each prediction location.
+The quantile maps (Figure \ref{fig:intensity-quantiles}) are also difficult to interpret.
+The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data.
+One possible intensity surface (0.025 quantile map) that signifies a kind of lower bound on abundance and another surface (0.975 quantile map) that signifies higher abundance.
+However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.
+A similar problem holds for maps of the marginal probability of exceeding a threshold at each prediction location.
 
 Presenting independent quantiles jointly in a single image is risky.
-We believe these caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps if they are not careful.
+These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.
 To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they are intensity functions and integrate them to obtain an expected abundance estimate of approximately 2,800 for the 0.025 quantile map and 10,200 for the 0.975 quantile map.
 A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance.
-However, these abundance estimates are outside the support of the posterior for abundance (see \autoref{fig:realized-abundance-posterior}).
-This demonstrates that the tendency to interpret these maps as showing possible intensity surfaces that are consistent with the observed data can lead to interpretations that are inconsistent with the very model that generated the maps.
+However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).
+This shows that the tendency to interpret these maps as showing possible intensity surfaces can lead to interpretations that are inconsistent with the very model that generated the maps.
 
 We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult or impossible to interpret.
-We next present some alternatives that avoid some, but not all, of these problems.
+In the following suggestion we suggest an approach that avoids some, but not all, of the pitfalls above.
 

### 2021-09-10 paper.tex
edits up to section 7

@@
 
-We use a GMRF representation of $\xi(s)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(s)$ via a SPDE.
+We use a GMRF representation of $\xi(\bs)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(\bs)$ via a SPDE.
 We define a finite element mesh with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$.
 The GRF is represented as $\xi(\bs) = \sum_l \beta_l \phi_l(\bs)$.
-The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices (see Appendix [BLAH]).
+The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices (\cite{blangiardo_spatial_2013}, Equation 6.19).
 The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function as distance between locations increases.
-We used a finite element mesh with 3748 nodes defined over the entire study region and extended over a buffer region to avoid the effects of boundary conditions within the region of inference.
+We use a finite element mesh with 3748 nodes defined over the entire study region and extended over a buffer region to avoid the effects of boundary conditions within the region of inference.
 
@@
 The purpose of penalised complexity priors is to place sufficient prior probability on a simple model to allow the random effect to be essentially `turned off' if the model component is not supported by the data.
-
 To specify penalised complexity priors on the SPDE effect we set $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$.
@@
 
-Our solution to these problems with quantile and standard deviation maps is to suggest that consideration should be given \emph{a priori} to relevant values of the process and acceptable levels of uncertainty that are suitable given the context and aims of the analysis.
-These values can then be used to construct summary maps.
+One solution to the problem of interpreting quantile and exceedance probability maps is to use methods that explicitly consider the \textit{joint} probability of events across prediction locations, as opposed to the \textit{marginal} probability at each location.
+The approach taken here requires that we decide, \emph{a priori}, on relevant values of the process and acceptable levels of uncertainty that are suitable given the context and aims of the analysis.
+These values can then be used to construct summary maps that consider joint events across prediction locations.
+
 We demonstrate this perspective using excursions sets and excursion functions \citep{bolin_excursion_2015}.
-Although there are other possible approaches [[cite here?]], excursions methods are a natural choice for Gaussian random fields.
-These methods require the user to specify thresholds of interest for the random field and acceptable levels of uncertainty.
-For this reason they avoid the issue of a default colour scale potentially affecting our interpretation of results.
+By choosing acceptable levels of uncertainty, these methods the avoid the issue of the default colour scale for the SD and CV maps which may potentially affecting our interpretation of results.
 Excursion sets and excursion functions are also based on the joint probability of events across a set of locations.
@@
 
-The positive excursion set with level $u$ for a function $f(s)$ with domain $\Omega$ is $A_u^{+}(f) = \{ s \in \Omega ; f(s) > u \}$, i.e.
+We give the technical definition of excursion sets and functions below, using the same notation as \cite{bolin_excursion_2015}.
+For readers unfamiliar with these methods, the notation can take some getting used to.
+However, the precise mathematical definitions are useful when it comes to interpreting these methods.
+We try to, as much as possible, give an informal explanation alongside the mathematical description.
+
+The positive excursion set with level $u$ for a function $f(\bs)$ with domain $\Omega$ is $A_u^{+}(f) = \{ \bs \in \Omega ; f(\bs) > u \}$, i.e.
 the set of all locations in $\Omega$ where $f$ exceeds a threshold value $u$.
-For a stochastic process $\lambda(s)$ the positive excursion set with level $u$ and probability $1 - \alpha$ is
+For a random field, $\lambda(\bs)$, the positive excursion set with level $u$ and probability $1 - \alpha$ is
 \begin{equation*}
-E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \} \;\;\; .
+E_{u,\alpha}^{+}(\lambda) = \argmax_{D}\{\lvert D \rvert : \mathbb{P}\left[D \subset A_u^{+}(\lambda)\right] \geq 1 - \alpha \} .
 \end{equation*}
-Note that $A_u^{+}(f)$ specifies a set for which a function $f(s)$ exceeds a threshold value $u$ for \textit{every location} in the set.
-As such the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the largest such set for which a threshold is exceeded simultaneously for all locations in the set, with a chosen probability.
-Negative excursion sets are similarly defined.
-Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate probabilities.
-An implementation is available in the \texttt{excursions} package \citep{bolin_calculating_2018} available through the Comprehensive R Archive Network \citep{r_2017}.
+Note that $A_u^{+}(f)$ specifies a set of locations for which a function $f(\bs)$ exceeds a threshold value $u$ for \textit{every location} in the set.
+Therefore, the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the \textit{largest} such set of locations for which \textit{realisations} of $\lambda(\bs)$ exceed the threshold $u$, \textit{simultaneously} for all locations in the set, with probability $1-\alpha$.
+Note that we must decide on relevant choices for $u$ and $\alpha$ given the aims and context of the analysis.
+Negative excursion sets are similarly defined by considering the probability of being below a threshold value.
+Contour uncertainty can also be defined as the set difference between positive and negative excursion sets \citep{bolin_excursion_2015}.
 
-\autoref{fig:excursions} (left panel) shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95.
+Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate the required probabilities in a computationally efficient way.
+A full description of the methods is given in \cite{bolin_excursion_2015} and a software implementation in the \texttt{excursions} package \citep{bolin_calculating_2018}.
+
+\autoref{fig:excursions}A shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95 (hence $\alpha = 0.05$).
 This figure can be interpreted in a natural way as the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
@@
 	\caption{Left:  The positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.
-Right: The positive excursion function with a level corresponding to 1 bird per hectare}
+Right: The positive excursion function with a level corresponding to 1 bird per hectare.
+The scale shows the probability values $1-\alpha$.}
 	\label{fig:excursions}
 \end{figure} 
-To visualise multiple such maps we can use the excursion function $F_u^{+}(s) = \sup \{1 - \alpha ; s \in E_{u,\alpha}^+ \}$, which defines for each location largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.
+This map could be used to define, for example, a `core region' for the \akepa{} population, with particular conservation importance.
+Crucially, this map depends on a clear mathematical description of what a `core region' actually means in a statistical sense.
+While any particular definition would be up for debate, the advantage is that the statistical methodology is taken as close as possible to the aims of the analysis.
+Often a map of species density will be interpreted informally in many different ways (one of which is identifying hotspots for example).
+Here we present an example of how a clear statistical definition can be used to produce a map that incorporates relevant definitions and uncertainties, and can be interpreted clearly.
+
+To visualise how such excursion sets change with different levels of uncertainty we can use the excursion function $F_u^{+}(\bs) = \sup \{1 - \alpha ; \bs \in E_{u,\alpha}^+ \}$.
+This defines, for each location, the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.
 i.e.
-if we allow greater uncertainty, which locations would be included in the excursion set.
-The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursions} (right panel).
-This figure can also be interpreted naturally.
-It shows the largest possible probability for each location to be a member of a set in which the intensity exceeds a threshold simultaneously across all locations in the set.
-It is clear from the figure that regions on the edge of the excursion set would be included if the $\alpha$ value were allowed to increase slightly.
-However, for regions in the north there is essentially no probability level for which those locations would be included in $A_u^{+}(\lambda)$.
-We chose the threshold of 1 bird per hectare and an error probability of 0.05 to demonstrate the approach.  Multiple values for each of these can be considered as a simple extension.
+if we allow greater uncertainty, this function shows which locations would be included in the excursion set.
+
+The excursion function with a level corresponding to 1 bird per hectare is shown in \autoref{fig:excursions}B.
+It is clear from the figure that regions close to the edge of the excursion set with $\alpha = 0.05$ would be included if the $\alpha$ value were allowed to increase slightly.
+Similarly, the core of the high density region seems to exceed 1 bird per hectare even for $\alpha$ levels close to zero.
+There is a region to the north of the main population that would be included in the positive excursion set for 1 bird per hectare if the probability were allowed to be lower, around 0.5.
+However, for regions in the north there is essentially no probability level for which those locations would be included.
+This map could be used to identify, for example, secondary areas of potential importance beyond the core population that could be investigated further.
+
+We chose the threshold of 1 bird per hectare and an error probability of 0.05 to demonstrate the approach.
+However, multiple values for each of these can be considered and we imagine that, in most contexts, there would be value in considering various sets of thresholds and uncertainty levels.
 

### 2021-11-30 paper.tex
naphon tutorial
update discussion and add unmarked citations

@@
 \usepackage[a4paper]{geometry}   % otherwise statsoc class doesn't compile to A4 correctly
-
 %\usepackage{setspace}
@@
 Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.
-One-stage analyses of distance sampling data have tended to be Bayesian and use Markov chain Monte Carlo (MCMC) methods for inference, usually requiring data augmentation to model unobserved individuals or groups
-% problem with this Royle reference here.  Not sure what I was intending to cite.
-\citep{royle_HierarchicalModelSpatial_2008, schmidt_using_2012}.
-\citet{oedekoven_bayesian_2014} present a one-stage model that avoids data augmentation and incorporate model uncertainty using reversible jump MCMC.
+A popular one-stage maximum-likelihood method, introduced by \cite{royle_ModelingAbundanceEffects_2004} and implemented in the \texttt{R} package \texttt{unmarked} \citep{fiske_UnmarkedPackageFitting_2011}, is based on reformulating the model as a multinomial likelihood.
+This depends choosing discrete distance classes and binning the data into counts defined on discrete spatial units.
+Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, a Poisson likelihood that can be optimised to achieve one-stage inference.
+This approach rests heavily on the discretisation of both the distance data and the spatial location data.
+Below we present a point process perspective that does not require this discretisation step in either the distance or spatial location data.
+The \texttt{unmarked} package also does not implement spatially structured random effects, something that we use here and is also popular in the two-stage approach implemented using \texttt{mgcv}.
+
+Bayesian one-stage approaches have tended to use Markov chain Monte Carlo (MCMC) methods for inference along with data augmentation to model unobserved individuals or groups \citep{schmidt_using_2012}.
+\citet{oedekoven_bayesian_2014} present a one-stage model that avoids data augmentation by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty using reversible jump MCMC.
 
@@
 However, the point process viewpoint is not new and has been taken numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
-
-\vfill
 
@@
 
-The model fits 40 to 60 seconds on a laptop with 10 Gb RAM and a 2.50 GHz CPU when INLA is run using the `empirical Bayes' option which does not incorporate uncertainty in hyper parameters.
-The model fits in minutes in full INLA mode including hyper parameter uncertainty.
+The model fits in roughly 40 to 60 seconds on a laptop with 10 Gb RAM and a 2.50 GHz CPU when INLA is run using the `empirical Bayes' option which does not incorporate uncertainty in hyper parameters.
+The model fits in a few minutes in the full INLA mode, including hyper parameter uncertainty.
 We suggest using the `empirical Bayes' mode for model development and full INLA on a smaller subset of candidate models.
@@
 
-The SD values also reflect a modelling \textit{assumption} but not necessarily variability due to varying survey effort.
-The SD is linked to the predicted mean by the assumed log-Gaussian relationship which means the mean and variance are directly connected, with higher mean values necessitating higher SD values.
-This means that the SD map will produce a pattern similar to the mean intensity map, as is the case here (Figures 2A and 2B).
-This relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort.
-One might expect uncertainty to be lower in areas with greater survey effort, however for the \akepa{} data the opposite is the case when interpreting the SD map.
+The SD values also reflect a modelling assumption rather than variability due to, say, varying survey effort.
+The SD is linked to the predicted mean by the assumed log-Gaussian relationship.
+If X is normally distributed with mean $\mu$ and variance $\sigma^2$ then $Z = \exp(X)$ has a mean-variance relationship given by $\text{Var}(Z) = \exp(\sigma^2 - 1)\mathbb{E}(Z)^2$.
+Therefore the SD map will generally show a pattern similar to the mean intensity map (Figures 2A and 2B).
+Since this relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort, one should take care when interpreting the SD map.
+It should be noted however, that this modelling assumption affects uncertainty quantification for all outputs based on the posterior intensity field, not just the SD map.
 
@@
 
-This paper presents a new approach to analysing point transect distance sampling data that incorporates several innovative methods and considers the bulk of an applied statistician's workflow of model specification, inference procedure, model evaluation and communication of results.
-Although any of these areas on its own would be substantive enough to be the main focus of a paper, our brief presentation of issues in sequence along with our chosen solutions will be useful to statisticians who face similar questions and data structures.
+We have presented a new approach to analysing point transect distance sampling data in a Bayesian setting.
+We chose to do this in a relatively simplified setting, considering a single year of data, with no covariate effects and using a GRF to explain spatial variation in density.
+There are several natural extensions to the model.
+Spatio-temporal models can also be fitted using \texttt{R-INLA}, which would allow multiple years of data to be modelled jointly, as was the case in \citep{camp_dsm_2020}.
+Another extension is to develop the observation process model by considering other detection functions or including covariates, for example.
+Factors such as weather conditions, observer expertise, animal behaviour and morphological traits are likely to affect detectability.
+Future research should investigate the degree of complexity that can be well-approximated by the iterated INLA approach.
 
-There are several natural extensions to the model.
-For this analysis we only considered a single survey year.
-Spatio-temporal extensions to this are possible, for example by considering a space-time interaction models.
-\texttt{R-INLA} has the ability to fit space-time effects that can be represented as a Kronecker product of the relevant precision matrices \citep{blangiardo_spatial_2013, yuan_point_2017}.
-An example of this type of space-time interaction effect is an interaction between the spatial SPDE with an auto-regressive temporal process.
-\cite{blangiardo_spatial_2013} show how this can be done with numerous examples and code snippets.
-
-Another natural extension is to consider more complicated observation processes.
-There is the potential to include explanatory covariates on detection function parameters.
-For example, we did not make use of the detection type (audio or visual) in our analysis, although this was recorded and detectability may vary depending on whether a bird is seen or heard.
-Other factors could be considered, such as weather conditions, observer expertise, animal behaviour and morphological traits.
-Any additional parameters to the non-linear model components will be estimated using the outer-optimisation step of iterated INLA.
-[[Can I speculate here about how iterated INLA might perform as parameters are added to non-linear components?
-Is, higher-dimensional optimisation slower/infeasible/unstable at a certain point?]]
-
-The iterated INLA inference procedure is a general approach that could be applied to other non-linear model components that arise in the area of spatial ecology.
-We note two promising areas in which our flexible inference procedure may be useful:
+The iterated INLA inference procedure has the potential to be applied to other non-linear model components that arise spatial statistics.
+We note two promising areas within the field of spatial ecology that highlight this potential:
 
@@
 	\item \textbf{Selectivity analysis} In surveys of the marine environment, the size, length and weight of fish affects the likelihood of being caught in a trawl net.
-Thus any analysis of these data must account for this.
-The probability of capture is typically modelled as a logistic function with parameters that depend on morphological traits \citep{herrmann_understanding_2016, madsen_selectivity_2007, galbraith_demersal_1994}, which could be estimated jointly with the spatial distribution within our inference framework.
-	\item \textbf{Functional responses} In the area of understanding species-habitat or species-species interactions, the concept of a `functional response' captures the idea that the abundance of a resource or risk will affect the response to it \citep{holling_some_1959}.
+The probability of capture is typically modelled as a logistic function with parameters that depend on morphological traits \citep{herrmann_understanding_2016, madsen_selectivity_2007, galbraith_demersal_1994}.
+This function is analogous to the detection function in distance sampling models.
+Our framework could allow, for example, the joint modelling of selectivity and spatial data.
+	\item \textbf{Functional responses} The concept of a `functional response' captures the idea that the overall abundance of a resource or risk will affect a species' response to it \citep{holling_some_1959}.
 These varying responses can be described mathematically using parametric equations that have the desired properties, e.g.
-a horizontal asymptote so that beyond a certain threshold no extra amount of a resource will affect the response to it.
-These types of responses are typically not considered in species distribution models but our framework allows them to be estimated alongside the more common linear or quadratic fixed effects.
-This flexibility in specifying model components could also be more widely applicable in areas such as public health and spatial epidemiology, where responses to conditions may be non-linear.
+an increasing function with a horizontal asymptote, to capture the notion of diminishing returns.
+Such non-linear functions could be incorporated within our framework.
+This flexibility in specifying model components could also be more widely applicable in areas such as public health and spatial epidemiology, where responses to conditions may suitably be modelled using non-linear relationships.
+\end{itemize}
+There are no doubt many more application areas that could make use of the additional flexibility that comes with allowing non-linear predictors.
 
-\end{itemize}
-We note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analysis
-
-In our analysis we only briefly touch upon the problem of communicating uncertainty in maps and highlight the excursions methods as a new and positive addition to this area.
-In our example, the threshold was chosen to illustrate the method but we believe that discussions with relevant stakeholders will lead to agreement on a (possibly large) set of thresholds that are relevant given the aims and context of the analysis.
-This requires the input of both the relevant domain experts and statisticians so stakeholders understand the consequences of the thresholds and uncertainty levels they wish to consider.
-We also note the potential for excursions to be used to define the notion of occupancy (and thus map occupied areas, with a given probability of error).
-
-We end with a note on the similarity between what we have done and existing two stage approach.
-The one-stage thinned point process perspective can seem fundamentally different to a two-stage GAM approach fitted to count data.
-It can be tempting to think that, because the point process model does not require us to bin the data into counts, we can avoid some loss of information.
-However, we also made the same assumption of constant intensity within each transect as in more traditional distance sampling approaches.
-This leads to the same loss of information since we do not use the exact location of points within each transect, except to account for detectability.
-
-In summary, we have presented a novel framework for analysing point transect distance sampling data that introduces new approaches to model specification, inference procedure and communication of results.
-These developments will be of interest to those with similar data structures in spatial ecology as well as researchers in other areas who we hope will explore the full flexibility and generality of these developments in their own work.
-
-% spp has potential to provide general framework? [Janine comment]
-% Mention modelling on sphere is possible without projecting on to R^2?
+The popularity of using INLA for spatial modelling has grown, in large part, due to the computational efficiency of the approach.
+However, this efficiency comes at the cost of reduced flexibility in model specification over, say, a more general probabilistic programming languages such as \texttt{JAGS} \citep{plummer_jags_2017} or \texttt{Stan} \citep{stan_Rstan_2020}.
+The iterated INLA approach allows a wider class of models to be considered whilst retaining much of the computational benefits of INLA.
+In addition to computational efficiency, there are some features of INLA that are not generally available in other packages, such as: (i) the ability to model using GRFs directly on the sphere, avoiding projecting to $\mathbb{R}^2$ \citep{lindgren_explicit_2011}; (ii) GRFs that account for complex barriers such as coastlines and islands, to avoid inappropriately smoothing across these \cite{bakka_NonstationaryGaussianModels_2019}; and (iii) support for joint likelihood models, allowing modelling of, for example, marked point processes \citep{illian_FittingComplexEcological_2013}, and combining multiple data sources.
+These approaches, and many others we have not mentioned, that are implemented in INLA, can now readily be incorporated into distance sampling analyses, as well as broader spatial ecology and spatial statistics applications.
 

### 2021-11-30 paper.tex
add Camp 2020 comparison paragraph

@@
 
-The abundance estimates broadly agree with a standard two-stage analysis of the data \citep{camp_dsm_2020} but the attractive feature of the one-stage model is that, by using Monte Carlo samples from the joint posterior, model summaries naturally incorporate uncertainty in the detection model.
+The abundance estimates cannot be directly compared to the results in \cite{camp_dsm_2020} since their two-stage model also includes a temporal correlation and we include an area of Hakalau that they had to exclude.
+Converting the above abundance estimates to density per hectare for the whole study region allows for some comparison with Figure 2 in \cite{camp_dsm_2020} which suggests that \akepa{} density was roughly between 0.5 and 1.5 birds per hectare in 2002 (considering both the design-based estimate and the spatio-temporal GAM estimate).
+In our analysis we find a credible region of roughly 0.9 to 1.7 birds per hectare, which is slightly different but overlaps significantly with the results in \cite{camp_dsm_2020}.
+
+An attractive feature of our Bayesian one-stage model is that, by basing model outputs on posterior Monte Carlo samples, all the outputs presented in the above, the summaries of the random field as well as the posterior abundance, incorporate the uncertainty in the detection model in a natural way.
 There is no additional step required to incorporate detection function uncertainty beyond working with samples from the joint posterior.
-
-\vfill
 

### 2021-12-13 paper.tex
address Rick's comments

@@
 The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
-Here we present an analysis of wildlife survey data collected on a critically endangered \hawaii{}an forest bird that aims to meet this objective.
+Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.
 The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.
@@
 the `sociality' of the species).
-For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient for which no straightforward explanatory cause has been found \cite{camp_dsm_2020}.
+For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient for which no straightforward explanatory cause has been found \citep{camp_dsm_2020}.
 From a statistical perspective, this heterogeneity that cannot be explained by available covariates suggests the inclusion of spatially structured random effects in the model.
@@
 It is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.
-The \akepa{} data is collected with the clear objective of monitoring the population and informing conservation management decisions.
+The \akepa{} data are collected with the clear objective of monitoring the population and informing conservation management decisions.
 These decisions will be taken by stakeholders with a range of statistical expertise.
@@
 
-% Update to section refs
-The rest of the paper proceeds as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the GMRF random effect; and Sections \ref{sec-results} and \ref{sec-comms} present the results of the analysis and discuss model evaluation and communication.
+The rest of the paper proceeds as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the GMRF random effect; and Sections \ref{sec-results} presents results of the analysis and discusses model evaluation and effective communication of uncertainty.
 
@@
 
-The survey data in the open forest stratum is available for download \citep{camp_datarelease_2002} and the full time series data was analysed in a spatio-temporal context in \citep{camp_dsm_2020}.
+The survey data used in this analysis is available for download \citep{camp_datarelease_2002} and the full time series data was analysed in a spatio-temporal context in \citep{camp_dsm_2020}.
 For the purposes of our analysis we select a single survey year from the \akepa{} time series (2002) that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability.
-In 2002, there were a total of 289 point transects sampled within the 4,603 ha study area of Hakalau.
-(Fig.
-\ref{fig:2002studyareapointspt}) A total of 276 \akepa{} were detected on 121 point transects.
+In 2002, there were a total of 289 point transects sampled within the 4,603 ha study area of Hakalau (Fig.
+\ref{fig:2002studyareapointspt}).
+A total of 276 \akepa{} were detected on 121 point transects.
 The number of detections within each point transect ranged from zero to six.
@@
 
-Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and a parametric detection function to estimate the detectability of animals as a function of distance from transect or observer \citep{buckland_advanced_2004, buckland_distance_2015}.
+Distance sampling methods aim to estimate abundance by using a spatially explicit sampling design and a parametric detection function to estimate the detectability of animals as a function of distance from transect or observer \citep{buckland_distance_2015}.
 Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.
@@
 These methods allow animal density to be associated with spatially-indexed covariates and non-parametric random effects.
-Therefore, animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{johnson_model-based_2010, miller_spatial_2013, buckland_model-based_2016}.
+Therefore, animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{buckland_model-based_2016, miller_spatial_2013, johnson_model-based_2010}.
 As such, varying density estimates are not restricted by the stratification design required for Horvitz-Thompson-like estimators.
@@
 A key concern with the two-stage approach is the propagation of uncertainty from the detection model to the spatial model.
-Early attempts to address this focused on bootstrapping \citep{lahiri_resampling_2003, hedley_spatial_2004} but more recent work has pointed to potential difficulties with the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_VariancePropagationDensity_2021, williams_chilean_2011}.
+Early attempts to address this focused on bootstrapping \citep{hedley_spatial_2004, lahiri_resampling_2003} but more recent work has pointed to potential difficulties with the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_VariancePropagationDensity_2021, williams_chilean_2011}.
 Instead, \cite{bravington_VariancePropagationDensity_2021} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of the detection function at the first-stage maximum-likelihood estimate.
@@
 A popular one-stage maximum-likelihood method, introduced by \cite{royle_ModelingAbundanceEffects_2004} and implemented in the \texttt{R} package \texttt{unmarked} \citep{fiske_UnmarkedPackageFitting_2011}, is based on reformulating the model as a multinomial likelihood.
-This depends choosing discrete distance classes and binning the data into counts defined on discrete spatial units.
+This depends on choosing discrete distance classes and binning the data into counts defined on discrete spatial units.
 Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, a Poisson likelihood that can be optimised to achieve one-stage inference.
@@
 Representing the observation model as an SPDE also differs substantially from standard distance sampling models that use a parametric model for the detection function.
-This may have hindered a wider adoption of using INLA for distance sampling analyses, \cite{yuan_point_2017} remains the only example we are aware of.
 Furthermore, parametric forms of detection functions are readily extendible by including covariates and random effects to explain variation in detectability.
-It is not clear how to do this the SPDE formulation.
+It is not clear how to do this using the SPDE implementation in INLA.
 These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA and the benefits of a one-stage modelling process.
 
-Here we present an approach that is closely related to \citet{yuan_point_2017}.
+Here we present an approach that is related to \citet{yuan_point_2017} but with some key differences.
 We also fit the model using INLA and take advantage of the benefits the increased computational efficiency compared to MCMC.
@@
 Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
-However, the point process viewpoint is not new and has been taken numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
+However, the point process framework is not new and has been applied numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
 
@@
 The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(\bs)$.
-The usual approach for two stage models is to assume $\lambda(\bs)$ is constant within each spatial unit, however we are able to relax this assumption slightly to allow $\lambda(\bs)$ to be a linear function within in each transect.
+The usual approach for two stage models is to assume $\lambda(\bs)$ is constant within each spatial unit, however we are able to relax this assumption slightly to allow $\lambda(\bs)$ to be a linear function within each transect.
 Given such an assumption, any observed deviations from uniformity can be attributed to detectability and not to variation in the intensity.
@@
 In this section we describe this approximate likelihood and show how it can be written as a modified Poisson likelihood.
-We follow closely with the approach given in \cite{simpson_going_2016} for a fully observed log-Gaussian Cox process but adapted for the point transect distance sampling case.
+We adapt the approach given in \cite{simpson_going_2016} for a fully observed log-Gaussian Cox process, modified to account for the point transect distance sampling observation process.
 
 To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.
-In a slight abuse of notation $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes any location within the transect.
+In a slight abuse of notation $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes any location within the point transect.
 It should be clear from context which is intended by $\bs_k$.
@@
 
-The approach taken here is similar to the Berman-Turner device \citep{berman_approximating_1992, baddeley_practical_2000} although without the assumption that the locations of observed points form a part of the quadrature scheme.
+The approach taken here is similar to the Berman-Turner device \citep{baddeley_practical_2000, berman_approximating_1992} that is used to fit a wide variety of point process models using GLM software.
+The main difference with our approach is that we do not use the assumption that the locations of observed points form a part of the quadrature scheme.
 
@@
 
-We denote by $\bar{p}_{\bm{v}}$ the posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.
+Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.
 New linearisation points are found by considering the functional
@@
 This fixed point is identified via an iterative scheme.
-Given a current choice of linearisation point $\bm{v}$, estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.
+Given a current choice of linearisation point $\bm{v}$, we estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.
 We iterate this procedure until a fixed point is identified within a chosen tolerance.
@@
 This section describes the spatial model for the intensity process.  
-As mentioned above, the \akepa{} population exhibits spatially heterogeneity for which there is no known explanatory covariate.
+As mentioned above, the \akepa{} population exhibits spatial heterogeneity for which there is no known explanatory covariate.
 Therefore we use a spatially structured random effect to explain this variation.
@@
 Often this is due to parameterisations and approximations that lead to dense prior precision matrices which have an  $\mathcal{O}(k^3)$ cost to factorise for an effect with $k$ basis parameters.
-A key development in the development of computationally efficient methods is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs).
+A key approach in computationally efficient methods for spatially structured random effects is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs).
 These result in a sparse precision matrix which can be factorised with a cost of $\mathcal{O}(k^{3/2})$.
@@
 
-In order to specify priors on the Mat\'ern covariance below we use a reparameterisation of $\kappa$ and $\tau$ to a range and variance parameter.
+In order to specify priors on the Mat\'ern covariance we use a reparameterisation of $\kappa$ and $\tau$ to a range and variance parameter.
 When $\nu = 1$ and the domain is two dimensional then the reparameterisation is $\rho = \sqrt{8} / \kappa$ for the range and $\sigma^2 = 1 / (4\pi\kappa^2\tau^2)$ for the marginal variance (\cite{blangiardo_spatial_2013}, p196).
@@
 		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
-		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$.}
+		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$.  The colour scale is different for each panel.}
 		\label{fig:intensity-mean-cv}
@@
 		\includegraphics[scale=0.525]{figures/intensity_quantiles.png}
-		\caption{The predicted posterior intensity quantiles:  A) 0.025 quantile B) 0.975 quantile.  Both plots plots use the same colour scale}
+		\caption{The predicted posterior intensity quantiles:  A) 0.025 quantile B) 0.975 quantile.  Both plots use the same colour scale.}
 		\label{fig:intensity-quantiles}
@@
 We note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).
-This is because this is the correlation range on the log-intensity scale.
+This is because the correlation function is defined on the log-intensity scale.
 On the intensity scale the empirical range of correlation is shorter.
@@
 
-To plot the posterior abundance estimate for the \akepa{} in Hakalau reserve we use a Monte Carlo method based on sampling realisations of the posterior intensity field.
+To plot the posterior abundance estimate for the \akepa{} in Hakalau we use a Monte Carlo method based on sampling realisations of the posterior intensity field.
 Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(\bs)\mathrm{d}\bs$ its corresponding random variable within our Bayesian framework.
@@
 
-The abundance estimates cannot be directly compared to the results in \cite{camp_dsm_2020} since their two-stage model also includes a temporal correlation and we include an area of Hakalau that they had to exclude.
+The abundance estimates cannot be directly compared to the results in \cite{camp_dsm_2020} since their two-stage model also includes a temporal correlation and we include an area of Hakalau that the authors excluded.
 Converting the above abundance estimates to density per hectare for the whole study region allows for some comparison with Figure 2 in \cite{camp_dsm_2020} which suggests that \akepa{} density was roughly between 0.5 and 1.5 birds per hectare in 2002 (considering both the design-based estimate and the spatio-temporal GAM estimate).
@@
 
-Our presentation of the results above is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (\autoref{fig:intensity-mean-cv} and \autoref{fig:intensity-quantiles}) are not always provided.
+Our presentation of the results above is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (Figures \ref{fig:intensity-mean-cv} and \ref{fig:intensity-quantiles}) are not always provided.
 It is common to see maps of point estimates (e.g.
@@
 
-The most common method to communicate uncertainty in a maps of animal density is to produce maps of some measure of predictive uncertainty, such as the standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}.
+The most common method to communicate uncertainty in maps of animal density is to produce maps of some measure of predictive uncertainty, such as the standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}.
 These can be derived from the posterior predictive distribution for the model in a Bayesian framework and bootstrapping in a maximum likelihood context.
@@
 The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.
-However, CV values will be generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.
+However, CV values will generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.
 Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.
@@
 Therefore the SD map will generally show a pattern similar to the mean intensity map (Figures 2A and 2B).
-Since this relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort, one should take care when interpreting the SD map.
+Since this relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort.
 It should be noted however, that this modelling assumption affects uncertainty quantification for all outputs based on the posterior intensity field, not just the SD map.
@@
 We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult or impossible to interpret.
-In the following suggestion we suggest an approach that avoids some, but not all, of the pitfalls above.
+In the following section we suggest an approach that avoids some, but not all, of the pitfalls above.
 
@@
 We demonstrate this perspective using excursions sets and excursion functions \citep{bolin_excursion_2015}.
-By choosing acceptable levels of uncertainty, these methods the avoid the issue of the default colour scale for the SD and CV maps which may potentially affecting our interpretation of results.
+By choosing acceptable levels of uncertainty, these methods avoid the issue of the default colour scale for the SD and CV maps which may potentially affecting our interpretation of results.
 Excursion sets and excursion functions are also based on the joint probability of events across a set of locations.
@@
 	\includegraphics[scale=0.5]{figures/excursions.png}
-	\caption{Left:  The positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.
-Right: The positive excursion function with a level corresponding to 1 bird per hectare.
+	\caption{A) The positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.
+B) The positive excursion function with a level corresponding to 1 bird per hectare.
 The scale shows the probability values $1-\alpha$.}
@@
 This defines, for each location, the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.
-i.e.
+I.e.
 if we allow greater uncertainty, this function shows which locations would be included in the excursion set.
@@
 There are several natural extensions to the model.
-Spatio-temporal models can also be fitted using \texttt{R-INLA}, which would allow multiple years of data to be modelled jointly, as was the case in \citep{camp_dsm_2020}.
+Spatio-temporal models can also be fitted using \texttt{R-INLA}, which would allow multiple years of data to be modelled jointly, as was the case in \cite{camp_dsm_2020}.
 Another extension is to develop the observation process model by considering other detection functions or including covariates, for example.
@@
 The iterated INLA approach allows a wider class of models to be considered whilst retaining much of the computational benefits of INLA.
-In addition to computational efficiency, there are some features of INLA that are not generally available in other packages, such as: (i) the ability to model using GRFs directly on the sphere, avoiding projecting to $\mathbb{R}^2$ \citep{lindgren_explicit_2011}; (ii) GRFs that account for complex barriers such as coastlines and islands, to avoid inappropriately smoothing across these \cite{bakka_NonstationaryGaussianModels_2019}; and (iii) support for joint likelihood models, allowing modelling of, for example, marked point processes \citep{illian_FittingComplexEcological_2013}, and combining multiple data sources.
+In addition to computational efficiency, there are some features of INLA that are not generally available in other packages, such as: (i) the ability to model using GRFs directly on the sphere, avoiding projecting to $\mathbb{R}^2$ \citep{lindgren_explicit_2011}; (ii) GRFs that account for complex barriers such as coastlines and islands, to avoid inappropriately smoothing across these \citep{bakka_NonstationaryGaussianModels_2019}; and (iii) support for joint likelihood models, allowing modelling of, for example, marked point processes \citep{illian_FittingComplexEcological_2013}, and combining multiple data sources.
 These approaches, and many others we have not mentioned, that are implemented in INLA, can now readily be incorporated into distance sampling analyses, as well as broader spatial ecology and spatial statistics applications.
+
+\section*{Acknowledgements}
+
+The \hawaii \akepa{} data are provided by U.S.
+Fish and Wildlife Service, Hakalau Forest Unit National Wildlife Refuge.
+We are grateful to the many U.S.
+Fish and Wildlife Service staff and volunteers who helped collect the data.
+We thank Steve Buckland for helpful comments on an early version.
+Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S.
+Government.
 

### 2024-07-25 paper.tex
add inlabru reference + edit iinla section

@@
-\documentclass{stylefile16/statsoc}
+\documentclass{statsoc}
 \usepackage[a4paper]{geometry}   % otherwise statsoc class doesn't compile to A4 correctly
@@
 \usepackage{natbib}
-\setcitestyle{stylefile16/rss}
+%\setcitestyle{rss}
 
@@
 
-% shortcuts
+% new commands
+\newcommand{\wt}[1]{\widetilde{#1}}
+\newcommand{\ol}[1]{\overline{#1}}
+\newcommand{\wh}[1]{\widehat{#1}}
 \newcommand{\bbeta}{\boldsymbol{\beta}}
@@
 \author[Andrew E. Seaton {\it et al.}]{Andrew E. Seaton}
-\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, 
-	University of St Andrews, 
-	St Andrews,
+\address{School of Mathematics \& Statistics, 
+	University of Glasgow, 
+	Glasgow,
 	UK}
-\email{aes22@st-andrews.ac.uk}
+\email{andrew.seaton.2@glasgow.ac.uk}
 
@@
 This is a non-linear function of $\phi$ which cannot be fitted using standard GLM methods.
-To address this we use a novel approximate method based on a first-order Taylor approximation of the non-linear predictor.
+To address this we use a novel approximate-method based on an iterative model fitting scheme.
+This method is implemented in the \texttt{R} package \texttt{inlabru} which is a wrapper for and an extension of the \texttt{R-INLA} package \citep{lindgren_inlabru_2024}.
+Full details of the method can be found in the software paper, here we present a brief overview.
 
@@
 Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.
-New linearisation points are found by considering the functional
+New linearisation points are found by via the functional
 \begin{equation}
-	f(\bar{p}_{\bm{v}}) = \left\{ \argmax_{\bm{u}_i} \: \bar{p}_{\bm{v}}(u_i | \bm{Y}), \quad i = 1, \ldots, n_{\bm{u}} \right\}
+	f(\ol{p}_{\bm{v}}) = (\wh{\bm{\theta}}_{\bm{v}},\wh{\bm{u}}_{\bm{v}})
 \end{equation}
-where $n_{\bm{u}}$ is the length of $\bm{u}$.
-Each marginal posterior $\bar{p}_{\bm{v}} (u_i | \bm{Y})$ and their modes are estimated during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost for each fit of the linearised model.
-We then seek a fixed point of this functional so that $\bm{u}_0 = f(\bar{p}_{\bm{u}_0})$.
-This fixed point is identified via an iterative scheme.
-Given a current choice of linearisation point $\bm{v}$, we estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_v)$ to use as the new linearisation point.
-We iterate this procedure until a fixed point is identified within a chosen tolerance.
+where
+\begin{equation}
+\wh{\bm{\theta}}_{\bm{v}} = \argmax_{\bm{\theta}} \ol{p}_{\bm{v}} ( \bm{\theta} | \bm{y}),
+\end{equation}
+the posterior mode for $\bm{\theta}$, and
+\begin{equation}
+\wh{\bm{u}}_{\bm{v}} = \argmax_{\bm{u}} \ol{p}_{\bm{v}} (\bm{u} | \bm{y}, \wh{\bm{\theta}}_{\bm{v}}),
+\end{equation}
+the joint conditional posterior mode for $\bm{u}$.
+This functional defines a new linearisation point as the posterior mode for $\bm{\theta}$ under the linearised model configuration and the conditional mode for $\bm{u}$ given $\bm{\theta}$.
+These posteriors are computed automatically during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost for each fit of the linearised model.
 
-In practice the point $\bm{u}_*$ may be problematic and result in a linearised model configuration that is challenging to fit.
-Instead this point is considered only as an initial candidate point.
-The next linearisation point is chosen by setting $\bm{u}_{\alpha} = (1 - \alpha)\bm{v} + \alpha \bm{u}_*$ and finding the value of $\alpha$ that minimises $ \lVert \eta(\bm{u}_{\alpha}) - \bar{\eta}_{\bm{v}} \rVert$.
-This means that the new linearisation point remains close to a feasible region for fitting, assuming the current point $\bm{v}$ is reasonable.
-In practice, we have found that this requires some care when choosing starting values for the initial model fit.
-We estimate $\alpha$ using an approximate line search method that avoids many potentially expensive evaluations of the non-linear predictor.
-This iterative method is implemented in the R package \texttt{inlabru} \citep{bachl_inlabru_2019} which is both a wrapper and an extension to \texttt{R-INLA}.
+The `optimal' approximate model is defined as a fixed point of this functional.
+Given a current choice of linearisation point $\bm{v}$, we estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_{\bm{v}})$ to use as the new linearisation point.
+We iterate this procedure until a fixed point $\bm{v}^*$ is identified within a chosen tolerance such that $\bm{v}^* = f(\bar{p}_{\bm{v}^*})$.
 
@@
 \clearpage
-\bibliographystyle{stylefile16/rss}
+\bibliographystyle{rss}
 \bibliography{paper}

### 2024-10-24 paper.tex
good luck with the merge...

--- 2021-12-13 paper.tex
@@
 \author[Andrew E. Seaton {\it et al.}]{Andrew E. Seaton}
-\address{Centre for Research into Ecological \& Environmental Modelling and School of Mathematics \& Statistics, 
-	University of St Andrews, 
-	St Andrews,
+\address{School of Biodiversity, One Health and Veterinary Medicine, 
+	University of Glasgow, 
+	Glasgow,
 	UK}
-\email{aes22@st-andrews.ac.uk}
+\email{andrew.seaton.2@glasgow.ac.uk}
 
@@
 Distance sampling methods aim to estimate the size and spatial distribution of populations of wild animals.
-Here we present an analysis of point transect distance sampling data collected on an endangered tropical forest bird.
-We consider a workflow from model specification, inference method, model evaluation and communication of results, at each stage discussing the statistical challenges involved and our strategies to address them.
-We take a point process perspective on point transect data, viewing observations of birds as a thinned log-Gaussian Cox process with a spatially structured random effect describing the spatial distribution of animals specified via a computationally efficient Gaussian Markov random field approximation to Gaussian random field \citep{lindgren_explicit_2011}.
+In traditional analyses, spatial modelling is conducted in a second modelling stage, conditional on estimates of the detectability of animals.
+Here we present a one-stage analysis that simultaneously estimates the detectability and spatial distribution of an endangered tropical bird population.
+We take a point process perspective, modelling animal locations as a thinned log-Gaussian Cox process.
 
-We use a novel approximate Bayesian inference approach based on iterated fitting of latent Gaussian models using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009} to account for non-linear model components and jointly estimate detectability and spatial distribution, a process that is usually separated into two models.
-Outputs from the model are based on samples from the joint posterior of all model parameters and so naturally incorporate uncertainty due to detectability, a key advantage of the one-stage Bayesian approach.
-We discuss limitations of methods to represent uncertainty in mapped estimates of species distributions and advocate for the \textit{a priori} selection of relevant thresholds and uncertainty levels by using excursions methods \citep{bolin_excursion_2015}.
-
-We present an example of a computationally efficient one-stage Bayesian approach to analysing distance sampling data.
-However, the inference method presented is general, flexible and applicable in a wide range of areas in statistics beyond modelling species distributions.
+Since the thinning process is non-linear, we cannot use standard methods for inference.
+To address this we use a novel approximate Bayesian inference approach based on an iterative scheme of model fits using integrated nested Laplace approximations (INLA).
+This approach is general, flexible and applicable to a wide variety of non-linear model specifications.
+To illustrate the value of the one-stage approach we hypothesise potential model outputs that are relevant for conservation science.
+Since model outputs are based on samples from the joint posterior, they naturally account for the uncertainty in the detection process, a key advantage of the one-stage model.
 
@@
 
-\keywords{Distance sampling, Species distribution modelling, Estimating abundance, Stochastic partial differential equations, Integrated nested Laplace approximation, SPDE approach}
+\keywords{Distance sampling, Estimating abundance, Integrated nested Laplace approximation, Point processes, Spatial statistics, Species distribution modelling}
 
@@
 To address this we use a novel approximate method based on a first-order Taylor approximation of the non-linear predictor.
+We outline the approach here, full details can be found in \cite{lindgren_InlabruSoftwareFitting_2024}
 
@@
 \begin{equation}
-	f(\bar{p}_{\bm{v}}) = \left\{ \argmax_{\bm{u}_i} \: \bar{p}_{\bm{v}}(u_i | \bm{Y}), \quad i = 1, \ldots, n_{\bm{u}} \right\}
+	f(\bar{p}_{\bm{v}}) = \argmax_{\bm{u}} \bar{p}_{\bm{v}}(u | \bm{Y})
 \end{equation}
@@
 		\includegraphics[scale=0.525]{figures/N_posterior.png}
-		\caption{Posterior abundance.  Grey area marks two Monte-carlo standard errors above and below the mean}
+		\caption{Posterior abundance.  Grey area marks two Monte Carlo standard errors above and below the mean}
 		\label{fig:realized-abundance-posterior}

### 2024-10-26 paper.tex
bibtex edits

--- 2024-07-25 paper.tex
@@
 \usepackage{textcomp}    % for Hawaii characters
+\usepackage{url}
 
@@
 \author[Andrew E. Seaton {\it et al.}]{Andrew E. Seaton}
-\address{School of Mathematics \& Statistics, 
+\address{School of Biodiversity, One Health and Veterinary Medicine, 
 	University of Glasgow, 
@@
 However, the two methods are closely related and the GRF model component in our model plays a similar role to a penalised smoothing spline on space.
-\cite{yue_bayesian_2014} show how spline models can be viewed through the perspective of SPDEs and \cite{ miller_understanding_2019} do the reverse to show how to implement the Mat\'ern SPDE as a penalised spline model in \texttt{mgcv} \citep{wood_gam_2017}.
+\cite{yue_bayesian_2014} show how spline models can be viewed through the perspective of SPDEs and \cite{miller_UnderstandingStochasticPartial_2020} do the reverse to show how to implement the Mat\'ern SPDE as a penalised spline model in \texttt{mgcv} \citep{wood_gam_2017}.
 

### 2024-10-28 paper.tex
add bird data to data availability statement

@@
 
+\section*{Data Availability}
+
+Hawaii Akepa (Loxops coccineus) point-transect distance sampling data were provided by U.S.
+Fish and Wildlife Service, Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex.
+The 2002 survey data are available from the U.S.
+Geological Survey: \url{https://doi.org/10.5066/P9Q9UXMZ} \citep{camp_datarelease_2002}.
+The time data for 1987-2017 is also available  available from the U.S.
+Geological Survey: \url{https://doi.org/10.5066/P98IO297} \citep{camp_datarelease_all}
+
+Code to reproduce the figures and results in this paper are available in a git repository and can be accessed at [git repo link].
+
 \section*{Acknowledgements}
 
-The \hawaii \akepa{} data are provided by U.S.
-Fish and Wildlife Service, Hakalau Forest Unit National Wildlife Refuge.
 We are grateful to the many U.S.
-Fish and Wildlife Service staff and volunteers who helped collect the data.
-We thank Steve Buckland for helpful comments on an early version.
+Fish and Wildlife Service staff and volunteers who helped collect the data through the Hawaiian Forest Bird survey.
+We thank Steve Buckland for helpful comments on an early draft.
 Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S.

### 2024-10-28 paper.tex
add github to data availability statement

@@
 
-Code to reproduce the figures and results in this paper are available in a git repository and can be accessed at [git repo link].
+Code to reproduce the figures and results in this paper are available in a git repository and can be accessed at \url{https://github.com/ASeatonSpatial/point-transects-paper/}.
 
@@
 We are grateful to the many U.S.
-Fish and Wildlife Service staff and volunteers who helped collect the data through the Hawaiian Forest Bird survey.
+Fish and Wildlife Service staff and volunteers who helped collect the data through the \hawaii Forest Bird survey.
 We thank Steve Buckland for helpful comments on an early draft.

### 2024-10-29 analyses/compare_models.R
remove unnecessary files

@@ -1,115 +1,0 @@
-# compare two different fitted models
-# to investigate effects of different versions of
-# INLA
-
-library(INLA)
-library(inlabru)
-library(ggplot2)
-library(rgeos)
-library(sf)
-library(patchwork)
-
-theme_set(theme_classic())
-
-data_path = here::here("analyses", "data")
-study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
-samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
-mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
-
-#### Fitted model  ####
-model_path_new = here::here("analyses", "fitted_model.RDS")
-model_path_old = here::here("holding_bay", "akepa_2002_fitted_model_new_inlabru.RDS")
-fit_new = readRDS(model_path_new)
-fit_old = readRDS(model_path_old)
-
-summary(fit_new)
-summary(fit_old)
-
-# Prediction locations
-pxl = pixels(mesh, nx = 300, ny = 300, mask = study_area)
-
-# map units are per km^2 = 1,000,000 m^2
-# I want intensity per hectare = 10,000 m^2
-# so divide by 100
-pr.int_new <- predict(fit_new, pxl, ~ exp(grf + Intercept)/100)
-pr.int_old <- predict(fit_old, pxl, ~ exp(grf + Intercept)/100)
-
-# mean 
-vals = c(pr.int_new$mean, pr.int_old$mean)
-
-# things for colour scale 
-lower = min(vals)
-upper = max(vals)
-mean_breaks = c(lower, upper)
-mean_labels = c(signif(lower, digits = 3),
-                signif(upper, digits = 3))
-
-p_new = ggplot() +
-  gg(study_area) +
-  gg(pr.int_new["mean"]) +
-  scale_fill_viridis_c(breaks = mean_breaks,
-                       labels = mean_labels) +
-  xlab("Easting") +
-  ylab("Northing") +
-  ggtitle("New inlabru") + 
-  coord_equal() +
-  theme(legend.position = "bottom", legend.direction = "horizontal") +
-  guides(fill = guide_colourbar(title.vjust = 0.95,
-                                title.theme = element_text(size = 18),
-                                label.theme = element_text(size = 16)))
-
-p_old = ggplot() +
-  gg(study_area) +
-  gg(pr.int_old["mean"]) +
-  scale_fill_viridis_c(breaks = mean_breaks,
-                       labels = mean_labels) +
[…]

### 2024-10-29 analyses/non_paper_ideas/fit_hn_covariate.R

@@ -1,202 +1,0 @@
-library(inlabru)
-library(INLA)
-library(dplyr)
-
-init.tutorial()
-
-#### Load Data ####
-
-data_path = here::here("analyses", "data")
-realobs <- readRDS(here::here(data_path, "obs_extended_no_crs.RDS"))
-study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
-samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
-mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
-
-### Plot detections for different habitat types
-
-df = realobs@data
-df$HabCode = as.factor(df$HabCode)
-df$Stratum = as.factor(df$Stratum)
-
-### Plot detections by Stratum
-df %>%
-  group_by(Stratum) %>%
-  summarise(n = n())
-
-df %>%
-  ggplot() +
-  geom_histogram(aes(x = distance, y = after_stat(density)), bins = 15) +
-  facet_wrap(~ Stratum)
-
-# These look very slightly different.  Could try fitting a different detection
-# function to each
-
-# create binary variables
-realobs$stratum = ifelse(realobs$Stratum == "CF", 0, 1)
-samplers$stratum = ifelse(samplers$Stratum == "CF", 0, 1)
-
-realobs$habcode = ifelse(realobs$HabCode == "Mesic", 1, 0)
-samplers$habcode = ifelse(samplers$HabCode == "Mesic", 1, 0)
-
-#### Specify detection function  ####
-
-# x is a binary variable (2 level factor)
-# lsig_x ajdusts for when x == 1
-# lsig_int is for x == 0
-
-# Log half-normal
-log_hn = function(distance, x, lsig_int, lsig_x){
-  -0.5*(distance/exp(lsig_int + lsig_x*x))^2
-}
-
-# Half-normal
-hn <- function(distance, x, lsig_int, lsig_x) {
-  exp(log_hn(distance, x, lsig_int, lsig_x))
-}
-
-# Include r for switching to polar coords
-# 2*pi for not knowing angle theta
-dsamp = function(distance, x, lsig_int, lsig_x){
-  log(2*pi) + log(distance) + log_hn(distance, x, lsig_int, lsig_x)
-}
-
-
-# define distance domain and starting values
-# for detection scale param intercept
-W <- 58/1000   # transect radius
[…]

### 2024-10-29 analyses/non_paper_ideas/info.R

@@ -1,121 +1,0 @@
-# Evaluate fitted model
-library(INLA)
-library(inlabru)
-library(sp)
-library(raster)
-library(patchwork)
-library(rgeos)
-
-#### Fitted model  ####
-model_path = here::here("analyses", "akepa_2002_fitted_model_new_inlabru.RDS")
-data_path = here::here("analyses", "data")
-fit = readRDS(model_path)
-
-#### Where to save figures ####
-fig_path = here::here("figures")
-
-study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
-samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
-mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
-
-# generate samples from joint posterior of Intercept and grf st dev
-n.mc =  1000
-ps = generate(fit,
-              formula = ~ data.frame(Intercept_latent, Stdev_for_grf),
-              n.samples = n.mc)
-ps = do.call(rbind, ps)
-head(ps)
-range(ps$Intercept_latent)
-range(ps$Stdev_for_grf)
-hist(ps$Stdev_for_grf)
-
-e1 = exp(2*(ps[,1] + ps[,2]))
-e2 = exp(ps[,1] + ps[,2]/2)
-
-A0 = 1 / n.mc * sum(e1) - (1 / n.mc * sum(e2))^2
-A0
-
-# regular posterior variance
-pxl = pixels(mesh, mask = study_area)
-pred = predict(fit, pxl, ~ exp(Intercept + grf))
-A1 = pred$sd^2
-max(A1)
-
-# Info and plot it
-Info = 1 - A1 / A0
-pxl$Info_var = Info
-
-g1 = ggplot() +
-  gg(pxl, mapping = aes(fill = Info_var)) +
-  coord_equal() +
-  scale_fill_viridis_c(name = expression("I"["Var"])) +
-  theme_minimal() +
-  theme(legend.position = "right",
-        axis.text = element_text(size = 12),
-        axis.title = element_text(size = 12),
-        legend.title = element_text(size = 12, vjust = 0.75),
-        legend.text = element_text(size = 12),
-        legend.key.size = unit(2, "line"))
-
-g1
-
-# This looks like the inverse of the intensity map.  In the north
-# sampling effort is lower and almost no birds detected.
-# In the south sampling effort is higher and more birds detected
-
[…]

### 2024-10-29 analyses/non_paper_ideas/pair_correlation_function.R

@@ -1,38 +1,0 @@
-# Try to get the pair correlation function
-
-library(INLA)
-library(inlabru)
-
-# set ggplot theme
-theme_set(theme_minimal())
-
-fit = readRDS("akepa_2002_fitted_model.RDS")
-study_area = readRDS("data/study_area_extended_no_crs.RDS")
-samplers = readRDS("data/samplers_extended_no_crs.RDS")
-mesh = readRDS("data/mesh_extended_no_crs.RDS")
-
-summary(fit)
-
-# inla.matern.cov needs kappa
-rho.mean = fit$summary.hyperpar[1, "mean"]
-kappa.mean = sqrt(8) / rho.mean
-sigma.mean = fit$summary.hyperpar[2, "mean"]
-var.mean = sigma.mean^2
-
-h = seq(0.01, 20000, length.out = 250)
-
-llam_cov = var.mean * inla.matern.cov(nu = 1,
-                                      kappa = kappa.mean,
-                                      x = h,
-                                      d = 2)
-
-df = data.frame(h = h,
-                mat_cov = llam_cov)
-
-ggplot() +
-  geom_line(aes(x = h, y = mat_cov))
-
-lam_cov = exp(
-
-ggplot() +
-  geom_line(aes(x = h, y = pcf))

### 2024-10-29 analyses/non_paper_ideas/rick_things/gam_comparison.R

@@ -1,130 +1,0 @@
-# Evaluate fitted model
-library(INLA)
-library(inlabru)
-library(sp)
-
-# set ggplot theme
-theme_set(theme_minimal())
-
-set.seed(9701071)
-
-#### Fitted model  ####
-file_name = "akepa_2002_fitted_model.RDS"
-fit = readRDS(file_name)
-
-study_area = readRDS("data/study_area_extended_no_crs.RDS")
-samplers = readRDS("data/samplers_extended_no_crs.RDS")
-mesh = readRDS("data/mesh_extended_no_crs.RDS")
-
-# intensity
-
-# Rick's grid already clipped to study region
-# pred.grid <- expand.grid(EASTING=seq(from=255400, to=261200, by=200),
-#                          NORTHING=seq(from=2189000, to=2200800, by=200))
-# coordinates(pred.grid) = ~ EASTING + NORTHING
-# keep = over(pred.grid, study_area, byid = TRUE)
-# keep[is.na(keep)] = 0
-# keep = as.logical(keep)
-
-# using mgcv inSide() to match Rick's code
-# bnd_points = study_area@polygons[[1]]@Polygons[[1]]@coords
-# bnd_points = data.frame(EASTING = bnd_points[,1], NORTHING = bnd_points[,2])
-# EASTING = pred.grid@coords[,1]
-# NORTHING = pred.grid@coords[,2]
-# keep = mgcv::inSide(bnd_points, EASTING, NORTHING)
-# 
-# pred.grid.studyarea = pred.grid[keep,]
-
-# Rick's grid already clipped to study region
-pred.grid.studyarea = read.csv("data/rick-pred-grid.csv")
-coordinates(pred.grid.studyarea) = ~ EASTING + NORTHING
-
-# to get back as matrix for dsm do 
-# pred.grid.matrix = pred.grid.studyarea@coords
-
-ggplot() + gg(pred.grid.studyarea) + gg(study_area) + coord_equal()
-
-pred.pxl = SpatialPixels(pred.grid.studyarea)
-# pr.int <- predict(fit, pred.pxl, ~ exp(grf + Intercept))
-pr.int <- generate(fit, pred.pxl, ~ exp(grf + Intercept), n.samples = 1000)
-pr.int = do.call(cbind, pr.int)
-
-# p1 = ggplot() + 
-#   gg(study_area) +
[…]

### 2024-10-29 R/fit_model.R
reorganise files

@@ -4,10 +4,10 @@
 library(ggplot2)
 library(sp)
 library(inlabru)
-library(INLA)  
+library(INLA)
 
 #### Save Fitted Model Object? ####
-#### WARNING: THIS WILL OVERWRITE EXISTING FITTED MODEL OBJECT 
+#### WARNING: THIS WILL OVERWRITE EXISTING FITTED MODEL OBJECT
 to_save = TRUE
 model_path = here::here("analyses", "fitted_model.RDS")
 
@@ -19,7 +19,11 @@ data_path = here::here("analyses", "data")
 realobs <- readRDS(here::here(data_path, "obs_extended_no_crs.RDS"))
 study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
 samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
-mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
+mesh_path = here::here(data_path, "mesh_extended_no_crs.RDS")
+if (!file.exists(mesh_path)){
+  source(here::here("data", "create_extended_mesh.R"))
+}
+mesh = readRDS(mesh_path)
 
 #### Specify detection function  ####
 
@@ -63,7 +67,7 @@ fml <- coordinates + distance ~ grf +
   log(2*pi) +   # 2*pi offset for not knowing angle theta
   Intercept
 
-W <- 58/1000   # transect radius, units km 
+W <- 58/1000   # transect radius, units km
 distance_domain <- inla.mesh.1d(seq(.Machine$double.eps, W, length.out = 30))
 starting_values <- list(lsig = 3.36 - log(1000))
 

### 2024-10-29 R/data_scripts/process_AnalysisDataset2002.R
update obs + samplers to sf

@@ -1,15 +1,13 @@
-# convert AnalysisDataset2002.csv to appropriate SpatialPointDataFrame objects
+# process AnalysisDataset2002.csv
+# into points sf object
 
 library(dplyr)
-library(sp)
-library(inlabru)   # for quick plotting
+library(sf)
+library(ggplot2)
 
-data_path = here::here("analyses", "data")
+data_path = here::here("R", "data_raw")
 newdata = read.csv(here::here(data_path, "AnalysisDataset2002.csv"),
                    stringsAsFactors = FALSE)
-
-# old_obs = readRDS("obs_2002_no_crs.RDS")
-# old_samplers = readRDS("samplers_no_crs.RDS")
 
 # Stratum codes:
 # OF = open forest
@@ -24,7 +22,8 @@ newdata %>%
 samplers$weight = 1
 samplers$x = samplers$Easting/1000
 samplers$y = samplers$Northing/1000
-coordinates(samplers) = ~ x + y
+samplers_sf = st_as_sf(samplers,
+                       coords = c("x", "y"))
 
 obs = newdata %>%
   select(SampleLabel, Distance, Easting, Northing, Stratum, HabCode) %>%
@@ -35,13 +34,15 @@ obs$x = obs$Easting/1000
 obs$y = obs$Northing/1000
 obs$distance = obs$distance / 1000
 
-coordinates(obs) = ~ x + y
+obs_sf = st_as_sf(obs,
+                  coords = c("x", "y"))
 
 # check it looks okay
-
 ggplot() +
-  gg(samplers) +
-  gg(obs, col = "green")
+  geom_sf(data = samplers_sf) +
+  geom_sf(data = obs_sf, col = "green")
 
-saveRDS(samplers, file = here::here(data_path, "samplers_extended_no_crs.RDS"))
-saveRDS(obs, file = here::here(data_path, "obs_extended_no_crs.RDS"))
+# write to disk
+out_path = here::here("R", "data")
+saveRDS(samplers_sf, file = here::here(data_path, "samplers.RDS"))
+saveRDS(obs_sf, file = here::here(data_path, "obs.RDS"))

### 2024-10-29 R/data_scripts/process_AnalysisDataset2002.R
fix out_path

@@ -44,5 +44,5 @@ ggplot() +
 
 # write to disk
 out_path = here::here("R", "data")
-saveRDS(samplers_sf, file = here::here(data_path, "samplers.RDS"))
-saveRDS(obs_sf, file = here::here(data_path, "obs.RDS"))
+saveRDS(samplers_sf, file = here::here(out_path, "samplers.RDS"))
+saveRDS(obs_sf, file = here::here(out_path, "obs.RDS"))

### 2024-10-30 R/data_scripts/create_study_area.R
update study area to sf

@@ -1,36 +1,26 @@
-# create polygon defining study area
-# use simple convex hull and combindation with open forest polygon for now
+# create polygon object defining study area
 
-library(sp)
-library(rgeos)
-library(inlabru)
+library(sf)
+library(ggplot2)
 
-data_path = here::here("analyses", "data")
-samplers <- readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
-samplersnew = samplers[samplers$Stratum == "CF",]
-# old_study_area = readRDS("study_area_no_crs.RDS")
-# chull = gConvexHull(samplersnew)
-# chull = gBuffer(chull, width = 500, capStyle = "SQUARE")
-#
-# ggplot() +
-#   gg(old_study_area) +
-#   gg(chull) +
-#   gg(samplers) +
-#   gg(samplersnew, colour = "red") +
-#   coord_equal()
-#
-# study_area_new = gUnion(old_study_area, chull)
+data_raw_path = here::here("R", "data_raw")
+data_path = here::here("R", "data")
+samplers <- readRDS(here::here(data_path, "samplers.RDS"))
 
-# bnd.csv = read.csv("bnd_forest2.csv")
-bnd.csv = read.csv(here::here(data_path, "OpenClosedForestBoundary.csv"))
+bnd.csv = read.csv(here::here(data_raw_path, "OpenClosedForestBoundary.csv"))
 bnd.xy = data.frame(x = bnd.csv$EASTING / 1000,
                     y = bnd.csv$NORTHING / 1000)
-study_area_new <- SpatialPolygons(list(Polygons(list(Polygon(bnd.xy)), ID=1)))
 
+# convert to sf polygon
+bnd.poly <- st_polygon(list(as.matrix(bnd.xy)))
+bnd.sfc <- st_sfc(bnd.poly)
+bnd.sf <- st_sf(geometry = bnd.sfc)
+
+# plot to check
 ggplot() +
-  gg(study_area_new) +
-  gg(samplers) +
-  gg(samplersnew, colour = "red") +
-  coord_equal()
+  geom_sf(data = bnd.sf, alpha = 0.2) +
+  geom_sf(data = samplers)
 
-saveRDS(study_area_new, file = here::here(data_path, "study_area_extended_no_crs.RDS"))
+saveRDS(bnd.sf,
+        file = here::here(data_path,
+                          "study_area.RDS"))

### 2024-10-30 R/data_scripts/create_extended_mesh.R
update mesh creation to use fmesher

@@ -1,40 +1,0 @@
-# Load required packages
-library(inlabru)
-library(INLA)  # need testing version of inla
-library(rgeos)
-
-data_path = here::here("analyses", "data")
-point_transects <- readRDS(here::here(data_path, "samplers_no_crs.RDS"))
-study_area <- readRDS(here::here(data_path,"study_area_extended_no_crs.RDS"))
-outer_boundary = gBuffer(study_area, width = 20)
-
-# meshbuilder()
-
-## Generated by meshbuilder()
-
-## Build boundary information:
-## (fmesher supports SpatialPolygons, but this app is not (yet) intelligent enough for that.)
-boundary <- list(
-  as.inla.mesh.segment(study_area),
-  as.inla.mesh.segment(outer_boundary))
-
-## Build the mesh:
-mesh <- inla.mesh.2d(boundary=boundary,
-                     max.edge=c(0.2, 4.1),
-                     min.angle=c(27, 25),
-                     max.n=c(48000, 16000), ## Safeguard against large meshes.
-                     max.n.strict=c(128000, 128000), ## Don't build a huge mesh!
-                     cutoff=0.076, ## Filter away adjacent points.
-                     offset=c(0.1, 0.3)) ## Offset for extra boundaries, if needed.
-
-
-# plot whole mesh
-ggplot() + gg(mesh) + gg(point_transects, colour = "red") +
-  coord_equal()
-
-# zoom in a bit
-ggplot() + gg(mesh) +
-  coord_equal() + xlim(c(250, 267)) + ylim(c(2183, 2210))
-
-# save the mesh
-saveRDS(object = mesh, file = here::here(data_path,"mesh_extended_no_crs.RDS"))

### 2024-10-30 R/data_scripts/create_mesh.R

@@ -1,0 +1,47 @@
+# Load required packages
+library(inlabru)
+library(INLA)  # need testing version of inla
+library(sf)
+library(ggplot2)
+
+data_path = here::here("R", "data")
+point_transects <- readRDS(here::here(data_path, "samplers.RDS"))
+study_area <- readRDS(here::here(data_path,"study_area.RDS"))
+outer_boundary = st_buffer(study_area, dist = 10)
+
+# meshbuilder()
+
+## Generated by meshbuilder()
+
+## Build boundary information:
+## (fmesher supports SpatialPolygons, but this app is not (yet) intelligent enough for that.)
+boundary <- list(
+  fm_as_segm(study_area),
+  fm_as_segm(outer_boundary)
+)
+
+## Build the mesh:
+mesh <- fm_mesh_2d_inla(boundary=boundary,
+                        max.edge=c(0.2, 4.1),
+                        min.angle=c(27, 25),
+                        max.n=c(48000, 16000), ## Safeguard against large meshes.
+                        max.n.strict=c(128000, 128000), ## Don't build a huge mesh!
+                        cutoff=0.076, ## Filter away adjacent points.
+                        offset=c(0.1, 0.3)) ## Offset for extra boundaries, if needed.
+
+
+# plot whole mesh
+ggplot() +
+  gg(mesh) +
+  geom_sf(data = point_transects,
+          colour = "red")
+
+# zoom in a bit
+ggplot() +
+  gg(mesh) +
+  xlim(c(250, 267)) +
+  ylim(c(2183, 2210))
+
+# save the mesh
+saveRDS(object = mesh,
+        file = here::here(data_path,"mesh.RDS"))

### 2024-10-30 R/fit_model.R
update fit_model.R

@@ -2,27 +2,26 @@
 
 # Load required packages
 library(ggplot2)
-library(sp)
+library(sf)
 library(inlabru)
 library(INLA)
 
 #### Save Fitted Model Object? ####
 #### WARNING: THIS WILL OVERWRITE EXISTING FITTED MODEL OBJECT
 to_save = TRUE
-model_path = here::here("analyses", "fitted_model.RDS")
+model_path = here::here("R",
+                        "fitted_models",
+                        "fitted_model.RDS")
 
 #### Produce intermediate plots? ####
-to_plot = FALSE
+to_plot = TRUE
 
 #### Load Data ####
-data_path = here::here("analyses", "data")
-realobs <- readRDS(here::here(data_path, "obs_extended_no_crs.RDS"))
-study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
-samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
-mesh_path = here::here(data_path, "mesh_extended_no_crs.RDS")
-if (!file.exists(mesh_path)){
-  source(here::here("data", "create_extended_mesh.R"))
-}
+data_path = here::here("R", "data")
+obs <- readRDS(here::here(data_path, "obs.RDS"))
+study_area = readRDS(here::here(data_path, "study_area.RDS"))
+samplers = readRDS(here::here(data_path, "samplers.RDS"))
+mesh_path = here::here(data_path, "mesh.RDS")
 mesh = readRDS(mesh_path)
 
 #### Specify detection function  ####
@@ -44,10 +43,9 @@ dsamp = function(distance, lsig){
 if (to_plot){
   g1 <- ggplot() +
     gg(mesh) +
-    gg(study_area) +
-    gg(samplers) +
-    gg(realobs, colour = "green") +
-    coord_equal()
+    geom_sf(data = study_area) +
+    geom_sf(data = samplers) +
+    geom_sf(data = obs, colour = "green")
   g1
 }
 
@@ -58,30 +56,33 @@ matern <- inla.spde2.pcmatern(mesh,
                               prior.sigma = c(2, 0.01),
                               prior.range = c(300/1000, 0.01))
 
-cmp <- ~ grf(main = coordinates, model = matern) +
+cmp <- ~ grf(main = geometry, model = matern) +
   lsig(1) + Intercept(1)
 
 # Predictor formula:
-fml <- coordinates + distance ~ grf +
[…]

### 2024-10-30 R/eval_model.R
update figures using latest fit

@@ -5,27 +5,27 @@ library(inlabru)
 library(ggplot2)
 library(scales)
 library(cowplot)
-library(rgeos)
-
-# set ggplot theme
-theme_set(theme_minimal())
+library(sf)
 
 set.seed(9701071)
 
 #### Fitted model  ####
-model_path = here::here("analyses", "fitted_model.RDS")
+model_path = here::here("R",
+                        "fitted_models",
+                        "fitted_model.RDS")
 fit = readRDS(model_path)
 
 #### Other things ####
-data_path = here::here("analyses", "data")
-study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
-samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
-mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
+data_path = here::here("R", "data")
+study_area = readRDS(here::here(data_path, "study_area.RDS"))
+samplers = readRDS(here::here(data_path, "samplers.RDS"))
+mesh = readRDS(here::here(data_path, "mesh.RDS"))
 
 #### Where to save figures ####
 fig_path = here::here("figures")
 
 #### ggplot theme ####
+# theme_set(theme_minimal())
 theme_set(theme_classic())
 
 #### Specify detection functions  ####
@@ -42,7 +42,6 @@ hn <- function(distance, lsig) exp(log_hn(distance, lsig))
 dsamp = function(distance, lsig){
   log(distance) + log_hn(distance, lsig)
 }
-
 
 #### Model evaluation ####
 summary(fit)
@@ -87,29 +86,38 @@ dev.off()
 ##### Posterior intensity plots #####
 
 # Prediction locations
-pxl = pixels(mesh, nx = 300, ny = 300, mask = study_area)
-nrow(pxl@coords)
+pxl = fm_pixels(mesh,
+                dims = c(300, 300),
+                mask = study_area)
 
 # map units are per km^2 = 1,000,000 m^2
 # I want intensity per hectare = 10,000 m^2
 # so divide by 100
-pr.int <- predict(fit, pxl, ~ exp(grf + Intercept)/100)
+pr.int <- predict(fit, pxl, ~ exp(grf + Intercept)/100,
+                  n.samples = 1000)
+
+# add x and y columns for geom_tile
+pr.int.coords = st_coordinates(pr.int)
+pr.int$x = pr.int.coords[,1]
[…]

### 2024-10-30 paper.tex
update fake abundance estimates

@@
 These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.
-To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they are intensity functions and integrate them to obtain an expected abundance estimate of approximately 2,800 for the 0.025 quantile map and 10,200 for the 0.975 quantile map.
+To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they are intensity functions and integrate them to obtain an expected abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.
 A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance.

### 2024-10-30 paper.tex
update pairwise distances

@@
 		\includegraphics[scale=0.6]{figures/post_pp_distances.png}
-		\caption{Boxplots of the frequency of pairwise distances (tick mark values for end of each bin) based on 1,000 simulated datasets.
+		\caption{Boxplots of the frequency of pairwise distances (tick mark values for end of each bin) based on 500 simulated datasets from the posterior predictive distribution.
 Red circles indicate the frequency of pairwise distances in the observed data.}

### 2024-10-30 R/eval_spde.R

@@ -1,10 +1,9 @@
-# Summarise posterior sampled point patterns 
+# Summarise posterior sampled point patterns
 # using pairwise distances
 
 library(INLA)
 library(inlabru)
 library(ggplot2)
-library(rgeos)
 library(sf)
 
 # set ggplot theme
@@ -12,23 +11,24 @@ theme_set(theme_minimal())
 set.seed(9701071)
 
 #### Fitted model  ####
-model_path = here::here("analyses", "fitted_model.RDS")
-# model_path = here::here("holding_bay", "akepa_2002_fitted_model_new_inlabru.RDS")
+model_path = here::here("R",
+                        "fitted_models",
+                        "fitted_model.RDS")
 fit = readRDS(model_path)
 
-data_path = here::here("analyses", "data")
-obs <- readRDS(here::here(data_path, "obs_extended_no_crs.RDS"))
-# obs <- readRDS(here::here(data_path, "obs_2002_no_crs.RDS"))
-study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
-samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
+data_path = here::here("R", "data")
+obs <- readRDS(here::here(data_path, "obs.RDS"))
+study_area = readRDS(here::here(data_path, "study_area.RDS"))
+samplers = readRDS(here::here(data_path, "samplers.RDS"))
+mesh = readRDS(here::here(data_path, "mesh.RDS"))
 
-# polygon samplers 
+# polygon samplers
+samplers$ID = 1:nrow(samplers)
 W = 58/1000
-samplers_buffered = gBuffer(samplers, width = W, byid = TRUE)
-mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
+samplers_buffered = st_buffer(samplers, dist = W)
 
 ### Generate posterior point patterns
-n.pp = 100
+n.pp = 500
 
 # simulate intensities and detection functions
 # each column of llam is a realisation of the log-intensity at each mesh node
@@ -64,37 +64,26 @@ hn <- function(distance, lsig) exp(log_hn(distance, lsig))
 ## Build boundary information:
 ## (fmesher supports SpatialPolygons, but this app is not (yet) intelligent enough for that.)
 boundary <- list(
-  as.inla.mesh.segment(study_area),
+  fm_as_segm(study_area),
   NULL)
 
 ## Build the mesh:
[…]

### 2024-10-30 R/excursions.R
update excursions figures

@@ -4,8 +4,8 @@ library(inlabru)
 library(ggplot2)
 library(excursions)
 library(cowplot)
-source("gg.R")   # small edit of gg.SpatialPixelsDataFrame
-
+library(sf)
+source(here::here("R", "gg.R"))   # small edit of gg.SpatialPixelsDataFrame
 
 # set ggplot theme
 theme_set(theme_minimal())
@@ -13,29 +13,31 @@ theme_set(theme_minimal())
 set.seed(1525)
 
 #### Fitted model  ####
-model_path = here::here("analyses", "fitted_model.RDS")
+model_path = here::here("R",
+                        "fitted_models",
+                        "fitted_model.RDS")
 fit = readRDS(model_path)
 
 #### Other things ####
-data_path = here::here("analyses", "data")
-study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
-samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
-mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
+data_path = here::here("R", "data")
+study_area = readRDS(here::here(data_path, "study_area.RDS"))
+samplers = readRDS(here::here(data_path, "samplers.RDS"))
+mesh = readRDS(here::here(data_path, "mesh.RDS"))
 
 #### Where to save figures ####
 fig_path = here::here("figures")
 
 #### monte-carlo samples of intensity ####
 
-# Does excursions work with an inlabru object?
-
-# There must be a way to get the actual values from
-# predict with n.samples > 1 ?
-
-# yes update this to use generate
-
 n.mc = 500     # lower this when fiddling with things
-pxl = pixels(mesh, nx = 300, ny = 300, mask = study_area)
+n.mc = 25     # lower this when fiddling with things
+pxl = fm_pixels(mesh,
+                dims = c(50, 50),
+                mask = study_area)
+
+pxl.coords = st_coordinates(pxl)
+pxl$x = pxl.coords[,1]
+pxl$y = pxl.coords[,2]
 
 X = generate(fit,
              pxl,
@@ -43,8 +45,6 @@ X = generate(fit,
              n.samples = n.mc)
 
 #### excursions ####
-# Note:  I use excursions.mc() because I am not sure
-# if excursions() knows how to work with fitted bru objects
 
 # re-scale X (per km) to per hectare:
 Xhec = X/100
[…]

### 2024-10-30 R/posterior_N.R
update abundance figure

@@ -1,252 +1,64 @@
-# Investigate effet of different integration schemes
-# on the posterior for abundance
+# Posterior abundance plot
 
 library(INLA)
 library(inlabru)
 library(ggplot2)
-library(rgeos)
+library(sf)
 library(patchwork)
 library(dplyr)
-source(here::here("analyses", "gg.R"))   # some small edits to inlabru gg methods
+source(here::here("R", "gg.R"))   # some small edits to inlabru gg methods
 
 # set ggplot theme
 theme_set(theme_minimal())
 
 #### Fitted model  ####
-model_path = here::here("analyses", "fitted_model.RDS")
+model_path = here::here("R",
+                        "fitted_models",
+                        "fitted_model.RDS")
 fit = readRDS(model_path)
 
 #### Other things ####
-data_path = here::here("analyses", "data")
-study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
-samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
-obs = readRDS(here::here(data_path, "obs_extended_no_crs.RDS"))
-mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
+data_path = here::here("R", "data")
+study_area = readRDS(here::here(data_path, "study_area.RDS"))
+samplers = readRDS(here::here(data_path, "samplers.RDS"))
+obs = readRDS(here::here(data_path, "obs.RDS"))
+mesh = readRDS(here::here(data_path, "mesh.RDS"))
 
 #### Where to save figures ####
 fig_path = here::here("figures")
 
-# Total surveyed region as % of study area?
-W = 58 / 1000
-nrow(samplers@data)*pi*W^2  / gArea(study_area) * 100
+# integration points
+ips = fm_int(domain = mesh,
+             samplers = study_area)
 
-# Buffer samplers points df to create transects
-samplers_buffered = gBuffer(samplers, width = W, byid = TRUE)
-
-# Did this buffer produce something of the right area?
-abs(gArea(samplers_buffered) - nrow(samplers@coords)*pi*W^2) / nrow(samplers@data)*pi*W^2 * 100
-# seems pretty close.
-
-# Remove transects from study area
-study_area_no_samplers = gDifference(study_area, samplers_buffered)
-abs(gArea(study_area) - (gArea(study_area_no_samplers) + gArea(samplers_buffered))) / gArea(study_area) * 100
[…]

### 2024-10-30 paper.tex
update IUCN ref

@@
 Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the HFBS \citep{scott_HFBS_1986}.
-Information from the HFBS is used to update the listing and delisting of endangered species, and establish preserves that coincided with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau).
-Hakalau is the first wildlife refuge established with the primary purpose to protect and manage native forests for the conservation of threatened and endangered bird and plant species.
+Information from the HFBS is used to update the International Union for Conservation of Nature redlist \citep{iucn_redlist_2022}, where the \akepa{} is currently listed as endangered.
+HFBS data is also used to establish preserves that coincide with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau).
+Hakalau is the first wildlife refuge established in \hawaii{} with the primary purpose to protect and manage native forests for the conservation of threatened and endangered bird and plant species.
 Due to its broad-scale coverage and robust design, the HFBS has become an invaluable dataset used to determine changes in bird species distributions, population sizes and trends in density patterns over time.

### 2024-10-30 R/session_info.R
re-run session_info.R

@@ -1,15 +1,15 @@
 # session info for results in paper
+
 library(ggplot2)
 library(sp)
 library(inlabru)
-library(INLA)  
+library(INLA)
 library(scales)
 library(cowplot)
-library(rgeos)
 library(sf)
 library(patchwork)
 library(dplyr)
 library(excursions)
 
 writeLines(capture.output(devtools::session_info()),
-           here::here("analyses", "session_info.txt"))
+           here::here("R", "session_info.txt"))

### 2024-10-30 paper.tex
start converting figures to pdf + final tweaks

@@
 \usepackage{url}
+
+% to find textwidth
+%\usepackage{layouts}
 
@@
 \section{Introduction}
+
+Text width here we go:
+%\printinunitsof{cm}\prntlen{\textwidth}
 
@@
 	\centering
-	\includegraphics[scale=0.5]{figures/study_area_design.png}
-	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.  Eastings and Northings are in kilometres.}
+	\includegraphics{figures/study_area_design.pdf}
+	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.
+The red outline marks the study area boundary as defined in \cite{camp_dsm_2020}.}
 	\label{fig:2002studyareapointspt}
@@
 	\begin{center}
-		\includegraphics[scale=0.525]{figures/intensity_mean_cv_sd.png}
+		\includegraphics[width=\textwidth]{figures/intensity_mean_cv_sd.pdf}
 		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$.  The colour scale is different for each panel.}
@@
 	\begin{center}
-		\includegraphics[scale=0.525]{figures/detfn_and_matern.png}
+		\includegraphics[width=\textwidth]{figures/detfn_and_matern.pdf}
 		\caption{A) Posterior half-normal detection function.

### 2024-10-30 R/eval_model.R

@@ -9,6 +9,13 @@ library(sf)
 
 set.seed(9701071)
 
+# statsoc doc class textwidth in cm
+tw = 14.69785
+twi = 5.7865551181  # pdf() needs inches
+
+# axis title text size
+ax.size = 12
+
 #### Fitted model  ####
 model_path = here::here("R",
                         "fitted_models",
@@ -58,10 +65,10 @@ hnpred <- predict(fit, distdf, ~ hn(distance, lsig + log(1000)),
 
 ghn = ggplot() +
         gg(hnpred) +
-        ylab("probability of detection\n") +
+        ylab("detection probability\n") +
         xlab("\ndistance (m)") +
-        theme(axis.title = element_text(size = 16),
-        axis.text = element_text(size = 16))
+        theme(axis.title = element_text(size = ax.size),
+        axis.text = element_text(size = ax.size))
 ghn
 
 ##### Posterior Matern plot ####
@@ -70,13 +77,13 @@ gmat = ggplot() +
   gg(blah) +
   ylab("correlation\n") +
   xlab("\ndistance (km)") +
-  theme(axis.title = element_text(size = 16, vjust = 0.3),
-        axis.text = element_text(size = 16))
+  theme(axis.title = element_text(size = ax.size, vjust = 0.3),
+        axis.text = element_text(size = ax.size))
 
 gmat
 
-png(filename = here::here(fig_path, "detfn_and_matern.png"),
-    width = 10, height = 5, units = "in", res = 100)
+pdf(file = here::here(fig_path, "detfn_and_matern.pdf"),
+    width = twi, height = twi/2)
 plot_grid(NULL, ghn, NULL, gmat, NULL,
           labels = c("", "A", "", "B", ""),
           rel_widths = c(0.05, 1, 0.15, 1, 0.05),
@@ -120,9 +127,9 @@ p1 = ggplot() +
   ylab("Northing") +
   theme_void() +
   theme(legend.position = "bottom", legend.direction = "horizontal") +
-  guides(fill = guide_colourbar(title.vjust = 0.95,
-                                title.theme = element_text(size = 18),
-                                label.theme = element_text(size = 16)))
+  guides(fill = guide_colourbar(title.vjust = 0.85,
+                                title.theme = element_text(size = ax.size),
+                                label.theme = element_text(size = ax.size)))
 
 p1
 
@@ -146,9 +153,9 @@ p2 = ggplot() +
   ylab("Northing") +
   theme_void() +
[…]

### 2024-10-30 R/make_study_area_fig.R

@@ -1,0 +1,42 @@
+# Evaluate fitted model
+library(INLA)
+library(inlabru)
+library(sf)
+library(ggplot2)
+
+theme_set(theme_minimal())
+
+# statsoc doc class textwidth in cm
+tw = 14.69785
+twi = 5.7865551181  # pdf() needs inches
+ax.size = 12 # axis text size
+
+data_path = here::here("R", "data")
+study_area = readRDS(here::here(data_path, "study_area.RDS"))
+samplers = readRDS(here::here(data_path, "samplers.RDS"))
+study_area_boundaries <- sf::st_cast(study_area, "LINESTRING")
+
+pdf(file = here::here("figures", "study_area_design.pdf"),
+    width = twi/2, height = twi/2)
+
+ggplot() +
+  geom_sf(data = study_area_boundaries,
+          # fill = NA,
+          size = 2,
+          colour = "red") +
+  geom_sf(data = samplers) +
+  xlab("Easting (km)") +
+  ylab("Northing (km)") +
+  scale_x_continuous(limits = c(254, 262),
+                     breaks = seq(254, 262, by = 2)) +
+  theme(
+    legend.position = "bottom",
+    axis.text = element_text(size = ax.size),
+    axis.text.x = element_text(angle = -45, vjust = 0, size = ax.size), # Rotate x-axis labels
+    axis.title = element_text(size = ax.size),
+    legend.title = element_text(size = ax.size, vjust = 0.75),
+    legend.text = element_text(size = ax.size),
+    legend.key.size = unit(2, "line")
+  )
+
+dev.off()

### 2024-10-30 figures/make_study_area_fig.R

@@ -1,28 +1,0 @@
-# Evaluate fitted model
-library(INLA)
-library(inlabru)
-
-# set ggplot theme
-theme_set(theme_minimal())
-
-study_area = readRDS("../analyses/data/study_area_extended_no_crs.RDS")
-samplers = readRDS("../analyses/data/samplers_extended_no_crs.RDS")
-
-png(filename = "study_area_design.png",
-    width = 10, height = 6, units = "in", res = 100)
-
-ggplot() +
-  gg(study_area) +
-  gg(samplers) +
-  coord_equal() +
-  xlab("Easting") +
-  ylab("Northing") +
-  theme(legend.position = "bottom",
-        axis.text = element_text(size = 12),
-        axis.title = element_text(size = 12),
-        legend.title = element_text(size = 12, vjust = 0.75),
-        legend.text = element_text(size = 12),
-        legend.key.size = unit(2, "line"))
-
-
-dev.off()

### 2024-10-30 R/eval_model.R
fix map legend clipping

@@ -120,16 +120,19 @@ p1 = ggplot() +
   geom_tile(data = pr.int,
             mappin = aes(x = x,
                          y = y,
-                         fill = mean)) +
+                         fill = mean,
+                         colour = mean)) +
   scale_fill_viridis_c(breaks = mean_breaks,
                        labels = mean_labels) +
+  scale_colour_viridis_c(breaks = mean_breaks,
+                         labels = mean_labels) +
   xlab("Easting") +
   ylab("Northing") +
   theme_void() +
   theme(legend.position = "bottom", legend.direction = "horizontal") +
   guides(fill = guide_colourbar(title.vjust = 0.85,
-                                title.theme = element_text(size = ax.size),
-                                label.theme = element_text(size = ax.size)))
+                                title.theme = element_text(size = ax.size-2),
+                                label.theme = element_text(size = ax.size-2)))
 
 p1
 
@@ -146,16 +149,19 @@ p2 = ggplot() +
   geom_tile(data = pr.int,
             mapping = aes(x = x,
                          y = y,
-                         fill = cv)) +
+                         fill = cv,
+                         colour = cv)) +
   scale_fill_viridis_c(breaks = cv_breaks,
+                       labels = cv_labels) +
+  scale_colour_viridis_c(breaks = cv_breaks,
                        labels = cv_labels) +
   xlab("Easting") +
   ylab("Northing") +
   theme_void() +
   theme(legend.position = "bottom", legend.direction = "horizontal") +
   guides(fill = guide_colourbar(title.vjust = 0.85,
-                                title.theme = element_text(size = ax.size),
-                                label.theme = element_text(size = ax.size)))
+                                title.theme = element_text(size = ax.size-2),
+                                label.theme = element_text(size = ax.size-2)))
 
 p2
 
@@ -164,26 +170,62 @@ lower = min(pr.int$sd)
 upper = max(pr.int$sd)
 sd_breaks = c(lower, upper)
 sd_labels = format(c(signif(lower, digits = 3),
-                     signif(upper, digits = 3)),
-                   scientific = TRUE)
+                     signif(upper, digits = 3)))
 
 p3 = ggplot() +
   geom_sf(data = study_area) +
   geom_tile(data = pr.int,
             mapping = aes(x = x,
                          y = y,
[…]

### 2024-10-30 R/eval_model.R
intensity figure tweaks + pdf conversion

@@ -252,10 +252,16 @@ p2 = ggplot() +
   geom_tile(data = pr.int,
             mapping = aes(x = x,
                          y = y,
-                         fill = q0.025)) +
-  scale_fill_viridis_c(limits = q_breaks,
+                         fill = q0.025,
+                         colour = q0.025)) +
+  scale_fill_viridis_c("",
+                       limits = q_breaks,
                        breaks = q_breaks,
                        labels = q_labels)+
+  scale_colour_viridis_c(limits = q_breaks,
+                         breaks = q_breaks,
+                         labels = q_labels,
+                         guide = "none")+
   xlab("Easting") +
   ylab("Northing") +
   theme_void() +
@@ -272,10 +278,16 @@ p3 = ggplot() +
   geom_tile(data = pr.int,
             mapping = aes(x = x,
                          y = y,
-                         fill = q0.975)) +
-  scale_fill_viridis_c(limits = q_breaks,
+                         fill = q0.975,
+                         colour = q0.975)) +
+  scale_fill_viridis_c("",
+                       limits = q_breaks,
                        breaks = q_breaks,
                        labels = q_labels)+
+  scale_colour_viridis_c(limits = q_breaks,
+                         breaks = q_breaks,
+                         labels = q_labels,
+                         guide = "none")+
   xlab("Easting") +
   ylab("Northing") +
   theme_void() +
@@ -287,12 +299,39 @@ p3 = ggplot() +
 
 p3
 
-png(filename = here::here(fig_path, "intensity_quantiles.png"),
-    width = 10, height = 5, units = "in", res = 100)
-plot_grid(NULL, p2, NULL, p3, NULL,
-          labels = c("", "A", "", "B", ""),
-          rel_widths = c(0.05, 1, 0.15, 1, 0.05),
-          ncol = 5)
+p2 = p2 +
+  theme(legend.margin = margin(t = 0, r = 10, b = 0, l = 10)) +
+  guides(fill = guide_colourbar(
+    title.vjust = 0.85,
+    title.theme = element_text(size = ax.size-2),
+    label.theme = element_text(size = ax.size-2),
+    barwidth = unit(2, "cm"),  # Consistent width for the color bar
+    barheight = unit(0.5, "cm") # Consistent height for the color bar
+  ))
+
+p3 = p3 +
[…]

### 2024-10-31 paper.tex
intensity figure tweaks + pdf conversion
figures converted to pdf

@@
 	\begin{center}
-		\includegraphics[scale=0.525]{figures/intensity_quantiles.png}
-		\caption{The predicted posterior intensity quantiles:  A) 0.025 quantile B) 0.975 quantile.  Both plots use the same colour scale.}
+		\includegraphics{figures/intensity_quantiles.pdf}
+		\caption{Pointwise posterior quantiles for the intensity:  A) 0.025 quantile B) 0.975 quantile}
 		\label{fig:intensity-quantiles}
@@
 	\begin{center}
-		\includegraphics[scale=0.6]{figures/post_pp_distances.png}
-		\caption{Boxplots of the frequency of pairwise distances (tick mark values for end of each bin) based on 500 simulated datasets from the posterior predictive distribution.
-Red circles indicate the frequency of pairwise distances in the observed data.}
+		\includegraphics{figures/post_pp_distances.pdf}
+		\caption{Boxplots of the frequency of pairwise distances based on 500 simulated datasets from the posterior predictive distribution.
+Axis labels mark the largest value in each bin.
+Red circles indicate the observed frequency of pairwise distances within each bin.}
 		\label{fig:post-pp-distances}
@@
 	\begin{center}
-		\includegraphics[scale=0.525]{figures/N_posterior.png}
-		\caption{Posterior abundance.  Grey area marks two Monte Carlo standard errors above and below the mean}
+		\includegraphics{figures/N_posterior.pdf}
+		\caption{Posterior abundance.  Grey area marks two Monte Carlo standard errors above and below the mean.}
 		\label{fig:realized-abundance-posterior}
@@
 \begin{figure}[!htb]
-	\includegraphics[scale=0.525]{figures/intensity_realized.png}
+	\includegraphics{figures/intensity_realized.pdf}
 	\caption{Three realizations of the posterior intensity field}
@@
 \begin{figure}[!htb]
-	\includegraphics[scale=0.5]{figures/excursions.png}
-	\caption{A) The positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.
+	\includegraphics{figures/excursions.pdf}
+	\caption{A) The blue area is the positive excursion set with a level corresponding to 1 bird per hectare and probability 0.95.
 B) The positive excursion function with a level corresponding to 1 bird per hectare.

### 2024-10-31 R/eval_model.R

@@ -343,7 +343,7 @@ sum(pr.int$q0.975*cell_area)
 
 ### plot three realisations of posterior intensity field ####
 
-set.seed(1982139)
+set.seed(19822139)
 draw1 = predict(fit, pxl, ~ exp(grf + Intercept), n.samples = 1)
 draw2 = predict(fit, pxl, ~ exp(grf + Intercept), n.samples = 1)
 draw3 = predict(fit, pxl, ~ exp(grf + Intercept), n.samples = 1)
@@ -382,7 +382,8 @@ p2 = ggplot() +
   geom_tile(data = draw2,
             mapping = aes(x = x,
                          y = y,
-                         fill = mean)) +
+                         fill = mean,
+                         colour = mean)) +
   viridisscale +
   scale_colour_viridis_c("",
                          limits = range(draw_all),
@@ -397,7 +398,8 @@ p3 = ggplot() +
   geom_tile(data = draw3,
             mapping = aes(x = x,
                          y = y,
-                         fill = mean)) +
+                         fill = mean,
+                         colour = mean)) +
   viridisscale +
   scale_colour_viridis_c("",
                          limits = range(draw_all),

### 2024-10-31 R/eval_spde.R

@@ -6,6 +6,10 @@ library(inlabru)
 library(ggplot2)
 library(sf)
 
+# statsoc doc class textwidth in cm
+tw = 14.69785
+twi = 5.7865551181  # pdf() needs inches
+
 # set ggplot theme
 theme_set(theme_minimal())
 set.seed(9701071)
@@ -138,8 +142,8 @@ obs.ds.vec = as.numeric(obs.ds)
 count.obs = hist(obs.ds.vec, breaks = breaks, plot = FALSE)$counts
 
 # plot
-png(filename = here::here("figures", "post_pp_distances.png"),
-    width = 7, height = 4, units = "in", res = 100)
+pdf(file = here::here("figures", "post_pp_distances.pdf"),
+    width = twi, height = twi/2)
 
 boxplot(counts,
         xlab = "distance (km)",
@@ -151,4 +155,5 @@ points(1:(length(breaks)-1), count.obs, col = "red")
 # xticks = floor(breaks[-1] - (breaks[2] - breaks[1])/2)
 xticks = breaks[-1]
 axis(1, at = 1:(length(breaks)-1), labels = xticks)
+
 dev.off()

### 2024-10-31 R/excursions.R

@@ -7,6 +7,9 @@ library(cowplot)
 library(sf)
 source(here::here("R", "gg.R"))   # small edit of gg.SpatialPixelsDataFrame
 
+# statsoc doc class textwidth in cm
+tw = 14.69785
+twi = 5.7865551181  # pdf() needs inches
 # set ggplot theme
 theme_set(theme_minimal())
 
@@ -91,9 +94,10 @@ p1 = ggplot() +
                           fill = E)) +
   # ggtitle("Excursion set for > 1 bird per hectare,\nalpha = 0.05") +
   scale_fill_manual(values = c("blue"),
-                    labels = c("Excursion set")) +
+                    labels = c("Excursion set"),
+                    guide = "none") +
   theme_void() +
-  theme(legend.position = "bottom",
+  theme(legend.position = "none",
         legend.direction = "vertical",
         plot.margin = margin(0.1, 0, 0.3, 0, "in"),
         legend.box.margin = margin(-4,0,0,0)) +
@@ -106,8 +110,8 @@ p1
 # ggsave(filename = "../figures/excursion_set.png",
 #        width = 5, height = 6, units = "in")
 
-png(filename = here::here(fig_path, "excursions.png"),
-    width = 10, height = 5, units = "in", res = 100)
+pdf(file = here::here(fig_path, "excursions.pdf"),
+    width = twi, height = twi/2)
 plot_grid(NULL, p1, NULL, p2, NULL,
           labels = c("", "A", "", "B", ""),
           rel_widths = c(0.05, 1, 0.15, 1, 0.05),

### 2024-10-31 R/posterior_N.R

@@ -8,6 +8,9 @@ library(patchwork)
 library(dplyr)
 source(here::here("R", "gg.R"))   # some small edits to inlabru gg methods
 
+# statsoc doc class textwidth in cm
+tw = 14.69785
+twi = 5.7865551181  # pdf() needs inches
 # set ggplot theme
 theme_set(theme_minimal())
 
@@ -65,10 +68,11 @@ N_post %>%
                   ymin = mean - 2 * sd / sqrt(n.mc),
                   ymax = mean + 2 * sd / sqrt(n.mc)),
               alpha = 0.2) +
-  ylab("p(N)")
+  xlab("Abundance") +
+  ylab("posterior probabililty")
 
-ggsave(filename = here::here(fig_path, "N_posterior.png"),
-       width = 5, height = 5, units = "in")
+ggsave(filename = here::here(fig_path, "N_posterior.pdf"),
+       width = twi/2, height = twi/2, units = "in")
 
 Nweight = (N_max - N_min) / length(N_seq)
 sum(Nweight * N_post$mean) # should be close to 1
@@ -94,6 +98,6 @@ N_post %>%
   ylab("posterior probability") +
   xlab("akepa density per hectare")
 
-ggsave(filename = here::here(fig_path, "density_posterior.png"),
-       width = 5, height = 5, units = "in")
+ggsave(filename = here::here(fig_path, "density_posterior.pdf"),
+       width = twi/2, height = twi/2, units = "in")
 

### 2024-10-31 R/excursions.R
update study area figure

@@ -69,10 +69,15 @@ p2 = ggplot() +
   geom_tile(data = Fpxl,
             mapping = aes(x = x,
                           y = y,
-                          fill = F)) +
+                          fill = F,
+                          colour = F)) +
   scale_fill_viridis_c(limits = c(0,1),
                        breaks = c(0, 0.5, 1),
                        labels = c(0, 0.5, 1))+
+  scale_colour_viridis_c(limits = c(0,1),
+                         breaks = c(0, 0.5, 1),
+                         labels = c(0, 0.5, 1),
+                         guide = "none")+
   theme_void() +
   theme(legend.position = "bottom",
         legend.direction = "horizontal") +
@@ -91,9 +96,13 @@ p1 = ggplot() +
   geom_tile(data = Epxl,
             mapping = aes(x = x,
                           y = y,
-                          fill = E)) +
+                          fill = E,
+                          colour = E)) +
   # ggtitle("Excursion set for > 1 bird per hectare,\nalpha = 0.05") +
   scale_fill_manual(values = c("blue"),
+                    labels = c("Excursion set"),
+                    guide = "none") +
+  scale_colour_manual(values = c("blue"),
                     labels = c("Excursion set"),
                     guide = "none") +
   theme_void() +

### 2024-10-31 paper.tex
intro edits

@@
 
-Text width here we go:
-%\printinunitsof{cm}\prntlen{\textwidth}
+The estimation of the abundance and spatial distribution of wild populations of animals is a critical objective in ecology and conservation \citep{schwarz_estimating_1999}.This paper presents a statistical analysis of survey data collected on a critically endangered Hawaiian forest bird, offering insights into its population size and spatial dynamics, which crucial to guide effective conservation management decisions.
+The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.
+The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution through time.
 
-The estimation of the size and spatial distribution of wild populations of animals is a critical objective within ecology and conservation \citep{schwarz_estimating_1999}.
-Here we present an analysis of wildlife survey data collected on a critically endangered Hawaiian forest bird that aims to meet this objective.
-The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.
-The remaining population is the focus of sustained conservation efforts and monitoring is required to inform decision-makers about changes in the overall abundance and spatial distribution.
-This information is critical when decisions about conservation strategies are made.
-However, estimating the abundance and spatial distribution of wild populations of animals presents many statistical challenges.
+However, estimating the abundance and spatial distribution of wild populations of animals presents various statistical challenges.
+Traditional methods are usually split into two stages, where the spatial distribution is estimated in a second-stage spatial model, conditional on estimates of the probability of detecting animals given they are present.
+This detectability is estimated in a first-stage distance sampling model \citep{millerExtendingDensitySurface2021, miller_spatial_2013}.
+The two-stage approach requires either the loss of uncertainty from the first stage, or the use of approximate methods to propagate uncertainty between the modelling stages \citep{bravington_VariancePropagationDensity_2021}.
+In this paper, we present the first one-stage analysis of point transect distance sampling data that simultaneously estimates the non-linear detectability and spatial distribution of the \akepa{} population.
+In order to achieve this, we take a point process perspective on distance sampling survey data which requires the use of a novel approach to inference that allows estimation of non-linear parametric model components within a generalized additive model framework.
+This approach is based on an iterative scheme of successive fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009} that is implemented in the software package \texttt{inlabru} \citep{lindgren_inlabru_2024}.
+This approach is flexible, general and applicable in a wide range of contexts beyond modelling species distributions.
 
-First, as in many ecological surveys, it is impossible to undertake a full census (i.e.
-the complete enumeration of all individuals within a defined study region).
-The existing population numbers in the thousands and lives in dense tropical forest where logistical and ecological challenges mean a census is not feasible.
-For this reason, a monitoring survey must sub-sample appropriately in space and time.
-For the \akepa{} this takes the form of the \hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986}, which is a large-scale, quantitative survey of Hawaiian forest birds.
-Annual surveys of the \akepa{} study-region consist of a number of point transects located along randomly located line transects.
+Since the existing \akepa{} population numbers in the thousands and lives in dense tropical forest where logistical and ecological challenges mean a full census of the population is not feasible.
+The \hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986} is a large-scale, quantitative survey of Hawaiian forest birds.
+Annual surveys of the \akepa{} study-region consist of a number of point transects located at approximately regular intervals along randomly located line transects.
+At these surveyed locations, the detectability of animals is unknown and must be estimated in order to convert observed counts into estimates of the total population size.
+The \akepa{} survey estimates detectability using a point transect distance sampling approach \citep{buckland_distance_2015} where, for each observation, the distance to the observer is recorded.
 
-Second, even at surveyed locations, the detectability of animals is unknown.
-The \akepa{} survey estimates detectability using a point transect distance sampling approach \citep{buckland_distance_2015} where, for each observation, the distance to the observer is recorded.
-The probability of detecting a bird is modelled using a parametric detection function that decays with increasing distance.
-The detection function parameters are estimated from the observed distances by assuming the true density is uniform with respect to distance and attributing deviations from uniformity to imperfect detectability.
-The combination of spatio-temporal subsampling and unknown detectability means statistical methods to analyse distance sampling data must model a complex observation process alongside the spatial distribution of animals.
+Additionally, a key aim of the analysis is to predict animal density at unsurveyed locations.
+If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed location.
+However, in many ecological contexts there is good reason to believe there are drivers of the spatial distribution for which we have no suitable explanatory covariates available.
+For the \akepa{} data, there is a clear example of an unexplained driver of the spatial distribution where the population favours the southern region of the preserve.
+No straightforward explanatory cause has been found for this preference \citep{camp_dsm_2020}.
+From a statistical perspective, we address this unexplained heterogeneity by including a spatially structured random effect in the model.
 
-Third, a key aim of the analysis is to predict animal density at unsurveyed locations.
-If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed location.
-However, in many ecological contexts there can be good reason to believe there are drivers of the spatial distribution for which we have no explanatory covariates available or for which, in principle, no covariate could be constructed (e.g.
-the `sociality' of the species).
-For the \akepa{} data, one example of an unexplained driver of the spatial distribution is a north-south gradient for which no straightforward explanatory cause has been found \citep{camp_dsm_2020}.
-From a statistical perspective, this heterogeneity that cannot be explained by available covariates suggests the inclusion of spatially structured random effects in the model.
-Such effects require a careful consideration of the model structure and inference methods to ensure that model fitting remains computationally feasible.
+These challenges in analysing distance sampling survey data mean that the resulting statistical model is necessarily complex.
+Since the \akepa{} data are collected with the clear objective of monitoring the population and informing conservation management decisions, there is a challenge for statisticians to effectively communicate with stakeholders who have a wide range of statistical expertise.
+We believe the Bayesian one-stage approach is a useful step to support statisticians in this task.
+Since model outputs are based on Monte Carlo samples from the joint posterior, summary outputs naturally marginalise over the uncertainty in the observation process in a way that would be challenging in the traditional maximum-likelihood two-stage context.
+We discuss various challenges associated with communicating uncertainty in mapped estimates of animal density and propose some illustrative alternatives that make use of the Bayesian one-stage perspective.
 
-Lastly, the above challenges mean the resulting statistical model is necessarily complex.
-It is therefore challenging to communicate the results of the analysis to non-statistically trained audiences.
-The \akepa{} data are collected with the clear objective of monitoring the population and informing conservation management decisions.
-These decisions will be taken by stakeholders with a range of statistical expertise.
-This final step of communicating results should therefore be considered with as much care as the rest of the analysis.
-This requires the input of statisticians to ensure uncertainty is appropriately captured and communicated in the key outputs of the analysis upon which decisions will be based.
-In particular, in the context of species distribution modelling, we note the challenge of communicating uncertainty in maps of predicted animal density.
-
-We present an analysis that addresses these statistical challenges by using the following approaches:
-
-\begin{enumerate}[(a)]
-	\item A spatial point process perspective on point transect distance sampling and representing the detection model as a thinning of a point processes.
-	\item A one-stage approximate Bayesian approach to inference based on repeated model fits using integrated nested Laplace approximations (INLA) \citep{rue_approximate_2009}.
-This allows for the simultaneous estimation of the spatial distribution and detectability of animals.
-	\item A computationally efficient Gaussian Markov random field (GMRF) spatially structured random effect specified through a stochastic partial differential equation (SPDE) approach \citep{lindgren_explicit_2011} to account for unobserved drivers of the spatial distribution.
-	\item A discussion on model evaluation and communication of results that takes advantage of the one-stage approach by sampling from the joint posterior of all model parameters.
-We advocate for excursions methods \citep{bolin_excursion_2015} as a way to handle uncertainty in spatial predictions of density.
-\end{enumerate}
-
-We demonstrate several novel contributions to the problem of species distribution modelling under unknown detectability.
-This is, to the best of our knowledge, the first analysis of point transect distance sampling data using a point process perspective.
-This requires a modified intensity function that accounts for incomplete location information for animal sightings.
-This is also the first application of a novel approximate Bayesian inference method to estimate the detection function jointly with the spatial model.
-This inference method is readily generalisable to other applications beyond distance sampling models.
-As such, this analysis will be of interest to the broad applied statistics community in general, as well as those specifically interested in modelling species distributions.
-
-The rest of the paper proceeds as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the GMRF random effect; and Sections \ref{sec-results} presents results of the analysis and discusses model evaluation and effective communication of uncertainty.
+The paper is structured as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the spatially-structured random effect; and Sections \ref{sec-results} presents results of the analysis, model evaluation and effective communication of uncertainty.
 

### 2024-10-31 R/make_study_area_fig.R

@@ -17,14 +17,15 @@ samplers = readRDS(here::here(data_path, "samplers.RDS"))
 study_area_boundaries <- sf::st_cast(study_area, "LINESTRING")
 
 pdf(file = here::here("figures", "study_area_design.pdf"),
-    width = twi/2, height = twi/2)
+    width = twi/1.5, height = twi/1.5)
 
 ggplot() +
   geom_sf(data = study_area_boundaries,
           # fill = NA,
           size = 2,
           colour = "red") +
-  geom_sf(data = samplers) +
+  geom_sf(data = samplers,
+          cex = 0.4) +
   xlab("Easting (km)") +
   ylab("Northing (km)") +
   scale_x_continuous(limits = c(254, 262),

### 2024-10-31 paper.tex
more edits

@@
 
-\section{Study design}
+\section{\Akepa{} survey design}
 \label{sec-study-design}
@@
 Hakalau is the first wildlife refuge established in \hawaii{} with the primary purpose to protect and manage native forests for the conservation of threatened and endangered bird and plant species.
-Due to its broad-scale coverage and robust design, the HFBS has become an invaluable dataset used to determine changes in bird species distributions, population sizes and trends in density patterns over time.
 
@@
 Maintaining and expanding the \akepa{} population at Hakalau is a primary conservation concern.
-Unbiased and precise abundance estimates are therefore required by land and resource managers for the purposes of evaluating management actions and developing future management plans.
+Unbiased and precise abundance estimates are required by land and resource managers for the purposes of evaluating management actions and developing future management plans.
 
-\subsection{Study area and survey design}
+\subsection{Point transect surveys}
 
@@
 Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength greater than Beaufort scale 3).
-During 8-minute counts, trained observers recorded the species, distance to the nearest metre and detection type for each bird detected, along with the sampling conditions cloud cover, rain, wind strength, gust strength, and time of day each point was surveyed.
+During 8-minute counts, trained observers recorded the species, distance to the nearest metre for each bird detected.
 \cite{camp_population_2010,camp_statespace_2016} provide a detailed description of Hakalau, the study area and the bird surveys.
 
-The survey data used in this analysis is available for download \citep{camp_datarelease_2002} and the full time series data was analysed in a spatio-temporal context in \citep{camp_dsm_2020}.
-For the purposes of our analysis we select a single survey year from the \akepa{} time series (2002) that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability.
+To demonstrate our approach we select a single survey year from the \akepa{} time series that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability.
 In 2002, there were a total of 289 point transects sampled within the 4,603 ha study area of Hakalau (Fig.
@@
 We selected data from a single year to demonstrate our approach in a simplified setting that does not require a temporal model component.
-The choice of year is somewhat arbitrary and the model has also been tested analysing data from other years with similar results to those presented here.
-However, our approach is readily extendible to a multi-year analysis, similar to that in \cite{camp_dsm_2020}, which we discuss in more detail in Section \ref{sec-discussion}.
+The choice of year is somewhat arbitrary and our model has also been applied to data from other years with similar results to those presented here.
+Our approach is readily extendible to a multi-year analysis, similar to that in \cite{camp_dsm_2020}, which we discuss in more detail in Section \ref{sec-discussion}.
 
@@
 Standard distance sampling approaches use a hybrid of design- and model-based inference to estimate population size.
-The probability of detection is modelled using a parametric equation and, given estimates of detectability, a randomized sampling design allows for the construction of Horvitz-Thompson-like estimators of animal density \citep{ buckland_advanced_2004, horvitz_generalization_1952}.
+The probability of detection is modelled using a parametric function and, given estimates of detectability, a randomized sampling design allows for the construction of Horvitz-Thompson-like estimators of animal density \citep{ buckland_advanced_2004, horvitz_generalization_1952}.
 Under this approach, animal density is uniform within each strata of the survey.
@@
 These methods allow animal density to be associated with spatially-indexed covariates and non-parametric random effects.
-Therefore, animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{buckland_model-based_2016, miller_spatial_2013, johnson_model-based_2010}.
-As such, varying density estimates are not restricted by the stratification design required for Horvitz-Thompson-like estimators.
-This additional flexibility to make predictions for smaller, user-specified geographic units, and extrapolate relationships to unsurveyed regions has made model-based distance sampling a common approach in the literature \citep{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}.
+Under the model-based paradigm, animal density can, in principle, be estimated or predicted for any subregion within the study area \citep{buckland_model-based_2016, miller_spatial_2013, johnson_model-based_2010} and spatially-varying density estimates are not restricted by the stratification design that is required for Horvitz-Thompson-like estimators.
+This additional flexibility allows for predictions at user-specified geographic units, such as sub-regions of nature reserves, and also for extrapolating covariate relationships to unsurveyed regions.
+These features have made the model-based distance sampling paradigm a common approach in the literature \citep{garciabaron_modelling_2019, herr_aerial_2019, breen_new_2017, williams_chilean_2011, stokes_monitoring_2010, williams_modeling_2006}.
 
 Model-based distance sampling has most commonly been implemented in a two-stage modelling framework.
-In the first stage, detectability is estimated using the frequency of observed distances to fit the detection function.
-In the second stage, detectability estimates from the first stage are used as an offset in a generalized additive model framework.
-Detections are are binned into counts within user-defined sampling units (usually point transects or segments of line transects) and an appropriate response distribution is chosen to model these counts.
-The \texttt{dsm} package \citep{miller_spatial_2013}, for example, provides tools to do this using the \texttt{R} package \texttt{mgcv} \citep{wood_gam_2017} to implement the spatial count model.
-Due to the often sparse nature of wildlife survey data, this may require consideration of over- or under-dispersion and zero-inflated distributions and the negative-binomial or Tweedie distributions are common choices.
-This two-stage approach has come under the name \textit{density surface models} and \cite{miller_spatial_2013} provide a review.
+In the first stage, detectability is estimated using binned counts of observed distances to fit a detection function.
+In the second stage, detectability estimates from the first stage are used as an offset term in a generalized additive model framework.
+Detections are are also binned into counts within user-defined sampling units (point transects or segments of line transects) and an appropriate response distribution is chosen to model these counts.
+The \texttt{dsm} package \citep{miller_spatial_2013}, for example, provides tools to do this using the \texttt{R} package \texttt{mgcv} \citep{wood_gam_2017} to implement the spatially-explicit count model.
+This two-stage model is known as a \textit{density surface model} and \cite{miller_spatial_2013} provide a review.
 
 A key concern with the two-stage approach is the propagation of uncertainty from the detection model to the spatial model.
-Early attempts to address this focused on bootstrapping \citep{hedley_spatial_2004, lahiri_resampling_2003} but more recent work has pointed to potential difficulties with the bootstrapping approach, noting the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_VariancePropagationDensity_2021, williams_chilean_2011}.
-Instead, \cite{bravington_VariancePropagationDensity_2021} propose avoiding bootstrapping by propagating error based on a second-order Taylor approximation of the detection function at the first-stage maximum-likelihood estimate.
+Early attempts to address this focused on bootstrapping \citep{hedley_spatial_2004, lahiri_resampling_2003}.
+More recently, the bootstrap approach has been brought into question due to the difficulty of choosing the resampling units when combining spatially structured random effects and spatial bootstraps \citep{bravington_VariancePropagationDensity_2021, williams_chilean_2011}.
+Instead, \cite{bravington_VariancePropagationDensity_2021} propose avoiding bootstrapping completely by propagating error based on a second-order Taylor approximation of the detection function at the first-stage maximum-likelihood estimate.
 
@@
 This depends on choosing discrete distance classes and binning the data into counts defined on discrete spatial units.
-Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, a Poisson likelihood that can be optimised to achieve one-stage inference.
+Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, the authors derive a Poisson likelihood to which standard methods maximum-likelihood methods can be applied.
 This approach rests heavily on the discretisation of both the distance data and the spatial location data.
-Below we present a point process perspective that does not require this discretisation step in either the distance or spatial location data.
-The \texttt{unmarked} package also does not implement spatially structured random effects, something that we use here and is also popular in the two-stage approach implemented using \texttt{mgcv}.
 
-Bayesian one-stage approaches have tended to use Markov chain Monte Carlo (MCMC) methods for inference along with data augmentation to model unobserved individuals or groups \citep{schmidt_using_2012}.
+Our approach, which we outline below, is to instead adopt a point process perspective that does not require a discretisation step in either the distance or spatial location data.
+Although discretisation is required, it occurs at the level of a latent Gaussian random field, i.e.
+the model predictor vector, and not at the level of binning observations.
+
+In contrast to the maximum likelihood approach, Bayesian one-stage approaches have tended to use data augmentation to model unobserved individuals or groups \citep{schmidt_using_2012}, with inference via Markov chain Monte Carlo (MCMC) methods.
 \citet{oedekoven_bayesian_2014} present a one-stage model that avoids data augmentation by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty using reversible jump MCMC.
 
-The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use INLA \citep{rue_approximate_2009} and present an application on line transect distance sampling data.
-\citet{yuan_point_2017} take a point process perspective and formulate the detection model as a thinning of a point pattern.
-Key to their approach is formulating the detection model as the solution to an SPDE.
-An approximate solution the SPDE is constructed using a B-spline basis and finite element methods which results in a sparse prior precision matrix for the detection model parameters which can be incorporated into the model in INLA.
+The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use INLA \citep{rue_approximate_2009} and present an application to line transect distance sampling data.
+\citet{yuan_point_2017} also take a point process perspective and formulate the detection model as a thinning of a point pattern.
+Key to their approach is formulating the detection model as the solution to an SPDE, a formulation that is unfamiliar to most distance sampling practitioners.
 
-A major downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, usually a key feature of detection functions.
+A downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, which is usually a key feature of detection functions.
 The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful.
-Representing the observation model as an SPDE also differs substantially from standard distance sampling models that use a parametric model for the detection function.
-Furthermore, parametric forms of detection functions are readily extendible by including covariates and random effects to explain variation in detectability.
-It is not clear how to do this using the SPDE implementation in INLA.
-These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA and the benefits of a one-stage modelling process.
+These differences with traditional distance sampling methods have perhaps lead to low uptake amongst practitioners of this approach, despite the computational advantages of INLA for spatial modelling.
+Here, we present an approach that is inspired by \citet{yuan_point_2017} but with some key differences.
+We also fit the model using INLA and take advantage of the benefits of the increased computational efficiency compared to MCMC.
+However, we use a well-known parametric family of detection functions, that is familiar to users of classical distance sampling methods.
+This results in components of the model predictor that are non-linear in their parameters, making model fitting infeasible using standard INLA methods \citep{rue_approximate_2009}.
+We address this by using a novel method of iterated model fits based on a first-order Taylor expansion of the non-linear model components.
+Our approach is implemented in the \texttt{inlabru} package \citep{lindgren_inlabru_2024, bachl_inlabru_2019}, which also has support for spatial data objects.
 
-Here we present an approach that is related to \citet{yuan_point_2017} but with some key differences.
-We also fit the model using INLA and take advantage of the benefits the increased computational efficiency compared to MCMC.
-However, we avoid the issues with the SPDE detection model, instead allowing the user to specify a parametric family of detection functions, such as the half-normal detection function, as will be more familiar to users of classical distance sampling methods.
-However, this parametric form results in components of the additive predictor that are non-linear in their parameters, making model fitting infeasible in the \texttt{R-INLA} implementation of INLA \citep{rue_approximate_2009}.
-We address this by using a method of iterated model fits based on a first-order Taylor expansion of the non-linear model components and implemented in the \texttt{inlabru} package \citep{bachl_inlabru_2019}.
+Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a realisation of a thinned point process.
+This requires a slightly more involved derivation of the relevant intensity functions compared to the line transect case in \cite{yuan_point_2017}.
+However, the point process perspective on distance sampling is not new and has been applied numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
 
-We fit the model to point transect \akepa{} data, which differ from line transect methods in that there is a need to account for the increasing area surveyed as distance increases from the transect.
-Our analysis is, to the best of our knowledge, the first analysis of point transect distance sampling data formulated as a thinned point process.
-However, the point process framework is not new and has been applied numerous times in analyses of line transect data \citep{buckland_model-based_2016, niemi_bayesian_2010, johnson_model-based_2010, waagepetersen_likelihood-based_2006, hedley_spatial_2004,  hogmander_random_1991, stoyan_remark_1982}.
+\subsection{The point process model}
 
-\subsection{Model specification}
-
-This section describes the statistical model for the spatial pattern of observed animal locations.
-We assume the location of animals are a point pattern that follows a log-Gaussian Cox process with intensity process $\lambda$, i.e.
-the intensity at location $\bs$ is a random variable denoted $\lambda(\bs)$.
+This section describes the statistical model for the spatial pattern of observed animal locations and distances to observers.
+We assume the location of all animals, detected or not, form a point pattern distributed as a log-Gaussian Cox process with intensity process $\lambda(\bs)$.
 The log-Gaussian Cox process is a flexible model that can include spatial covariates to model the mean intensity as well a mean-zero spatially structured random effect to account for unexplained heterogeneity not captured by the covariates \citep{moller_log_1998}.
@@
 
-Distance sampling models specify the thinning function $g(\bs)$ as a function that decays with increasing distance.
+Distance sampling models specify the thinning function $g(\bs)$ as a function that decays as distance increases.
 The type of distance measured depends on the type of survey.
@@
 For the remainder of the paper we assume a point transect survey design and hence horizontal distance to an observer located at the centre of each point transect.
-
 The thinning probability function is specified as a parametric family of functions.
-For example, if $r(\bs)$ denotes the distance of a point at $\bs$ from an observer, the half-normal thinning probability function is $g(\bs | \sigma) = \exp(-r(\bs)^2 / 2\sigma^2)$, where $\sigma^2 > 0$ is a variance parameter to be estimated.
+We denote the distance of a point at $\bs$ from an observer as $r(\bs)$.
+The \textit{half-normal} detection function is $g(\bs | \sigma) = \exp(-r(\bs)^2 / 2\sigma^2)$, where $\sigma^2 > 0$ is a variance parameter to be estimated.
 Other types of detection function, such as hazard-rate, negative exponential, can also be used.
@@
 In addition to these parametric forms, additional adjustments have been proposed to increase the flexibility of such models by including trigonometric series expansions \citep{buckland_distance_2015}.
+For simplicity, $\sigma$ is modelled using only an intercept parameter, but our approach can readily be extended to include covariates.
 
-In this analysis we assume a half-normal detection function with no covariates on detection parameters to demonstrate the approach in a simplified setting.
-The detection function parameters can only be estimated if an assumption is made about $\lambda(\bs)$ otherwise $g(\bs)$ and the intensity are confounded.
+The detection function parameters can only be estimated if a uniformity assumption is made about the distribution of animals with respect to $r(\bs)$.
+Without such an assumption, the detection probability and intensity are confounded.
 The standard assumption in distance sampling is that the intensity is constant with respect to changes in $r(\bs)$.
-The usual approach for two stage models is to assume $\lambda(\bs)$ is constant within each spatial unit, however we are able to relax this assumption slightly to allow $\lambda(\bs)$ to be a linear function within each transect.
-Given such an assumption, any observed deviations from uniformity can be attributed to detectability and not to variation in the intensity.
+The usual approach for two stage models is to assume $\lambda(\bs)$ is constant within each spatial unit.
+With the point process perspective we are able to relax this assumption slightly to allow $\lambda(\bs)$ to be a linear function within each transect.
 
-A point transect distance sampling survey consists of a set of $K$ point transects.
-We denote $k$th subset of space covered by the $k$th point transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.
-For simplicity, we assume that all point transects are non-overlapping disks with equal radius $W$, as is the case in the HFBS survey.
-It would be straightforward to relax this assumption.
-The probability of observing a point at location $\bs \in \Omega_k$ given it is at $\bs \in \Omega_k$ we denote $g_k(\bs)$.
+To describe the model in general we assume a point transect distance sampling survey consists of a set of $K$ point transects.
+We denote $k$-th transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.
+In the HFBS the point transects are non-overlapping disks all with with equal radius $W$ and centres at locations $\bs_k$ for $k = 1, \ldots, K$.
+The probability of observing a point at location $\bs \in \Omega_k$, given there is a point at that location, we denote by $g_k(\bs)$.
 The probability of observing a point outside the surveyed region is zero.
-Since the point transects are non-overlapping, each location $\bs \in \Omega$ is unambiguously associated with a single thinning probability function $g_k$.
-For example, letting $\bs_k$ denote the centre of transect $\Omega_k$, the half-normal thinning probability function for points in $\Omega_k$ is $g_k(\bs) = \exp(-\lVert \bs - \bs_k \rVert_2^2 / 2\sigma^2)$.
-The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation.
-The thinning probability function for any $\bs \in \Omega$ is then given by $g(\bs) = g_{k(\bs)}(\bs)$ where $k(\bs)$ is an indexing function such that $k(\bs) = k$ for $\bs \in \Omega_k$.
+Since the point transects are non-overlapping, each location $\bs \in \Omega$ is unambiguously associated with a single observer and, therefore, a single thinning probability function $g_k$.
+The half-normal detection function for points in $\Omega_k$ is therefore $g_k(\bs) = \exp(-\lVert \bs - \bs_k \rVert_2^2 / 2\sigma^2)$.
+The assumption of non-overlapping survey regions could be relaxed by including extra information such as the time of each observation and observer indentifiers.
+The thinning probability function for any $\bs \in \Omega$ is then given by $g(\bs) = g_{k(\bs)}(\bs)$ where $k(\bs)$ is an indexing function such that $k(\bs) = k$ whenever $\bs \in \Omega_k$.
 
@@
 A common approach for fitting log-Gaussian Cox processes in practice is to approximate the likelihood by replacing the integral with a weighted sum.
-In this section we describe this approximate likelihood and show how it can be written as a modified Poisson likelihood.
+In this section we describe this approximate likelihood and show how it can be written as a modified Poisson likelihood which allows us to use INLA for inference.
 We adapt the approach given in \cite{simpson_going_2016} for a fully observed log-Gaussian Cox process, modified to account for the point transect distance sampling observation process.
@@
 To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.
-In a slight abuse of notation $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes any location within the point transect.
-It should be clear from context which is intended by $\bs_k$.
+In a slight abuse of notation $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes a location within the $k$-th point transect.
 The thinning function $g_k(\bs_k(r, \theta))$ depends only on $r$ and not on $\theta$ and so we use the shorthand notation $ g(\bs_k(r, \theta)) = g_k(r)$.
-
 The integral in the likelihood can be simplified using an assumption that $\lambda(\bs)$ is linear within each sampling unit.
@@
 We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.
-A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piecewise linear basis functions used for the GMRF spatial effect (Section \ref{sec-gmrf}).
+A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piecewise linear basis functions used for the spatially-structured random effect described in Section \ref{sec-gmrf}.
 
 It follows that $\int \lambda(\bs)g(\bs) \mathrm{d}\bs = \sum_{k=1}^K 2\pi \lambda(\bs_k) \int_0^W r g_k(r)\mathrm{d}r$ by applying the change of variables formula for integration and substituting in \eqref{eq-linear-intensity}.
-Note that this integral now only involves the one dimensional integral with respect to $r$.
+This integral now only involves the one-dimensional integral with respect to $r$, greatly simplifying the integration required to evaluate the likelihood.
 For each sampling unit we approximate this integral using a midpoint integration method with $M$ integration locations $r_{k1}, \ldots, r_{kM}$ and associated weights $\alpha_{k1}, \ldots, \alpha_{kM}$.
@@
 In the above, we assume the data are complete records of animal locations.
-That is, each observation has a known position $\bs_{k(\bs_i)} = \bs_{k(\bs_i)}(r_i, \theta_i)$.
+That is, each observation has a known position $\bm{y}_i = \bs_{k(\bs_i)}(r_i, \theta_i)$.
 However, in many distance sampling surveys only the location of the observer and the distance to the observer are recorded.
-Data of this type can be analysed within a point process framework by deriving the appropriate intensity function.
-We consider the case where $r$ is recorded but not $\theta$.
-It follows that the intensity for points at a distance $r$ observed within sampling unit $\Omega_k$ is
+Data of this type can be analysed within a point process framework by deriving the appropriate intensity function for this censored data.
+If only $r$ is recorded and $\theta$ is unknown, the intensity for points at a distance $r$ observed within point transect $\Omega_k$ is
 \begin{align}
@@
 where $c_k(r)$ is a circle of radius $r$ centred at $\bs_k$ and the second line follows from change of variables and the assumption of linear intensity within $\Omega_k$.
-This intensity differs from the full data case through the $2\pi$  term that accounts for the fact that we do not observe $\theta$ and the additional $r$ term that accounts for the increasing area surveyed at larger distances.
+This intensity differs from the full data case through the $2\pi$  term that accounts for the fact that we do not observe $\theta$ and the Jacobian term $r$ that accounts for the increasing area surveyed at larger distances.
 We note that a similar equation is derived from the perspective of a triangular distribution for point transect distance sampling in \cite{buckland_advanced_2004}.

### 2024-10-31 paper.tex
more edits

@@
 
-\subsection{Writing the model as a modified Poisson likelihood}
+\subsection{The modified Poisson likelihood}
+\label{sec-lgcp}
 
@@
 
-The half-normal detection function depends on the strictly positive parameter $\sigma^2$.
-Since it is not possible to restrict the range of parameters to be strictly positive within \texttt{R-INLA} this model cannot be fitted using the standard INLA methodology.
-Instead, we choose a log link and consider $g(\bs | \phi)$, where $\log\sigma^2 = \phi$.
-This is a non-linear function of $\phi$ which cannot be fitted using standard GLM methods.
-To address this we use a novel approximate-method based on an iterative model fitting scheme.
+The half-normal detection function depends on the strictly positive parameter $\sigma^2$ and so this model cannot be fitted using the standard INLA methodology.
+We choose a log link $\log\sigma^2 = \phi$ and use a novel approximate-method based on an iterative model fitting scheme.
 This method is implemented in the \texttt{R} package \texttt{inlabru} which is a wrapper for and an extension of the \texttt{R-INLA} package \citep{lindgren_inlabru_2024}.
-Full details of the method can be found in the software paper, here we present a brief overview.
+We briefly summarise the approach here, full details of the method can be found in \cite{lindgren_inlabru_2024}.
 
-Let $\eta(\bm{u}) = \log \tilde{\lambda}(\bm{u})$ where $\bm{u}$ is the parameter vector of the predictor (which includes $\phi$ as well as any parameters for fixed and random effects).
-Denote by $\bar{\eta}_{\bm{u}_0}$ the first-order Taylor approximation of $\eta(\bm{u})$ at the point $\bm{u}_0$.
-That is
+The family of models that can be fitted using the INLA methodology are known as latent Gaussian models.
+These depend on a latent Gaussian vector of parameters which we denote $\bm{u}$.
+Parameters are related to observations via a predictor $\eta(\bm{u})$ which, for the LGCP model, is the log-intensity evaluated at the observed point locations and the integration scheme locations as described in Section \ref{sec-lgcp}.
+Since $\phi$ is an element of $\bm{u}$, $\eta(\bm{u})$ is a non-linear function with respect to $\bm{u}$.
+
+To motivate the approach, we consider a linearised approximation of $\eta(\bm{u})$ around point $\bm{u}_0$:
 \begin{equation*}
@@
 where $\bm{B}_{\bm{u}_0}$ is the derivative matrix of $\eta(\bm{u})$ evaluated at $\bm{u}_0$.
-Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the linear Taylor approximation $\bar{\eta}_{\bm{u_0}}$.
+Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the first-order Taylor approximation $\bar{\eta}_{\bm{u_0}}$.
+This model configuration has $\bm{B}_{\bm{u}_0}$ as the model design matrix and $\eta(\bm{u}_0} - \bm{B}_{\bm{u}_0}\bm{u}_0$ as a known offset term.
 
-The question then is how to choose the linearisation point $\bm{u}_0$.
-To find an `optimal' linearisation point we use a fixed point iteration scheme.
-There are various possible choices to define an `optimal' choice and this affects the properties of the approximate inference method.
-Here we describe the criteria used when fitting the model that produces the results presented in this paper.
-However, more recently, other options have been added to \texttt{inlabru}, the software package that implements these methods.
-
+The `optimal' linearisation point is defined via a fixed point iteration scheme.
 Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.
-New linearisation points are found by via the functional
+A new linearisation point is found by via the functional
 \begin{equation}
@@
 the joint conditional posterior mode for $\bm{u}$.
-This functional defines a new linearisation point as the posterior mode for $\bm{\theta}$ under the linearised model configuration and the conditional mode for $\bm{u}$ given $\bm{\theta}$.
-These posteriors are computed automatically during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost for each fit of the linearised model.
+In other words, this functional defines a new linearisation point as the posterior mode for $\bm{\theta}$ under the linearised model configuration and the conditional mode for $\bm{u}$ given $\bm{\theta}$.
+These posteriors are computed automatically during the model fitting using the standard INLA method and so this functional can be evaluated at no extra cost.
 

### 2024-10-31 paper.tex
edits up to results section

@@
 Our approach is readily extendible to a multi-year analysis, similar to that in \cite{camp_dsm_2020}, which we discuss in more detail in Section \ref{sec-discussion}.
-
-\vfill
 
@@
 Since $\bar{\eta}_{\bm{u}_0}$ is linear in $\bm{u}$ it is possible to fit an approximate model in INLA by replacing the non-linear $\eta$ with the first-order Taylor approximation $\bar{\eta}_{\bm{u_0}}$.
-This model configuration has $\bm{B}_{\bm{u}_0}$ as the model design matrix and $\eta(\bm{u}_0} - \bm{B}_{\bm{u}_0}\bm{u}_0$ as a known offset term.
+This approximate model configuration has design matrix $\bm{B}_{\bm{u}_0}$ and known offset term $\eta(\bm{u}_0) - \bm{B}_{\bm{u}_0}\bm{u}_0$.
 
@@
 
-The `optimal' approximate model is defined as a fixed point of this functional.
+The final linearisation point is defined as the fixed point of this functional.
 Given a current choice of linearisation point $\bm{v}$, we estimate $\bar{p}_{\bm{v}}$ and set $\bm{u}_* :=f(\bar{p}_{\bm{v}})$ to use as the new linearisation point.
 We iterate this procedure until a fixed point $\bm{v}^*$ is identified within a chosen tolerance such that $\bm{v}^* = f(\bar{p}_{\bm{v}^*})$.
+In other words, we iteratively search for a location in the parameter space for $\bm{u}$ where the Taylor expansion is taken about the posterior mode.
 
-\section{Spatially Structured Random Effect}
+\subsection{The intensity process model}
 \label{sec-gmrf}
 
-This section describes the spatial model for the intensity process.  
 As mentioned above, the \akepa{} population exhibits spatial heterogeneity for which there is no known explanatory covariate.
-Therefore we use a spatially structured random effect to explain this variation.
+To address this we use a sparse Gaussian Markov random field (GMRF) representation of Gaussian random field (GRF) with Mat\'ern covariance based on a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.
 
-Using spatially-structured random effects can be computationally expensive.
-Often this is due to parameterisations and approximations that lead to dense prior precision matrices which have an  $\mathcal{O}(k^3)$ cost to factorise for an effect with $k$ basis parameters.
-A key approach in computationally efficient methods for spatially structured random effects is the use of sparse Gaussian Markov random field (GMRF) representations of Gaussian random fields (GRFs).
-These result in a sparse precision matrix which can be factorised with a cost of $\mathcal{O}(k^{3/2})$.
-We use a GMRF representation based on a stochastic partial differential equation approach \citep{lindgren_explicit_2011}.
-This representation is constructed by applying finite element methods to solve the SPDE numerically.
-
-Since only the distance to the observer was recorded, and not the exact location of \akepa{}, we used the adjusted intensity given in equation \eqref{intensity-incomplete}.
-The log-intensity of animal locations, detected or undetected, is given by
+We use the adjusted intensity given in equation \eqref{intensity-incomplete} for incomplete data.  The log-intensity of animal locations, detected or undetected, is given by
 \begin{equation*}
@@
 \end{equation*}
-where $\beta_0$ is an intercept parameter and $\xi(\bs)$ is a mean zero GRF with Mat\'ern covariance.  The Mat\'ern covariance function in two dimensions can be written as 
+where $\beta_0$ is an intercept parameter and $\xi(\bs)$ is a mean zero GRF with Mat\'ern covariance
 \begin{equation}
-C(\bs_1,\bs_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|\bs_1-\bs_2\|)^{\nu}K_\nu(\kappa \|\bs_1-\bs_2\|)
+C(\bs_1,\bs_2) = \frac{2^{1-\nu}}{4\pi\kappa^2\tau^2\Gamma(\nu)}(\kappa \|\bs_1-\bs_2\|)^{\nu}K_\nu(\kappa \|\bs_1-\bs_2\|),
 \end{equation}
 where \(\nu, \kappa, \tau\) are parameters and \(K_{\nu}\) is the modified Bessel function of the second kind.
-The three parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and it is conventional to assume a value for $\nu$ which specifies the mean-square differentiability of the process.
+The three covariance function parameters are not simultaneously identifiable \citep{zhang_inconsistent_2004} and the SPDE approach requires us to assume a value for $\nu$ which specifies the mean-square differentiability of the process.
 We set $\nu = 1$ which is the default value in \texttt{R-INLA}.
 
-We use a GMRF representation of $\xi(\bs)$ following the approach of \cite{lindgren_explicit_2011} based on specifying $\xi(\bs)$ via a SPDE.
 We define a finite element mesh with $L$ nodes and associated piece-wise linear basis functions $\phi_1, \ldots, \phi_L$.
@@
 The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices (\cite{blangiardo_spatial_2013}, Equation 6.19).
+See \cite{lindgren_explicit_2011} for full details.
+
 The parameters $\tau$ and $\kappa$ control the shape and rate of decay of the Mat\'ern covariance function as distance between locations increases.
-We use a finite element mesh with 3748 nodes defined over the entire study region and extended over a buffer region to avoid the effects of boundary conditions within the region of inference.
-
 In order to specify priors on the Mat\'ern covariance we use a reparameterisation of $\kappa$ and $\tau$ to a range and variance parameter.
@@
 This parameterisation is used to set penalised complexity priors \citep{simpson_penalising_2017} on $\rho$ and $\sigma^2$ which allow the GRMF effect to shrink towards zero.
-For the Mat\'ern field, the base model is the limiting case with $\sigma^2 \rightarrow 0$ and $\rho \rightarrow \infty$, which defines a random field that is almost surely zero everywhere.
-The purpose of penalised complexity priors is to place sufficient prior probability on a simple model to allow the random effect to be essentially `turned off' if the model component is not supported by the data.
-To specify penalised complexity priors on the SPDE effect we set $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$.
+The base model is the limiting case with $\sigma^2 \rightarrow 0$ and $\rho \rightarrow \infty$, which defines a random field that is almost surely zero everywhere.
+We set $\mathbb{P}(\sigma > 2) = 0.01$ and $\mathbb{P}(\rho < 130) = 0.01$.
 The value of 130 for the range parameter was selected based on the minimum distance between sampling locations.
 
-The SPDE approach can at first appear to be fundamentally different  other approaches such as penalised regression splines.
-However, the two methods are closely related and the GRF model component in our model plays a similar role to a penalised smoothing spline on space.
-\cite{yue_bayesian_2014} show how spline models can be viewed through the perspective of SPDEs and \cite{miller_UnderstandingStochasticPartial_2020} do the reverse to show how to implement the Mat\'ern SPDE as a penalised spline model in \texttt{mgcv} \citep{wood_gam_2017}.
+The SPDE approach can at first appear to be fundamentally different to other spatial effects such as penalised regression splines, which are more typically used in the two-stage approach.
+However, these two methods are closely related and the GRF model component in our model plays a similar role to a penalised smoothing spline.
+\cite{yue_bayesian_2014} show how spline models can be viewed through the perspective of SPDEs.
+\cite{miller_UnderstandingStochasticPartial_2020} present the reverse perspective and show how to implement the Mat\'ern SPDE as a penalised spline model in \texttt{mgcv} \citep{wood_gam_2017}.
 

### 2024-10-31 paper.tex
combine detfn matern and pairwise plots

@@
 
-The posterior detection function (\autoref{fig:det-matern}) shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres and also broadly agrees with a two-stage analysis.
+The posterior detection function (\autoref{fig:dist-mat-pairs}) shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres and also broadly agrees with a two-stage analysis.
 \begin{figure}[!htb]
 	\begin{center}
-		\includegraphics[width=\textwidth]{figures/detfn_and_matern.pdf}
+		\includegraphics[width=\textwidth]{figures/dist_mat_pairs.pdf}
 		\caption{A) Posterior half-normal detection function.
 B) Posterior Mat\'ern correlation function.
-The black line is the mean posterior estimate and the grey shaded area is 95\% credible interval in both plots}
-		\label{fig:det-matern}
+The black line is the posterior mean and the grey shaded area is 95\% credible interval in both plots.
+C) Boxplots of the frequency of pairwise distances based on 500 simulated datasets from the posterior predictive distribution.
+Axis labels mark the largest value in each 1km bin.
+Red circles indicate the observed count of pairwise distances within each bin.}
+		\label{fig:dist-mat-pairs}
 	\end{center}
 \end{figure}
-The posterior Mat\'ern correlation function is shown in \autoref{fig:det-matern}; the grey regions show the 95\% credible interval.
+The posterior Mat\'ern correlation function is shown in \autoref{fig:dist-mat-pairs}B; the grey regions show the 95\% credible interval.
 We note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).
@@
 We calculated the pairwise distances between all observations for each of these datasets and compared the frequency of pairwise distances with those in the observed data.
-\begin{figure}[!htb]
-	\begin{center}
-		\includegraphics{figures/post_pp_distances.pdf}
-		\caption{Boxplots of the frequency of pairwise distances based on 500 simulated datasets from the posterior predictive distribution.
-Axis labels mark the largest value in each bin.
-Red circles indicate the observed frequency of pairwise distances within each bin.}
-		\label{fig:post-pp-distances}
-	\end{center}
-\end{figure}
-Figure \ref{fig:post-pp-distances} shows that the SPDE effect broadly captures the spatial clustering in the observed data.
+
+Figure  shows that the SPDE effect broadly captures the spatial clustering in the observed data.
 There is some evidence of under-predicting the strength of the clustering at shorter distances between 1,000 and 2,500 m, with the simulated data showing some evidence of negative bias.
@@
 
-Figure \ref{fig:post-pp-distances} is similar in spirit to Ripley's $K$-function \citep{ripley_SecondorderAnalysisStationary_1976}, which is a common approach to investigating spatial structure in point patterns.
+Figure \autoref{fig:dist-mat-pairs}C is similar in spirit to Ripley's $K$-function \citep{ripley_SecondorderAnalysisStationary_1976}, which is a common approach to investigating spatial structure in point patterns.
 The pairwise distances in the observed data is a function of spatial heterogeneity as well as the spatially restricted sampling scheme.

### 2024-10-31 R/eval_model.R

@@ -53,43 +53,6 @@ dsamp = function(distance, lsig){
 #### Model evaluation ####
 summary(fit)
 
-#### Posterior half-normal plot ####
-
-W = 58/1000
-Wm = 58    # transect radius in metres
-distdf <- data.frame(distance = seq(.Machine$double.eps, Wm, length=100))
-
-# adjust for change of units (km in model fit, metres here)
-hnpred <- predict(fit, distdf, ~ hn(distance, lsig + log(1000)),
-                  n.samples = 100)
-
-ghn = ggplot() +
-        gg(hnpred) +
-        ylab("detection probability\n") +
-        xlab("\ndistance (m)") +
-        theme(axis.title = element_text(size = ax.size),
-        axis.text = element_text(size = ax.size))
-ghn
-
-##### Posterior Matern plot ####
-blah = spde.posterior(fit, "grf", what = "matern.correlation")
-gmat = ggplot() +
-  gg(blah) +
-  ylab("correlation\n") +
-  xlab("\ndistance (km)") +
-  theme(axis.title = element_text(size = ax.size, vjust = 0.3),
-        axis.text = element_text(size = ax.size))
-
-gmat
-
-pdf(file = here::here(fig_path, "detfn_and_matern.pdf"),
-    width = twi, height = twi/2)
-plot_grid(NULL, ghn, NULL, gmat, NULL,
-          labels = c("", "A", "", "B", ""),
-          rel_widths = c(0.05, 1, 0.15, 1, 0.05),
-          ncol = 5)
-dev.off()
-
 ##### Posterior intensity plots #####
 
 # Prediction locations
@@ -417,4 +380,179 @@ plot_grid(NULL, p1, NULL, p2, NULL, p3, NULL,
           ncol = 7)
 dev.off()
 
+#### Posterior half-normal plot ####
 
+W = 58/1000
+Wm = 58    # transect radius in metres
+distdf <- data.frame(distance = seq(.Machine$double.eps, Wm, length=100))
+
+# adjust for change of units (km in model fit, metres here)
+hnpred <- predict(fit, distdf, ~ hn(distance, lsig + log(1000)),
+                  n.samples = 100)
+
+ghn = ggplot() +
+        gg(hnpred) +
+        ylab("detection probability\n") +
+        xlab("\ndistance (m)") +
+        theme(axis.title = element_text(size = ax.size),
+        axis.text = element_text(size = ax.size))
+ghn
+
+##### Posterior Matern plot ####
[…]

### 2024-10-31 R/eval_spde.R

@@ -1,159 +1,0 @@
-# Summarise posterior sampled point patterns
-# using pairwise distances
-
-library(INLA)
-library(inlabru)
-library(ggplot2)
-library(sf)
-
-# statsoc doc class textwidth in cm
-tw = 14.69785
-twi = 5.7865551181  # pdf() needs inches
-
-# set ggplot theme
-theme_set(theme_minimal())
-set.seed(9701071)
-
-#### Fitted model  ####
-model_path = here::here("R",
-                        "fitted_models",
-                        "fitted_model.RDS")
-fit = readRDS(model_path)
-
-data_path = here::here("R", "data")
-obs <- readRDS(here::here(data_path, "obs.RDS"))
-study_area = readRDS(here::here(data_path, "study_area.RDS"))
-samplers = readRDS(here::here(data_path, "samplers.RDS"))
-mesh = readRDS(here::here(data_path, "mesh.RDS"))
-
-# polygon samplers
-samplers$ID = 1:nrow(samplers)
-W = 58/1000
-samplers_buffered = st_buffer(samplers, dist = W)
-
-### Generate posterior point patterns
-n.pp = 500
-
-# simulate intensities and detection functions
-# each column of llam is a realisation of the log-intensity at each mesh node
-post.sample = generate(fit,
-                formula = ~ {c(
-                  llam = grf_latent + Intercept_latent,
-                  lsig = lsig_latent)
-                },
-                n.samples = n.pp)
-
-llam = post.sample[1:mesh$n, ]
-lsig.post = post.sample["lsig",]
-
-# create distance bins and matrix to store counts
-breaks = 0:12
-counts = matrix(NA, nrow = n.pp, ncol = length(breaks)-1)
-
-# detection functions
-# Log half-normal
-log_hn = function(distance, lsig){
-  -0.5*(distance/exp(lsig))^2
-}
-
-# Half-normal
-hn <- function(distance, lsig) exp(log_hn(distance, lsig))
-
-# Need to create a new mesh just in the study area and
-# project llam to this. Boundary was giving very large intensities
-# when trying with full mesh and samplers argument
-
-## Generated by meshbuilder()
-
-## Build boundary information:
-## (fmesher supports SpatialPolygons, but this app is not (yet) intelligent enough for that.)
[…]

### 2024-11-05 paper.tex
final edits

@@
 
-The model fits in roughly 40 to 60 seconds on a laptop with 10 Gb RAM and a 2.50 GHz CPU when INLA is run using the `empirical Bayes' option which does not incorporate uncertainty in hyper parameters.
-The model fits in a few minutes in the full INLA mode, including hyper parameter uncertainty.
-We suggest using the `empirical Bayes' mode for model development and full INLA on a smaller subset of candidate models.
-
-The predicted mean of the posterior intensity field is shown in \autoref{fig:intensity-mean-cv}A along with a map of the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid with each prediction cell area approximately 1.7 hectares.
+\autoref{fig:intensity-mean-cv}A shows a map of the posterior mean intensity field alongside the coefficient of variation (CV) (\autoref{fig:intensity-mean-cv}B) and standard deviation of the posterior intensity field (\autoref{fig:intensity-mean-cv}C) on a regular prediction grid.
 \begin{figure}[!htb]
@@
 		\includegraphics[width=\textwidth]{figures/intensity_mean_cv_sd.pdf}
-		\caption{The A) mean, B) CV and C) SD of the posterior intensity field.  Units are per m$^2$.  The colour scale is different for each panel.}
+		\caption{The A) mean, B) coefficient of variation and C) standard deviation of the posterior intensity field.  For the mean and standard deviaion the units are number of birds per m$^2$.}
 		\label{fig:intensity-mean-cv}
@@
 \end{figure}
-This shows a region of high intensity in the south and much lower intensity in the north, agreeing with a standard two-stage analysis of the same data \citep{camp_dsm_2020}.
+This shows a region of higher intensity in the south and lower intensity in the north, agreeing with the broad patterns observed in a standard two-stage analysis of the same data \citep{camp_dsm_2020}.
 The CV plot shows that the CV in the posterior intensity field is lower in areas with greater sampling effort and higher intensity.
-There is a clear indication of preferential sampling with more survey effort in the south, where densities are higher, compared to the north.
-In the north, estimated intensities are lower, which also contributes to larger CV values and makes the CV map hard to interpret.
-The standard deviation map shows stronger overall posterior variability in the south where intensity is larger, most likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.
+The standard deviation map shows stronger overall posterior variability in the south where intensity is larger.
+Although survey effort is higher in the south, the increased uncertainty is likely due to the assumed log-normal relationship between the random field and the Poisson rate parameter.
+For this reason we suggest using both the CV and SD to visualise uncertainty.
 
-We also map the lower and upper quantiles corresponding to a 95\% credible interval for each prediction location (Figure \ref{fig:intensity-quantiles}).
-As we discuss below, these quantiles are independent summary statistics of the posterior intensity field at each prediction location.
+\autoref{fig:dist-mat-pairs} summarises the estimates of spatial structure in the model.
+The posterior detection function (\autoref{fig:dist-mat-pairs}A) shows that the detection probability drops to around 0.25 at the maximum observable distance of 58 metres.
+\begin{figure}[!htb]
+	\begin{center}
+		\includegraphics[width=\textwidth]{figures/dist_mat_pairs.pdf}
+		\caption{A) Posterior half-normal detection function.
+B) Posterior Mat\'ern correlation function for the GMRF spatial effect.
+The black line is the posterior mean and the grey shaded area marks the 95\% credible interval.
+C) Boxplots of the frequency of pairwise distances based on 500 simulated datasets from the posterior predictive distribution.
+Red circles indicate the observed count of pairwise distances within each bin.}
+		\label{fig:dist-mat-pairs}
+	\end{center}
+\end{figure}
+The posterior Mat\'ern correlation function is shown in \autoref{fig:dist-mat-pairs}B, this represents the degree of spatial smoothing in the posterior intensity field.
+We note that this correlation function shows positive correlation on the log-scale.
+On the intensity scale the empirical range of correlation is shorter.
+\autoref{fig:dist-mat-pairs}C summarises the model's ability to replicate the spatial structure in the intensity and detection process simultaneously by comparing the frequency of pairwise distances between all observations and those on data simulated from the joint posterior of the fitted model.
+This is based on 500 Monte Carlo samples and is essentially an adjusted version of Ripley's $K$-function \citep{ripley_SecondorderAnalysisStationary_1976}, which accounts for the distance sampling observation process.
+The model shows a good ability to reconstruct the spatial structure due to the distance to observe and spatially varying intensity surface.
+
+To visualise the posterior abundance for the \akepa{} in Hakalau we use a Monte Carlo method based on sampling realisations of the posterior intensity field.
+Let $N = \int_{\Omega}\lambda(\bs)\mathrm{d}\bs$, the integral of the intensity field over the study region.
+A point estimate can be calculated by integrating the posterior mean intensity.
+To generate a full posterior estimate for abundance we approximate the posterior as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)}, \bm{Y})$ where $\lambda^{(1)}, \ldots, \lambda^{(m)}$ are samples from the posterior intensity field.
+Due to the properties of the log-Gaussian Cox process, each $\pi(N | \lambda = \lambda^{(i)}, \bm{Y})$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda^{(i)}(\bs)\mathrm{d}\bs$.
+\begin{figure}[!htb]
+	\begin{center}
+		\includegraphics{figures/N_posterior.pdf}
+		\caption{Posterior abundance.  Grey area marks two Monte Carlo standard errors above and below the mean.}
+		\label{fig:realized-abundance-posterior}
+	\end{center}
+\end{figure}
+\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
+This allows us to estimate the probability for any specific value of realised abundance $n$ which may be of interest to conservation managers.
+This approach can also be used to generate credible intervals and other posterior summaries of interest.
+
+Our abundance estimates cannot be directly compared to the results in \cite{camp_dsm_2020} since their two-stage model also includes a temporal correlation effect and we include an area of Hakalau that the authors in \cite{camp_dsm_2020} excluded.
+Converting the above abundance estimates to density per hectare for the whole study region allows for some comparison with Figure 2 in \cite{camp_dsm_2020} which suggests that \akepa{} density was roughly between 0.5 and 1.5 birds per hectare in 2002 (considering both the design-based estimate and the spatio-temporal GAM estimate).
+In our analysis we find a credible region of roughly 0.9 to 1.7 birds per hectare which overlaps significantly with the interval found in \cite{camp_dsm_2020}.
+
+An attractive feature of the Bayesian one-stage model is that, by basing model outputs on posterior Monte Carlo samples, all the outputs incorporate the uncertainty in the detection model in a natural way.
+There is no additional step required to incorporate detection function uncertainty beyond working with samples from the joint posterior.
+
+\subsection{Limitations of mapped summaries}
+
+All figures that summarise the posterior intensity field have potential weaknesses that should be considered when communicating with stakeholders.
+This sections discusses these weaknesses and presents and example of one approach that tries to address them.
+The aim of this section is to show that our approach is readily extendible to produce bespoke model outputs.
+
+Our presentation of the results above is broadly consistent with approaches taken in the species distribution modelling literature, although it is frustratingly common to see maps of point estimates without any accompanying communication of uncertainty.
+The most common method to communicate uncertainty in maps of animal density is to produce maps of measures of predictive uncertainty, such as the standard deviation, CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}, posterior quantiles or the probability of exceeding certain thresholds at each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.
+Whilst such maps are useful, they all mask certain properties of the intensity field that may be important when communicating the results of the analysis.
+For example, the posterior mean will always be smoother than realisations of the field.
+If there is interest in understanding patterns of spatial clustering, the posterior mean map presents a misleadingly homogeneous picture.
+This highlights the fact that even a relatively innocuous map of the posterior mean can mask certain features of the random field.
+\autoref{fig:intensity-realizations} shows three realisations of the posterior field, all of which show more heterogeneity in the high intensity region compared to the posterior mean.
+\begin{figure}[!htb]
+	\includegraphics{figures/intensity_realized.pdf}
+	\caption{Three realizations of the posterior intensity field}
+	\label{fig:intensity-realizations}
+\end{figure}
+
+The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.
+However, CV values will generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.
+Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.
+However, there is also variable sampling effort within the study region, with higher CV values in the region with lower effort.
+It is not clear what impact this has on the CV values and this makes the map hard to interpret, even for statisticians who understand the model well.
+The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values also intended to indicate uncertainty.
+However, a default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.
+Whether these relative differences matter is dependent on the overall context of the analysis.
+Some levels of variability may be acceptable in some contexts and unacceptable in others.
+
+Quantile maps (Figure \ref{fig:intensity-quantiles}) are also difficult to interpret.
+The temptation is to perceive these maps as showing possible intensity surfaces that could have produced the observed data.
+Stake holders may interpret the 0.025 quantile map as a kind of lower bound intensity surfaces and the 0.975 quantile map as an upper bound.
+However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.
+A similar problem holds for maps of the marginal probability of exceeding a threshold at each prediction location.
 \begin{figure}[!htb]
@@
 \end{figure}
+These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.
+To demonstrate the misleading nature of these maps, if we (incorrectly) treat the lower and upper quantile plots as though they are intensity surfaces and integrate them, then we obtain an abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.
+However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).
 
-The posterior detection function (\autoref{fig:dist-mat-pairs}) shows that detectability drops to just under 0.25 at the maximum observable distance of 58 metres and also broadly agrees with a two-stage analysis.
-\begin{figure}[!htb]
-	\begin{center}
-		\includegraphics[width=\textwidth]{figures/dist_mat_pairs.pdf}
-		\caption{A) Posterior half-normal detection function.
-B) Posterior Mat\'ern correlation function.
-The black line is the posterior mean and the grey shaded area is 95\% credible interval in both plots.
-C) Boxplots of the frequency of pairwise distances based on 500 simulated datasets from the posterior predictive distribution.
-Axis labels mark the largest value in each 1km bin.
-Red circles indicate the observed count of pairwise distances within each bin.}
-		\label{fig:dist-mat-pairs}
-	\end{center}
-\end{figure}
-The posterior Mat\'ern correlation function is shown in \autoref{fig:dist-mat-pairs}B; the grey regions show the 95\% credible interval.
-We note that this correlation function shows positive correlation even at distances of 10,000 metres, wider than the full length of the study region (approximately 8,000 metres).
-This is because the correlation function is defined on the log-intensity scale.
-On the intensity scale the empirical range of correlation is shorter.
-
-To evaluate the ability of the SPDE effect to capture the spatial heterogeneity in the observed data we generated 1,000 point patterns using the posterior intensity field and thinned them with the posterior detection function to create 1,000 simulated datasets to compare with the observed data.
-We calculated the pairwise distances between all observations for each of these datasets and compared the frequency of pairwise distances with those in the observed data.
-
-Figure  shows that the SPDE effect broadly captures the spatial clustering in the observed data.
-There is some evidence of under-predicting the strength of the clustering at shorter distances between 1,000 and 2,500 m, with the simulated data showing some evidence of negative bias.
-This could be due to over-smoothing in some areas which can happen if using a stationary model.
-Large areas of the study region having low density, which has a larger correlation range than higher density areas, and so the effect will average between these.
-
-Figure \autoref{fig:dist-mat-pairs}C is similar in spirit to Ripley's $K$-function \citep{ripley_SecondorderAnalysisStationary_1976}, which is a common approach to investigating spatial structure in point patterns.
-The pairwise distances in the observed data is a function of spatial heterogeneity as well as the spatially restricted sampling scheme.
-Other methods of comparison could also be considered such as the average difference between predicted and observed counts at each point transect.
-Working with posterior predictive datasets is a flexible approach to model evaluation that allows for innovative methods of comparison with observed data that can focus on particular features of the model.
-
-To plot the posterior abundance estimate for the \akepa{} in Hakalau we use a Monte Carlo method based on sampling realisations of the posterior intensity field.
-Let $n$ denote the abundance of the population within the region of interest and $N = \int_{\Omega}\lambda(\bs)\mathrm{d}\bs$ its corresponding random variable within our Bayesian framework.
-Integrating the mean of the posterior intensity field provides a point estimate for the abundance, i.e.
-the expected abundance.
-To generate a full posterior estimate for abundance we approximate the posterior $\pi(N | \bm{Y})$ by taking $m$ Monte Carlo samples of the posterior intensity field, $\lambda^{(1)}, \ldots, \lambda^{(m)}$ and approximate the posterior for the abundance as $\pi(N | \bm{Y}) \approx 1 / m \sum_{i=1}^m \pi (N | \lambda = \lambda^{(i)}, \bm{Y})$.
-Each $\pi(N | \lambda = \lambda^{(i)}, \bm{Y})$ is a Poisson probability mass function with rate parameter $\int_{\Omega}\lambda^{(i)}(\bs)\mathrm{d}\bs$.
-\begin{figure}[!htb]
-	\begin{center}
-		\includegraphics{figures/N_posterior.pdf}
-		\caption{Posterior abundance.  Grey area marks two Monte Carlo standard errors above and below the mean.}
-		\label{fig:realized-abundance-posterior}
-	\end{center}
-\end{figure}
-
-\autoref{fig:realized-abundance-posterior} shows the approximate posterior for $N$ with $m = 20,000$.
-This allows us to estimate the probability for any specific value of realised abundance $n$ which may be of interest to conservation managers.
-This approach can also be used to generate credible intervals and other posterior summaries of interest.
-
-The abundance estimates cannot be directly compared to the results in \cite{camp_dsm_2020} since their two-stage model also includes a temporal correlation and we include an area of Hakalau that the authors excluded.
-Converting the above abundance estimates to density per hectare for the whole study region allows for some comparison with Figure 2 in \cite{camp_dsm_2020} which suggests that \akepa{} density was roughly between 0.5 and 1.5 birds per hectare in 2002 (considering both the design-based estimate and the spatio-temporal GAM estimate).
-In our analysis we find a credible region of roughly 0.9 to 1.7 birds per hectare, which is slightly different but overlaps significantly with the results in \cite{camp_dsm_2020}.
-
-An attractive feature of our Bayesian one-stage model is that, by basing model outputs on posterior Monte Carlo samples, all the outputs presented in the above, the summaries of the random field as well as the posterior abundance, incorporate the uncertainty in the detection model in a natural way.
-There is no additional step required to incorporate detection function uncertainty beyond working with samples from the joint posterior.
-
-\subsection{Limitations of mapped summaries}
-
-Our presentation of the results above is broadly consistent with approaches taken in the species distribution modelling literature, although maps of predictive uncertainty (Figures \ref{fig:intensity-mean-cv} and \ref{fig:intensity-quantiles}) are not always provided.
-It is common to see maps of point estimates (e.g.
-a map of expected density) without any accompanying communication of uncertainty.
-It is also common to report a point estimate of the expected abundance along with uncertainty in the point estimate.
-However, uncertainty in a point estimate will have lower variance than the variance of the posterior random variable.
-For this reason we chose to present the full abundance posterior in \autoref{fig:realized-abundance-posterior} as opposed to uncertainty of a `best guess' for abundance.
-Our intention in this section is to highlight further limitations with the model outputs presented above and to suggest ways to address these limitations.
-
-The most common method to communicate uncertainty in maps of animal density is to produce maps of some measure of predictive uncertainty, such as the standard deviation or CV \citep{fuller_novel_2018, vallejo_responses_2017,bradbury_mapping_2014}.
-These can be derived from the posterior predictive distribution for the model in a Bayesian framework and bootstrapping in a maximum likelihood context.
-Another approach is to map quantiles or the probability of exceeding certain thresholds at each prediction location \citep{russell_avoidance_2016, wilson_hierarchical_2010}.
-In Section \ref{sec-results}, we present maps of the CV, standard deviation, and the 0.025 and 0.975 quantiles of the posterior intensity field (Figures \ref{fig:intensity-mean-cv}B, \ref{fig:intensity-mean-cv}C, \ref{fig:intensity-quantiles}A, and \ref{fig:intensity-quantiles}B, respectively) that are all intended to communicate uncertainty in the estimate of the \akepa{} spatial distributions.
-Whilst such maps that summarise spatial predictions are useful, they all mask certain properties that may be important when communicating the results of the analysis.
-
-A map showing the posterior predicted mean across the study region (Figure \ref{fig:intensity-mean-cv}A) is often the key output of a species distribution model.
-However, the posterior mean will always be smoother than realisations of the field itself, highlighting that even this relatively innocuous summary statistic can mask important features of the random field.
-\autoref{fig:intensity-realizations} shows three such realisations from the posterior intensity field.
-Note that each realisation has a finer-grained spatial structure than is shown in the posterior mean (\autoref{fig:intensity-mean-cv}A).
-Our interpretation is that the clustering of animals is stronger than might be expected if we only looked at the map of the posterior mean.
-\begin{figure}[!htb]
-	\includegraphics{figures/intensity_realized.pdf}
-	\caption{Three realizations of the posterior intensity field}
-	\label{fig:intensity-realizations}
-\end{figure}
-
-For model evaluation we recommend plotting multiple realisations of the posterior field as these will more closely resemble the spatial structure of the observed data and thus what ecologists might encounter in the field.
-The mean can give a sense of homogeneity, particularly in the high density area in the south, when compared to realisations of the mean.
-Presenting multiple realisations, perhaps as an animation, is also an effective way to communicate uncertainty.
-\cite{bowman_GraphicsUncertainty_2019a} provide an interpolation method that preserves the mean and covariance structure of realisations that smoothly interpolates between realisations.
-This avoids abrupt changes that can occur when animating the raw realisations.
-
-The CV map (\autoref{fig:intensity-mean-cv}B) is also intended to communicate uncertainty.
-However, CV values will generally be higher in regions of low predicted intensity, particularly if the posterior standard deviation is relatively consistent across the study region or positively correlated with the intensity.
-Higher CV values in Figure \ref{fig:intensity-mean-cv}B represent areas of lower intensity but not necessarily higher uncertainty.
-However, there is also variable sampling effort within the study region, with higher CV values in the region with lower effort.
-It is not clear what impact this has on the CV values and this makes the map hard to interpret.
-
-The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values also intended to indicate uncertainty.
-However, care should be taken as to what the mapped values imply in the context of a given analysis.
-A default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.
-However, whether these relative differences matter is dependent on the overall context of the analysis.
-Some levels of variability may be acceptable in some contexts and unacceptable in others.
-
-The SD values also reflect a modelling assumption rather than variability due to, say, varying survey effort.
-The SD is linked to the predicted mean by the assumed log-Gaussian relationship.
-If X is normally distributed with mean $\mu$ and variance $\sigma^2$ then $Z = \exp(X)$ has a mean-variance relationship given by $\text{Var}(Z) = \exp(\sigma^2 - 1)\mathbb{E}(Z)^2$.
-Therefore the SD map will generally show a pattern similar to the mean intensity map (Figures 2A and 2B).
-Since this relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort.
-It should be noted however, that this modelling assumption affects uncertainty quantification for all outputs based on the posterior intensity field, not just the SD map.
-
-The quantile maps (Figure \ref{fig:intensity-quantiles}) are also difficult to interpret.
-The temptation is to perceive the maps as showing possible intensity surfaces that could have produced the observed data.
-One possible intensity surface (0.025 quantile map) that signifies a kind of lower bound on abundance and another surface (0.975 quantile map) that signifies higher abundance.
-However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.
-A similar problem holds for maps of the marginal probability of exceeding a threshold at each prediction location.
-
-Presenting independent quantiles jointly in a single image is risky.
-These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.
-To demonstrate the possible consequences of this we (incorrectly) treat the lower and upper quantile plots as though they are intensity functions and integrate them to obtain an expected abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.
-A naive (and tempting) interpretation of these numbers is as lower and upper limits of the 95\% credible interval for abundance.
-However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).
-This shows that the tendency to interpret these maps as showing possible intensity surfaces can lead to interpretations that are inconsistent with the very model that generated the maps.
-
+Broadly speaking, we advocate for moving beyond maps of uncertainty presented for their own sake and towards model outputs that directly relate to the wider objectives of the analysis.
 We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult or impossible to interpret.
@@
 
-\subsection{Excursion sets and excursion functions}
+\subsection{Identification of hotspots}
 
-One solution to the problem of interpreting quantile and exceedance probability maps is to use methods that explicitly consider the \textit{joint} probability of events across prediction locations, as opposed to the \textit{marginal} probability at each location.
-The approach taken here requires that we decide, \emph{a priori}, on relevant values of the process and acceptable levels of uncertainty that are suitable given the context and aims of the analysis.
-These values can then be used to construct summary maps that consider joint events across prediction locations.
+In this section we consider the hypothetical challenge of using our model to identify population hotspots within Hakalau refuge.
+This acts as a focus for communication that directs our choices when producing model outputs.
 
-We demonstrate this perspective using excursions sets and excursion functions \citep{bolin_excursion_2015}.
-By choosing acceptable levels of uncertainty, these methods avoid the issue of the default colour scale for the SD and CV maps which may potentially affecting our interpretation of results.
-Excursion sets and excursion functions are also based on the joint probability of events across a set of locations.
-For this reason they avoid the interpretability issues of the quantile maps.
+A solution to the problem of interpreting quantile and exceedance probability maps is to use methods that explicitly consider the joint probability of events across all prediction locations.
+One option is to use excursions sets and excursion functions \citep{bolin_excursion_2015}.
+This requires users to choose thresholds of interest (e.g.
+the definition of a hotspot), and to select an acceptable level of uncertainty in knowing whether the field exceeds this threshold.
+This forces us to consider this explicitly and leaves us less at the whim of an arbitrary colour scale.
+Crucially, excursion sets and functions can be estimated using Monte Carlo samples and so fit well within our one-stage Bayesian approach.
 
-We give the technical definition of excursion sets and functions below, using the same notation as \cite{bolin_excursion_2015}.
+We provide the technical definition of excursion sets and functions below, using the same notation as \cite{bolin_excursion_2015}.
 For readers unfamiliar with these methods, the notation can take some getting used to.
@@
 We try to, as much as possible, give an informal explanation alongside the mathematical description.
-
-The positive excursion set with level $u$ for a function $f(\bs)$ with domain $\Omega$ is $A_u^{+}(f) = \{ \bs \in \Omega ; f(\bs) > u \}$, i.e.
+The positive excursion set with level $u$ for a deterministic function $f(\bs)$ with domain $\Omega$ is $A_u^{+}(f) = \{ \bs \in \Omega ; f(\bs) > u \}$, i.e.
 the set of all locations in $\Omega$ where $f$ exceeds a threshold value $u$.
@@
 \end{equation*}
-Note that $A_u^{+}(f)$ specifies a set of locations for which a function $f(\bs)$ exceeds a threshold value $u$ for \textit{every location} in the set.
-Therefore, the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the \textit{largest} such set of locations for which \textit{realisations} of $\lambda(\bs)$ exceed the threshold $u$, \textit{simultaneously} for all locations in the set, with probability $1-\alpha$.
+In other words, the positive excursion set $E_{u,\alpha}^{+}(\lambda)$ is the \textit{largest} set of locations for which realisations of $\lambda(\bs)$ exceed the threshold $u$, \textit{simultaneously}, for all locations in the set, with probability $1-\alpha$.
 Note that we must decide on relevant choices for $u$ and $\alpha$ given the aims and context of the analysis.
-Negative excursion sets are similarly defined by considering the probability of being below a threshold value.
-Contour uncertainty can also be defined as the set difference between positive and negative excursion sets \citep{bolin_excursion_2015}.
-
+Negative excursion sets are similarly defined.
 Excursion sets can be estimated by considering candidate sets for $D$ of increasing size and a sequential integration scheme to estimate the required probabilities in a computationally efficient way.
@@
 \autoref{fig:excursions}A shows the positive excursion set with a level corresponding to 1 bird per hectare with probability 0.95 (hence $\alpha = 0.05$).
+Note that we use this threshold as the definition of a hotspot, and proceed with generating model results that incorporate this definition.
 This figure can be interpreted in a natural way as the largest region for which the intensity is greater than 1 bird per hectare for every location within the region, with probability 0.95.
@@
 While any particular definition would be up for debate, the advantage is that the statistical methodology is taken as close as possible to the aims of the analysis.
-Often a map of species density will be interpreted informally in many different ways (one of which is identifying hotspots for example).
-Here we present an example of how a clear statistical definition can be used to produce a map that incorporates relevant definitions and uncertainties, and can be interpreted clearly.
-
+Often a map of species density will be interpreted informally in many different ways (one of which is identifying hotspots).
 To visualise how such excursion sets change with different levels of uncertainty we can use the excursion function $F_u^{+}(\bs) = \sup \{1 - \alpha ; \bs \in E_{u,\alpha}^+ \}$.
@@
 Similarly, the core of the high density region seems to exceed 1 bird per hectare even for $\alpha$ levels close to zero.
-There is a region to the north of the main population that would be included in the positive excursion set for 1 bird per hectare if the probability were allowed to be lower, around 0.5.
-However, for regions in the north there is essentially no probability level for which those locations would be included.
 This map could be used to identify, for example, secondary areas of potential importance beyond the core population that could be investigated further.
-
-We chose the threshold of 1 bird per hectare and an error probability of 0.05 to demonstrate the approach.
-However, multiple values for each of these can be considered and we imagine that, in most contexts, there would be value in considering various sets of thresholds and uncertainty levels.
 
@@
 
-We have presented a new approach to analysing point transect distance sampling data in a Bayesian setting.
-We chose to do this in a relatively simplified setting, considering a single year of data, with no covariate effects and using a GRF to explain spatial variation in density.
-There are several natural extensions to the model.
+There are several natural extensions to the model presented here.
 Spatio-temporal models can also be fitted using \texttt{R-INLA}, which would allow multiple years of data to be modelled jointly, as was the case in \cite{camp_dsm_2020}.
@@
 Factors such as weather conditions, observer expertise, animal behaviour and morphological traits are likely to affect detectability.
-Future research should investigate the degree of complexity that can be well-approximated by the iterated INLA approach.
 
@@
 This function is analogous to the detection function in distance sampling models.
-Our framework could allow, for example, the joint modelling of selectivity and spatial data.
+Our framework allows for the joint modelling of selectivity and spatial data.
 	\item \textbf{Functional responses} The concept of a `functional response' captures the idea that the overall abundance of a resource or risk will affect a species' response to it \citep{holling_some_1959}.
@@
 an increasing function with a horizontal asymptote, to capture the notion of diminishing returns.
-Such non-linear functions could be incorporated within our framework.
-This flexibility in specifying model components could also be more widely applicable in areas such as public health and spatial epidemiology, where responses to conditions may suitably be modelled using non-linear relationships.
+Such non-linear functions could be incorporated within our framework
 \end{itemize}
-There are no doubt many more application areas that could make use of the additional flexibility that comes with allowing non-linear predictors.
 
@@
 In addition to computational efficiency, there are some features of INLA that are not generally available in other packages, such as: (i) the ability to model using GRFs directly on the sphere, avoiding projecting to $\mathbb{R}^2$ \citep{lindgren_explicit_2011}; (ii) GRFs that account for complex barriers such as coastlines and islands, to avoid inappropriately smoothing across these \citep{bakka_NonstationaryGaussianModels_2019}; and (iii) support for joint likelihood models, allowing modelling of, for example, marked point processes \citep{illian_FittingComplexEcological_2013}, and combining multiple data sources.
-These approaches, and many others we have not mentioned, that are implemented in INLA, can now readily be incorporated into distance sampling analyses, as well as broader spatial ecology and spatial statistics applications.
+These approaches can now readily be incorporated into distance sampling analyses, as well as broader spatial ecology and spatial statistics applications.
 
@@
 Geological Survey: \url{https://doi.org/10.5066/P9Q9UXMZ} \citep{camp_datarelease_2002}.
-The time data for 1987-2017 is also available  available from the U.S.
-Geological Survey: \url{https://doi.org/10.5066/P98IO297} \citep{camp_datarelease_all}
-
-Code to reproduce the figures and results in this paper are available in a git repository and can be accessed at \url{https://github.com/ASeatonSpatial/point-transects-paper/}.
+The data for 1987-2017 are also available  available from the U.S.
+Geological Survey: \url{https://doi.org/10.5066/P98IO297} \citep{camp_datarelease_all} Code to reproduce the figures and results in this paper are available in a git repository and can be accessed at \url{https://github.com/ASeatonSpatial/point-transects-paper/}.
 

### 2024-11-12 paper.tex
update acknowledgements
H + R comments

@@
 
-The estimation of the abundance and spatial distribution of wild populations of animals is a critical objective in ecology and conservation \citep{schwarz_estimating_1999}.This paper presents a statistical analysis of survey data collected on a critically endangered Hawaiian forest bird, offering insights into its population size and spatial dynamics, which crucial to guide effective conservation management decisions.
+The estimation of the abundance and spatial distribution of wild populations of animals is a critical objective in ecology and conservation \citep{schwarz_estimating_1999}.This paper presents a statistical analysis of survey data collected on a critically endangered Hawaiian forest bird, offering insights into its population size and spatial dynamics, which are crucial to guide effective conservation management decisions.
 The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.
@@
 
-Since the existing \akepa{} population numbers in the thousands and lives in dense tropical forest where logistical and ecological challenges mean a full census of the population is not feasible.
+Since the existing \akepa{} population numbers in the thousands and lives in dense tropical forest, logistical and ecological challenges mean a full census of the population is not feasible.
 The \hawaii Forest Bird Survey (HFBS) \citep{scott_HFBS_1986} is a large-scale, quantitative survey of Hawaiian forest birds.
@@
 Additionally, a key aim of the analysis is to predict animal density at unsurveyed locations.
-If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed location.
+If animal density is modelled using spatial covariates then the covariate effects estimated in the surveyed region can be extrapolated to unsurveyed locations.
 However, in many ecological contexts there is good reason to believe there are drivers of the spatial distribution for which we have no suitable explanatory covariates available.
@@
 
-The paper is structured as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} and \ref{sec-gmrf} describe the iterated INLA fitting procedure and the spatially-structured random effect; and Sections \ref{sec-results} presents results of the analysis, model evaluation and effective communication of uncertainty.
+The paper is structured as follows:  Section \ref{sec-study-design} describes the \akepa{} study region and survey design; Section \ref{sec-ds-pp} reviews current approaches to distance sampling and presents the point process perspective; Sections \ref{sec-iinla} describes the iterated INLA fitting procedure and the spatially-structured random effect; and Sections \ref{sec-results} presents results of the analysis, model evaluation and effective communication of uncertainty.
+Section \ref{sec-discussion} discusses potential extensions of the model and wider uses for iterative INLA.
 
@@
 
-The \hawaii{} \akepa{} is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to \hawaii{} Island, USA.
+The \hawaii{} \akepa{} is an internationally and federally endangered Hawaiian honeycreeper (\citealp{usfws_akepa_1970, birdlife_akepa_2016}) that is endemic to the island of \hawaii{}, USA.
 Large-scale, quantitative surveys of Hawaiian forest birds and their habitat commenced in the mid-1970s through the HFBS \citep{scott_HFBS_1986}.
 Information from the HFBS is used to update the International Union for Conservation of Nature redlist \citep{iucn_redlist_2022}, where the \akepa{} is currently listed as endangered.
-HFBS data is also used to establish preserves that coincide with native bird hotspots, including Hakalau Forest National Wildlife Refuge on \hawaii{} Island (hereafter Hakalau).
+HFBS data is also used to establish preserves that coincide with native bird hotspots, including Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{} (hereafter Hakalau).
 Hakalau is the first wildlife refuge established in \hawaii{} with the primary purpose to protect and manage native forests for the conservation of threatened and endangered bird and plant species.
@@
 We limit our study area to the open-forest and closed-forest strata of Hakalau (\autoref{fig:2002studyareapointspt}), an extension of the area considered in \cite{camp_population_2010, camp_statespace_2016}, who omitted the closed-forest stratum because it was not sampled in the early years of the surveys.
-For our analysis, we use data from a later year in which the closed-forest stratum was sampled and so we are able to include it.
+For our analysis, we use data from a later year in which the closed-forest stratum was sampled and so broaden the extent of spatial inference compared to previous studies.
 The open-forest stratum was previously heavily grazed, and, since the removal of cattle in 1988, regeneration has proceeded naturally \citep{maxfield_hakalau_1998}.
@@
 	\includegraphics{figures/study_area_design.pdf}
-	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest National Wildlife Refuge, \hawaii{} Island.
+	\caption{Study area showing the 2002 survey points (black dots) in Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{}.
 The red outline marks the study area boundary as defined in \cite{camp_dsm_2020}.}
@@
 
-The survey follows a point-transect distance sampling design, recording the horizontal distances from observers to detected birds.
+The survey follows a point transect distance sampling design, recording the horizontal distances from observers to detected birds.
 Surveys commenced at dawn and continued until 11:00 or halted when weather conditions exceeded prescribed conditions that hindered detecting birds (light rain, and wind and gust strength greater than Beaufort scale 3).
@@
 
-To demonstrate our approach we select a single survey year from the \akepa{} time series that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability.
+To demonstrate our approach we selected a single survey year from the \akepa{} time series that contains a broad sampling of the study area with sufficient numbers of detections to estimate detectability.
 In 2002, there were a total of 289 point transects sampled within the 4,603 ha study area of Hakalau (Fig.
 \ref{fig:2002studyareapointspt}).
-A total of 276 \akepa{} were detected on 121 point transects.
+A total of 276 \akepa{} was detected on 121 point transects.
 The number of detections within each point transect ranged from zero to six.
@@
 This depends on choosing discrete distance classes and binning the data into counts defined on discrete spatial units.
-Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, the authors derive a Poisson likelihood to which standard methods maximum-likelihood methods can be applied.
+Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, the authors derive a Poisson likelihood to which standard maximum-likelihood methods can be applied.
 This approach rests heavily on the discretisation of both the distance data and the spatial location data.
@@
 
-The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use INLA \citep{rue_approximate_2009} and present an application to line transect distance sampling data.
+The only Bayesian one-stage analysis that does not use MCMC is, to the best of our knowledge, \citet{yuan_point_2017}, who use INLA and present an application to line transect distance sampling data.
 \citet{yuan_point_2017} also take a point process perspective and formulate the detection model as a thinning of a point pattern.
-Key to their approach is formulating the detection model as the solution to an SPDE, a formulation that is unfamiliar to most distance sampling practitioners.
+Key to their approach is formulating the detection model as the solution to an stochastic partial differential equation (SPDE), a formulation that is unfamiliar to most distance sampling practitioners.
 
-A downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, which is usually a key feature of detection functions.
+A downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, which is a critical assumption of detection functions \citep{buckland_distance_2015}.
 The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful.
@@
 To describe the model in general we assume a point transect distance sampling survey consists of a set of $K$ point transects.
-We denote $k$-th transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.
-In the HFBS the point transects are non-overlapping disks all with with equal radius $W$ and centres at locations $\bs_k$ for $k = 1, \ldots, K$.
+We denote $k$-th point transect as $\Omega_k \subset \mathbb{R}^2$ and the total surveyed region is $\Omega = \cup_{k=1}^K \Omega_k$.
+In the HFBS the point transects are non-overlapping disks all with equal radius $W$ and centres at locations $\bs_k$ for $k = 1, \ldots, K$.
 The probability of observing a point at location $\bs \in \Omega_k$, given there is a point at that location, we denote by $g_k(\bs)$.
@@
 To evaluate the integral we use polar coordinates notation $\bs_k(r, \theta) = \bs_k + r\left[\cos\theta, \sin\theta \right]^T$ with $r \in [0, W]$ and $\theta \in [0, 2\pi]$ to represent locations in each sampling unit $\Omega_k$.
-In a slight abuse of notation $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes a location within the $k$-th point transect.
+In a slight abuse of notation, $\bs_k$ denotes the centre of the point transect and $\bs_k(r,\theta)$ describes a location within the $k$-th point transect.
 The thinning function $g_k(\bs_k(r, \theta))$ depends only on $r$ and not on $\theta$ and so we use the shorthand notation $ g(\bs_k(r, \theta)) = g_k(r)$.
@@
 These depend on a latent Gaussian vector of parameters which we denote $\bm{u}$.
-Parameters are related to observations via a predictor $\eta(\bm{u})$ which, for the LGCP model, is the log-intensity evaluated at the observed point locations and the integration scheme locations as described in Section \ref{sec-lgcp}.
+Parameters are related to observations via a predictor $\eta(\bm{u})$ which, for the log-Gaussian Cox process model, is the log-intensity evaluated at the observed point locations and the integration scheme locations as described in Section \ref{sec-lgcp}.
 Since $\phi$ is an element of $\bm{u}$, $\eta(\bm{u})$ is a non-linear function with respect to $\bm{u}$.
@@
 Let $\bar{p}_{\bm{v}}$ denote the INLA approximate posterior distribution of the model parameters given the linearised model configuration at linearisation point $\bm{v}$.
-A new linearisation point is found by via the functional
+A new linearisation point is found via the functional
 \begin{equation}
@@
 The GRF is represented as $\xi(\bs) = \sum_l \beta_l \phi_l(\bs)$.
-The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices (\cite{blangiardo_spatial_2013}, Equation 6.19).
+The parameters $\beta_1, \ldots, \beta_L$ form a GMRF with sparse prior precision matrix $\bm{Q} = \frac{1}{\tau^2}\left(\kappa^4\bm{C} + 2\kappa^2\bm{G_1} + \bm{G_2}\right)$ where $\bm{C}$, $\bm{G_1}$, $\bm{G_2}$ are all sparse matrices that depend on a triangulation of the spatial domain (\cite{blangiardo_spatial_2013}, Equation 6.19).
 See \cite{lindgren_explicit_2011} for full details.
@@
 All figures that summarise the posterior intensity field have potential weaknesses that should be considered when communicating with stakeholders.
-This sections discusses these weaknesses and presents and example of one approach that tries to address them.
+This sections discusses these weaknesses and presents an example of one approach that tries to address them.
 The aim of this section is to show that our approach is readily extendible to produce bespoke model outputs.
@@
 It is not clear what impact this has on the CV values and this makes the map hard to interpret, even for statisticians who understand the model well.
-The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values also intended to indicate uncertainty.
+The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values that are also intended to indicate uncertainty.
 However, a default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.
@@
 The temptation is to perceive these maps as showing possible intensity surfaces that could have produced the observed data.
-Stake holders may interpret the 0.025 quantile map as a kind of lower bound intensity surfaces and the 0.975 quantile map as an upper bound.
+Stakeholders may interpret the 0.025 quantile map as a kind of lower bound intensity surfaces and the 0.975 quantile map as an upper bound.
 However, presenting these quantiles together in a single map obscures the fact that it is vanishingly unlikely for the random field to \textit{simultaneously} achieve the 0.025 or 0.975 quantiles at all prediction locations.
@@
 These caveats make the map difficult to use for most non-statistically trained audiences and even trained statisticians may misinterpret these quantile maps.
-To demonstrate the misleading nature of these maps, if we (incorrectly) treat the lower and upper quantile plots as though they are intensity surfaces and integrate them, then we obtain an abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.
+To demonstrate the misleading nature of these maps, if we were to (incorrectly) treat the lower and upper quantile plots as though they were intensity surfaces and integrate them, then we would obtain an abundance estimate of approximately 2,900 for the 0.025 quantile map and 10,700 for the 0.975 quantile map.
 However, these abundance estimates are not supported by the posterior for abundance (\autoref{fig:realized-abundance-posterior}).
+In other words, these maps tempt an interpretation that is inconsistent with the very model that produced the maps.
 
 Broadly speaking, we advocate for moving beyond maps of uncertainty presented for their own sake and towards model outputs that directly relate to the wider objectives of the analysis.
-We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult or impossible to interpret.
+We conclude that, although all useful in their own way, each of the mapped summaries of the posterior intensity field suffer from some weaknesses, whether through masking certain properties of the random field or through being difficult to interpret.
 In the following section we suggest an approach that avoids some, but not all, of the pitfalls above.
@@
 This defines, for each location, the largest possible probability $1 -\alpha$ for which that location would be in the excursion set defined using probability $1 - \alpha$.
-I.e.
-if we allow greater uncertainty, this function shows which locations would be included in the excursion set.
+That is, if we allow greater uncertainty, this function shows which locations would be included in the excursion set.
 
@@
 
-The iterated INLA inference procedure has the potential to be applied to other non-linear model components that arise spatial statistics.
+The iterated INLA inference procedure has the potential to be applied to other non-linear model components that arise in spatial statistics.
 We note two promising areas within the field of spatial ecology that highlight this potential:
@@
 The popularity of using INLA for spatial modelling has grown, in large part, due to the computational efficiency of the approach.
-However, this efficiency comes at the cost of reduced flexibility in model specification over, say, a more general probabilistic programming languages such as \texttt{JAGS} \citep{plummer_jags_2017} or \texttt{Stan} \citep{stan_Rstan_2020}.
+However, this efficiency comes at the cost of reduced flexibility in model specification over, say, a more general probabilistic programming language such as \texttt{JAGS} \citep{plummer_jags_2017} or \texttt{Stan} \citep{stan_Rstan_2020}.
 The iterated INLA approach allows a wider class of models to be considered whilst retaining much of the computational benefits of INLA.
@@
 
-Hawaii Akepa (Loxops coccineus) point-transect distance sampling data were provided by U.S.
+Hawaii Akepa (Loxops coccineus) point transect distance sampling data were provided by U.S.
 Fish and Wildlife Service, Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex.
@@
 The data for 1987-2017 are also available  available from the U.S.
-Geological Survey: \url{https://doi.org/10.5066/P98IO297} \citep{camp_datarelease_all} Code to reproduce the figures and results in this paper are available in a git repository and can be accessed at \url{https://github.com/ASeatonSpatial/point-transects-paper/}.
+Geological Survey: \url{https://doi.org/10.5066/P98IO297} \citep{camp_datarelease_all}.
+Code to reproduce the figures and results in this paper are available in a git repository and can be accessed at \url{https://github.com/ASeatonSpatial/point-transects-paper/}.
 
@@
 We thank Steve Buckland for helpful comments on an early draft.
+We also thank Dave Miller for helpful discussions on density surface models and how they relate to point process models.
 Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S.

### 2024-11-12 paper.tex
add alt text to figures

@@
 \usepackage{url}
+\usepackage{subcaption}
 
@@
 The red outline marks the study area boundary as defined in \cite{camp_dsm_2020}.}
+	\subcaption*{Alt text: A map of the Hakalau study region with the locations of point transects.
+The point transects are arranged approximately evenly spaced along straight lines moving roughly east to west.
+The southern portion has higher survey effort comopared to the northern portion of the study area.}
 	\label{fig:2002studyareapointspt}
@@
 		\caption{The A) mean, B) coefficient of variation and C) standard deviation of the posterior intensity field.  For the mean and standard deviaion the units are number of birds per m$^2$.}
+		\subcaption*{Alt text:  A figure of 3 maps of the study region that summarise the posterior intensity field.
+Map A shows the posterior mean showing a hot spot in the southern portion of the map.
+Map B shows the posterior coefficient of variation, showing higher values in the northern portion of the map.
+Map C shows the posterior standard deviation, showing higher standard deviation values in the southern portion of the map.}
 		\label{fig:intensity-mean-cv}
@@
 Red circles indicate the observed count of pairwise distances within each bin.}
+		\subcaption*{Alt text:  A figure summarising the estimated spatial structure in the point transect distance sampling data.
+Sub-figure A shows the posterior detection function with associated uncertainty.
+Detectability drops to just above 0.25 at the maximum observable distance.
+Sub-figure B shows the posterior Mat\`ern correlation function for the random effect.
+It shows a decreasing correlation as distance increases, dropping to a correlation around 0.1 at a distance of 10 kilometres.
+Sub-figure C shows boxplots that summarise the posterior distribution of pairwise distances between observations simulated from the posterior predictive distribution.
+It shows a very close match between the posterior mean and pairwise distances observed in the real observations.}
 		\label{fig:dist-mat-pairs}
@@
 		\caption{Posterior abundance.  Grey area marks two Monte Carlo standard errors above and below the mean.}
+		\subcaption*{Alt text:  A figure summarising the posterior for abundance.  The posterior is approximately bell-shaped with the bulk of the posterior mass between 4000 and 8000 birds.}
 		\label{fig:realized-abundance-posterior}
@@
 	\caption{Three realizations of the posterior intensity field}
+	\subcaption*{Alt text: A figure showing three realisations of the posterior intensity field.
+The realisations all show broadly the same trend as the posterior mean but with more variability and a shorter effective range of correlation.}
 	\label{fig:intensity-realizations}
@@
 		\caption{Pointwise posterior quantiles for the intensity:  A) 0.025 quantile B) 0.975 quantile}
+		\subcaption*{Alt text: 2 maps summarising pointwise prediction quantiles for the posterior intensity field. Map A shows the 0.025 quantile, and map B shows the 0.975 quantile.}
 		\label{fig:intensity-quantiles}
@@
 The scale shows the probability values $1-\alpha$.}
+	\subcaption*{Alt text: Two maps showing estimates of excursion set and excursion functions.
+Map A shows the excursion set with a level corresponding to 1 bird per hectare and probability 0.95.
+The map shows a region in the south and a very small region slightly north of this that are within the excusion set.
+Map B shows the excursion function, showing that most of the hotspot would remain in the excursion set even under probability values very close to 1.
+A few areas additional areas would be included if the probability were lower than 0.95 but this are on the fringe and are not substantially larger excursion sets than the 0.95 example.}
 	\label{fig:excursions}

### 2025-01-08 paper.tex
change author order
most of Janine comments

@@
 
+% seems like this document class uses short author from the last author
+% macro in the preamble?  
+\author{Steve J. Kendall}
+\address{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}
+
 \author{Richard J. Camp}
 \address{U. S. Geological Survey, Pacific Island Ecosystems Research Center, P.O. Box 44, \hawaii{} National Park, HI 96718, U.S.A.}
-
-% seems like this document class uses short author from the last author
-% macro in the preamble?  
-\author[Andrew E. Seaton \textit{et al.}]{Steve J. Kendall}
-\address{U. S. Fish and Wildlife, Big Island National Wildlife Refuge Complex, 60 Nowelo St., Suite 100, Hilo, HI  96720, U.S.A.}
 
@@
 
-The estimation of the abundance and spatial distribution of wild populations of animals is a critical objective in ecology and conservation \citep{schwarz_estimating_1999}.This paper presents a statistical analysis of survey data collected on a critically endangered Hawaiian forest bird, offering insights into its population size and spatial dynamics, which are crucial to guide effective conservation management decisions.
+The estimation of the abundance and spatial distribution of wild populations of animals is a critical objective in ecology and conservation \citep{schwarz_estimating_1999}.
+This paper is motivated by the need to analyse wildlife survey data to achieve this aim.
+Such data necessarily has spatial and temporal dependencies, as well as complex observation processes, that need to be accounted for in a statistical analysis.
+In particular, we present analysis of data collected on a critically endangered Hawaiian forest bird, offering insights into its population size and spatial dynamics, which are crucial to guide effective conservation management decisions.
 The \hawaii{} \akepa{} (hereafter \akepa{}; \textit{Loxops coccineus}; nomenclature according to \citealp{usfws_akepa_1970}) is an endemic species whose population declined dramatically during the 20th century  \citep{usfws_revised_2006, judge_akepa_2018}.
@@
 Information from the HFBS is used to update the International Union for Conservation of Nature redlist \citep{iucn_redlist_2022}, where the \akepa{} is currently listed as endangered.
-HFBS data is also used to establish preserves that coincide with native bird hotspots, including Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{} (hereafter Hakalau).
+HFBS data are also used to establish preserves that coincide with native bird hotspots, including Hakalau Forest Unit of the Big Island National Wildlife Refuge Complex on the island of \hawaii{} (hereafter Hakalau).
 Hakalau is the first wildlife refuge established in \hawaii{} with the primary purpose to protect and manage native forests for the conservation of threatened and endangered bird and plant species.
@@
 The point transects are arranged approximately evenly spaced along straight lines moving roughly east to west.
-The southern portion has higher survey effort comopared to the northern portion of the study area.}
+The southern portion has higher survey effort compared to the northern portion of the study area.}
 	\label{fig:2002studyareapointspt}
@@
 The probability of detection is modelled using a parametric function and, given estimates of detectability, a randomized sampling design allows for the construction of Horvitz-Thompson-like estimators of animal density \citep{ buckland_advanced_2004, horvitz_generalization_1952}.
-Under this approach, animal density is uniform within each strata of the survey.
+Under this approach, animal density is uniform within each stratum of the survey.
 
@@
 In the second stage, detectability estimates from the first stage are used as an offset term in a generalized additive model framework.
-Detections are are also binned into counts within user-defined sampling units (point transects or segments of line transects) and an appropriate response distribution is chosen to model these counts.
+Detections are also binned into counts within user-defined sampling units (point transects or segments of line transects) and an appropriate response distribution is chosen to model these counts.
 The \texttt{dsm} package \citep{miller_spatial_2013}, for example, provides tools to do this using the \texttt{R} package \texttt{mgcv} \citep{wood_gam_2017} to implement the spatially-explicit count model.
@@
 
-A downside to this approach is that the solutions to the detection model SPDE are not necessarily monotonically decreasing, which is a critical assumption of detection functions \citep{buckland_distance_2015}.
+A downside to this approach is that the solutions to the SPDE are not necessarily monotonically decreasing, which is a critical assumption of detection functions \citep{buckland_distance_2015}.
 The authors' suggestion to reject any non-monotonically decreasing functions when sampling from the posterior is potentially computationally wasteful.
@@
 Distance sampling models specify the thinning function $g(\bs)$ as a function that decays as distance increases.
-The type of distance measured depends on the type of survey.
-For line transects the perpendicular distance to the transect line is used whereas for point transects it is the horizontal distance to the observer.
+The distance measured depends on the type of survey transect.
+For line transects the perpendicular distance to the transect line is used, whereas for point transects the distance refers to the distance to the observer at a central, unmoving location.
 For the remainder of the paper we assume a point transect survey design and hence horizontal distance to an observer located at the centre of each point transect.
@@
 \end{align}
-We note that this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.
+As mentioned above, this is a relaxation of the traditional assumption in distance sampling that the intensity is constant within each sampling unit.
 A similar relaxation is also possible in the case of line transects \citep{yuan_point_2017} and this assumption complements the choice of piecewise linear basis functions used for the spatially-structured random effect described in Section \ref{sec-gmrf}.
@@
 \end{equation}
-where $C$ is a constant.  This approximate likelihood is implemented in \texttt{inlabru} as a \texttt{"cp"} likelihood.
+where $C$ is a constant.
 
-The approach taken here is similar to the Berman-Turner device \citep{baddeley_practical_2000, berman_approximating_1992} that is used to fit a wide variety of point process models using GLM software.
+The approach taken here is similar to the Berman-Turner device \citep{baddeley_practical_2000, berman_approximating_1992} that is used to fit a wide variety of point process models using GLM-type fitting software.
 The main difference with our approach is that we do not use the assumption that the locations of observed points form a part of the quadrature scheme.
+This approximate likelihood is implemented in \texttt{inlabru} as a \texttt{"cp"} likelihood.
 
@@
 The standard deviation (SD) map (\autoref{fig:intensity-mean-cv}C) shows spatially varying SD values that are also intended to indicate uncertainty.
-However, a default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.
-Whether these relative differences matter is dependent on the overall context of the analysis.
+
+While these measures of uncertainty do convey some information, our main issue with plots like these is that a default colour scale will nearly always show some spatial variation in uncertainty, with some regions of relatively high and low uncertainty.
+Whether these relative differences matter is a separate question, one that is dependent on the overall context of the analysis.
 Some levels of variability may be acceptable in some contexts and unacceptable in others.
+It is often unclear what these uncertainty maps actually mean to us in the wider context of an analysis.
+Below we advocate for an approache that requires more thought about what levels of uncertainty are acceptable for a given aim, and to produce model outputs that incorporate this thinking.
+The Bayesian one-stage approach is well suited to support this kind of bespoke model output.
 
@@
 One option is to use excursions sets and excursion functions \citep{bolin_excursion_2015}.
-This requires users to choose thresholds of interest (e.g.
+These approaches require users to choose thresholds of interest (e.g.
 the definition of a hotspot), and to select an acceptable level of uncertainty in knowing whether the field exceeds this threshold.
-This forces us to consider this explicitly and leaves us less at the whim of an arbitrary colour scale.
-Crucially, excursion sets and functions can be estimated using Monte Carlo samples and so fit well within our one-stage Bayesian approach.
+This is in contrast to the issues we mentioned above regarding default colour scales and unclear meanings of uncertainty maps.
+Crucially, excursion sets and functions can be estimated using Monte Carlo samples and so fit well within the one-stage Bayesian approach.
 

### 2025-12-10 R/data_scripts/create_mesh.R
Code polish; needs rerunning

@@ -1,47 +1,70 @@
 # Load required packages
 library(inlabru)
-library(INLA)  # need testing version of inla
+library(INLA) # need testing version of inla
 library(sf)
 library(ggplot2)
 
-data_path = here::here("R", "data")
+data_path <- here::here("R", "data")
 point_transects <- readRDS(here::here(data_path, "samplers.RDS"))
-study_area <- readRDS(here::here(data_path,"study_area.RDS"))
-outer_boundary = st_buffer(study_area, dist = 10)
+study_area <- readRDS(here::here(data_path, "study_area.RDS"))
+outer_boundary <- fm_nonconvex_hull(study_area, convex = 10)
 
 # meshbuilder()
 
-## Generated by meshbuilder()
+## Originally generated by meshbuilder(), then modified.
 
 ## Build boundary information:
-## (fmesher supports SpatialPolygons, but this app is not (yet) intelligent enough for that.)
 boundary <- list(
-  fm_as_segm(study_area),
-  fm_as_segm(outer_boundary)
+  study_area,
+  outer_boundary
 )
 
 ## Build the mesh:
-mesh <- fm_mesh_2d_inla(boundary=boundary,
-                        max.edge=c(0.2, 4.1),
-                        min.angle=c(27, 25),
-                        max.n=c(48000, 16000), ## Safeguard against large meshes.
-                        max.n.strict=c(128000, 128000), ## Don't build a huge mesh!
-                        cutoff=0.076, ## Filter away adjacent points.
-                        offset=c(0.1, 0.3)) ## Offset for extra boundaries, if needed.
+mesh_old <- fm_mesh_2d_inla(
+  boundary = boundary,
+  max.edge = c(0.2, 4.1),
+  min.angle = c(27, 25),
+  max.n = c(48000, 16000), ## Safeguard against large meshes.
+  max.n.strict = c(128000, 128000), ## Don't build a huge mesh!
+  cutoff = 0.076, ## Filter away adjacent points.
+  offset = c(0.1, 0.3)
+) ## Offset for extra boundaries, if needed.
+mesh_hex <- fm_mesh_2d_inla(
+  loc = fm_hexagon_lattice(study_area, edge_len = 0.15),
+  boundary = boundary,
+  max.edge = c(0.15 * 1.1, 4.1),
+  min.angle = c(27, 25),
[…]

### 2025-12-10 R/data_scripts/create_study_area.R

@@ -3,13 +3,15 @@
 library(sf)
 library(ggplot2)
 
-data_raw_path = here::here("R", "data_raw")
-data_path = here::here("R", "data")
+data_raw_path <- here::here("R", "data_raw")
+data_path <- here::here("R", "data")
 samplers <- readRDS(here::here(data_path, "samplers.RDS"))
 
-bnd.csv = read.csv(here::here(data_raw_path, "OpenClosedForestBoundary.csv"))
-bnd.xy = data.frame(x = bnd.csv$EASTING / 1000,
-                    y = bnd.csv$NORTHING / 1000)
+bnd.csv <- read.csv(here::here(data_raw_path, "OpenClosedForestBoundary.csv"))
+bnd.xy <- data.frame(
+  x = bnd.csv$EASTING / 1000,
+  y = bnd.csv$NORTHING / 1000
+)
 
 # convert to sf polygon
 bnd.poly <- st_polygon(list(as.matrix(bnd.xy)))
@@ -22,5 +24,8 @@ ggplot() +
   geom_sf(data = samplers)
 
 saveRDS(bnd.sf,
-        file = here::here(data_path,
-                          "study_area.RDS"))
+  file = here::here(
+    data_path,
+    "study_area.RDS"
+  )
+)

### 2025-12-10 R/data_scripts/process_AnalysisDataset2002.R

@@ -5,9 +5,10 @@ library(dplyr)
 library(sf)
 library(ggplot2)
 
-data_path = here::here("R", "data_raw")
-newdata = read.csv(here::here(data_path, "AnalysisDataset2002.csv"),
-                   stringsAsFactors = FALSE)
+data_path <- here::here("R", "data_raw")
+newdata <- read.csv(here::here(data_path, "AnalysisDataset2002.csv"),
+  stringsAsFactors = FALSE
+)
 
 # Stratum codes:
 # OF = open forest
@@ -19,23 +20,25 @@ newdata %>%
   filter(Stratum %in% c("OF", "CF")) %>%
   distinct() -> samplers
 
-samplers$weight = 1
-samplers$x = samplers$Easting/1000
-samplers$y = samplers$Northing/1000
-samplers_sf = st_as_sf(samplers,
-                       coords = c("x", "y"))
+samplers$weight <- 1
+samplers$x <- samplers$Easting / 1000
+samplers$y <- samplers$Northing / 1000
+samplers_sf <- st_as_sf(samplers,
+  coords = c("x", "y")
+)
 
-obs = newdata %>%
+obs <- newdata %>%
   select(SampleLabel, Distance, Easting, Northing, Stratum, HabCode) %>%
   filter(Stratum %in% c("OF", "CF"), !is.na(Distance)) %>%
   rename(distance = Distance)
 
-obs$x = obs$Easting/1000
-obs$y = obs$Northing/1000
-obs$distance = obs$distance / 1000
+obs$x <- obs$Easting / 1000
+obs$y <- obs$Northing / 1000
+obs$distance <- obs$distance / 1000
 
-obs_sf = st_as_sf(obs,
-                  coords = c("x", "y"))
+obs_sf <- st_as_sf(obs,
+  coords = c("x", "y")
+)
 
 # check it looks okay
 ggplot() +
@@ -43,6 +46,6 @@ ggplot() +
   geom_sf(data = obs_sf, col = "green")
 
 # write to disk
-out_path = here::here("R", "data")
+out_path <- here::here("R", "data")
 saveRDS(samplers_sf, file = here::here(out_path, "samplers.RDS"))
 saveRDS(obs_sf, file = here::here(out_path, "obs.RDS"))

### 2025-12-10 R/data_summary.R

@@ -3,10 +3,10 @@
 library(INLA)
 library(inlabru)
 
-data_path = here::here("analyses", "data")
-study_area = readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
-samplers = readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
-mesh = readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
+data_path <- here::here("analyses", "data")
+study_area <- readRDS(here::here(data_path, "study_area_extended_no_crs.RDS"))
+samplers <- readRDS(here::here(data_path, "samplers_extended_no_crs.RDS"))
+mesh <- readRDS(here::here(data_path, "mesh_extended_no_crs.RDS"))
 
 # number of transects
 nrow(samplers@coords)
@@ -19,7 +19,7 @@ length(unique(obs$SampleLabel))
 
 # range of counts at each sampling unit
 library(dplyr)
-hmm = data.frame(SampleLabel = obs$SampleLabel)
+hmm <- data.frame(SampleLabel = obs$SampleLabel)
 hmm %<>%
   group_by(SampleLabel) %>%
   summarise(n = n())

### 2025-12-10 R/eval_model.R

@@ -10,26 +10,28 @@ library(sf)
 set.seed(9701071)
 
 # statsoc doc class textwidth in cm
-tw = 14.69785
-twi = 5.7865551181  # pdf() needs inches
+tw <- 14.69785
+twi <- 5.7865551181 # pdf() needs inches
 
 # axis title text size
-ax.size = 12
+ax.size <- 12
 
 #### Fitted model  ####
-model_path = here::here("R",
-                        "fitted_models",
-                        "fitted_model.RDS")
-fit = readRDS(model_path)
+model_path <- here::here(
+  "R",
+  "fitted_models",
+  "fitted_model.RDS"
+)
+fit <- readRDS(model_path)
 
 #### Other things ####
-data_path = here::here("R", "data")
-study_area = readRDS(here::here(data_path, "study_area.RDS"))
-samplers = readRDS(here::here(data_path, "samplers.RDS"))
-mesh = readRDS(here::here(data_path, "mesh.RDS"))
+data_path <- here::here("R", "data")
+study_area <- readRDS(here::here(data_path, "study_area.RDS"))
+samplers <- readRDS(here::here(data_path, "samplers.RDS"))
+mesh <- readRDS(here::here(data_path, "mesh.RDS"))
 
 #### Where to save figures ####
-fig_path = here::here("figures")
+fig_path <- here::here("figures")
 
 #### ggplot theme ####
 # theme_set(theme_minimal())
@@ -38,15 +40,15 @@ theme_set(theme_classic())
 #### Specify detection functions  ####
 
 # Log half-normal
-log_hn = function(distance, lsig){
-  -0.5*(distance/exp(lsig))^2
+log_hn <- function(distance, lsig) {
+  -0.5 * (distance / exp(lsig))^2
 }
 
 # Half-normal
 hn <- function(distance, lsig) exp(log_hn(distance, lsig))
 
 # Include r for switching to polar coords
-dsamp = function(distance, lsig){
+dsamp <- function(distance, lsig) {
   log(distance) + log_hn(distance, lsig)
 }
 
@@ -56,103 +58,139 @@ summary(fit)
 ##### Posterior intensity plots #####
 
 # Prediction locations
-pxl = fm_pixels(mesh,
-                dims = c(300, 300),
-                mask = study_area)
+pxl <- fm_pixels(mesh,
+  dims = c(300, 300),
+  mask = study_area
+)
 
 # map units are per km^2 = 1,000,000 m^2
 # I want intensity per hectare = 10,000 m^2
 # so divide by 100
[…]

### 2025-12-10 R/eval_spde.R

@@ -1,0 +1,260 @@
+# Produce summary figures of the posterior intensity field
+
+library(INLA)
+library(inlabru)
+library(ggplot2)
+library(scales)
+library(cowplot)
+library(sf)
+library(dplyr)
+library(magrittr)
+library(tidyr)
+
+set.seed(9701071)
+
+# statsoc doc class textwidth in cm
+tw <- 14.69785
+twi <- 5.7865551181 # pdf() needs inches
+
+# axis title text size
+ax.size <- 12
+
+#### Fitted model  ####
+model_path <- here::here(
+  "R",
+  "fitted_models",
+  "fitted_model.RDS"
+)
+fit <- readRDS(model_path)
+
+#### Other things ####
+data_path <- here::here("R", "data")
+study_area <- readRDS(here::here(data_path, "study_area.RDS"))
+samplers <- readRDS(here::here(data_path, "samplers.RDS"))
+mesh <- readRDS(here::here(data_path, "mesh.RDS"))
+
+#### Where to save figures ####
+fig_path <- here::here("figures")
+
+#### ggplot theme ####
+theme_set(theme_minimal())
+# theme_set(theme_classic())
+
+#### Specify detection functions  ####
+
+# Log half-normal
+log_hn <- function(distance, lsig) {
+  -0.5 * (distance / exp(lsig))^2
+}
+
+# Half-normal
+hn <- function(distance, lsig) exp(log_hn(distance, lsig))
+
+# Include r for switching to polar coords
+dsamp <- function(distance, lsig) {
+  log(distance) + log_hn(distance, lsig)
+}
+
+
+#### Posterior half-normal plot ####
+
+W <- 58 / 1000
+Wm <- 58 # transect radius in metres
+distdf <- data.frame(distance = seq(.Machine$double.eps, Wm, length = 100))
+
+# adjust for change of units (km in model fit, metres here)
+hnpred <- predict(fit, distdf, ~ hn(distance, lsig + log(1000)),
+                  n.samples = 100
+)
+
+ghn <- ggplot() +
+  gg(hnpred) +
+  ylim(c(0, 1)) +
+  ylab("detection probability\n") +
+  xlab("\ndistance (m)") +
+  theme(
+    axis.title = element_text(size = ax.size),
+    axis.text = element_text(size = ax.size)
+  )
+ghn
+
+#### Posterior Matern plot ####
[…]

### 2025-12-10 R/excursions.R

@@ -5,124 +5,152 @@ library(ggplot2)
 library(excursions)
 library(cowplot)
 library(sf)
-source(here::here("R", "gg.R"))   # small edit of gg.SpatialPixelsDataFrame
 
 # statsoc doc class textwidth in cm
-tw = 14.69785
-twi = 5.7865551181  # pdf() needs inches
+tw <- 14.69785
+twi <- 5.7865551181 # pdf() needs inches
 # set ggplot theme
 theme_set(theme_minimal())
 
 set.seed(1525)
 
 #### Fitted model  ####
-model_path = here::here("R",
-                        "fitted_models",
-                        "fitted_model.RDS")
-fit = readRDS(model_path)
+model_path <- here::here(
+  "R",
+  "fitted_models",
+  "fitted_model.RDS"
+)
+fit <- readRDS(model_path)
 
 #### Other things ####
-data_path = here::here("R", "data")
-study_area = readRDS(here::here(data_path, "study_area.RDS"))
-samplers = readRDS(here::here(data_path, "samplers.RDS"))
-mesh = readRDS(here::here(data_path, "mesh.RDS"))
+data_path <- here::here("R", "data")
+study_area <- readRDS(here::here(data_path, "study_area.RDS"))
+samplers <- readRDS(here::here(data_path, "samplers.RDS"))
+mesh <- readRDS(here::here(data_path, "mesh.RDS"))
 
 #### Where to save figures ####
-fig_path = here::here("figures")
+fig_path <- here::here("figures")
 
 #### monte-carlo samples of intensity ####
 
-n.mc = 500     # lower this when fiddling with things
-n.mc = 25     # lower this when fiddling with things
-pxl = fm_pixels(mesh,
-                dims = c(50, 50),
-                mask = study_area)
+n.mc <- 500 # lower this when fiddling with things
+n.mc <- 25 # lower this when fiddling with things
+pxl <- fm_pixels(mesh,
+  dims = c(50, 50),
+  mask = study_area
+)
 
-pxl.coords = st_coordinates(pxl)
-pxl$x = pxl.coords[,1]
-pxl$y = pxl.coords[,2]
-
-X = generate(fit,
-             pxl,
-             ~ exp(grf + Intercept),
-             n.samples = n.mc)
+X <- generate(fit,
+  pxl,
+  ~ exp(grf + Intercept),
+  n.samples = n.mc
+)
 
 #### excursions ####
 
[…]

### 2025-12-10 R/fit_model.R

@@ -8,44 +8,47 @@ library(INLA)
 
 #### Save Fitted Model Object? ####
 #### WARNING: THIS WILL OVERWRITE EXISTING FITTED MODEL OBJECT
-to_save = TRUE
-model_path = here::here("R",
-                        "fitted_models",
-                        "fitted_model.RDS")
+to_save <- TRUE
+model_path <- here::here(
+  "R",
+  "fitted_models",
+  "fitted_model.RDS"
+)
 
 #### Produce intermediate plots? ####
-to_plot = TRUE
+to_plot <- TRUE
 
 #### Load Data ####
-data_path = here::here("R", "data")
+data_path <- here::here("R", "data")
 obs <- readRDS(here::here(data_path, "obs.RDS"))
-study_area = readRDS(here::here(data_path, "study_area.RDS"))
-samplers = readRDS(here::here(data_path, "samplers.RDS"))
-mesh_path = here::here(data_path, "mesh.RDS")
-mesh = readRDS(mesh_path)
+study_area <- readRDS(here::here(data_path, "study_area.RDS"))
+samplers <- readRDS(here::here(data_path, "samplers.RDS"))
+mesh_path <- here::here(data_path, "mesh.RDS")
+mesh <- readRDS(mesh_path)
 
 #### Specify detection function  ####
 
 # Log half-normal
-log_hn = function(distance, lsig){
-  -0.5*(distance/exp(lsig))^2
+log_hn <- function(distance, lsig) {
+  -0.5 * (distance / exp(lsig))^2
 }
 
 # Half-normal
 hn <- function(distance, lsig) exp(log_hn(distance, lsig))
 
 # Include r for switching to polar coords
-dsamp = function(distance, lsig){
+dsamp <- function(distance, lsig) {
   log(distance) + log_hn(distance, lsig)
 }
 
 # Plot the data
-if (to_plot){
+if (to_plot) {
   g1 <- ggplot() +
     gg(mesh) +
     geom_sf(data = study_area) +
     geom_sf(data = samplers) +
-    geom_sf(data = obs, colour = "green")
+    geom_sf(data = obs, colour = "green") +
+    coord_sf()
   g1
 }
 
@@ -53,8 +56,9 @@ if (to_plot){
 
 # SPDE:
 matern <- inla.spde2.pcmatern(mesh,
-                              prior.sigma = c(2, 0.01),
-                              prior.range = c(300/1000, 0.01))
[…]

### 2025-12-10 R/gg.R

@@ -1,99 +1,0 @@
-# library(ggpolypath)
-
-# update gg.SpatialPixelsDataFrame to play well with factor
-# this could be improved by allowing pixels to have multiple columns, only do default
-# behaviour if length(names(data)) == 1, otherwise make user specify
-
-# the long lat default name behaviour comes from the fortify.sp function
-# by default this labels coords long lat, regardless of crs
-# I assume this is why Fabian just kept long lat as default for sp objects
-# That this behaviour comes from fortify means that we cannot use the aes(...) argument
-# to provide the names of the coordinates.  The created dataframe will have columns
-# long and lat.
-# So seems like easiest thing is to just just xlab and ylab to get labels you want
-
-gg.SpatialPixelsDataFrame <- function(data,
-                                      mapping = NULL,
-                                      alpha = NULL,
-                                      crs = NULL,
-                                      mask = NULL, ...) {
-  if (!is.null(crs)) {
-    data <- spTransform(data, crs)
-  }
-  if (!is.null(mask)) {
-    data <- data[as.vector(!is.na(over(data, mask))), ]
-  }
-
-  df <- as.data.frame(data)
-
-  if (!is.factor(data[,c(names(data)[[1]])])){
-  dmap <- aes_string(
-    x = coordnames(data)[1],
-    y = coordnames(data)[2],
-    fill = names(data)[[1]]
-  )} else
-    dmap <- aes_string(
-      x = coordnames(data)[1],
-      y = coordnames(data)[2],
-      group = names(data)[[1]]
-    )
-
-  if (!is.null(mapping)) {
-    dmap <- modifyList(dmap, mapping)
-  }
-  if (!is.null(alpha)) dmap <- modifyList(dmap, aes_string(alpha = alpha))
-  gm <- geom_tile(data = df, mapping = dmap, ...)
-
-  # If data is not discrete (factor), add default color palette
-  # if (!inherits(data[[deparse(dmap$fill)]], "factor")) {
-  #  gm = c(gm, scale_fill_gradientn(colours = bru.pal()))
-  # }
-  gm
-}
-
-
-#' # Update this so plotting identifies holes in polygons
-#' # replace geom_polygon with ggpolypath::geom_polypath
-#'
-#' #' Geom for SpatialPolygons objects
-#' #'
[…]

### 2025-12-10 R/make_study_area_fig.R

@@ -7,29 +7,36 @@ library(ggplot2)
 theme_set(theme_minimal())
 
 # statsoc doc class textwidth in cm
-tw = 14.69785
-twi = 5.7865551181  # pdf() needs inches
-ax.size = 12 # axis text size
+tw <- 14.69785
+twi <- 5.7865551181 # pdf() needs inches
+ax.size <- 12 # axis text size
 
-data_path = here::here("R", "data")
-study_area = readRDS(here::here(data_path, "study_area.RDS"))
-samplers = readRDS(here::here(data_path, "samplers.RDS"))
-study_area_boundaries <- sf::st_cast(study_area, "LINESTRING")
+data_path <- here::here("R", "data")
+study_area <- readRDS(here::here(data_path, "study_area.RDS"))
+samplers <- readRDS(here::here(data_path, "samplers.RDS"))
 
-pdf(file = here::here("figures", "study_area_design.pdf"),
-    width = twi/1.5, height = twi/1.5)
+pdf(
+  file = here::here("figures", "study_area_design.pdf"),
+  width = twi / 1.5, height = twi / 1.5
+)
 
 ggplot() +
-  geom_sf(data = study_area_boundaries,
-          # fill = NA,
-          size = 2,
-          colour = "red") +
-  geom_sf(data = samplers,
-          cex = 0.4) +
+  geom_sf(
+    data = study_area,
+    fill = NA,
+    size = 2,
+    colour = "red"
+  ) +
+  geom_sf(
+    data = samplers,
+    cex = 0.4
+  ) +
   xlab("Easting (km)") +
   ylab("Northing (km)") +
-  scale_x_continuous(limits = c(254, 262),
-                     breaks = seq(254, 262, by = 2)) +
+  scale_x_continuous(
+    limits = c(254, 262),
+    breaks = seq(254, 262, by = 2)
+  ) +
   theme(
     legend.position = "bottom",
     axis.text = element_text(size = ax.size),

### 2025-12-10 R/posterior_N.R

@@ -6,98 +6,114 @@ library(ggplot2)
 library(sf)
 library(patchwork)
 library(dplyr)
-source(here::here("R", "gg.R"))   # some small edits to inlabru gg methods
 
 # statsoc doc class textwidth in cm
-tw = 14.69785
-twi = 5.7865551181  # pdf() needs inches
+tw <- 14.69785
+twi <- 5.7865551181 # pdf() needs inches
 # set ggplot theme
 theme_set(theme_minimal())
 
 #### Fitted model  ####
-model_path = here::here("R",
-                        "fitted_models",
-                        "fitted_model.RDS")
-fit = readRDS(model_path)
+model_path <- here::here(
+  "R",
+  "fitted_models",
+  "fitted_model.RDS"
+)
+fit <- readRDS(model_path)
 
 #### Other things ####
-data_path = here::here("R", "data")
-study_area = readRDS(here::here(data_path, "study_area.RDS"))
-samplers = readRDS(here::here(data_path, "samplers.RDS"))
-obs = readRDS(here::here(data_path, "obs.RDS"))
-mesh = readRDS(here::here(data_path, "mesh.RDS"))
+data_path <- here::here("R", "data")
+study_area <- readRDS(here::here(data_path, "study_area.RDS"))
+samplers <- readRDS(here::here(data_path, "samplers.RDS"))
+obs <- readRDS(here::here(data_path, "obs.RDS"))
+mesh <- readRDS(here::here(data_path, "mesh.RDS"))
 
 #### Where to save figures ####
-fig_path = here::here("figures")
+fig_path <- here::here("figures")
 
 # integration points
-ips = fm_int(domain = mesh,
-             samplers = study_area)
+ips <- fm_int(
+  domain = mesh,
+  samplers = study_area
+)
 
 # test to get range for N posterior
-test = generate(fit,
-                newdata = ips,
-                formula = ~ sum(weight * exp(Intercept + grf)),
-                n.samples = 500)
+test <- generate(fit,
+  newdata = ips,
+  formula = ~ sum(weight * exp(Intercept + grf)),
+  n.samples = 500
+)
 
 # define range of N
-N_min = round(mean(test) - 4.5*sd(test))
-N_max = round(mean(test) + 4.5*sd(test))
-by = round((N_max - N_min)/100)
-N_seq = seq(N_min, N_max, by = by) # should be roughly length 100
[…]

### 2025-12-10 R/session_info.R

@@ -11,5 +11,7 @@ library(patchwork)
 library(dplyr)
 library(excursions)
 
-writeLines(capture.output(devtools::session_info()),
-           here::here("R", "session_info.txt"))
+writeLines(
+  capture.output(devtools::session_info()),
+  here::here("R", "session_info.txt")
+)

### 2025-12-10 R/eval_model.R
Add code for hazard-rate, to compare

@@ -59,7 +59,7 @@ summary(fit)
 
 # Prediction locations
 pxl <- fm_pixels(mesh,
-  dims = c(300, 300),
+  dims = c(150, 300),
   mask = study_area
 )
 

### 2025-12-10 R/eval_spde.R

@@ -100,21 +100,6 @@ samplers_buffered <- st_buffer(samplers, dist = W)
 ### Generate posterior point patterns
 n.pp <- 500
 
-# simulate intensities and detection functions
-# each column of llam is a realisation of the log-intensity at each mesh node
-post.sample <- generate(fit,
-  formula = ~ {
-    c(
-      llam = grf_latent + Intercept_latent,
-      lsig = lsig_latent
-    )
-  },
-  n.samples = n.pp
-)
-
-llam <- post.sample[seq_len(fm_dof(mesh)), ]
-lsig.post <- post.sample["lsig", ]
-
 # create distance bins and matrix to store counts
 breaks <- 0:12
 counts <- matrix(NA, nrow = n.pp, ncol = length(breaks) - 1)
@@ -153,64 +138,98 @@ inner_mesh <- fm_mesh_2d_inla(
   offset = 0.1
 ) ## Offset for extra boundaries, if needed.
 
-A <- fm_basis(mesh, loc = inner_mesh$loc)
+# simulate intensities and detection functions
+# each column of llam is a realisation of the log-intensity at each
+# inner_mesh node
+post.sample <- generate(
+  fit,
+  st_sf(geometry = fm_as_sfc(inner_mesh, format = "loc")),
+  formula = ~ {
+    c(llam = grf + Intercept, lsig = lsig_latent)
+  },
+  n.samples = n.pp
+)
 
-inner_llam <- A %*% llam
+inner_llam <- post.sample[seq_len(NROW(inner_mesh$loc)), ]
+lsig.post <- post.sample["lsig", ]
 
-# Note:  I think there is a way to use inner_mesh in the generate()
-# call above to avoid having to create A and project afterwards.
-# Have not tested this though.
-
-for (i in 1:n.pp) {
-  cat("\nGenerating point pattern ", i, "\n")
-  a.pp <- sample.lgcp(
-    mesh = inner_mesh,
-    loglambda = inner_llam[, i],
-    samplers = study_area
-  )
-  a.pp <- st_as_sf(a.pp)
-  # keep only detectable points
-  a.pp <- a.pp[a.pp %>%
-    st_within(samplers_buffered) %>%
-    lengths() > 0, ]
-  a.lsig <- lsig.post[i]
-
-  cat("\nThinning point pattern ", i, "\n")
-
-  # add transect ID for each point
[…]

### 2025-12-10 R/fit_model.R

@@ -32,13 +32,24 @@ mesh <- readRDS(mesh_path)
 log_hn <- function(distance, lsig) {
   -0.5 * (distance / exp(lsig))^2
 }
+log_hr <- function(distance, sig, gam) {
+  log(1-exp(-(distance / sig)^(-gam)))
+}
 
 # Half-normal
 hn <- function(distance, lsig) exp(log_hn(distance, lsig))
+hr <- function(distance, sig, gam) {
+  1-exp(-(distance / sig)^(-gam))
+}
 
 # Include r for switching to polar coords
 dsamp <- function(distance, lsig) {
   log(distance) + log_hn(distance, lsig)
+}
+
+# Include r for switching to polar coords
+dsamp2 <- function(distance, sig,gam) {
+  log(distance) + log_hr(distance, sig, gam)
 }
 
 # Plot the data
@@ -61,17 +72,24 @@ matern <- inla.spde2.pcmatern(mesh,
 )
 
 cmp <- ~ grf(main = geometry, model = matern) +
-  lsig(1) + Intercept(1)
+  lsig(1) + Intercept(1) +
+  sig(1, prec.linear = 1, marginal = bm_marginal(qexp, rate = 1/0.02)) +
+  gam(1, prec.linear = 1, marginal = bm_marginal(qexp, rate = 1))
 
 # Predictor formula:
 fml <- geometry + distance ~ grf +
   dsamp(distance, lsig) +
   log(2 * pi) + # 2*pi offset for not knowing angle theta
   Intercept
+fml2 <- geometry + distance ~ grf +
+  dsamp2(distance, sig, gam) +
+  log(2 * pi) + # 2*pi offset for not knowing angle theta
+  Intercept
 
 W <- 58 / 1000 # transect radius, units km
 distance_domain <- fm_mesh_1d(W * seq(.Machine$double.eps, 1, length.out = 30)^2)
 starting_values <- list(lsig = 3.36 - log(1000))
+starting_values2 <- list(sig = 3.36 - log(1000),gam = 0)
 
 lik <- bru_obs(
   data = obs,
@@ -84,16 +102,40 @@ lik <- bru_obs(
   formula = fml
 )
 
+lik2 <- bru_obs(
+  data = obs,
+  family = "cp",
+  samplers = samplers,
+  domain = list(
+    geometry = mesh,
+    distance = distance_domain
+  ),
+  formula = fml2
+)
+
 fit <- bru(
   components = cmp,
   lik,
   options = list(
[…]
